### 2015-03-13 background.tex
copy pasted text from the workshop paper (of course lots of things have to change)

\subsection{Concurrency in the Linux Kernel}
\label{concurrency}

Nowadays, multicore architectures have become the mainstream, leading to an ever-increasing demand for concurrency in operating systems. For example, modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently to perform device related operations; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging). All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:

\begin{definition}
\label{definition:datarace}
A data race occurs when two distinct threads access a shared memory location, at least one of the accesses modifies the location, and there is no mechanism in place to prevent these accesses from being simultaneous.
\end{definition}

Listing~\ref{data_race_example} is an example of a data race between two entry points of a Linux driver. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

%\begin{figure}[!h]
%\centering
%\noindent\begin{minipage}{.95\textwidth}
%\noindent\begin{minipage}{.485\textwidth}
%\begin{lstlisting}[caption={Example of a simple data race in a Linux device driver.},
%label=data_race_example]
%static void ep1 (struct shared *dev) {
%  dev->resource = 5;
%}
%
%static void ep2 (struct shared *dev) {
%  if (dev->resource != 5)
%    dev->resource = 3;
%}
%\end{lstlisting}
%\end{minipage}\hfill
%\noindent\begin{minipage}{.485\textwidth}
%\begin{lstlisting}[caption={Data race eliminated by using the highlighted locks.},
%label=lock_example]
%static void ep1 (struct shared *dev) {
%@lock(&dev->lock);@
%  dev->resource = 5;
%@unlock(&dev->lock);@
%}
%
%static void ep2 (struct shared *dev) {
%@lock(&dev->lock);@
%  if (dev->resource != 5)
%    dev->resource = 3;
%@unlock(&dev->lock);@
%}
%\end{lstlisting}
%\end{minipage}
%\end{minipage}
%\end{figure}

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e. source code that cannot be executed by more than one thread simultaneously. Listing~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.

Lock-based synchronisation is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread. Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we choose to focus on reasoning about locks.

\subsection{Limitations in Reasoning about Thread Interleavings}

A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during a concurrent execution of a program. The happens-before relation can be defined as follows:

\begin{definition}
\label{definition:datarace}
Let $E_1$ and $E_2$ be two events during a program's concurrent execution. If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
\end{definition}

Happens-before is transitive, i.e. for any three events $E_1$, $E_2$ and $E_3$, if $E_1 \rightarrow E_2$ and $E_2 \rightarrow E_3$, then $E_1 \rightarrow E_3$. If two events $E_1$ and $E_2$ occur in different threads, and neither $E_1 \rightarrow E_2$, nor $E_2 \rightarrow E_1$ holds, then $E_1$ and $E_2$ happen concurrently, which indicates a data race if the two events access the same shared resource and at least one of the accesses is a modifier. Figure~\ref{happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points of a driver. Each of the three statements \{A, B, C\} in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation. The same applies for each of the four statements \{D, E, F, G\} in thread 2. Statement C in thread 1 releases the lock, and subsequently allows D in thread 2 to acquire the same lock. Thus, $C \rightarrow D$. This means that for this specific thread interleaving the set of statements \{A, B, C\} happens before the set of statements \{D, E, F, G\}.

\begin{figure}[htbp]
\centering
\includegraphics[width=.95\linewidth]{img/happens_before.pdf}
\caption{Example showcasing the main limitation of happens-before.}
\label{happens_before}
\end{figure}

The above example showcases the main limitation of happens-before. If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements B and G, because $B \rightarrow G$ in this specific schedule. Thus, a tool based on happens-before requires a scheduler that produces as many different thread interleavings as possible to increase execution path coverage~\cite{savage1997eraser}. Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs. Furthermore, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}. Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.

### 2015-03-13 conclusion.tex
copy pasted text from the workshop paper (of course lots of things have to change)

In this paper, we presented \emph{static pair-wise lock set analysis}, a novel sound verification technique for proving the absence of data races in device drivers. In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving. Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated. We have prototyped this approach in \textsc{Whoop}, a practical tool for automatic concurrency verification of Linux drivers.

Compared to traditional data race detection tools that typically attempt to explore as many thread interleavings as feasible, and thus face scalability issues in the presence of realistic concurrent programs, \textsc{Whoop} not only entirely avoids reasoning about thread interleavings, but also allows the reuse of existing successful sequential verification techniques. The main limitation of our approach is that it can potentially report many false positives, as we over-approximate the device driver shared state. To tackle this problem, we plan to investigate (i) invariant generation for taming our coarse abstraction and (ii) counterexample feasibility checking to evaluate if a reported bug is real or spurious.

Finally, we plan to perform a large-scale evaluation of the technique against the state-of-the-art as soon as the prototype is complete. For this purpose, we are already gathering a large number of publicly available Linux device drivers.

### 2015-03-13 implementation.tex
copy pasted text from the workshop paper (of course lots of things have to change)

We have prototyped static pair-wise lock set analysis in \textsc{Whoop}, a practical tool for automatic verification of data race freedom in Linux drivers written in C~\cite{kernighan1988c}. The tool's architecture is depicted in Figure~\ref{figure:toolchain}, and the whole toolchain can be summarised as follows:

\begin{figure}[tbp]
\centering
\includegraphics[width=.99\linewidth]{img/toolchain.pdf}
\caption{The \textsc{Whoop} architecture, which is powered by the Clang/LLVM and Boogie frameworks.}
\label{figure:toolchain}
\end{figure}

Initially, the device driver source code, together with a Linux environmental model\footnote{Stub header files required for compiling and subsequently analysing a device driver.}, is compiled to LLVM-IR using Clang/LLVM~\cite{lattner2004llvm}. Then, the LLVM-IR is compiled to the Boogie~\cite{barnett2006boogie} verification language using SMACK~\cite{rakamaric2008automatic}, an open source LLMV-IR to Boogie translator which can efficiently model heap manipulating programs. Next, the Boogie program is transformed into a sequentialised abstract program, using pair-wise sequentialisation and lock set instrumentation as discussed in Section~\ref{method}. Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover. Successful verification implies that the original driver is free of data races, while an error denotes a \emph{potential} data race.

\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution). Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation. We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious. Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

### 2015-03-13 intro.tex
copy pasted text from the workshop paper (of course lots of things have to change)

Device drivers exist at the thin boundary between hardware and software. They are complex pieces of system-level software responsible for the interaction between an operating system and any hardware devices that are attached to a computer. Device drivers are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}. This has a negative impact in hardware product releases, as time-to-market is commonly dominated by device driver development, verification and validation schedules~\cite{yavatkar2012era}.

Even after a device driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

A major reason behind these concurrency bugs is that drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}: modern operating systems are able to invoke a device driver from multiple concurrent threads; a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context; and the user may decide to remove a device while its driver is still running some operation. All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms are in place to control concurrent access to shared resources.

Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system. Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races. Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on finding bugs related to operating system interfacing and/or proving other generic sequential program properties. The few of these techniques that attempt to deal with data races are usually either \emph{unsound} (i.e. can miss real bugs) or \emph{imprecise} (i.e. can report false bugs). Furthermore, most previous techniques that attempt to reason about thread interleavings face significant scalability issues, because of the exponentially large state-space of realistic concurrent programs~\cite{musuvathi2008finding}.

In this paper, we present \emph{static pair-wise lock set analysis}, a novel technique for automatically verifying the absence of data races in device drivers. The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lock set analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

In summary, our paper makes the following two contributions: (i) a novel concurrency verification technique for device drivers based on lock set analysis, sequentialisation and automated theorem proving; and (ii) a prototype tool, \textsc{Whoop}, that implements our technique and can be applied to Linux drivers for automatic concurrency verification.

### 2015-03-13 related.tex
copy pasted text from the workshop paper (of course lots of things have to change)

Using static analysis to detect data races is a promising alternative to dynamic concurrency bug finding, which is inherently limited by scalability and execution path coverage issues. Notable previous work includes the static analysers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}. However, all these three tools heavily rely on user annotations, and thus require a lot of manual effort that limits their applicability. \textsc{Whoop}, on the other hand, does not require any source code modifications, and thus can be applied with zero effort.

Most related to our work are the static lock set analysers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have significant limitations. The authors of RELAY found 5022 warnings when analysing the Linux kernel, with only 25 of them being true data races. To limit these false positives, RELAY employs post-analysis unsound filters, but these can also filter out true races. Although Locksmith successfully detected data races in several small Pthreads applications and 7 medium-sized Linux device drivers, it also reported a significant number of false positives. The authors also reported that Locksmith was unable to run on several large programs, showcasing its limited scalability.

### 2015-03-13 theory.tex
copy pasted text from the workshop paper (of course lots of things have to change)

Our technique resolves around reasoning about the \emph{lock sets} of a device driver. Lock set analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that consistently protect a memory location during program execution. If that lock set ever becomes empty, the tool reports a potential data race. This is because an inconsistent lock set suggests that a memory location can be accessed simultaneously by two or more threads. Lock set analysis avoids reasoning about arbitrary thread interleavings, and thus can scale well in realistic concurrent programs, but suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores. To counter these limitations, we apply the idea of Eraser's lock set analysis in a static verification context.

We propose \emph{static pair-wise lock set analysis}, a novel verification technique for proving data race freedom in device drivers. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lock set analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

The proof sketch behind our static lock set analysis is given in Section~\ref{proof}. The verification technique is described in more detail in Section~\ref{method}. The prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.

\subsection{Static lock set analysis proof sketch}
\label{proof}

We now formalise our approach to verifying data race freedom in a device driver by statically analysing its lock sets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all device drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.

For each entry point $\mathit{ep}_{i}$, we compute its lock set $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$. Furthermore, for each entry point $\mathit{ep}_{i}$, we compute $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which maps each $m$ to true if and only if $\mathit{ep}_{i}$ writes to $m$ during some execution.

\begin{theorem}
\label{theorem:locksets}
For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the device driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
\end{theorem}

A sketch of how the above theorem can be proved is as follows. Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$. By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$. By the definition of a lock set, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$. As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

\subsection{Verification method}
\label{method}

\begin{table*}
\small
\begin{center}
\begin{tabular}{lll}
\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
\noalign{\smallskip}\toprule
\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
\noalign{\smallskip}
                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
\noalign{\smallskip}
                                       & W$\lbrack $ptr$\rbrack := true;$ &
\end{tabular}
\end{center}
\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
\label{tab:instrumentation}
\end{table*}

Our approach begins by performing \emph{two-thread reduction}, an abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent driver, and then performs \emph{pair-wise sequentialisation}, which combines the two arbitrary threads in a single sequential pair. This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point. This process repeats until all pairs of entry points have been sequentialised. To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value. This over-approximates any effects from all the unmodeled threads on the device driver shared state.

Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point. The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}. The two regions directly contribute to the static lock set analysis and race checking as follows: the logger is responsible for recording all lock sets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lock set $\mathit{LS}$ from the logger. This whole process is summarised below:

\begin{description}
  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lock set $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false. Also, the logger's current lock set $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty. Each time the logger performs a read or write access to $m$, it updates its lock set $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$. This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.

  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lock set $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger. Also, the logger's current lock set $\mathit{LS}_{CR}$ is initially empty. Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty. If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$. In this case a counterexample is generated and reported to the user. To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
\end{description}

### 2015-03-13 intro.tex
updated intro

@@
 
-A major reason behind these concurrency bugs is that drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}: modern operating systems are able to invoke a device driver from multiple concurrent threads; a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context; and the user may decide to remove a device while its driver is still running some operation.
-All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms are in place to control concurrent access to shared resources.
+A major reason behind these concurrency bugs is that drivers, as well as other kernel modules in general, operate inside a complicated and hostile environment~\cite{corbet2005linux}.
+For example the Linux kernel is able to invoke its modules from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.
+All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.
 
@@
 Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
-Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on finding bugs related to operating system interfacing and/or proving other generic sequential program properties.
+Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs.
 The few of these techniques that attempt to deal with data races are usually either \emph{unsound} (i.e.
@@
 
-In this paper, we present \emph{static pair-wise lock set analysis}, a novel technique for automatically verifying the absence of data races in device drivers.
-The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lock set analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
+In this paper, we present \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g.
+device drivers and filesystems).
+The key idea behind our approach is that a module can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
 The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
+We then show that by exploiting the race-freedom guarantees that our technique provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers.
+We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules from the Linux distribution.
 
-In summary, our paper makes the following two contributions: (i) a novel concurrency verification technique for device drivers based on lock set analysis, sequentialisation and automated theorem proving; and (ii) a prototype tool, \textsc{Whoop}, that implements our technique and can be applied to Linux drivers for automatic concurrency verification.
+To summarize, our contributions are as follows:
+\begin{itemize}
+\item We present static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux kernel modules.
+\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug finder.
+\item We develop \whoop, a prototype tool that implements our approach and we use it to analyse 10 Linux kernel modules.
+\end{itemize}

### 2015-03-13 background.tex
update in background section

@@
 
-Nowadays, multicore architectures have become the mainstream, leading to an ever-increasing demand for concurrency in operating systems.
-For example, modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently to perform device related operations; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
+Nowadays, multicore architectures have become mainstream, leading to an ever-increasing demand for concurrency in operating systems.
+Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same module concurrently; a running module process can block and be replaced by another process in the same module; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
 All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:
@@
 
-Listing~\ref{data_race_example} is an example of a data race between two entry points of a Linux driver.
+Figure~\ref{data_race_example} is an example of a data race between two entry points of a Linux module.
 The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}.
@@
 
-%\begin{figure}[!h]
-%\centering
-%\noindent\begin{minipage}{.95\textwidth}
-%\noindent\begin{minipage}{.485\textwidth}
-%\begin{lstlisting}[caption={Example of a simple data race in a Linux device driver.},
-%label=data_race_example]
-%static void ep1 (struct shared *dev) {
-%  dev->resource = 5;
-%}
-%
-%static void ep2 (struct shared *dev) {
-%  if (dev->resource != 5)
-%    dev->resource = 3;
-%}
-%\end{lstlisting}
-%\end{minipage}\hfill
-%\noindent\begin{minipage}{.485\textwidth}
-%\begin{lstlisting}[caption={Data race eliminated by using the highlighted locks.},
-%label=lock_example]
-%static void ep1 (struct shared *dev) {
-%@lock(&dev->lock);@
-%  dev->resource = 5;
-%@unlock(&dev->lock);@
-%}
-%
-%static void ep2 (struct shared *dev) {
-%@lock(&dev->lock);@
-%  if (dev->resource != 5)
-%    dev->resource = 3;
-%@unlock(&dev->lock);@
-%}
-%\end{lstlisting}
-%\end{minipage}
-%\end{minipage}
-%\end{figure}
+\begin{figure}[!h]
+\centering
+\begin{lstlisting}
+static void ep1 (struct shared *dev) {
+  dev->resource = 5;
+}
+
+static void ep2 (struct shared *dev) {
+  if (dev->resource != 5)
+    dev->resource = 3;
+}
+\end{lstlisting}
+\caption{Simple data race in a Linux module}
+\label{data_race_example}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\noindent\begin{minipage}{.95\textwidth}
+\begin{lstlisting}
+static void ep1 (struct shared *dev) {
+lock(&dev->lock);
+  dev->resource = 5;
+unlock(&dev->lock);
+}
+
+static void ep2 (struct shared *dev) {
+lock(&dev->lock);
+  if (dev->resource != 5)
+    dev->resource = 3;
+unlock(&dev->lock);
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Using locks to eliminate the data race}
+\label{lock_example}
+\end{figure}
 
@@
 source code that cannot be executed by more than one thread simultaneously.
-Listing~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.
+Figure~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.
 

### 2015-03-13 intro.tex
update in background section

@@
 The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
-We then show that by exploiting the race-freedom guarantees that our technique provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers.
-We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules from the Linux distribution.
+
+We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers.
+We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules available in the Linux distribution.
 

### 2015-03-14 background.tex
updates to background

@@
 \subsection{Concurrency in the Linux Kernel}
-\label{concurrency}
+\label{bg:concurrency}
 
@@
 
-Figure~\ref{data_race_example} is an example of a data race between two entry points of a Linux module.
+Figure~\ref{fig:data_race_example} is an example of a data race between two entry points of a Linux module.
 The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}.
@@
 \caption{Simple data race in a Linux module}
-\label{data_race_example}
+\label{fig:data_race_example}
 \end{figure}
@@
 \caption{Using locks to eliminate the data race}
-\label{lock_example}
+\label{fig:lock_example}
 \end{figure}
@@
 source code that cannot be executed by more than one thread simultaneously.
-Figure~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.
+Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in the previous example.
 
@@
 On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread.
-Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we choose to focus on reasoning about locks.
+Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we focus on locks.
 
-\subsection{Limitations in Reasoning about Thread Interleavings}
+\subsection{Reasoning about Thread Interleavings}
+\label{bg:happensbefore}
 
-A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during a concurrent execution of a program.
-The happens-before relation can be defined as follows:
+A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during the execution of a concurrent program.
+Happens-before can be defined as follows:
 
@@
 Let $E_1$ and $E_2$ be two events during a program's concurrent execution.
-If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
+If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism (e.g.
+locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
 \end{definition}
@@
 If two events $E_1$ and $E_2$ occur in different threads, and neither $E_1 \rightarrow E_2$, nor $E_2 \rightarrow E_1$ holds, then $E_1$ and $E_2$ happen concurrently, which indicates a data race if the two events access the same shared resource and at least one of the accesses is a modifier.
-Figure~\ref{happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points of a driver.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=1\linewidth]{img/happens_before.pdf}
+\caption{Example showcasing the main limitation of happens-before}
+\label{fig:happens_before}
+\end{figure}
+
+Figure~\ref{fig:happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points.
 Each of the three statements \{A, B, C\} in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation.
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.95\linewidth]{img/happens_before.pdf}
-\caption{Example showcasing the main limitation of happens-before.}
-\label{happens_before}
-\end{figure}
-
-The above example showcases the main limitation of happens-before.
+The above example also showcases the main limitation of happens-before.
 If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements B and G, because $B \rightarrow G$ in this specific schedule.
@@
 Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs.
-Furthermore, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}.
+Furthermore, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}, a problem known as state space explosion.
 Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.
+
+\subsection{Reasoning about Locks}
+\label{bg:lockset}
+
+An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behaviour of concurrent programs.
+Lockset analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution.
+If that lockset ever becomes empty, the analysis reports a potential data race on that memory location.
+This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=1\linewidth]{img/lockset.pdf}
+\caption{Applying lockset analysis on two threads}
+\label{fig:locksets}
+\end{figure}
+
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis on two concurrently running threads.
+
+Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs.
+The technique, though, suffers from imprecision (i.e.
+can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
+Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores.

### 2015-03-14 intro.tex
updates to background

@@
 
-We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers.
+We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse device drivers.
 We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules available in the Linux distribution.

### 2015-03-14 theory.tex
updates to background

@@
-Our technique resolves around reasoning about the \emph{lock sets} of a device driver.
-Lock set analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that consistently protect a memory location during program execution.
-If that lock set ever becomes empty, the tool reports a potential data race.
-This is because an inconsistent lock set suggests that a memory location can be accessed simultaneously by two or more threads.
-Lock set analysis avoids reasoning about arbitrary thread interleavings, and thus can scale well in realistic concurrent programs, but suffers from imprecision (i.e.
-can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
-Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores.
-To counter these limitations, we apply the idea of Eraser's lock set analysis in a static verification context.
+To counter these limitations, we apply the idea of Eraser's lockset analysis in a static context.
 
-We propose \emph{static pair-wise lock set analysis}, a novel verification technique for proving data race freedom in device drivers.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lock set analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
+We propose \emph{static pair-wise lockset analysis}, a novel verification technique for proving data race freedom in device drivers.
+The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
 The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
 
-The proof sketch behind our static lock set analysis is given in Section~\ref{proof}.
+The proof sketch behind our static lockset analysis is given in Section~\ref{proof}.
 The verification technique is described in more detail in Section~\ref{method}.
@@
 
-\subsection{Static lock set analysis proof sketch}
+\subsection{Static lockset analysis proof sketch}
 \label{proof}
 
-We now formalise our approach to verifying data race freedom in a device driver by statically analysing its lock sets.
+We now formalise our approach to verifying data race freedom in a device driver by statically analysing its locksets.
 Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver.
@@
 
-For each entry point $\mathit{ep}_{i}$, we compute its lock set $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
+For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
 Furthermore, for each entry point $\mathit{ep}_{i}$, we compute $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which maps each $m$ to true if and only if $\mathit{ep}_{i}$ writes to $m$ during some execution.
@@
 By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
-By the definition of a lock set, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$.
+By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$.
 As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
@@
 The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}.
-The two regions directly contribute to the static lock set analysis and race checking as follows: the logger is responsible for recording all lock sets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lock set $\mathit{LS}$ from the logger.
+The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger.
 This whole process is summarised below:
@@
 \begin{description}
-  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lock set $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false.
-Also, the logger's current lock set $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty.
-Each time the logger performs a read or write access to $m$, it updates its lock set $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$.
+  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false.
+Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty.
+Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$.
 This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.
 
-  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lock set $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger.
-Also, the logger's current lock set $\mathit{LS}_{CR}$ is initially empty.
+  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger.
+Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty.
 Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty.

### 2015-03-14 intro.tex
updates to intro

@@
-Device drivers exist at the thin boundary between hardware and software.
-They are complex pieces of system-level software responsible for the interaction between an operating system and any hardware devices that are attached to a computer.
-Device drivers are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
-This has a negative impact in hardware product releases, as time-to-market is commonly dominated by device driver development, verification and validation schedules~\cite{yavatkar2012era}.
+Linux kernel modules (e.g.
+device drivers and filesystems) exist at the thin boundary between hardware and software.
+They are complex pieces of system-level software responsible for the interaction between a user application, the operating system and any hardware devices that are attached to a computer.
+Kernel modules are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
+Especially in the case of drivers, this has a negative impact in hardware product releases, as time-to-market is commonly dominated by device driver development, verification and validation schedules~\cite{yavatkar2012era}.
 
-Even after a device driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
+Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
 Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance.
@@
 
-A major reason behind these concurrency bugs is that drivers, as well as other kernel modules in general, operate inside a complicated and hostile environment~\cite{corbet2005linux}.
+A major reason behind these concurrency bugs is that Linux kernel modules operate inside a complicated and hostile environment~\cite{corbet2005linux}.
 For example the Linux kernel is able to invoke its modules from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.

### 2015-03-14 background.tex
update background on lockset analysis

@@
 Figure~\ref{fig:happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points.
-Each of the three statements \{A, B, C\} in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation.
-The same applies for each of the four statements \{D, E, F, G\} in thread 2.
-Statement C in thread 1 releases the lock, and subsequently allows D in thread 2 to acquire the same lock.
+Each of the three statements $\{A, B, C\}$ in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation.
+The same applies for each of the four statements $\{D, E, F, G\}$ in thread 2.
+Statement $C$ in thread 1 releases the lock, and subsequently allows $D$ in thread 2 to acquire the same lock.
 Thus, $C \rightarrow D$.
-This means that for this specific thread interleaving the set of statements \{A, B, C\} happens before the set of statements \{D, E, F, G\}.
+This means that for this specific thread interleaving the set of statements $\{A, B, C\}$ happens before the set of statements $\{D, E, F, G\}$.
 
 The above example also showcases the main limitation of happens-before.
-If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements B and G, because $B \rightarrow G$ in this specific schedule.
+If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements $B$ and $G$, because $B \rightarrow G$ in this specific schedule.
 Thus, a tool based on happens-before requires a scheduler that produces as many different thread interleavings as possible to increase execution path coverage~\cite{savage1997eraser}.
@@
 
-\subsection{Reasoning about Locks}
+\subsection{Reasoning about Locksets}
 \label{bg:lockset}
@@
 \includegraphics[width=1\linewidth]{img/lockset.pdf}
-\caption{Applying lockset analysis on two threads}
+\caption{Applying lockset analysis on a thread of a concurrent program}
 \label{fig:locksets}
@@
 
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis on two concurrently running threads.
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a thread of a concurrent program.
+Lockset analysis starts by creating one lockset for each thread $T$ in the program, called current lockset of $T$ or $CLS_T$, and one lockset for each shared variable $v$, called lockset of $v$ or $LS_v$.
+In this example we analyse only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$).
+Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\{ \}$.
+Likewise, $LS_A$ starts as the set of all existing locks in the program, which in this case is $\{M, N\}$.
+Lockset analysis then proceeds as follows:
+
+\begin{enumerate}
+\item In each \emph{lock} and \emph{unlock} operation, update ($CLS$) with the currently held locks.
+\item In each shared variable access, update ($LS_A$) with the intersection of the latest ($CLS$) and the ($LS_A$) of the previous statement.
+\item If ($LS_A$) ever becomes empty, then report that access to $A$ is unprotected.
+\end{enumerate}
 
@@
 Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores.
+To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

### 2015-03-14 theory.tex
update background on lockset analysis
update on theory section

@@
-To counter these limitations, we apply the idea of Eraser's lockset analysis in a static context.
-
-We propose \emph{static pair-wise lockset analysis}, a novel verification technique for proving data race freedom in device drivers.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
+We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
+The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
+The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
 
@@
 
-We now formalise our approach to verifying data race freedom in a device driver by statically analysing its locksets.
-Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver.
-Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all device drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one).
+We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets.
+Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module.
+Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one).
 Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
+The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
 
 For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
-Furthermore, for each entry point $\mathit{ep}_{i}$, we compute $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which maps each $m$ to true if and only if $\mathit{ep}_{i}$ writes to $m$ during some execution.
+Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution.
+In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
+\PDComment{any concrete reference that quantifiers might be harder for SMT solving?}
 
@@
 \label{theorem:locksets}
-For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the device driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
+For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
 \end{theorem}
@@
 By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
-By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$.
+By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$.
 As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
@@
 
-Our approach begins by performing \emph{two-thread reduction}, an abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent driver, and then performs \emph{pair-wise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
+Our approach begins by performing \emph{two-thread reduction}, an abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent driver, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
 This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point.

### 2015-03-14 intro.tex
various updates on abstract and intro

@@
-Linux kernel modules (e.g.
-device drivers and filesystems) exist at the thin boundary between hardware and software.
+Linux kernel modules, such as device drivers and filesystems, exist at the thin boundary between hardware and software.
 They are complex pieces of system-level software responsible for the interaction between a user application, the operating system and any hardware devices that are attached to a computer.
-Kernel modules are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
-Especially in the case of drivers, this has a negative impact in hardware product releases, as time-to-market is commonly dominated by device driver development, verification and validation schedules~\cite{yavatkar2012era}.
+Linux kernel modules are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
+Especially in the case of drivers, this has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
 
@@
 
-In this paper, we present \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g.
+In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux kernel modules written in the C language and (ii) accelerating external bug finders that can easily plug in our toolchain.
+\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g.
 device drivers and filesystems).
@@
 
-We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse device drivers.
-We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules available in the Linux distribution.
+We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse device drivers and other concurrent programs.
+Using \whoop and \corral we analysed 10 modules available in the Linux distribution.
 
@@
 \begin{itemize}
-\item We present static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux kernel modules.
+\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux kernel modules and accelerating a plugged-in precise bug finder.
+\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux kernel modules.
 \item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug finder.
-\item We develop \whoop, a prototype tool that implements our approach and we use it to analyse 10 Linux kernel modules.
+\item We analyse 10 Linux kernel modules, including various drivers and a filesystem.
 \end{itemize}

### 2015-03-14 theory.tex
various updates on abstract and intro

@@
 The verification technique is described in more detail in Section~\ref{method}.
-The prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
+The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}.
+Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.
+Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
 
-\subsection{Static lockset analysis proof sketch}
+\subsection{Static Lockset Analysis Proof Sketch}
 \label{proof}
@@
 
-\subsection{Verification method}
+\subsection{Verification Method}
 \label{method}
@@
 
-Our approach begins by performing \emph{two-thread reduction}, an abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent driver, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
+Our verification technique begins by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
 This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point.
@@
 \end{description}
+
+\subsection{Watchdog Summarisation}
+\label{summarisation}
+
+Draft.
+
+\subsection{Optimisations}
+\label{optimisations}
+
+Draft.

### 2015-03-14 technique.tex
updated the toolchain figure and some text

--- 2015-03-14 theory.tex
@@
-We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
+We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in precise bug finder.
+Figure~\ref{fig:whoop}
+
+\begin{figure*}
+\centering
+\includegraphics[width=.99\linewidth]{img/whoop.pdf}
+\caption{The \textsc{Whoop} infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and Corral) tools}
+\label{fig:whoop}
+\end{figure*}
 
@@
 Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
+
+\subsection{Static Pairwise Lockset Analysis}
+\label{spla}
+
+We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
+The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
+The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
 

### 2015-03-14 background.tex
update on theory section
wrote text regarding the translation from the original program to Boogie IVL

@@
 
-Figure~\ref{fig:data_race_example} is an example of a data race between two entry points of a Linux module.
+Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux module.
 The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}.
@@
 
-\begin{figure}[!h]
-\centering
-\begin{lstlisting}
+\begin{lstlisting}[caption = Simple data race in a Linux module, label = fig:data_race_example]
 static void ep1 (struct shared *dev) {
@@
 \end{lstlisting}
-\caption{Simple data race in a Linux module}
-\label{fig:data_race_example}
-\end{figure}
 
-\begin{figure}[!h]
-\centering
-\noindent\begin{minipage}{.95\textwidth}
-\begin{lstlisting}
+The most common technique for eliminating data races is using a \emph{locking} mechanism.
+Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e.
+source code that cannot be executed by more than one thread simultaneously.
+Example~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Example~\ref{fig:data_race_example}.
+
+\begin{lstlisting}[caption = Using locks to eliminate the data race, label = fig:lock_example]
 static void ep1 (struct shared *dev) {
@@
 \end{lstlisting}
-\end{minipage}
-\caption{Using locks to eliminate the data race}
-\label{fig:lock_example}
-\end{figure}
-
-The most common technique for eliminating data races is using a \emph{locking} mechanism.
-Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e.
-source code that cannot be executed by more than one thread simultaneously.
-Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in the previous example.
 
@@
 \includegraphics[width=1\linewidth]{img/happens_before.pdf}
-\caption{Example showcasing the main limitation of happens-before}
+\caption{Applying happens-before in two concurrently running threads}
 \label{fig:happens_before}
@@
 In this example we analyse only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$).
-Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\{ \}$.
+Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\varnothing$.
 Likewise, $LS_A$ starts as the set of all existing locks in the program, which in this case is $\{M, N\}$.

### 2015-03-14 technique.tex
wrote text regarding the translation from the original program to Boogie IVL

--- 2015-03-14 theory.tex
@@
+We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug finder.
+
+\begin{figure*}
+\centering
+\includegraphics[width=.99\linewidth]{img/whoop.pdf}
+\caption{The \whoop infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
+\label{fig:whoop}
+\end{figure*}
+
+Figure~\ref{fig:whoop} depicts the \whoop infrastructure.
+Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model.
+The environmental model (i.e.
+stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed.
+Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{tech:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Next, the analysis phase of \whoop runs (see Section~\ref{tech:analysis}), detecting all potential data races in the abstract program.
+Finally, \whoop uses information from the race analysis phase to accelerate bug finding with \corral (see Section~\ref{tech:bugfinding}).
+
+\subsection{Translation to Boogie IVL}
+\label{tech:translation}
+
+The main aim of this phase is to take an input Linux kernel module written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
+Boogie and \corral).
+To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.
+
+First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given module, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points.
+This function acts as the main entry point to the module after the latter has been registered with the kernel.
+This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{tech:analysis} and~\ref{tech:bugfinding}).
+The output is a file that contains information regarding the entry points and a file with the rewritten program.
+
+Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
+SMACK then translates the LLVM-IR to Boogie IVL.
+SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL.
+To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.
+
+\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
+static void entrypoint(struct net_device *dev) {
+  struct shared *tp = netdev_priv(dev);
+  mutex_lock(&tp->wk.mutex);
+  tp->field = 4;
+  mutex_unlock(&tp->wk.mutex);
+}
+\end{lstlisting}
+
+Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK.
+The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource.
+Notice that functional calls (e.g.
+to lock and unlock a mutex) are preserved in the translation.
+
+\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
+procedure entrypoint(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    $p0 := dev;
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call mutex_lock($p3);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    $M.0[$p4] := 4;
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call mutex_unlock($p6);
+    return;
+}
+\end{boogie}
+
+\subsection{Static Pairwise Lockset Analysis}
+\label{tech:analysis}
+
+%Next, the Boogie program is transformed into a sequentialised abstract program, using pair-wise sequentialisation and lock set instrumentation as discussed in Section~\ref{method}.
+Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover.
+Successful verification implies that the original driver is free of data races, while an error denotes a \emph{potential} data race.
+
 We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
@@
 Draft.
+
+\subsection{Assumptions and Limitations}
+\label{limitations}
+
+\subsection{Bug Finding Phase}
+\label{tech:bugfinding}
+
+%\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution).
+Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation.
+We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious.
+Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

### 2015-03-14 technique.tex
fix in the text

@@
 Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{tech:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
-Next, the analysis phase of \whoop runs (see Section~\ref{tech:analysis}), detecting all potential data races in the abstract program.
-Finally, \whoop uses information from the race analysis phase to accelerate bug finding with \corral (see Section~\ref{tech:bugfinding}).
+Next, \whoop performs its static lockset analysis (see Section~\ref{tech:analysis}) to detect all potential data races in the abstract program.
+Finally, \whoop uses information from the static analysis phase to accelerate bug finding with \corral (see Section~\ref{tech:bugfinding}).
 

### 2015-03-14 bugfinding.tex
split the monolithic technique section into finer grained sections

--- 2015-03-14 technique.tex
@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in precise bug finder.
-Figure~\ref{fig:whoop}
-
-\begin{figure*}
-\centering
-\includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\caption{The \textsc{Whoop} infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and Corral) tools}
-\label{fig:whoop}
-\end{figure*}
-
-The proof sketch behind our static lockset analysis is given in Section~\ref{proof}.
-The verification technique is described in more detail in Section~\ref{method}.
-The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}.
-Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.
-Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
-
-\subsection{Static Pairwise Lockset Analysis}
-\label{spla}
-
-We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
-
-\subsection{Static Lockset Analysis Proof Sketch}
-\label{proof}
-
-We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets.
-Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module.
-Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one).
-Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
-The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
-
-For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
-Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution.
-In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
-\PDComment{any concrete reference that quantifiers might be harder for SMT solving?}
-
-\begin{theorem}
-\label{theorem:locksets}
-For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
-\end{theorem}
-
-A sketch of how the above theorem can be proved is as follows.
-Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$.
-By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
-By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$.
-As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
-
-\subsection{Verification Method}
-\label{method}
-
-\begin{table*}
-\small
-\begin{center}
-\begin{tabular}{lll}
-\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
-\noalign{\smallskip}\toprule
-\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
-\noalign{\smallskip}
-                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
-\noalign{\smallskip}
-                                       & W$\lbrack $ptr$\rbrack := true;$ &
-\end{tabular}
-\end{center}
-\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
-\label{tab:instrumentation}
-\end{table*}
-
-Our verification technique begins by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point.
-This process repeats until all pairs of entry points have been sequentialised.
-To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value.
-This over-approximates any effects from all the unmodeled threads on the device driver shared state.
-
-Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point.
-The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}.
-The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger.
-This whole process is summarised below:
-
-\begin{description}
-  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false.
-Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty.
-Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$.
-This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.
-
-  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger.
-Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty.
-Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty.
-If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$.
-In this case a counterexample is generated and reported to the user.
-To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
-\end{description}
-
-\subsection{Watchdog Summarisation}
-\label{summarisation}
-
-Draft.
-
-\subsection{Optimisations}
-\label{optimisations}
-
-Draft.
+%\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution).
+Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation.
+We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious.
+Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

### 2015-03-14 staticanalysis.tex
split the monolithic technique section into finer grained sections

--- 2015-03-14 technique.tex
@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in precise bug finder.
-Figure~\ref{fig:whoop}
-
-\begin{figure*}
-\centering
-\includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\caption{The \textsc{Whoop} infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and Corral) tools}
-\label{fig:whoop}
-\end{figure*}
+After the original program has been translated to the abstract Boogie program, the core \whoop tool is invoked, which performs \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
+The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
+The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
 
@@
 Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.
-Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
-
-\subsection{Static Pairwise Lockset Analysis}
-\label{spla}
-
-We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
 
@@
 
-\subsection{Verification Method}
+\subsection{Instrumentation and Analysis}
 \label{method}
+
+\whoop Engine is invoked. In this phase, \whoop instruments the program for \emph{static pairwise lockset analysis}
 
@@
 Draft.
+
+\subsection{Assumptions and Limitations}
+\label{limitations}
+
+%Next, the Boogie program is transformed into a sequentialised abstract program, using pair-wise sequentialisation and lock set instrumentation as discussed in Section~\ref{method}.
+Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover.
+Successful verification implies that the original driver is free of data races, while an error denotes a \emph{potential} data race.

### 2015-03-14 translation.tex
split the monolithic technique section into finer grained sections

--- 2015-03-14 technique.tex
@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug finder.
-
-\begin{figure*}
-\centering
-\includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\caption{The \whoop infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
-\label{fig:whoop}
-\end{figure*}
-
-Figure~\ref{fig:whoop} depicts the \whoop infrastructure.
-Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model.
-The environmental model (i.e.
-stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed.
-Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{tech:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
-Next, the analysis phase of \whoop runs (see Section~\ref{tech:analysis}), detecting all potential data races in the abstract program.
-Finally, \whoop uses information from the race analysis phase to accelerate bug finding with \corral (see Section~\ref{tech:bugfinding}).
-
-\subsection{Translation to Boogie IVL}
-\label{tech:translation}
-
 The main aim of this phase is to take an input Linux kernel module written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
@@
 \end{boogie}
-
-\subsection{Static Pairwise Lockset Analysis}
-\label{tech:analysis}
-
-%Next, the Boogie program is transformed into a sequentialised abstract program, using pair-wise sequentialisation and lock set instrumentation as discussed in Section~\ref{method}.
-Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover.
-Successful verification implies that the original driver is free of data races, while an error denotes a \emph{potential} data race.
-
-We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
-
-The proof sketch behind our static lockset analysis is given in Section~\ref{proof}.
-The verification technique is described in more detail in Section~\ref{method}.
-The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}.
-Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.
-Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
-
-\subsection{Static Lockset Analysis Proof Sketch}
-\label{proof}
-
-We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets.
-Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module.
-Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one).
-Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
-The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
-
-For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
-Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution.
-In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
-\PDComment{any concrete reference that quantifiers might be harder for SMT solving?}
-
-\begin{theorem}
-\label{theorem:locksets}
-For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
-\end{theorem}
-
-A sketch of how the above theorem can be proved is as follows.
-Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$.
-By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
-By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$.
-As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
-
-\subsection{Verification Method}
-\label{method}
-
-\begin{table*}
-\small
-\begin{center}
-\begin{tabular}{lll}
-\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
-\noalign{\smallskip}\toprule
-\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
-\noalign{\smallskip}
-                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
-\noalign{\smallskip}
-                                       & W$\lbrack $ptr$\rbrack := true;$ &
-\end{tabular}
-\end{center}
-\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
-\label{tab:instrumentation}
-\end{table*}
-
-Our verification technique begins by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point.
-This process repeats until all pairs of entry points have been sequentialised.
-To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value.
-This over-approximates any effects from all the unmodeled threads on the device driver shared state.
-
-Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point.
-The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}.
-The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger.
-This whole process is summarised below:
-
-\begin{description}
-  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false.
-Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty.
-Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$.
-This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.
-
-  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger.
-Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty.
-Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty.
-If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$.
-In this case a counterexample is generated and reported to the user.
-To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
-\end{description}
-
-\subsection{Watchdog Summarisation}
-\label{summarisation}
-
-Draft.
-
-\subsection{Optimisations}
-\label{optimisations}
-
-Draft.
-
-\subsection{Assumptions and Limitations}
-\label{limitations}
-
-\subsection{Bug Finding Phase}
-\label{tech:bugfinding}
-
-%\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution).
-Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation.
-We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious.
-Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

### 2015-03-14 whoop.tex
split the monolithic technique section into finer grained sections

--- 2015-03-14 technique.tex
@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in precise bug finder.
-Figure~\ref{fig:whoop}
+We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug finder.
 
@@
 \includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\caption{The \textsc{Whoop} infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and Corral) tools}
+\caption{The \whoop infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
 \label{fig:whoop}
@@
 
-The proof sketch behind our static lockset analysis is given in Section~\ref{proof}.
-The verification technique is described in more detail in Section~\ref{method}.
-The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}.
-Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.
-Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.
-
-\subsection{Static Pairwise Lockset Analysis}
-\label{spla}
-
-We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
-The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
-
-\subsection{Static Lockset Analysis Proof Sketch}
-\label{proof}
-
-We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets.
-Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module.
-Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one).
-Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
-The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
-
-For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
-Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution.
-In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
-\PDComment{any concrete reference that quantifiers might be harder for SMT solving?}
-
-\begin{theorem}
-\label{theorem:locksets}
-For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
-\end{theorem}
-
-A sketch of how the above theorem can be proved is as follows.
-Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$.
-By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
-By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$.
-As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
-
-\subsection{Verification Method}
-\label{method}
-
-\begin{table*}
-\small
-\begin{center}
-\begin{tabular}{lll}
-\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
-\noalign{\smallskip}\toprule
-\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
-\noalign{\smallskip}
-                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
-\noalign{\smallskip}
-                                       & W$\lbrack $ptr$\rbrack := true;$ &
-\end{tabular}
-\end{center}
-\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
-\label{tab:instrumentation}
-\end{table*}
-
-Our verification technique begins by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point.
-This process repeats until all pairs of entry points have been sequentialised.
-To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value.
-This over-approximates any effects from all the unmodeled threads on the device driver shared state.
-
-Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point.
-The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}.
-The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger.
-This whole process is summarised below:
-
-\begin{description}
-  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false.
-Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty.
-Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$.
-This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.
-
-  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger.
-Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty.
-Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty.
-If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$.
-In this case a counterexample is generated and reported to the user.
-To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
-\end{description}
-
-\subsection{Watchdog Summarisation}
-\label{summarisation}
-
-Draft.
-
-\subsection{Optimisations}
-\label{optimisations}
-
-Draft.
+Figure~\ref{fig:whoop} depicts the \whoop infrastructure.
+Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model.
+The environmental model (i.e.
+stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed.
+Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program.
+Finally, \whoop uses information from the static analysis phase to accelerate bug finding with \corral (see Section~\ref{bugfinding}).

### 2015-03-15 staticanalysis.tex
text about the Whoop parsing phase

@@
 
-The proof sketch behind our static lockset analysis is given in Section~\ref{proof}.
-The verification technique is described in more detail in Section~\ref{method}.
-The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}.
-Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.
+The proof sketch behind our static lockset analysis is given in Section~\ref{tech:proof}.
+The instrumentation and analysis is described in more detail in Section~\ref{tech:method}.
+The summarisation that we perform to achieve scalability is discussed in Section~\ref{tech:summarisation}.
+Various optimisations that we perform to achieve precision are described in Section~\ref{tech:optimisations}.
+Finally, assumption and limitations are discussed in Section~\ref{tech:limitations}.
 
 \subsection{Static Lockset Analysis Proof Sketch}
-\label{proof}
+\label{tech:proof}
 
@@
 \subsection{Instrumentation and Analysis}
-\label{method}
+\label{tech:method}
 
-\whoop Engine is invoked. In this phase, \whoop instruments the program for \emph{static pairwise lockset analysis}
+The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentialising the abstract Boogie program.
+During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g.
+the following call to \texttt{mutex\_lock()}:
+
+\begin{boogie}
+call mutex_lock($p3);
+\end{boogie}
+
+Will be rewritten to:
+
+\begin{boogie}
+const {:lock} unique lock$0: int;
+
+call mutex_lock(lock$0);
+\end{boogie}
+
+If \whoop cannot infer a lock it will exit with a warning.
+A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks).
+It is relatively straightforward to enhance \whoop, though, with new locking primitives.
+\whoop will also exit with a warning if it detects ``improper'' use of locks (e.g.
+dynamic lock creation).
+The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing modules for the Linux kernel.
+
+When the lock rewriting finishes, \whoop traverses the Boogie IVL code, uses the information extracted by Chauffeur and separates each independent entry point call-graph from the rest of the program.
+To achieve this, it duplicates helper functions and renames them accordingly: e.g.
+for entry point \texttt{ep1}, a helper function \texttt{foo} will be renamed to \texttt{foo\$ep1}.
+This allows \whoop to perform entry point sensitive instrumentation and analysis.
+Although the program can potentially become much larger, this does not affect the analysis as discussed later.
+
+%\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
+%call mutex_lock($p3);
+%\end{boogie}
 
@@
 \subsection{Watchdog Summarisation}
-\label{summarisation}
+\label{tech:summarisation}
 
@@
 \subsection{Optimisations}
-\label{optimisations}
+\label{tech:optimisations}
 
@@
 \subsection{Assumptions and Limitations}
-\label{limitations}
+\label{tech:limitations}
 

### 2015-03-15 translation.tex
some fixes

@@
 This function acts as the main entry point to the module after the latter has been registered with the kernel.
-This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{tech:analysis} and~\ref{tech:bugfinding}).
+This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{staticanalysis} and~\ref{bugfinding}).
 The output is a file that contains information regarding the entry points and a file with the rewritten program.

### 2015-03-15 staticanalysis.tex
added text for locking instrumentation

@@
 Although the program can potentially become much larger, this does not affect the analysis as discussed later.
+
+Next, \whoop instruments the program with global variables for lockset analysis and race checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:
+
+\begin{boogie}
+var {:current_lockset} lock$0_in_CLS_$ep1: bool;
+var {:lockset} lock$0_in_LS_$M.0_$ep1: bool;
+var {:access_checking} WRITTEN_$M.0_$ep1: bool;
+var {:access_checking} READ_$M.0_$ep1: bool;
+\end{boogie}
+
+\whoop also instruments a global watchdog variable per memory region that is common to all entry points:
+
+\begin{boogie}
+const {:watchdog} WATCHED_ACCESS_$M.0: int;
+\end{boogie}
+
+The above instrumented variables represent the follows: \texttt{lock\$0\_in\_CLS\_\$ep1} is the current lockset (CLS) of \texttt{ep1}; \texttt{WATCHED\_ACCESS\_\$M.0} is an \emph{unconstrained} constant representing the offset to memory region \texttt{\$M.0} that should be checked for races; \texttt{lock\$0\_in\_LS\_\$M.0\_\$ep1} is the lockset for \texttt{\$M.0}; \texttt{WRITTEN\_\$M.0\_\$ep1} is the write-set for \texttt{\$M.0}; and \texttt{READ\_\$M.0\_\$ep1} is the read-set for \texttt{\$M.0}.
+Verification involves proving that two entry points cannot race at the watched offset of every memory region.
+The arbitrary watched offset implies that every offset of each memory region is race-free.
+Watchdog race checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
+
+\whoop proceeds by instrumenting each entry point call-graph for lockset analysis by replacing each call to a locking or unlocking function, with an instrumented version of the original lock and unlock functions.
+This means that Example~\ref{fig:smack_translation} will be instrumented as:
+
+\begin{boogie}%[caption = Program instrumented with lock/unlock functions, label = fig:lock_instrumentation]
+procedure entrypoint(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    $p0 := dev;
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep(lock$0, true);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    $M.0[$p4] := 4;
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep(lock$0, false);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_UPDATE\_CLS\_\$ep} is a special \whoop function that updates the CLS global variable to either true or false, if the lock is held or released accordingly.
+
+%The alternative, which we call watchdog race checking, uses a single, unconstrained constant representing an offset with respect to which races should be checked: the ?watched offset?.
+Verification involves proving for every array that a data race at the watched offset is impossible.
+Because the watched offset is arbitrary, this implies that every offset of each array is race-free.
+For each array, two Booleans are introduced to record whether a read from or write to the watched offset has occurred.
+Initially these Booleans are false, and they are reset at each barrier.
+Thread s sets the ?read?
+Boolean to true whenever it reads from the watched offset, and similarly for the ?write?
+Boolean.
+A race between s and t is reported if thread t reads from the watched offset and the ?write?
+Boolean is true, or if thread t writes to the watched offset and either the ?read?
+or ?write?
+Boolean is true.
+The non-deterministic choice per array access is eliminated.
 

### 2015-03-15 staticanalysis.tex
updates to instrumentation text

--- 2015-03-15 staticanalysis.tex
@@
 
+Next, \whoop instruments the program with global variables for lockset analysis and race checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:
+
+\begin{boogie}
+var {:current_lockset} lock$0_in_CLS_$ep1: bool;
+var {:lockset} lock$0_in_LS_$M.0_$ep1: bool;
+var {:access_checking} WRITTEN_$M.0_$ep1: bool;
+var {:access_checking} READ_$M.0_$ep1: bool;
+\end{boogie}
+
+\whoop also instruments a global watchdog variable per memory region that is common to all entry points:
+
+\begin{boogie}
+const {:watchdog} WATCHED_ACCESS_$M.0: int;
+\end{boogie}
+
+The above instrumented variables represent the follows: \texttt{lock\$0\_in\_CLS\_\$ep1} is the current lockset (CLS) of \texttt{ep1}; \texttt{WATCHED\_ACCESS\_\$M.0} is an \emph{unconstrained} constant representing the offset to memory region \texttt{\$M.0} that should be checked for races; \texttt{lock\$0\_in\_LS\_\$M.0\_\$ep1} is the lockset for \texttt{\$M.0}; \texttt{WRITTEN\_\$M.0\_\$ep1} is the write-set for \texttt{\$M.0}; and \texttt{READ\_\$M.0\_\$ep1} is the read-set for \texttt{\$M.0}.
+Verification involves proving that two entry points cannot race at the watched offset of every memory region.
+The arbitrary watched offset implies that every offset of each memory region is race-free.
+Watchdog race checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
+
+\whoop proceeds by instrumenting each entry point call-graph for lockset analysis by replacing each call to a locking or unlocking function with a \whoop-specific lock and unlock functions.
+This means that Example~\ref{fig:smack_translation} will become:
+
+\begin{boogie}
+procedure ep1(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    $p0 := dev;
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, true);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    $M.0[$p4] := 4;
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, false);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_UPDATE\_CLS\_\$ep1()} is a special function that updates the CLS global variable to either true or false, if the lock is held or released by \texttt{ep1} accordingly:
+
+\begin{boogie}
+procedure {:inline 1} _UPDATE_CLS_$ep1(lock: int,
+    isLocked: bool);
+  modifies lock$0_in_CLS_$ep1;
+
+implementation {:inline 1} _UPDATE_CLS_$ep1(
+    lock: int, isLocked: bool) {
+  _UPDATE:
+    lock$0_in_CLS_$ep1 := (if lock == lock$0 then
+      isLocked else lock$0_in_CLS_$ep1);
+  return;
+}
+\end{boogie}
+
+Next, \whoop replaces every write and read access in each entry point call-graph with a \whoop-specific function:
+
+\begin{boogie}
+procedure ep1(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    \textbf{$p0 := dev;}
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, true);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    call _WRITE_LS_$M.0_$ep1($p4);
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, false);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_WRITE\_LS\_\$M.0\_\$ep1()} is a special function that computes the intersection of the CLS and the lockset of \texttt{\$M.0} (see Section~\ref{bg:lockset}).
+It also updates the write-set of \texttt{\$M.0} to true, because there was a write access by \texttt{ep1}:
+
+\begin{boogie}
+procedure {:inline 1} _WRITE_LS_$M.0_$ep1(
+    ptr: int);
+  modifies lock$0_in_LS_$M.0_$ep1,
+    WRITTEN_$M.0_$ep1;
+
+implementation {:inline 1} _WRITE_LS_$M.0_$ep1(
+    ptr: int) {
+  _WRITE:
+    goto anon1_Then, anon1_Else;
+
+  anon1_Then:
+    assume {:partition} WATCHED_ACCESS_$M.0 ==
+      ptr && DEVICE_IS_REGISTERED_$ep;
+    lock$0_in_LS_$M.0_$ep := lock$0_in_CLS_$ep && lock$0_in_LS_$M.0_$ep;
+    WRITTEN_$M.0_$ep := true;
+    return;
+  anon1_Else:
+    assume {:partition} !(WATCHED_ACCESS_$M.0 ==
+      ptr && DEVICE_IS_REGISTERED_$ep);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_READ\_LS\_\$M.0\_\$ep1()} updates the lockset of \texttt{\$M.0} and the read-set accordingly for a read access.
+
+To be sound, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}) for each read access to a shared memory location.
+This over-approximates any effects from all the unmodeled threads on the shared state, but can and will result into false positives.
+
+%For each array, two Booleans are introduced to record whether a read from or write to the watched offset has occurred.
+Initially these Booleans are false, and they are reset at each barrier.
+Thread s sets the ?read?
+Boolean to true whenever it reads from the watched offset, and similarly for the ?write?
+Boolean.
+A race between s and t is reported if thread t reads from the watched offset and the ?write?
+Boolean is true, or if thread t writes to the watched offset and either the ?read?
+or ?write?
+Boolean is true.
+The non-deterministic choice per array access is eliminated.
+
 %\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
@@
 %\end{boogie}
-
-\begin{table*}
-\small
-\begin{center}
-\begin{tabular}{lll}
-\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
-\noalign{\smallskip}\toprule
-\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
-\noalign{\smallskip}
-                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
-\noalign{\smallskip}
-                                       & W$\lbrack $ptr$\rbrack := true;$ &
-\end{tabular}
-\end{center}
-\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
-\label{tab:instrumentation}
-\end{table*}
 

### 2015-03-15 translation.tex
updates to instrumentation text

@@
 \begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
-static void entrypoint(struct net_device *dev) {
+static void ep1(struct net_device *dev) {
   struct shared *tp = netdev_priv(dev);
@@
 \begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
-procedure entrypoint(dev: int) modifies $M.0; {
+procedure ep1(dev: int) modifies $M.0; {
   var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;

### 2015-03-15 staticanalysis.tex
updates to instrumentation text

@@
 
-To be sound, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}) for each read access to a shared memory location.
-This over-approximates any effects from all the unmodeled threads on the shared state, but can and will result into false positives.
+\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
+This is achieved by
 
-%For each array, two Booleans are introduced to record whether a read from or write to the watched offset has occurred.
-Initially these Booleans are false, and they are reset at each barrier.
-Thread s sets the ?read?
-Boolean to true whenever it reads from the watched offset, and similarly for the ?write?
-Boolean.
-A race between s and t is reported if thread t reads from the watched offset and the ?write?
-Boolean is true, or if thread t writes to the watched offset and either the ?read?
-or ?write?
-Boolean is true.
-The non-deterministic choice per array access is eliminated.
+This process repeats until all pairs of entry points have been sequentialised.
+To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}).
+This over-approximates any effects from all the unmodeled threads on the shared state.
 
-%\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
-%call mutex_lock($p3);
-%\end{boogie}
-
-Our verification technique begins by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point.
-This process repeats until all pairs of entry points have been sequentialised.
-To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value.
-This over-approximates any effects from all the unmodeled threads on the device driver shared state.
-
-Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point.
-The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}.
-The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger.
-This whole process is summarised below:
-
-\begin{description}
-  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false.
-Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty.
-Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$.
-This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.
-
-  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger.
-Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty.
 Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty.
@@
 To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
-\end{description}
 

### 2015-03-15 background.tex
some fixes
updated the instrumentation and analysis part

@@
 static void ep1 (struct shared *dev) {
-  lock(&dev->lock);
+  mutex_lock(&dev->lock);
   dev->resource = 5;
-  unlock(&dev->lock);
+  mutex_unlock(&dev->lock);
 }
@@
 static void ep2 (struct shared *dev) {
-  lock(&dev->lock);
+  mutex_lock(&dev->lock);
   if (dev->resource != 5)
     dev->resource = 3;
-  unlock(&dev->lock);
+  mutex_unlock(&dev->lock);
 }
@@
 \item In each shared variable access, update ($LS_A$) with the intersection of the latest ($CLS$) and the ($LS_A$) of the previous statement.
-\item If ($LS_A$) ever becomes empty, then report that access to $A$ is unprotected.
+\item If ($LS_A$) ever becomes empty, then report that the access to $A$ is unprotected.
 \end{enumerate}

### 2015-03-15 staticanalysis.tex
updated the instrumentation and analysis part

@@
 \whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by
+This is achieved by creating a checker function that calls the two instrumented entry points and then performs race checking assertions.
+This process repeats until all possible pairs of entry points have been sequentialised.
+To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}).
+This over-approximates any effects from all the unmodeled threads on the shared state of the pair.
+Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.
 
-This process repeats until all pairs of entry points have been sequentialised.
-To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}).
-This over-approximates any effects from all the unmodeled threads on the shared state.
+After each checker function calls a pair of entry points, it asserts that each shared memory location is protected by at least one common lock (see Section~\ref{bg:lockset}).
+The assertion check is triggered iff the memory location was accessed by both entry points and at least one of the accesses was a write.
+We do not check for read-read races, which are inherently benign.
+The following is an example of a checker function for entry points \texttt{ep1} and \texttt{ep2}:
 
-Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty.
-If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$.
-In this case a counterexample is generated and reported to the user.
-To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
+\begin{boogie}
+procedure check$ep1$ep2(dev: int);
+  requires !lock$0_in_CLS_$ep1;
+  requires !lock$0_in_CLS_$ep2;
+  requires lock$0_in_LS_$M.0_$ep1;
+  requires lock$0_in_LS_$M.0_$ep2;
+  requires !WRITTEN_$M.0_$ep1;
+  requires !WRITTEN_$M.0_$ep2;
+  requires !READ_$M.0_$ep1;
+  requires !READ_$M.0_$ep2;
+  modifies lock$0_in_CLS_$ep1, lock$0_in_CLS_$ep2,
+    lock$0_in_LS_$M.0_$ep1, lock$0_in_LS_$M.0_$ep2,
+    WRITTEN_ $M.0_$ep1,WRITTEN_$M.0_$ep2,
+    READ_$M.0_$ep1, READ_$M.0_$ep2;
+
+implementation check$ep1$ep2(dev: int) {
+  _CHECK:
+    call ep1(dev);
+    call ep2(dev);
+    assert (WRITTEN_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
+      || (WRITTEN_$M.0_$ep1 && READ_$M.0_$ep2)
+      || (READ_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
+      ==> lock$0_in_LS_$M.0_$ep1 &&
+      lock$0_in_LS_$M.0_$ep2;
+    return;
+}
+\end{boogie}
+
+Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover (Boogie supports Z3~\cite{de2008z3} and CVC4~\cite{barrett2011cvc4}).
+Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
+To improve usability \whoop has a built-in error reporter that matches Boogie counterexamples to source code.
+The following is an example of a reported error:
+
+\begin{lstlisting}
+foo.c: error: potential write-write race:
+
+write by entry point ep1, foo.c:36:2
+  shared->resource = 4;
+
+write by entry point ep2, foo.c:52:2
+  shared->resource = 5;
+\end{lstlisting}
 
@@
 \label{tech:limitations}
-
-%Next, the Boogie program is transformed into a sequentialised abstract program, using pair-wise sequentialisation and lock set instrumentation as discussed in Section~\ref{method}.
-Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover.
-Successful verification implies that the original driver is free of data races, while an error denotes a \emph{potential} data race.

### 2015-03-15 translation.tex
updated the instrumentation and analysis part

@@
   mutex_lock(&tp->wk.mutex);
-  tp->field = 4;
+  tp->resource = 4;
   mutex_unlock(&tp->wk.mutex);

### 2015-03-15 conclusion.tex
update conclucsions

@@
-In this paper, we presented \emph{static pair-wise lock set analysis}, a novel sound verification technique for proving the absence of data races in device drivers.
+In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux kernel modules and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis.
 In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving.
 Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.
-We have prototyped this approach in \textsc{Whoop}, a practical tool for automatic concurrency verification of Linux drivers.
 
@@
 The main limitation of our approach is that it can potentially report many false positives, as we over-approximate the device driver shared state.
-To tackle this problem, we plan to investigate (i) invariant generation for taming our coarse abstraction and (ii) counterexample feasibility checking to evaluate if a reported bug is real or spurious.
-
-Finally, we plan to perform a large-scale evaluation of the technique against the state-of-the-art as soon as the prototype is complete.
-For this purpose, we are already gathering a large number of publicly available Linux device drivers.
+To tackle this problem, we use the race-related information that \whoop generates to speed up \corral, a state-of-the-art precise bug finder for concurrent programs.

### 2015-03-15 staticanalysis.tex
update conclucsions
initial text for summarisation

@@
 
-Draft.
+Early on during the development of \whoop we realised that scalability would be a serious issue.
+When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code), \whoop was running out of memory.
+One reason was that the translation produced hundreds of thousands of Boogie IVL code.
+Another reason was that some entry points of the driver have recursion (which does not work with inlining).
+Because of these reasons we decided to investigate summarisation techniques that would allow us to analyse Linux kernel modules in a scalable fashion, but without destroying precision and reporting countless false positives.
+
+Towards this, we came up with \emph{watchdog summarisation}, a novel summarisation technique that combines watchdog race checking with pre- and post-conditions and loop invariants for increased precision.
 
@@
 
-Draft.
+Kernel imposed locks
+Sound reduction of checked memory regions
+Modelling kernel methods
+Controlling SMACK's coarse over-approximation of function pointers
 
@@
 \label{tech:limitations}
+
+Draft.

### 2015-03-15 staticanalysis.tex
updates in summarisation

--- 2015-03-15 staticanalysis.tex
@@
 
-Draft.
+Early on during the development of \whoop we realised that scalability would be a serious issue.
+When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code), we were running out of memory.
+One reason was that the translation produced hundreds of thousands of Boogie IVL code, which were slow to parse and fit in memory.
+Another reason was that some entry points of the driver recursively call helper functions.
+Inherently, recursion does not work with inlining.
+
+To tackle the above issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision and reporting countless false positives.
+Watchdog summarisation combines watchdog race checking with pre- and post-conditions and loop invariants for increased precision.
 

### 2015-03-15 staticanalysis.tex
updates in summarisation

@@
 Inherently, recursion does not work with inlining.
+To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision and reporting countless false positives.
 
-To tackle the above issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision and reporting countless false positives.
-Watchdog summarisation combines watchdog race checking with pre- and post-conditions and loop invariants for increased precision.
+Watchdog summarisation is based on watchdog race checking and uses the Houdini~\cite{flanagan2001houdini} algorithm to infer pre- and post-conditions and loop invariants from a pool of automatically generated candidate invariants.
 

### 2015-03-15 staticanalysis.tex
updates in summarisation

@@
 
-Early on during the development of \whoop we realised that scalability would be a serious issue.
-When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code), we were running out of memory.
-One reason was that the translation produced hundreds of thousands of Boogie IVL code, which were slow to parse and fit in memory.
-Another reason was that some entry points of the driver recursively call helper functions.
+Early on during the development of \whoop, we realised that scalability would be a serious issue.
+When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory.
+One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory.
+Another reason was that some entry points recursively call helper functions.
 Inherently, recursion does not work with inlining.
@@
 
-\subsection{Optimisations}
+\subsection{Optimisations for Precision}
 \label{tech:optimisations}

### 2015-03-15 staticanalysis.tex
updates in summarisation

@@
 Inherently, recursion does not work with inlining.
-To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision and reporting countless false positives.
+To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision.
 
-Watchdog summarisation is based on watchdog race checking and uses the Houdini~\cite{flanagan2001houdini} algorithm to infer pre- and post-conditions and loop invariants from a pool of automatically generated candidate invariants.
+Watchdog summarisation is based on our lockset and watchdog race checking instrumentation and uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available in Boogie) to automatically compute summaries from a pool of automatically generated candidate invariants.
 

### 2015-03-15 staticanalysis.tex
updates in summarisation

@@
 
-Watchdog summarisation is based on our lockset and watchdog race checking instrumentation and uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available in Boogie) to automatically compute summaries from a pool of automatically generated candidate invariants.
+Watchdog summarisation uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
+
+The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{tech:method}).
+The key idea behind watchdog summarisation is that we automatically generate candidate invariants in the granularity of watched accesses.
+This means that for e.g.
+watched access \texttt{\$M.0[\$p0]} the entry point has performed a write access, but for \texttt{\$M.0[\$p1]} it has not.
+This reduces the false positives from reporting errors if only e.g.
+a field of a struct is racing, but the rest of the fields are not.
+An example of candidate post-conditions is the following:
+
+\begin{boogie}
+procedure foo$ep1(dev: int);
+  ...
+  ensures _b$ls$entrypoint1$0
+    ==> WATCHED_ACCESS_$M.0 == dev + 4
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ls$entrypoint1$1
+    ==> WATCHED_ACCESS_$M.0 == dev + 8
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$0
+    ==> WATCHED_ACCESS_$M.0 == dev + 4
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$1
+    ==> WATCHED_ACCESS_$M.0 == dev + 8
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ...
+\end{boogie}
 

### 2015-03-15 staticanalysis.tex
updates in summarisation

@@
 The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{tech:method}).
-The key idea behind watchdog summarisation is that we automatically generate candidate invariants in the granularity of watched accesses.
-This means that for e.g.
-watched access \texttt{\$M.0[\$p0]} the entry point has performed a write access, but for \texttt{\$M.0[\$p1]} it has not.
+The key idea behind watchdog summarisation is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
 This reduces the false positives from reporting errors if only e.g.
-a field of a struct is racing, but the rest of the fields are not.
-An example of candidate post-conditions is the following:
+a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not.
+An example of watchdog-based candidate post-conditions is the following:
 
@@
 
+To generate the candidate invariants, \whoop performs an intra-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
+To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
+For example the previous \texttt{\$M.0} candidates would now be:
+
+\begin{boogie}
+procedure foo$ep1(dev: int);
+  ...
+  ensures _b$ls$entrypoint1$0
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ls$entrypoint1$1
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$0
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$1
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ...
+\end{boogie}
+
 \subsection{Optimisations for Precision}

### 2015-03-15 staticanalysis.tex
text about optimisations

@@
 
-\subsection{Optimisations for Precision}
+\subsection{Modelling and Optimisations}
 \label{tech:optimisations}
 
-Kernel imposed locks
-Sound reduction of checked memory regions
-Modelling kernel methods
-Controlling SMACK's coarse over-approximation of function pointers
+A common practise for statically analysing software is to ``close'' the environment by abstracting away the low level implementation details.
+Any abstract environmental model, though, can and will ultimately result in decreased precision and thus false positives.
+This is one of the main limitations of statically analysing real life software that use complex external APIs.
+In our case, we developed a simple environmental model for the Linux kernel that consists of stub header files.
+Because we only focus on lockset-based properties during the race-checking phase of \whoop, we can get away with abstracting a lot of the underlying functionality.
+Our model, though, can still result in a lot of false positives.
+
+To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various modules.
+This is an ongoing manual effort: the more kernel modules we study, the more domain-specific properties we discover, which we then use for increased precision.
+
+In our experience, one of the most effective precision optimisations is to enrich \whoop with information regarding kernel imposed locks.
+The Linux kernel is able to serialise a number of entry point calls, which thus can never run concurrently with each other.
+An an example, a large number of networking and ethernet entry points are mutually serialised with RTNL, a network-specific kernel lock.
+\whoop will not create pairs for entry points that cannot run concurrently.
+
+Another optimisation, is to soundly reduce the checked memory locations.
+If a memory location is accessed only by one entry point in the pair, then it is safe to not check it.
+This can potentially decrease the amount of generated candidate invariants and speed up the verification process.
+
+Modelling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux.
+For example, the \texttt{register\_netdev()} function registers a network driver with the kernel.
+Before this function is called, the kernel is not able to invoke any of the driver's entry points.
+Modelling this functionality can reduce false positives when race checking accesses before the driver's registration.
+
+Finally, we also optimise the use of function pointers.
+LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site.
+To increase precision, Chauffeur analyses the AST of the original program and attempts to infer context-sensitive function pointer use information.
+This information is the used by \whoop to increase precision.
+To be sound, this optimisation is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.
 

### 2015-03-15 staticanalysis.tex
updates in the static analysis section

@@
 The instrumentation and analysis is described in more detail in Section~\ref{tech:method}.
-The summarisation that we perform to achieve scalability is discussed in Section~\ref{tech:summarisation}.
-Various optimisations that we perform to achieve precision are described in Section~\ref{tech:optimisations}.
-Finally, assumption and limitations are discussed in Section~\ref{tech:limitations}.
+The summarisation that we perform to achieve scalability is presented in Section~\ref{tech:summarisation}.
+Information about modelling, optimisations and assumptions is given in Section~\ref{tech:optimisations}.
+Finally, the limitations of our static lockset analysis are discussed in Section~\ref{tech:limitations}.
 
@@
 
-\subsection{Modelling and Optimisations}
+\subsection{Modelling, Optimisations and Assumptions}
 \label{tech:optimisations}
@@
 
-Finally, we also optimise the use of function pointers.
+We also optimise the use of function pointers.
 LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site.
@@
 
-\subsection{Assumptions and Limitations}
+Finally, we assume that the formal parameters of an entry point do not alias, and thus do not race.
+This is a potentially unsound feature that can be turned off using a command line option.
+In our experience so far, though, we have not missed any data races because of this assumption.
+We also assume that our environmental model and domain-specific knowledge are correct and do not cause any bugs to be missed.
+
+\subsection{Limitations}
 \label{tech:limitations}
 
-Draft.
+The main limitation of \whoop is that it is based on static lockset analysis.
+As an example of this limitation is that we do not check for dynamically created locks (or for locks that are provided by external libraries).
+Because we perform two thread reduction, we also over-approximate the read accesses, which is another source of false positives.
+Environmental modelling is also another limitation of \whoop.
+As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.
+
+We argue that further increasing the precision of our models is orthogonal to the contributions of this paper.
+Moreover, despite unavoidable imprecisions during the sound static data race analysis, \whoop can use the reported information to significantly speedup precise bug-finders for concurrent programs, as seen in Sections~\ref{bugfinding} and~\ref{evaluation}.

### 2015-03-16 staticanalysis.tex
updates in the static analysis section

@@
 In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
-\PDComment{any concrete reference that quantifiers might be harder for SMT solving?}
 
@@
 
-The main limitation of \whoop is that it is based on static lockset analysis.
-As an example of this limitation is that we do not check for dynamically created locks (or for locks that are provided by external libraries).
+\whoop inherits all limitations of lockset analysis (as discussed in Section~\ref{bg:lockset}): a violation of the locking discipline does not always correspond to a real error (e.g.
+\whoop does not reasons about lockfree data structures in a special way, it just over-approximates their read accesses); and we do not check for dynamically created locks (or for locks that are provided by external libraries).
 Because we perform two thread reduction, we also over-approximate the read accesses, which is another source of false positives.
-Environmental modelling is also another limitation of \whoop.
+
+Environmental modelling is another limitation of \whoop.
 As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.
-
 We argue that further increasing the precision of our models is orthogonal to the contributions of this paper.

### 2015-03-16 bugfinding.tex
draft text for speeding up corral
initial text for bug finding section

@@
+\whoop is a sound but potentially imprecise static data race analyser.
+In our experience, developers appreciate soundness but do not want to get overwhelmed by countless false positive reports.
+Towards this we allow an external precise bug-finder to be plugged in the \whoop infrastructure.
+The race-related information generated by \whoop can then be used to accelerate the bug-finder a bug-finder for concurrent programs that can be easily plugged in our infrastructure.
+Currently we only support \corral, a state-of-the-art bug-finder used in Microsoft to analyse Windows device drivers.
+We chose \corral because it is open-source and because it accepts Boogie IVL, similar to \whoop.
+Our technique though should be able to accelerate an arbitrary bug-finder for concurrent programs.
+
+Sound partial order reduction
+
+Yield instrumentation
+
+Granularities:
+
+Full-force
+
+Coarse
+
+MemoryRegion
+
+Speeding up Corral
+
 %\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution).
@@
 We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious.
-Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.
+Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug-finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

### 2015-03-16 conclusion.tex
initial text for bug finding section

@@
 The main limitation of our approach is that it can potentially report many false positives, as we over-approximate the device driver shared state.
-To tackle this problem, we use the race-related information that \whoop generates to speed up \corral, a state-of-the-art precise bug finder for concurrent programs.
+To tackle this problem, we use the race-related information that \whoop generates to speed up \corral, a state-of-the-art precise bug-finder for concurrent programs.

### 2015-03-16 intro.tex
initial text for bug finding section

@@
 
-In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux kernel modules written in the C language and (ii) accelerating external bug finders that can easily plug in our toolchain.
+In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux kernel modules written in the C language and (ii) accelerating external bug-finders that can easily plug in our toolchain.
 \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g.
@@
 
-We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse device drivers and other concurrent programs.
+We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers and other concurrent programs.
 Using \whoop and \corral we analysed 10 modules available in the Linux distribution.
@@
 \begin{itemize}
-\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux kernel modules and accelerating a plugged-in precise bug finder.
+\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux kernel modules and accelerating a plugged-in precise bug-finder.
 \item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux kernel modules.
-\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug finder.
+\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
 \item We analyse 10 Linux kernel modules, including various drivers and a filesystem.

### 2015-03-16 whoop.tex
initial text for bug finding section

@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug finder.
+We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug-finder.
 
@@
 Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program.
-Finally, \whoop uses information from the static analysis phase to accelerate bug finding with \corral (see Section~\ref{bugfinding}).
+Finally, \whoop uses information from the static analysis phase to accelerate bug-finding with \corral (see Section~\ref{bugfinding}).

### 2015-03-16 bugfinding.tex
updates in bug finding section

@@
 The race-related information generated by \whoop can then be used to accelerate the bug-finder a bug-finder for concurrent programs that can be easily plugged in our infrastructure.
-Currently we only support \corral, a state-of-the-art bug-finder used in Microsoft to analyse Windows device drivers.
+Currently we only support \corral~\cite{lal2012corral}, a state-of-the-art bug-finder used in Microsoft to analyse Windows device drivers.
 We chose \corral because it is open-source and because it accepts Boogie IVL, similar to \whoop.
-Our technique though should be able to accelerate an arbitrary bug-finder for concurrent programs.
+Our technique, though, should be able to accelerate any arbitrary bug-finder for concurrent programs.
 
-Sound partial order reduction
+\subsection{Bug-Finding with \corral}
+\label{bf:corral}
 
-Yield instrumentation
+\corral is a verifier for Boogie IVL that uses SMT solvers (e.g.
+Z3 or CVC4) underneath to reason about program behaviour and reports a counterexample if an assertion violation is found.
+\corral attempts to prove reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analysing a complex concurrent program.
+Towards this, \corral performs two abstractions: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that impact the generated verification conditions; and (ii) stratified inlining, where it attempts to identify a minimal unrolling of loops and recursion that also impact the generated verification conditions.
 
-Granularities:
+Because \corral performs bounded model checking, it is inherently unsound, in other words it can miss real bugs.
+Assuming a precise environmental model, though, \corral will only report true bugs.
+\whoop takes advantage of this precision to report only feasible data races.
 
-Full-force
+\subsection{Sound Partial Order Reduction}
+\label{bf:reduction}
 
-Coarse
+By default, \corral instruments a \texttt{yield} statement (which denotes a preemption) before every \emph{visible operation} (i.e.
+synchronisation operations and shared memory accesses) of a concurrently running thread.
+It then sequentialises the concurrent program with a schedule that explores all possible interleavings up to a pre-defined bound.
 
-MemoryRegion
+Our idea for accelerating \corral is simple: we simplify the sequentialisation that \corral performs by removing \texttt{yield} statements from all shared memory accesses that \whoop has soundly proved to not be racy.
+We basically perform a partial order reduction that can significantly reduce the interleavings that \corral needs to explore, and thus increase performance and scalability.
 
-Speeding up Corral
+We now show that our partial order reduction is sound, in other words it does not miss any more bugs than the bugs that \corral would miss even without taking advantage of the \whoop race-related information.
+Let \corral run an $n$ amount of entry point pairs with yields in all visible operations.
+Let this configuration be known as $C$.
+Each entry point pair is running in a separate \corral process.
+\corral will miss all bugs that are triggered by more than 3 bugs running concurrently and all bugs that are triggered by investigating a larger than $r$ recursion bound and larger than $k$ context switch bound.
+Let \corral then run the same $n$ amount of pairs, but this time with all \texttt{yield} statements in non-racy shared memory accesses (according to \whoop) removed.
+Let this configuration be known as $W+C$.
+Because these shared memory accesses cannot race, performing a \texttt{yield} will not introduce any new program behaviours.
+Thus, \corral will find the same bugs using the configuration $W+C$, as using the configuration $C$.
 
-%\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution).
-Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation.
-We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious.
-Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug-finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.
+\subsection{Race Checking Instrumentation}
+\label{bf:racechecking}
+
+We use a simple, but effective, race checking instrumentation with \corral. This instrumentation is optional and we use it as a feasibility checking for the potential data races that \whoop reports.

### 2015-03-16 evaluation.tex
updates in bug finding section

\subsection{Granularity of Yield Instrumentation}
\label{eval:granularity}

Granularities: Full-force; Coarse; MemoryRegion

### 2015-03-16 bugfinding.tex
updates in bug finding section

@@
 
-We use a simple, but effective, race checking instrumentation with \corral. This instrumentation is optional and we use it as a feasibility checking for the potential data races that \whoop reports.
+We use a simple, but effective, race checking instrumentation with \corral.
+This instrumentation, which is based on previous work~\cite{}, can be enabled using a command line option and we use it as a feasibility checking for the potential data races that \whoop reports.
+Whenever there is a write access, e.g.
+\texttt{\$M.13 := 0}, we instrument the Boogie IVL code as follows:
+
+\begin{boogie}
+$M.13 := 0;
+yield;
+assert $M.13 == 0;
+\end{boogie}
+
+Likewise, for a read access, e.g. \texttt{\$p0 := \$M.13}, we instrument the Boogie IVL code as follows:
+
+\begin{boogie}
+$p0 := $M.13;
+yield;
+$rv0 := $M.13;
+assert $p0 == $rv0;
+\end{boogie}
+
+Our instrumentation tolerates benign races, as it will never report an assertion violation if there is a write-write race, where the two write accesses update the shared memory location with the same value.

### 2015-03-16 bugfinding.tex
updates in bug finding section

@@
 yield;
-$rv0 := $M.13;
+$rv0 := $M.13; // $rv0 is an auxiliary variable
 assert $p0 == $rv0;
@@
 
-Our instrumentation tolerates benign races, as it will never report an assertion violation if there is a write-write race, where the two write accesses update the shared memory location with the same value.
+Our instrumentation tolerates benign races, as it will never report a write-write race, if the two write accesses update the shared memory location with the same value.

### 2015-03-17 intro.tex
added comment in intro

@@
 Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system.
-Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
+\PDComment{maybe also mention that races in C lead to undefined behaviour and potential issues with compilers?} Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
 Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs.

### 2015-03-18 evaluation.tex
draft table
updates to table

@@
 
-Granularities: Full-force; Coarse; MemoryRegion
+Granularities: Yield-All (\corral without \whoop); \yieldcoarse; \yieldmr
+
+\newcommand{\colspacing}{\hspace{1.8em}}
+\begin{table*}[t]
+\small
+\centering
+\begin{tabular}{l rrr rr rr rr rr}
+\centering
+\input{experiments/tables/granularity.tex}
+\end{tabular}
+\caption{Results ...}
+\label{tab:granularity}
+\end{table*}
+
+The machzwd watchdog driver explodes when running on vanilla \corral, because in some entry point pairs it attempts to analyse a large number of memory regions.
+For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyse.
+The same entry point pair is found race-free by \whoop.
+Using this information in \corral with the \yieldmr instrumentation, the pair takes less than a second to analyse.
+The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.

### 2015-03-18 evaluation.tex
update to table

@@
+\subsection{Experimental Setup}
+\label{eval:setup}
+
+We conducted all experiments on ... We use the following software versions: Z3 4.1; SMACK 1.4; Clang/LLVM 3.5; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
+
+\subsection{Benchmarks}
+\label{eval:benchmarks}
+
+We evaluate our methodology against 10 modules taken from the Linux kernel (currently a mix of latest and older distributions).
+
 \subsection{Granularity of Yield Instrumentation}
 \label{eval:granularity}
+
+We evaluate the scalability of our acceleration technique with respect to our benchmark sets.
 
@@
 \centering
-\begin{tabular}{l rrr rr rr rr rr}
+\begin{tabular}{l rrr rr rr r r}
 \centering

### 2015-03-18 evaluation.tex
update to table

@@
 The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.
+
+Our experiments used the default \corral context switch bound \texttt{/k:2}.
+A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialised program.
+Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings.
+\PDComment{We should take a few measurements with higher k -- no need to be in the table -- just to verify this argument.}

### 2015-03-18 bugfinding.tex
update to table

@@
 
-By default, \corral instruments a \texttt{yield} statement (which denotes a preemption) before every \emph{visible operation} (i.e.
+By default, \corral instruments a \texttt{yield} statement (which denotes a context-switch) before each \emph{visible operation} (i.e.
 synchronisation operations and shared memory accesses) of a concurrently running thread.
@@
 
-Our idea for accelerating \corral is simple: we simplify the sequentialisation that \corral performs by removing \texttt{yield} statements from all shared memory accesses that \whoop has soundly proved to not be racy.
-We basically perform a partial order reduction that can significantly reduce the interleavings that \corral needs to explore, and thus increase performance and scalability.
+Our idea for accelerating \corral is simple: we simplify the sequentialisation that \corral performs by instrumenting \texttt{yield} statements only in shared memory accesses that \whoop reported as potentially racy.
+This effectively acts as a partial order reduction that can significantly reduce the interleavings that \corral needs to explore, and thus increase performance and scalability.
 
@@
 Thus, \corral will find the same bugs using the configuration $W+C$, as using the configuration $C$.
+
+\subsection{Granularity of Context-Switches}
+\label{bf:granularity}
+
+As aforementioned, \corral instruments by default a \texttt{yield} statement in each visible operation.
+Using information from \whoop, though, we can avoid instrumenting context-switches in memory accesses that do not race.
+To achieve this, we use the \corral command line option \texttt{/cooperative}, which disables \texttt{yield} instrumentation in \corral, and then we instrument context-switches inside \whoop before sending these Boogie files to \corral.
+We have implemented two different levels of context-switch instrumentation granularity: \emph{\yieldcoarse} and \emph{\yieldmr}.
+
+\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop finds that a pair of entry points cannot race, then it will only instrument context-switches in synchronisation points of this pair; else if \whoop finds that a pair of entry points can race, then it will instrument context-switches in all visible operations of this pair.
+
+\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches in all synchronisation points, but only instruments context-switches in memory regions that can potentially race (regardless if a pair has not been fully verified as race-free).
+In our experiments (see Section~\ref{evaluation}), \yieldmr performs significantly faster than \yieldcoarse.
 

### 2015-03-18 evaluation.tex
update to table

@@
-\subsection{Experimental Setup}
-\label{eval:setup}
-
-We conducted all experiments on ... We use the following software versions: Z3 4.1; SMACK 1.4; Clang/LLVM 3.5; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
-
-\subsection{Benchmarks}
-\label{eval:benchmarks}
-
-We evaluate our methodology against 10 modules taken from the Linux kernel (currently a mix of latest and older distributions).
-
-\subsection{Granularity of Yield Instrumentation}
-\label{eval:granularity}
-
-We evaluate the scalability of our acceleration technique with respect to our benchmark sets.
-
-Granularities: Yield-All (\corral without \whoop); \yieldcoarse; \yieldmr
-
 \newcommand{\colspacing}{\hspace{1.8em}}
@@
 \caption{Results ...}
-\label{tab:granularity}
+\label{tab:races}
 \end{table*}
 
-The machzwd watchdog driver explodes when running on vanilla \corral, because in some entry point pairs it attempts to analyse a large number of memory regions.
+\subsection{Experimental Setup}
+\label{eval:setup}
+
+We conducted all experiments on ... We use the following software versions: Z3 4.1; SMACK 1.4; Clang/LLVM 3.5; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
+
+\subsection{Benchmarks}
+\label{eval:benchmarks}
+
+We evaluate our methodology against 10 modules taken from the Linux kernel (currently a mix of latest and older distributions). \PDComment{Say something about the benchmarks}
+
+\subsection{Context-Switch Granularity of Yield Instrumentation}
+\label{eval:granularity}
+
+We evaluate the scalability of our acceleration technique with respect to our benchmark sets.
+
+Our experiments use the default \corral context switch bound \texttt{/k:2}.
+A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialised program.
+Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings.
+\PDComment{We should take a few measurements with higher k -- no need to be in the table -- just to verify this argument.}
+
+When running \corral on its own, it can potentially report a spurious race (that \whoop does not report), because \corral is not enhanced with domain-specific information that \whoop uses to discard spurious races.
+Using \whoop + \corral, though, will inherently discard spurious data races (assuming a precise environmental model).
+
+In Table~\ref{tab:races}, we can notice that \whoop is much faster than \corral. For exampl, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
+
+When \corral on its own tries to analyse the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyse a large number of memory regions.
 For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyse.
@@
 The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.
-
-Our experiments used the default \corral context switch bound \texttt{/k:2}.
-A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialised program.
-Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings.
-\PDComment{We should take a few measurements with higher k -- no need to be in the table -- just to verify this argument.}

### 2015-03-18 staticanalysis.tex
fixed an example

@@
   $bb0:
-    \textbf{$p0 := dev;}
+    $p0 := dev;
     $p1 := $pa($p0, 32, 1);

### 2015-03-19 evaluation.tex
update table
updated table

@@
 
-We conducted all experiments on ... We use the following software versions: Z3 4.1; SMACK 1.4; Clang/LLVM 3.5; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
+We conducted all experiments on ... We use the following software versions: Clang/LLVM 3.5; SMACK 1.4; Z3 4.1; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
 
@@
 
-In Table~\ref{tab:races}, we can notice that \whoop is much faster than \corral. For exampl, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
+In Table~\ref{tab:races}, we can notice that \whoop is much faster than \corral. For example, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
 

### 2015-03-19 evaluation.tex
updated table

@@
 \centering
-\begin{tabular}{l rrr rr rr r r}
+\begin{tabular}{l rrr rrr rr r r}
 \centering
@@
 \end{tabular}
-\caption{Results ...}
+\caption{Data race analysis and scalability results.}
 \label{tab:races}
@@
 We conducted all experiments on ... We use the following software versions: Clang/LLVM 3.5; SMACK 1.4; Z3 4.1; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
+
+All reported times are in seconds and averages of N executions.
+We report: lines of C code (LoC); number of entry point pairs (\#Pairs); memory regions in the program (\#MRs); the number of racy memory regions that \whoop found (\#Racy MRs); the number of race reports that \whoop generated (\#Race Reports); the number of races that \corral found (same in all configurations).
 

### 2015-03-19 evaluation.tex
updated table

@@
 
-When running \corral on its own, it can potentially report a spurious race (that \whoop does not report), because \corral is not enhanced with domain-specific information that \whoop uses to discard spurious races.
+%When running \corral on its own, it can potentially report a spurious race (that \whoop does not report), because \corral is not enhanced with domain-specific information that \whoop uses to discard spurious races.
 Using \whoop + \corral, though, will inherently discard spurious data races (assuming a precise environmental model).
@@
 
-When \corral on its own tries to analyse the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyse a large number of memory regions.
+%When \corral on its own tries to analyse the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyse a large number of memory regions.
 For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyse.

### 2015-03-21 background.tex
initial text for bug finding section
many updates to refocus in drivers

@@
 Nowadays, multicore architectures have become mainstream, leading to an ever-increasing demand for concurrency in operating systems.
-Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same module concurrently; a running module process can block and be replaced by another process in the same module; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
+Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
 All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:
@@
 
-Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux module.
+Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver.
 The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}.
@@
 
-\begin{lstlisting}[caption = Simple data race in a Linux module, label = fig:data_race_example]
+\begin{lstlisting}[caption = Simple data race in a Linux driver, label = fig:data_race_example]
 static void ep1 (struct shared *dev) {
@@
 can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
-Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores.
+Furthermore, as a classic dynamic analyser, Eraser's bug-finding ability is limited to the execution paths that the tool explores.
 To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

### 2015-03-21 conclusion.tex
many updates to refocus in drivers

@@
-In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux kernel modules and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis.
+In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis.
 In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving.

### 2015-03-21 intro.tex
many updates to refocus in drivers

@@
-Linux kernel modules, such as device drivers and filesystems, exist at the thin boundary between hardware and software.
-They are complex pieces of system-level software responsible for the interaction between a user application, the operating system and any hardware devices that are attached to a computer.
-Linux kernel modules are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
-Especially in the case of drivers, this has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
+Device drivers exist at the thin boundary between hardware and software.
+They are complex pieces of system-level software responsible for the interaction between the operating system and any hardware devices that are attached to a computer.
+Drivers are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
+This has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
 
@@
 
-A major reason behind these concurrency bugs is that Linux kernel modules operate inside a complicated and hostile environment~\cite{corbet2005linux}.
-For example the Linux kernel is able to invoke its modules from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.
+A major reason behind these concurrency bugs is that device drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}.
+For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.
 All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.
@@
 
-In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux kernel modules written in the C language and (ii) accelerating external bug-finders that can easily plug in our toolchain.
-\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g.
+In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux drivers written in the C language and (ii) accelerating external bug-finders that can easily plug in our toolchain.
+\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers (e.g.
 device drivers and filesystems).
-The key idea behind our approach is that a module can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
+The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
 The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
@@
 We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers and other concurrent programs.
-Using \whoop and \corral we analysed 10 modules available in the Linux distribution.
+Using \whoop and \corral we analysed 10 drivers available in the Linux distribution.
 
@@
 \begin{itemize}
-\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux kernel modules and accelerating a plugged-in precise bug-finder.
-\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux kernel modules.
+\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux drivers and accelerating a plugged-in precise bug-finder.
+\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux drivers.
 \item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
-\item We analyse 10 Linux kernel modules, including various drivers and a filesystem.
+\item We analyse 10 Linux drivers, including various drivers and a filesystem.
 \end{itemize}

### 2015-03-21 staticanalysis.tex
many updates to refocus in drivers

@@
-After the original program has been translated to the abstract Boogie program, the core \whoop tool is invoked, which performs \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules.
+After the original program has been translated to the abstract Boogie program, the core \whoop tool is invoked, which performs \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux drivers.
 The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
@@
 
-We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets.
-Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module.
-Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one).
+We now formalise our approach to verifying data race freedom in a driver by statically analysing its locksets.
+Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver.
+Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one).
 Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
@@
 \label{theorem:locksets}
-For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
+For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
 \end{theorem}
@@
 dynamic lock creation).
-The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing modules for the Linux kernel.
+The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing drivers for the Linux kernel.
 
@@
 Inherently, recursion does not work with inlining.
-To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision.
+To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux drivers in a scalable fashion, but without destroying precision.
 
@@
 
-To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various modules.
-This is an ongoing manual effort: the more kernel modules we study, the more domain-specific properties we discover, which we then use for increased precision.
+To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various drivers.
+This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then use for increased precision.
 

### 2015-03-21 translation.tex
fix
many updates to refocus in drivers

@@
-The main aim of this phase is to take an input Linux kernel module written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
+The main aim of this phase is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
 Boogie and \corral).
@@
 
-First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given module, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points.
-This function acts as the main entry point to the module after the latter has been registered with the kernel.
+First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given driver, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points.
+This function acts as the main entry point to the driver after the latter has been registered with the kernel.
 This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{staticanalysis} and~\ref{bugfinding}).
@@
 The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource.
-Notice that functional calls (e.g.
+Notice that function calls (e.g.
 to lock and unlock a mutex) are preserved in the translation.

### 2015-03-21 whoop.tex
many updates to refocus in drivers

@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug-finder.
+We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux drivers for data races and (ii) accelerate a plugged-in bug-finder.
 
@@
 Figure~\ref{fig:whoop} depicts the \whoop infrastructure.
-Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model.
+Initially, \whoop accepts a driver written in C, together with a Linux environmental model.
 The environmental model (i.e.
-stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed.
-Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+stub Linux header files) is required to ``close'' the driver and allow it to be translated and subsequently analysed.
+Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
 Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program.

### 2015-04-17 bugfinding.tex
Working on the Corral section.

@@
-\whoop is a sound but potentially imprecise static data race analyser.
-In our experience, developers appreciate soundness but do not want to get overwhelmed by countless false positive reports.
-Towards this we allow an external precise bug-finder to be plugged in the \whoop infrastructure.
-The race-related information generated by \whoop can then be used to accelerate the bug-finder a bug-finder for concurrent programs that can be easily plugged in our infrastructure.
-Currently we only support \corral~\cite{lal2012corral}, a state-of-the-art bug-finder used in Microsoft to analyse Windows device drivers.
-We chose \corral because it is open-source and because it accepts Boogie IVL, similar to \whoop.
-Our technique, though, should be able to accelerate any arbitrary bug-finder for concurrent programs.
+\whoop is a sound but potentially imprecise static data race analyser.  While
+developers do appreciate soundness, it is more important to ensure they are not
+overwhelmed by countless reports of false bugs.  Hence, to improve precision, we
+enable an external precise bug-finder for concurrent programs to be plugged into
+the \whoop infrastructure. The race-related information generated by \whoop can
+then be used to accelerate the bug-finder.  Currently, we support
+\corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by
+Microsoft to analyse Windows device drivers. \corral accepts Boogie IVL and
+hence it was easy to integerate into our tool flow. Our technique, though, is
+general and capable of accelerating any arbitrary bug-finder for concurrent
+programs.
 
@@
 
-\corral is a verifier for Boogie IVL that uses SMT solvers (e.g.
-Z3 or CVC4) underneath to reason about program behaviour and reports a counterexample if an assertion violation is found.
-\corral attempts to prove reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analysing a complex concurrent program.
-Towards this, \corral performs two abstractions: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that impact the generated verification conditions; and (ii) stratified inlining, where it attempts to identify a minimal unrolling of loops and recursion that also impact the generated verification conditions.
+\corral is a symbolic bounded verifier for Boogie IVL that uses SMT solvers
+(e.g., Z3 or CVC4) to statically reason about program behaviours. It checks for
+violations of provided assertions, and reports a precise counterexample if an
+assertion violation is found. \corral performs bounded exploration of a
+concurrent program in two steps. First, given a bound on the number of allowed
+context-switches, the concurrent program is appropriately \emph{sequentialized},
+and the generated sequential version preserves reachable states of the original
+concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.  Then, \corral
+attempts to prove bounded (in terms of the number of loop iterations and
+recursion depth) sequential reachability of a bug in a goal-directed, lazy
+fashion to postpone state space explosion when analysing a large program.  It
+performs two abstractions to achieve this: (i) variable abstraction, where it
+attempts to identify a minimal set of shared variables that have to be precisely
+tracked in order to discharge all assertions; and (ii) stratified inlining,
+where it attempts to inline procedures on-demand as they are required for
+proving program assertions.
 
-Because \corral performs bounded model checking, it is inherently unsound, in other words it can miss real bugs.
-Assuming a precise environmental model, though, \corral will only report true bugs.
-\whoop takes advantage of this precision to report only feasible data races.
+Because \corral performs bounded verification both in terms of the number of
+allowed context switches and loop/recursion unrollings, it is inherently
+unsound; in other words, it can miss real bugs. On the other hand, \corral is
+precise, and assuming a precise model of the program environment it will only
+report true bugs. \whoop takes advantage of this precision to report only
+feasible data races.
 
@@
 
-By default, \corral instruments a \texttt{yield} statement (which denotes a context-switch) before each \emph{visible operation} (i.e.
-synchronisation operations and shared memory accesses) of a concurrently running thread.
-It then sequentialises the concurrent program with a schedule that explores all possible interleavings up to a pre-defined bound.
+By default, \corral instruments a \texttt{yield} statement (which denotes a
+context-switch) before each \emph{visible operation} (i.e. synchronisation
+operations and shared memory accesses) of a concurrently running thread. It then
+sequentialises the concurrent program with a schedule that explores all possible
+interleavings up to a pre-defined bound.
 

### 2015-04-17 bugfinding.tex
Done for now with the Corral section.

@@
 By default, \corral instruments a \texttt{yield} statement (which denotes a
-context-switch) before each \emph{visible operation} (i.e. synchronisation
-operations and shared memory accesses) of a concurrently running thread. It then
-sequentialises the concurrent program with a schedule that explores all possible
-interleavings up to a pre-defined bound.
+potential nondeterministic context-switch) before each \emph{visible operation}
+(i.e., synchronisation operations and shared memory accesses) of every thread.
+It then leverages sequentialization to explore all possible thread interleavings
+up to a pre-defined bound.
 
-Our idea for accelerating \corral is simple: we simplify the sequentialisation that \corral performs by instrumenting \texttt{yield} statements only in shared memory accesses that \whoop reported as potentially racy.
-This effectively acts as a partial order reduction that can significantly reduce the interleavings that \corral needs to explore, and thus increase performance and scalability.
+Our approach to accelerating \corral is simple and yet effective: we simplify
+the sequentialisation that \corral performs by instrumenting \texttt{yield}
+statements only at shared memory accesses that \whoop reported as potentially
+racy. This effectively acts as a partial order reduction that can significantly
+reduce the number of interleavings that \corral needs to explore, and thus
+increase performance and scalability.
 
-We now show that our partial order reduction is sound, in other words it does not miss any more bugs than the bugs that \corral would miss even without taking advantage of the \whoop race-related information.
-Let \corral run an $n$ amount of entry point pairs with yields in all visible operations.
-Let this configuration be known as $C$.
-Each entry point pair is running in a separate \corral process.
-\corral will miss all bugs that are triggered by more than 3 bugs running concurrently and all bugs that are triggered by investigating a larger than $r$ recursion bound and larger than $k$ context switch bound.
-Let \corral then run the same $n$ amount of pairs, but this time with all \texttt{yield} statements in non-racy shared memory accesses (according to \whoop) removed.
-Let this configuration be known as $W+C$.
-Because these shared memory accesses cannot race, performing a \texttt{yield} will not introduce any new program behaviours.
-Thus, \corral will find the same bugs using the configuration $W+C$, as using the configuration $C$.
+\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
+about entry point pairs, or could we prove it just for any pair?}
+We now show that our partial order reduction is sound: it does not miss any more
+bugs than what \corral would miss even without taking advantage of the \whoop
+race-related information. Let \corral run an $n$ amount of entry point pairs
+with yields in all visible operations. Let this configuration be known as $C$.
+Each entry point pair is running in a separate \corral process. \corral will
+miss all bugs that are triggered by more than 3 bugs running concurrently and
+all bugs that are triggered by investigating a larger than $r$ recursion bound
+and larger than $k$ context switch bound. Let \corral then run the same $n$
+amount of pairs, but this time with all \texttt{yield} statements in non-racy
+shared memory accesses (according to \whoop) removed. Let this configuration be
+known as $W+C$. Because these shared memory accesses cannot race, performing a
+\texttt{yield} will not introduce any new program behaviours. Thus, \corral will
+find the same bugs using the configuration $W+C$, as using the configuration
+$C$.
 
@@
 
-As aforementioned, \corral instruments by default a \texttt{yield} statement in each visible operation.
-Using information from \whoop, though, we can avoid instrumenting context-switches in memory accesses that do not race.
-To achieve this, we use the \corral command line option \texttt{/cooperative}, which disables \texttt{yield} instrumentation in \corral, and then we instrument context-switches inside \whoop before sending these Boogie files to \corral.
-We have implemented two different levels of context-switch instrumentation granularity: \emph{\yieldcoarse} and \emph{\yieldmr}.
+\ZRComment{I think this paragraph is not needed. You already talk about
+how \corral by default instruments everything in the beginning of your previous
+section. The command line switch /cooperative is I think an implementation
+detail we do not have to mention. More generally, I would merge this
+subsection with the previous one and I would keep the descriptions of
+granularities.}
+As aforementioned, \corral instruments by default a \texttt{yield} statement in
+each visible operation. Using information from \whoop, though, we can avoid
+instrumenting context-switches in memory accesses that do not race. To achieve
+this, we use the \corral command line option \texttt{/cooperative}, which
+disables \texttt{yield} instrumentation in \corral, and then we instrument
+context-switches inside \whoop before sending these Boogie files to \corral. We
+have implemented two different levels of context-switch instrumentation
+granularity: \emph{\yieldcoarse} and \emph{\yieldmr}.
 
-\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop finds that a pair of entry points cannot race, then it will only instrument context-switches in synchronisation points of this pair; else if \whoop finds that a pair of entry points can race, then it will instrument context-switches in all visible operations of this pair.
+\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
+finds that a pair of entry points cannot race, then it will only instrument
+context-switches in synchronisation points of this pair; else if \whoop finds
+that a pair of entry points can race, then it will instrument context-switches
+in all visible operations of this pair.
 
-\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches in all synchronisation points, but only instruments context-switches in memory regions that can potentially race (regardless if a pair has not been fully verified as race-free).
-In our experiments (see Section~\ref{evaluation}), \yieldmr performs significantly faster than \yieldcoarse.
+\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
+in all synchronisation points, but only instruments context-switches in memory
+regions that can potentially race (regardless if a pair has not been fully
+verified as race-free). In our experiments (see Section~\ref{evaluation}),
+\yieldmr performs significantly faster than \yieldcoarse.
 
@@
 
-We use a simple, but effective, race checking instrumentation with \corral.
-This instrumentation, which is based on previous work~\cite{}, can be enabled using a command line option and we use it as a feasibility checking for the potential data races that \whoop reports.
-Whenever there is a write access, e.g.
-\texttt{\$M.13 := 0}, we instrument the Boogie IVL code as follows:
-
+We employ a simple, but effective, encoding of data-race checks within \corral.
+For example, whenever there is a write access to a shared variable such as
+\texttt{\$M.13 := 0}, we instrument the Boogie IVL code to check for write-write
+races as follows:
+%
 \begin{boogie}
-$M.13 := 0;
-yield;
-assert $M.13 == 0;
+$M.13 := 0;        // original write
+yield;             // allow for a context-switch
+assert $M.13 == 0; // check written value
+\end{boogie}
+%
+Likewise, for a read access to a shared variable such as \texttt{\$p0 :=
+\$M.13}, we instrument the Boogie IVL code to check for read-write races as
+follows:
+\ZRComment{Why do you need this auxiliary variable?}
+%
+\begin{boogie}
+$p0 := $M.13;       // original read
+yield;              // allow for a context-switch
+$rv0 := $M.13;      // $rv0 is auxiliary variable
+assert $p0 == $rv0; // check read value
 \end{boogie}
 
-Likewise, for a read access, e.g. \texttt{\$p0 := \$M.13}, we instrument the Boogie IVL code as follows:
-
-\begin{boogie}
-$p0 := $M.13;
-yield;
-$rv0 := $M.13; // $rv0 is an auxiliary variable
-assert $p0 == $rv0;
-\end{boogie}
-
-Our instrumentation tolerates benign races, as it will never report a write-write race, if the two write accesses update the shared memory location with the same value.
+Note that our instrumentation conveniently tolerates most benign races, as it
+does not report a write-write race if the two write accesses update the shared
+memory location with the same value.  We introduce the instrumentation as a
+precise feasibility check for the potential data races that \whoop reports.

### 2015-04-20 intro.tex
updates in abstract and intro

@@
 Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system.
-\PDComment{maybe also mention that races in C lead to undefined behaviour and potential issues with compilers?} Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
+According to the C language standard, data races also lead to undefined behaviour, which can be the source of various compilation and optimisation issues \PDComment{which is the best cite here?}.
+Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
 Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs.
@@
 
-In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux drivers written in the C language and (ii) accelerating external bug-finders that can easily plug in our toolchain.
-\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers (e.g.
-device drivers and filesystems).
+In this paper, we present \whoop, a new approach that (i) statically analyses the concurrent behaviour of Linux device drivers written in the C language and (ii) uses the results from the static analysis to accelerate bug-finding.
+\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers.
 The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
@@
 
-We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers and other concurrent programs.
-Using \whoop and \corral we analysed 10 drivers available in the Linux distribution.
+We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers, as well as other concurrent programs.
+Using \whoop and \corral we analysed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
 

### 2015-04-20 intro.tex
small update in intro
restructuring of core section

@@
 
-In this paper, we present \whoop, a new approach that (i) statically analyses the concurrent behaviour of Linux device drivers written in the C language and (ii) uses the results from the static analysis to accelerate bug-finding.
+In this paper, we present \whoop, a new automatic approach that (i) statically analyses the concurrent behaviour of Linux device drivers written in the C language and (ii) exploits the results from the static analysis to accelerate bug-finding.
 \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers.
@@
 \item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
-\item We analyse 10 Linux drivers, including various drivers and a filesystem.
+\item We analyse \sizeOfBenchmarks Linux drivers to show that \whoop is efficient in finding data races.
 \end{itemize}

### 2015-04-20 whoop.tex
restructuring of core section

@@
-We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux drivers for data races and (ii) accelerate a plugged-in bug-finder.
+We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyse Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.
 
@@
 
-Figure~\ref{fig:whoop} depicts the \whoop infrastructure.
-Initially, \whoop accepts a driver written in C, together with a Linux environmental model.
-The environmental model (i.e.
-stub Linux header files) is required to ``close'' the driver and allow it to be translated and subsequently analysed.
-Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
-Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program.
-Finally, \whoop uses information from the static analysis phase to accelerate bug-finding with \corral (see Section~\ref{bugfinding}).
+The key idea behind our novel static analysis is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
+The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
+
+Figure~\ref{fig:whoop} depicts the \whoop toolchain.
+Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analysed.
+Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Next, \whoop performs \emph{static pairwise lockset analysis} to detect all potential data races in the abstract program.
+The proof sketch behind our static lockset analysis is given in Section~\ref{whoop:proof}.
+The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}.
+The summarisation that we perform to achieve scalability is presented in Section~\ref{whoop:summarisation}.
+After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{whoop:bugfinding})
+
+Finally, we discuss optimisations and assumptions in Section~\ref{whoop:optimisations}, and limitations of our technique in Section~\ref{whoop:limitations}.
+
+\subsection{Translation to Boogie IVL}
+\label{whoop:translation}
+
+The main aim of this phase is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
+Boogie and \corral).
+To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.
+
+First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given driver, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points.
+This function acts as the main entry point to the driver after the latter has been registered with the kernel.
+This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{whoop:method} and~\ref{whoop:bugfinding}).
+The output is a file that contains information regarding the entry points and a file with the rewritten program.
+
+Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
+SMACK then translates the LLVM-IR to Boogie IVL.
+SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL.
+To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.
+
+\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
+static void ep1(struct net_device *dev) {
+  struct shared *tp = netdev_priv(dev);
+  mutex_lock(&tp->wk.mutex);
+  tp->resource = 4;
+  mutex_unlock(&tp->wk.mutex);
+}
+\end{lstlisting}
+
+Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK.
+The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource.
+Notice that function calls (e.g.
+to lock and unlock a mutex) are preserved in the translation.
+
+\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
+procedure ep1(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    $p0 := dev;
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call mutex_lock($p3);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    $M.0[$p4] := 4;
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call mutex_unlock($p6);
+    return;
+}
+\end{boogie}
+
+\subsection{Static Lockset Analysis Proof Sketch}
+\label{whoop:proof}
+
+We now formalise our approach to verifying data race freedom in a driver by statically analysing its locksets.
+Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver.
+Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one).
+Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
+The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
+
+For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
+Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution.
+In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
+
+\begin{theorem}
+\label{theorem:locksets}
+For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
+\end{theorem}
+
+A sketch of how the above theorem can be proved is as follows.
+Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$.
+By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
+By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$.
+As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
+
+\subsection{Instrumentation and Analysis}
+\label{whoop:method}
+
+The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentialising the abstract Boogie program.
+During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g.
+the following call to \texttt{mutex\_lock()}:
+
+\begin{boogie}
+call mutex_lock($p3);
+\end{boogie}
+
+Will be rewritten to:
+
+\begin{boogie}
+const {:lock} unique lock$0: int;
+
+call mutex_lock(lock$0);
+\end{boogie}
+
+If \whoop cannot infer a lock it will exit with a warning.
+A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks).
+It is relatively straightforward to enhance \whoop, though, with new locking primitives.
+\whoop will also exit with a warning if it detects ``improper'' use of locks (e.g.
+dynamic lock creation).
+The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing drivers for the Linux kernel.
+
+When the lock rewriting finishes, \whoop traverses the Boogie IVL code, uses the information extracted by Chauffeur and separates each independent entry point call-graph from the rest of the program.
+To achieve this, it duplicates helper functions and renames them accordingly: e.g.
+for entry point \texttt{ep1}, a helper function \texttt{foo} will be renamed to \texttt{foo\$ep1}.
+This allows \whoop to perform entry point sensitive instrumentation and analysis.
+Although the program can potentially become much larger, this does not affect the analysis as discussed later.
+
+Next, \whoop instruments the program with global variables for lockset analysis and race checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:
+
+\begin{boogie}
+var {:current_lockset} lock$0_in_CLS_$ep1: bool;
+var {:lockset} lock$0_in_LS_$M.0_$ep1: bool;
+var {:access_checking} WRITTEN_$M.0_$ep1: bool;
+var {:access_checking} READ_$M.0_$ep1: bool;
+\end{boogie}
+
+\whoop also instruments a global watchdog variable per memory region that is common to all entry points:
+
+\begin{boogie}
+const {:watchdog} WATCHED_ACCESS_$M.0: int;
+\end{boogie}
+
+The above instrumented variables represent the follows: \texttt{lock\$0\_in\_CLS\_\$ep1} is the current lockset (CLS) of \texttt{ep1}; \texttt{WATCHED\_ACCESS\_\$M.0} is an \emph{unconstrained} constant representing the offset to memory region \texttt{\$M.0} that should be checked for races; \texttt{lock\$0\_in\_LS\_\$M.0\_\$ep1} is the lockset for \texttt{\$M.0}; \texttt{WRITTEN\_\$M.0\_\$ep1} is the write-set for \texttt{\$M.0}; and \texttt{READ\_\$M.0\_\$ep1} is the read-set for \texttt{\$M.0}.
+Verification involves proving that two entry points cannot race at the watched offset of every memory region.
+The arbitrary watched offset implies that every offset of each memory region is race-free.
+Watchdog race checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
+
+\whoop proceeds by instrumenting each entry point call-graph for lockset analysis by replacing each call to a locking or unlocking function with a \whoop-specific lock and unlock functions.
+This means that Example~\ref{fig:smack_translation} will become:
+
+\begin{boogie}
+procedure ep1(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    $p0 := dev;
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, true);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    $M.0[$p4] := 4;
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, false);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_UPDATE\_CLS\_\$ep1()} is a special function that updates the CLS global variable to either true or false, if the lock is held or released by \texttt{ep1} accordingly:
+
+\begin{boogie}
+procedure {:inline 1} _UPDATE_CLS_$ep1(lock: int,
+    isLocked: bool);
+  modifies lock$0_in_CLS_$ep1;
+
+implementation {:inline 1} _UPDATE_CLS_$ep1(
+    lock: int, isLocked: bool) {
+  _UPDATE:
+    lock$0_in_CLS_$ep1 := (if lock == lock$0 then
+      isLocked else lock$0_in_CLS_$ep1);
+  return;
+}
+\end{boogie}
+
+Next, \whoop replaces every write and read access in each entry point call-graph with a \whoop-specific function:
+
+\begin{boogie}
+procedure ep1(dev: int) modifies $M.0; {
+  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
+  $bb0:
+    $p0 := dev;
+    $p1 := $pa($p0, 32, 1);
+    $p2 := $pa($pa($p1, 0, 24), 12, 1);
+    $p3 := $pa($pa($p2, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, true);
+    $p4 := $pa($pa($p1, 0, 24), 4, 1);
+    call _WRITE_LS_$M.0_$ep1($p4);
+    $p5 := $pa($pa($p1, 0, 24), 12, 1);
+    $p6 := $pa($pa($p5, 0, 12), 0, 1);
+    call _UPDATE_CLS_$ep1(lock$0, false);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_WRITE\_LS\_\$M.0\_\$ep1()} is a special function that computes the intersection of the CLS and the lockset of \texttt{\$M.0} (see Section~\ref{bg:lockset}).
+It also updates the write-set of \texttt{\$M.0} to true, because there was a write access by \texttt{ep1}:
+
+\begin{boogie}
+procedure {:inline 1} _WRITE_LS_$M.0_$ep1(
+    ptr: int);
+  modifies lock$0_in_LS_$M.0_$ep1,
+    WRITTEN_$M.0_$ep1;
+
+implementation {:inline 1} _WRITE_LS_$M.0_$ep1(
+    ptr: int) {
+  _WRITE:
+    goto anon1_Then, anon1_Else;
+
+  anon1_Then:
+    assume {:partition} WATCHED_ACCESS_$M.0 ==
+      ptr && DEVICE_IS_REGISTERED_$ep;
+    lock$0_in_LS_$M.0_$ep := lock$0_in_CLS_$ep && lock$0_in_LS_$M.0_$ep;
+    WRITTEN_$M.0_$ep := true;
+    return;
+  anon1_Else:
+    assume {:partition} !(WATCHED_ACCESS_$M.0 ==
+      ptr && DEVICE_IS_REGISTERED_$ep);
+    return;
+}
+\end{boogie}
+
+The \texttt{\_READ\_LS\_\$M.0\_\$ep1()} updates the lockset of \texttt{\$M.0} and the read-set accordingly for a read access.
+
+\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
+This is achieved by creating a checker function that calls the two instrumented entry points and then performs race checking assertions.
+This process repeats until all possible pairs of entry points have been sequentialised.
+To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}).
+This over-approximates any effects from all the unmodeled threads on the shared state of the pair.
+Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.
+
+After each checker function calls a pair of entry points, it asserts that each shared memory location is protected by at least one common lock (see Section~\ref{bg:lockset}).
+The assertion check is triggered iff the memory location was accessed by both entry points and at least one of the accesses was a write.
+We do not check for read-read races, which are inherently benign.
+The following is an example of a checker function for entry points \texttt{ep1} and \texttt{ep2}:
+
+\begin{boogie}
+procedure check$ep1$ep2(dev: int);
+  requires !lock$0_in_CLS_$ep1;
+  requires !lock$0_in_CLS_$ep2;
+  requires lock$0_in_LS_$M.0_$ep1;
+  requires lock$0_in_LS_$M.0_$ep2;
+  requires !WRITTEN_$M.0_$ep1;
+  requires !WRITTEN_$M.0_$ep2;
+  requires !READ_$M.0_$ep1;
+  requires !READ_$M.0_$ep2;
+  modifies lock$0_in_CLS_$ep1, lock$0_in_CLS_$ep2,
+    lock$0_in_LS_$M.0_$ep1, lock$0_in_LS_$M.0_$ep2,
+    WRITTEN_ $M.0_$ep1,WRITTEN_$M.0_$ep2,
+    READ_$M.0_$ep1, READ_$M.0_$ep2;
+
+implementation check$ep1$ep2(dev: int) {
+  _CHECK:
+    call ep1(dev);
+    call ep2(dev);
+    assert (WRITTEN_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
+      || (WRITTEN_$M.0_$ep1 && READ_$M.0_$ep2)
+      || (READ_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
+      ==> lock$0_in_LS_$M.0_$ep1 &&
+      lock$0_in_LS_$M.0_$ep2;
+    return;
+}
+\end{boogie}
+
+Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover (Boogie supports Z3~\cite{de2008z3} and CVC4~\cite{barrett2011cvc4}).
+Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
+To improve usability \whoop has a built-in error reporter that matches Boogie counterexamples to source code.
+The following is an example of a reported error:
+
+\begin{lstlisting}
+foo.c: error: potential write-write race:
+
+write by entry point ep1, foo.c:36:2
+  shared->resource = 4;
+
+write by entry point ep2, foo.c:52:2
+  shared->resource = 5;
+\end{lstlisting}
+
+\subsection{Watchdog Summarisation}
+\label{whoop:summarisation}
+
+Early on during the development of \whoop, we realised that scalability would be a serious issue.
+When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory.
+One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory.
+Another reason was that some entry points recursively call helper functions.
+Inherently, recursion does not work with inlining.
+To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux drivers in a scalable fashion, but without destroying precision.
+
+Watchdog summarisation uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
+
+The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{whoop:method}).
+The key idea behind watchdog summarisation is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
+This reduces the false positives from reporting errors if only e.g.
+a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not.
+An example of watchdog-based candidate post-conditions is the following:
+
+\begin{boogie}
+procedure foo$ep1(dev: int);
+  ...
+  ensures _b$ls$entrypoint1$0
+    ==> WATCHED_ACCESS_$M.0 == dev + 4
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ls$entrypoint1$1
+    ==> WATCHED_ACCESS_$M.0 == dev + 8
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$0
+    ==> WATCHED_ACCESS_$M.0 == dev + 4
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$1
+    ==> WATCHED_ACCESS_$M.0 == dev + 8
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ...
+\end{boogie}
+
+To generate the candidate invariants, \whoop performs an intra-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
+To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
+For example the previous \texttt{\$M.0} candidates would now be:
+
+\begin{boogie}
+procedure foo$ep1(dev: int);
+  ...
+  ensures _b$ls$entrypoint1$0
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ls$entrypoint1$1
+    ==> lock$0_in_LS_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$0
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ensures _b$ac$entrypoint1$1
+    ==> !WRITTEN_$M.0_$entrypoint1;
+  ...
+\end{boogie}
+
+\subsection{Accelerating Bug-Finding}
+\label{whoop:bugfinding}
+
+\whoop is a sound but potentially imprecise static data race analyser.  While
+developers do appreciate soundness, it is more important to ensure they are not
+overwhelmed by countless reports of false bugs.  Hence, to improve precision, we
+enable an external precise bug-finder for concurrent programs to be plugged into
+the \whoop infrastructure. The race-related information generated by \whoop can
+then be used to accelerate the bug-finder.  Currently, we support
+\corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by
+Microsoft to analyse Windows device drivers. \corral accepts Boogie IVL and
+hence it was easy to integerate into our tool flow. Our technique, though, is
+general and capable of accelerating any arbitrary bug-finder for concurrent
+programs.
+
+\subsubsection{Bug-Finding with \corral}
+\label{bf:corral}
+
+\corral is a symbolic bounded verifier for Boogie IVL that uses SMT solvers
+(e.g., Z3 or CVC4) to statically reason about program behaviours. It checks for
+violations of provided assertions, and reports a precise counterexample if an
+assertion violation is found. \corral performs bounded exploration of a
+concurrent program in two steps. First, given a bound on the number of allowed
+context-switches, the concurrent program is appropriately \emph{sequentialized},
+and the generated sequential version preserves reachable states of the original
+concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.  Then, \corral
+attempts to prove bounded (in terms of the number of loop iterations and
+recursion depth) sequential reachability of a bug in a goal-directed, lazy
+fashion to postpone state space explosion when analysing a large program.  It
+performs two abstractions to achieve this: (i) variable abstraction, where it
+attempts to identify a minimal set of shared variables that have to be precisely
+tracked in order to discharge all assertions; and (ii) stratified inlining,
+where it attempts to inline procedures on-demand as they are required for
+proving program assertions.
+
+Because \corral performs bounded verification both in terms of the number of
+allowed context switches and loop/recursion unrollings, it is inherently
+unsound; in other words, it can miss real bugs. On the other hand, \corral is
+precise, and assuming a precise model of the program environment it will only
+report true bugs. \whoop takes advantage of this precision to report only
+feasible data races.
+
+\subsubsection{Sound Partial Order Reduction}
+\label{bf:reduction}
+
+By default, \corral instruments a \texttt{yield} statement (which denotes a
+potential nondeterministic context-switch) before each \emph{visible operation}
+(i.e., synchronisation operations and shared memory accesses) of every thread.
+It then leverages sequentialization to explore all possible thread interleavings
+up to a pre-defined bound.
+
+Our approach to accelerating \corral is simple and yet effective: we simplify
+the sequentialisation that \corral performs by instrumenting \texttt{yield}
+statements only at shared memory accesses that \whoop reported as potentially
+racy. This effectively acts as a partial order reduction that can significantly
+reduce the number of interleavings that \corral needs to explore, and thus
+increase performance and scalability.
+
+\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
+about entry point pairs, or could we prove it just for any pair?}
+We now show that our partial order reduction is sound: it does not miss any more
+bugs than what \corral would miss even without taking advantage of the \whoop
+race-related information. Let \corral run an $n$ amount of entry point pairs
+with yields in all visible operations. Let this configuration be known as $C$.
+Each entry point pair is running in a separate \corral process. \corral will
+miss all bugs that are triggered by more than 3 bugs running concurrently and
+all bugs that are triggered by investigating a larger than $r$ recursion bound
+and larger than $k$ context switch bound. Let \corral then run the same $n$
+amount of pairs, but this time with all \texttt{yield} statements in non-racy
+shared memory accesses (according to \whoop) removed. Let this configuration be
+known as $W+C$. Because these shared memory accesses cannot race, performing a
+\texttt{yield} will not introduce any new program behaviours. Thus, \corral will
+find the same bugs using the configuration $W+C$, as using the configuration
+$C$.
+
+\subsubsection{Granularity of Context-Switches}
+\label{bf:granularity}
+
+\ZRComment{I think this paragraph is not needed. You already talk about
+how \corral by default instruments everything in the beginning of your previous
+section. The command line switch /cooperative is I think an implementation
+detail we do not have to mention. More generally, I would merge this
+subsection with the previous one and I would keep the descriptions of
+granularities.}
+As aforementioned, \corral instruments by default a \texttt{yield} statement in
+each visible operation. Using information from \whoop, though, we can avoid
+instrumenting context-switches in memory accesses that do not race. To achieve
+this, we use the \corral command line option \texttt{/cooperative}, which
+disables \texttt{yield} instrumentation in \corral, and then we instrument
+context-switches inside \whoop before sending these Boogie files to \corral. We
+have implemented two different levels of context-switch instrumentation
+granularity: \emph{\yieldcoarse} and \emph{\yieldmr}.
+
+\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
+finds that a pair of entry points cannot race, then it will only instrument
+context-switches in synchronisation points of this pair; else if \whoop finds
+that a pair of entry points can race, then it will instrument context-switches
+in all visible operations of this pair.
+
+\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
+in all synchronisation points, but only instruments context-switches in memory
+regions that can potentially race (regardless if a pair has not been fully
+verified as race-free). In our experiments (see Section~\ref{evaluation}),
+\yieldmr performs significantly faster than \yieldcoarse.
+
+\subsubsection{Race Checking Instrumentation}
+\label{bf:racechecking}
+
+We employ a simple, but effective, encoding of data-race checks within \corral.
+For example, whenever there is a write access to a shared variable such as
+\texttt{\$M.13 := 0}, we instrument the Boogie IVL code to check for write-write
+races as follows:
+%
+\begin{boogie}
+$M.13 := 0;        // original write
+yield;             // allow for a context-switch
+assert $M.13 == 0; // check written value
+\end{boogie}
+%
+Likewise, for a read access to a shared variable such as \texttt{\$p0 :=
+\$M.13}, we instrument the Boogie IVL code to check for read-write races as
+follows:
+\ZRComment{Why do you need this auxiliary variable?}
+%
+\begin{boogie}
+$p0 := $M.13;       // original read
+yield;              // allow for a context-switch
+$rv0 := $M.13;      // $rv0 is auxiliary variable
+assert $p0 == $rv0; // check read value
+\end{boogie}
+
+Note that our instrumentation conveniently tolerates most benign races, as it
+does not report a write-write race if the two write accesses update the shared
+memory location with the same value.  We introduce the instrumentation as a
+precise feasibility check for the potential data races that \whoop reports.
+
+\subsection{Optimisations and Assumptions}
+\label{whoop:optimisations}
+
+A common practise for statically analysing software is to ``close'' the environment by abstracting away the low level implementation details.
+Any abstract environmental model, though, can and will ultimately result in decreased precision and thus false positives.
+This is one of the main limitations of statically analysing real life software that use complex external APIs.
+In our case, we developed a simple environmental model for the Linux kernel that consists of stub header files.
+Because we only focus on lockset-based properties during the race-checking phase of \whoop, we can get away with abstracting a lot of the underlying functionality.
+Our model, though, can still result in a lot of false positives.
+
+To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various drivers.
+This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then use for increased precision.
+
+In our experience, one of the most effective precision optimisations is to enrich \whoop with information regarding kernel imposed locks.
+The Linux kernel is able to serialise a number of entry point calls, which thus can never run concurrently with each other.
+An an example, a large number of networking and ethernet entry points are mutually serialised with RTNL, a network-specific kernel lock.
+\whoop will not create pairs for entry points that cannot run concurrently.
+
+Another optimisation, is to soundly reduce the checked memory locations.
+If a memory location is accessed only by one entry point in the pair, then it is safe to not check it.
+This can potentially decrease the amount of generated candidate invariants and speed up the verification process.
+
+Modelling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux.
+For example, the \texttt{register\_netdev()} function registers a network driver with the kernel.
+Before this function is called, the kernel is not able to invoke any of the driver's entry points.
+Modelling this functionality can reduce false positives when race checking accesses before the driver's registration.
+
+We also optimise the use of function pointers.
+LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site.
+To increase precision, Chauffeur analyses the AST of the original program and attempts to infer context-sensitive function pointer use information.
+This information is the used by \whoop to increase precision.
+To be sound, this optimisation is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.
+
+Finally, we assume that the formal parameters of an entry point do not alias, and thus do not race.
+This is a potentially unsound feature that can be turned off using a command line option.
+In our experience so far, though, we have not missed any data races because of this assumption.
+We also assume that our environmental model and domain-specific knowledge are correct and do not cause any bugs to be missed.
+
+\subsection{Limitations}
+\label{whoop:limitations}
+
+\whoop inherits all limitations of lockset analysis (as discussed in Section~\ref{bg:lockset}): a violation of the locking discipline does not always correspond to a real error (e.g.
+\whoop does not reasons about lockfree data structures in a special way, it just over-approximates their read accesses); and we do not check for dynamically created locks (or for locks that are provided by external libraries).
+Because we perform two thread reduction, we also over-approximate the read accesses, which is another source of false positives.
+
+Environmental modelling is another limitation of \whoop.
+As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.
+We argue that further increasing the precision of our models is orthogonal to the contributions of this paper.
+Moreover, despite unavoidable imprecisions during the sound static data race analysis, \whoop can use the reported information to significantly speedup precise bug-finders for concurrent programs, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.
+
+\PDComment{Corral limitations ...}

### 2015-04-20 background.tex
various updates

@@
 Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5.
-A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.
+A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behavior.
 
@@
 
-Lock-based synchronisation is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}.
+Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}.
 Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency.
 On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread.
-Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we focus on locks.
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, in this work we focus on locks.
 
@@
 Let $E_1$ and $E_2$ be two events during a program's concurrent execution.
-If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism (e.g.
+If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronization mechanism (e.g.
 locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
@@
 
-An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behaviour of concurrent programs.
+An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behavior of concurrent programs.
 Lockset analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution.
@@
 Lockset analysis starts by creating one lockset for each thread $T$ in the program, called current lockset of $T$ or $CLS_T$, and one lockset for each shared variable $v$, called lockset of $v$ or $LS_v$.
-In this example we analyse only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$).
+In this example we analyze only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$).
 Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\varnothing$.
@@
 can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
-Furthermore, as a classic dynamic analyser, Eraser's bug-finding ability is limited to the execution paths that the tool explores.
+Furthermore, as a classic dynamic analyzer, Eraser's bug-finding ability is limited to the execution paths that the tool explores.
 To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

### 2015-04-20 conclusion.tex
various updates

@@
-In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis.
-In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving.
+In this paper, we presented \whoop, a novel infrastructure for (i) analyzing the concurrent behavior of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis.
+In the heart of our approach, we combine static lock set analysis with sound sequentialization and automated theorem proving.
 Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.

### 2015-04-20 evaluation.tex
many updates to refocus in drivers
restructuring of core section
various updates

@@
 
-We evaluate our methodology against 10 modules taken from the Linux kernel (currently a mix of latest and older distributions). \PDComment{Say something about the benchmarks}
+We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel. \PDComment{Say something more about the benchmarks}
 
@@
 Our experiments use the default \corral context switch bound \texttt{/k:2}.
-A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialised program.
+A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialized program.
 Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings.
@@
 
-%When \corral on its own tries to analyse the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyse a large number of memory regions.
-For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyse.
+%When \corral on its own tries to analyze the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyze a large number of memory regions.
+For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyze.
 The same entry point pair is found race-free by \whoop.
-Using this information in \corral with the \yieldmr instrumentation, the pair takes less than a second to analyse.
+Using this information in \corral with the \yieldmr instrumentation, the pair takes less than a second to analyze.
 The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.

### 2015-04-20 intro.tex
various updates

--- 2015-04-20 intro.tex
@@
 For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.
-All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.
+All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (such as locks) are in place to control concurrent access to shared resources.
 
 Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system.
-According to the C language standard, data races also lead to undefined behaviour, which can be the source of various compilation and optimisation issues \PDComment{which is the best cite here?}.
-Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
-Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs.
+According to the C language standard, data races also lead to undefined behavior, which can be the source of various compilation and optimization issues \PDComment{which is the best cite here?}.
+Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
+Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyze the program behavior of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs.
 The few of these techniques that attempt to deal with data races are usually either \emph{unsound} (i.e.
@@
 
-In this paper, we present \whoop, a new approach that (i) statically analyses the concurrent behaviour of Linux device drivers written in the C language and (ii) uses the results from the static analysis to accelerate bug-finding.
+In this paper, we present \whoop, a new automatic approach that (i) statically analyses the concurrent behavior of Linux device drivers written in the C language and (ii) exploits the results from the static analysis to accelerate bug-finding.
 \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers.
@@
 
-We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers, as well as other concurrent programs.
-Using \whoop and \corral we analysed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
+We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
+Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
 
@@
 \begin{itemize}
-\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux drivers and accelerating a plugged-in precise bug-finder.
+\item We present \whoop, a practical tool for analyzing the concurrent behavior of Linux drivers and accelerating a plugged-in precise bug-finder.
 \item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux drivers.
 \item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
-\item We analyse 10 Linux drivers, including various drivers and a filesystem.
+\item We analyze \sizeOfBenchmarks Linux drivers to show that \whoop is efficient in finding data races.
 \end{itemize}

### 2015-04-20 related.tex
various updates

@@
 Using static analysis to detect data races is a promising alternative to dynamic concurrency bug finding, which is inherently limited by scalability and execution path coverage issues.
-Notable previous work includes the static analysers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}.
+Notable previous work includes the static analyzers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}.
 However, all these three tools heavily rely on user annotations, and thus require a lot of manual effort that limits their applicability.
@@
 
-Most related to our work are the static lock set analysers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
+Most related to our work are the static lock set analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
 Both tools, though, have significant limitations.
-The authors of RELAY found 5022 warnings when analysing the Linux kernel, with only 25 of them being true data races.
+The authors of RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races.
 To limit these false positives, RELAY employs post-analysis unsound filters, but these can also filter out true races.

### 2015-04-20 whoop.tex
various updates

@@
-We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyse Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.
+We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyze Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.
 
@@
 Figure~\ref{fig:whoop} depicts the \whoop toolchain.
-Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analysed.
+Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analyzed.
 Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
@@
 The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}.
-The summarisation that we perform to achieve scalability is presented in Section~\ref{whoop:summarisation}.
+The summarization that we perform to achieve scalability is presented in Section~\ref{whoop:summarization}.
 After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{whoop:bugfinding})
 
-Finally, we discuss optimisations and assumptions in Section~\ref{whoop:optimisations}, and limitations of our technique in Section~\ref{whoop:limitations}.
+Finally, we discuss optimizations and assumptions in Section~\ref{whoop:optimizations}, and limitations of our technique in Section~\ref{whoop:limitations}.
 
@@
 Boogie and \corral).
-To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.
+To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur\footnote{https://github.com/mc-imperial/chauffeur}, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.
 
@@
 
-We now formalise our approach to verifying data race freedom in a driver by statically analysing its locksets.
+We now formalize our approach to verifying data race freedom in a driver by statically analyzing its locksets.
 Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver.
 Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one).
-Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.
+Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronization bugs that arise from careless fine-grained locking.
 The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
@@
 
-The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentialising the abstract Boogie program.
+The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentializing the abstract Boogie program.
 During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g.
@@
 dynamic lock creation).
-The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing drivers for the Linux kernel.
+The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
 
@@
 
-\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair.
+\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialization}, which combines the two arbitrary threads in a single sequential pair.
 This is achieved by creating a checker function that calls the two instrumented entry points and then performs race checking assertions.
-This process repeats until all possible pairs of entry points have been sequentialised.
+This process repeats until all possible pairs of entry points have been sequentialized.
 To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}).
@@
 
-Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover (Boogie supports Z3~\cite{de2008z3} and CVC4~\cite{barrett2011cvc4}).
+Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to the Z3~\cite{de2008z3} theorem prover.
 Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
-To improve usability \whoop has a built-in error reporter that matches Boogie counterexamples to source code.
+To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
 The following is an example of a reported error:
@@
 
-\subsection{Watchdog Summarisation}
-\label{whoop:summarisation}
+\subsection{Watchdog Summarization}
+\label{whoop:summarization}
 
-Early on during the development of \whoop, we realised that scalability would be a serious issue.
+Early on during the development of \whoop, we realized that scalability would be a serious issue.
 When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory.
@@
 Inherently, recursion does not work with inlining.
-To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux drivers in a scalable fashion, but without destroying precision.
+To tackle these issues we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze Linux drivers in a scalable fashion, but without destroying precision.
 
-Watchdog summarisation uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
+Watchdog summarization uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
 
 The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{whoop:method}).
-The key idea behind watchdog summarisation is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
+The key idea behind watchdog summarization is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
 This reduces the false positives from reporting errors if only e.g.
@@
 
-\whoop is a sound but potentially imprecise static data race analyser.  While
-developers do appreciate soundness, it is more important to ensure they are not
-overwhelmed by countless reports of false bugs.  Hence, to improve precision, we
-enable an external precise bug-finder for concurrent programs to be plugged into
-the \whoop infrastructure. The race-related information generated by \whoop can
-then be used to accelerate the bug-finder.  Currently, we support
-\corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by
-Microsoft to analyse Windows device drivers. \corral accepts Boogie IVL and
-hence it was easy to integerate into our tool flow. Our technique, though, is
-general and capable of accelerating any arbitrary bug-finder for concurrent
-programs.
+\whoop is a sound but potentially imprecise static data race analyzer.
+While developers do appreciate soundness, it is more important to ensure they are not overwhelmed by countless reports of false bugs.
+Hence, to improve precision, we enable an external precise bug-finder for concurrent programs to be plugged into \whoop.
+The race-related information generated by \whoop can then be used to accelerate the bug-finder.
+Currently, we support \corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers.
+\corral accepts Boogie IVL and hence it was easy to integrate into our tool flow.
+Our technique, though, is general and capable of accelerating any arbitrary bug-finder for concurrent programs.
 
-\subsubsection{Bug-Finding with \corral}
-\label{bf:corral}
+\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors.
+It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found.
+\corral performs bounded exploration of a concurrent program in two steps.
+First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.
+Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program.
+It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.
 
-\corral is a symbolic bounded verifier for Boogie IVL that uses SMT solvers
-(e.g., Z3 or CVC4) to statically reason about program behaviours. It checks for
-violations of provided assertions, and reports a precise counterexample if an
-assertion violation is found. \corral performs bounded exploration of a
-concurrent program in two steps. First, given a bound on the number of allowed
-context-switches, the concurrent program is appropriately \emph{sequentialized},
-and the generated sequential version preserves reachable states of the original
-concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.  Then, \corral
-attempts to prove bounded (in terms of the number of loop iterations and
-recursion depth) sequential reachability of a bug in a goal-directed, lazy
-fashion to postpone state space explosion when analysing a large program.  It
-performs two abstractions to achieve this: (i) variable abstraction, where it
-attempts to identify a minimal set of shared variables that have to be precisely
-tracked in order to discharge all assertions; and (ii) stratified inlining,
-where it attempts to inline procedures on-demand as they are required for
-proving program assertions.
-
-Because \corral performs bounded verification both in terms of the number of
-allowed context switches and loop/recursion unrollings, it is inherently
-unsound; in other words, it can miss real bugs. On the other hand, \corral is
-precise, and assuming a precise model of the program environment it will only
-report true bugs. \whoop takes advantage of this precision to report only
-feasible data races.
+Because \corral performs bounded verification both in terms of the number of allowed context switches and loop/recursion unrollings, it is inherently unsound; in other words, it can miss real bugs.
+On the other hand, \corral is precise, and assuming a precise model of the program environment it will only report true bugs.
+\whoop takes advantage of this precision to report only feasible data races.
 
@@
 potential nondeterministic context-switch) before each \emph{visible operation}
-(i.e., synchronisation operations and shared memory accesses) of every thread.
+(i.e., synchronization operations and shared memory accesses) of every thread.
 It then leverages sequentialization to explore all possible thread interleavings
@@
 Our approach to accelerating \corral is simple and yet effective: we simplify
-the sequentialisation that \corral performs by instrumenting \texttt{yield}
+the sequentialization that \corral performs by instrumenting \texttt{yield}
 statements only at shared memory accesses that \whoop reported as potentially
@@
 known as $W+C$. Because these shared memory accesses cannot race, performing a
-\texttt{yield} will not introduce any new program behaviours. Thus, \corral will
+\texttt{yield} will not introduce any new program behaviors. Thus, \corral will
 find the same bugs using the configuration $W+C$, as using the configuration
@@
 finds that a pair of entry points cannot race, then it will only instrument
-context-switches in synchronisation points of this pair; else if \whoop finds
+context-switches in synchronization points of this pair; else if \whoop finds
 that a pair of entry points can race, then it will instrument context-switches
@@
 \yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
-in all synchronisation points, but only instruments context-switches in memory
+in all synchronization points, but only instruments context-switches in memory
 regions that can potentially race (regardless if a pair has not been fully
@@
 
-\subsection{Optimisations and Assumptions}
-\label{whoop:optimisations}
+\subsection{Optimizations and Assumptions}
+\label{whoop:optimizations}
 
-A common practise for statically analysing software is to ``close'' the environment by abstracting away the low level implementation details.
+A common practice for statically analyzing software is to ``close'' the environment by abstracting away the low level implementation details.
 Any abstract environmental model, though, can and will ultimately result in decreased precision and thus false positives.
-This is one of the main limitations of statically analysing real life software that use complex external APIs.
+This is one of the main limitations of statically analyzing real life software that use complex external APIs.
 In our case, we developed a simple environmental model for the Linux kernel that consists of stub header files.
@@
 
-In our experience, one of the most effective precision optimisations is to enrich \whoop with information regarding kernel imposed locks.
-The Linux kernel is able to serialise a number of entry point calls, which thus can never run concurrently with each other.
-An an example, a large number of networking and ethernet entry points are mutually serialised with RTNL, a network-specific kernel lock.
+In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding kernel imposed locks.
+The Linux kernel is able to serialize a number of entry point calls, which thus can never run concurrently with each other.
+An an example, a large number of networking and ethernet entry points are mutually serialized with RTNL, a network-specific kernel lock.
 \whoop will not create pairs for entry points that cannot run concurrently.
 
-Another optimisation, is to soundly reduce the checked memory locations.
+Another optimization, is to soundly reduce the checked memory locations.
 If a memory location is accessed only by one entry point in the pair, then it is safe to not check it.
@@
 
-Modelling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux.
+Modeling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux.
 For example, the \texttt{register\_netdev()} function registers a network driver with the kernel.
 Before this function is called, the kernel is not able to invoke any of the driver's entry points.
-Modelling this functionality can reduce false positives when race checking accesses before the driver's registration.
+Modeling this functionality can reduce false positives when race checking accesses before the driver's registration.
 
-We also optimise the use of function pointers.
+We also optimize the use of function pointers.
 LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site.
@@
 This information is the used by \whoop to increase precision.
-To be sound, this optimisation is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.
+To be sound, this optimization is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.
 
@@
 
-Environmental modelling is another limitation of \whoop.
+Environmental modeling is another limitation of \whoop.
 As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.

### 2015-04-20 whoop.tex
updates

@@
 
-\subsubsection{Sound Partial Order Reduction}
+\subsubsection{Sound Partial-Order Reduction}
 \label{bf:reduction}
@@
 statements only at shared memory accesses that \whoop reported as potentially
-racy. This effectively acts as a partial order reduction that can significantly
+racy. This effectively acts as a partial-order reduction that can significantly
 reduce the number of interleavings that \corral needs to explore, and thus
@@
 \ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
-about entry point pairs, or could we prove it just for any pair?}
-We now show that our partial order reduction is sound: it does not miss any more
+about entry point pairs, or could we prove it just for any pair?}\PDComment{Lets also see what Ally thinks about this part.}
+We now show that our partial-order reduction is sound: it does not miss any more
 bugs than what \corral would miss even without taking advantage of the \whoop
@@
 amount of pairs, but this time with all \texttt{yield} statements in non-racy
-shared memory accesses (according to \whoop) removed. Let this configuration be
+shared memory accesses that are removed according to information from \whoop. Let this configuration be
 known as $W+C$. Because these shared memory accesses cannot race, performing a
@@
 
-\subsubsection{Granularity of Context-Switches}
-\label{bf:granularity}
-
-\ZRComment{I think this paragraph is not needed. You already talk about
-how \corral by default instruments everything in the beginning of your previous
-section. The command line switch /cooperative is I think an implementation
-detail we do not have to mention. More generally, I would merge this
-subsection with the previous one and I would keep the descriptions of
-granularities.}
-As aforementioned, \corral instruments by default a \texttt{yield} statement in
-each visible operation. Using information from \whoop, though, we can avoid
-instrumenting context-switches in memory accesses that do not race. To achieve
-this, we use the \corral command line option \texttt{/cooperative}, which
-disables \texttt{yield} instrumentation in \corral, and then we instrument
-context-switches inside \whoop before sending these Boogie files to \corral. We
-have implemented two different levels of context-switch instrumentation
-granularity: \emph{\yieldcoarse} and \emph{\yieldmr}.
+We have implemented two different levels of context-switch instrumentation granularity:
 
@@
 We employ a simple, but effective, encoding of data-race checks within \corral.
-For example, whenever there is a write access to a shared variable such as
+For example, whenever there is a write access to a shared variable, such as
 \texttt{\$M.13 := 0}, we instrument the Boogie IVL code to check for write-write
@@
 %
-Likewise, for a read access to a shared variable such as \texttt{\$p0 :=
+Likewise, for a read access to a shared variable, such as \texttt{\$p0 :=
 \$M.13}, we instrument the Boogie IVL code to check for read-write races as
 follows:
-\ZRComment{Why do you need this auxiliary variable?}
 %
@@
 yield;               // allow for a context-switch
-$rv0 := $M.13;      // $rv0 is auxiliary variable
-assert $p0 == $rv0; // check read value
+assert $p0 == $M.13; // check read value
 \end{boogie}
 
-Note that our instrumentation conveniently tolerates most benign races, as it
-does not report a write-write race if the two write accesses update the shared
-memory location with the same value.  We introduce the instrumentation as a
-precise feasibility check for the potential data races that \whoop reports.
+Note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
+We introduce the instrumentation as a precise feasibility check for the potential data races that \whoop reports.
 

### 2015-04-20 whoop.tex
updated toolchain figure

@@
 
-The main aim of this phase is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
+The aim of this phase (see Figure~\ref{fig:whoop} -- A) is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
 Boogie and \corral).
@@
 
-The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentializing the abstract Boogie program.
+The core \whoop tool (see Figure~\ref{fig:whoop} -- B) is responsible for parsing, instrumenting and sequentializing the abstract Boogie program.
 During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g.
@@
 Currently, we support \corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers.
-\corral accepts Boogie IVL and hence it was easy to integrate into our tool flow.
+\corral accepts Boogie IVL and hence it was easy to integrate into our toolchain (see Figure~\ref{fig:whoop} -- C).
 Our technique, though, is general and capable of accelerating any arbitrary bug-finder for concurrent programs.

### 2015-04-23 intro.tex
slight update in abstract/intro

@@
 
-In this paper, we present \whoop, a new automatic approach that (i) statically analyses the concurrent behavior of Linux device drivers written in the C language and (ii) exploits the results from the static analysis to accelerate bug-finding.
+In this paper, we present \whoop, a new fully automatic approach that (i) statically analyses Linux device drivers written in the C language for data races and (ii) exploits the race-related results from the static analysis to accelerate bug-finding.
 \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers.

### 2015-04-23 intro.tex
slight update in abstract/intro

--- 2015-04-20 intro.tex
@@
 
-In this paper, we present \whoop, a new automatic approach that (i) statically analyses the concurrent behavior of Linux device drivers written in the C language and (ii) exploits the results from the static analysis to accelerate bug-finding.
-\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers.
+In this paper, we present \whoop, a new fully automatic approach that (i) statically analyses device drivers written in the C language for data races and (ii) exploits the race-related results from the static analysis to accelerate bug-finding.
+\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in device drivers.
 The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
@@
 
+\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g.
+Windows) as well as other concurrent programs with similar structure (e.g.
+filesystems).
+
 To summarize, our contributions are as follows:
 \begin{itemize}
-\item We present \whoop, a practical tool for analyzing the concurrent behavior of Linux drivers and accelerating a plugged-in precise bug-finder.
-\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux drivers.
+\item We present \whoop, a new fully automatic approach for analyzing device drivers for data races and accelerating bug-finding.
+\item We discuss static pairwise lockset analysis,  a sound and scalable technique for automatically verifying the absence of data races in device drivers.
 \item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.

### 2015-04-25 evaluation.tex
fixed caption in table 1
updates to the evaluation

@@
 \newcommand{\colspacing}{\hspace{1.8em}}
+\begin{table}[t]
+\small
+\centering
+\setlength{\tabcolsep}{0.3em}
+\caption{Program statistics and data race detection results from running \whoop and \corral on our benchmarks. Corral uses a context switch bound of 2.}
+\label{tab:stats}
+\begin{tabular}{l rrr rr r}
+\centering
+\input{experiments/tables/statistics.tex}
+\end{tabular}
+\end{table}
+
 \begin{table*}[t]
@@
 \centering
+\caption{Runtime comparison with different yield instrumentation granularities and context-switch bounds (csb).}
+\label{tab:results}
 \begin{tabular}{l rrr rrr rr r r}
 \centering
-\input{experiments/tables/granularity.tex}
+\input{experiments/tables/scalability.tex}
 \end{tabular}
-\caption{Data race analysis and scalability results.}
-\label{tab:races}
 \end{table*}
+
+%\begin{table*}[t]
+%\small
+%\centering
+%\caption{Program statistics, analysis and scalability results. Corral uses a context switch bound of 2.}
+%\label{tab:races}
+%\begin{tabular}{l rrr rrr rr r r}
+%\centering
+%\input{experiments/tables/granularity.tex}
+%\end{tabular}
+%\end{table*}
+%
+%\begin{table*}[t]
+%\small
+%\centering
+%\caption{Analysis and scalability results. Corral uses a context switch bound of 5 and 9, respectively.}
+%\label{tab:races2}
+%\begin{tabular}{l rr r r rr r r}
+%\centering
+%\input{experiments/tables/granularity2.tex}
+%\end{tabular}
+%\end{table*}
 
@@
 
-We conducted all experiments on ... We use the following software versions: Clang/LLVM 3.5; SMACK 1.4; Z3 4.1; \corral commit 3aa62d7425b5; Boogie commit d6a7f2bd79c9.
-
-All reported times are in seconds and averages of N executions.
-We report: lines of C code (LoC); number of entry point pairs (\#Pairs); memory regions in the program (\#MRs); the number of racy memory regions that \whoop found (\#Racy MRs); the number of race reports that \whoop generated (\#Race Reports); the number of races that \corral found (same in all configurations).
+All experiments were performed on a ... machine running ..., LLVM 3.5; SMACK 1.5; Z3 4.3.2; Boogie rev. 4192 and \corral rev. 534.
 
@@
 
+Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs that \whoop identified (\#Racy Pairs); number of racy memory regions that \whoop reported (\#Racy MRs); and number of races that \corral discovered (\#Races Found).
+
 \subsection{Context-Switch Granularity of Yield Instrumentation}
@@
 
-We evaluate the scalability of our acceleration technique with respect to our benchmark sets.
+We evaluate the scalability of our acceleration technique with respect to our benchmark sets (Table~\ref{tab:results}).
+All reported times are in seconds and averages of three runs.
+\corral was configured with a time budget of 3000 seconds per tool invocation (one for each pair of entry points); \whoop did not use a timeout.
 
-Our experiments use the default \corral context switch bound \texttt{/k:2}.
+Our experiments use the \corral context switch bounds \texttt{/k:2}, \texttt{/k:5} and \texttt{/k:9}.
 A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialized program.
 Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings.
-\PDComment{We should take a few measurements with higher k -- no need to be in the table -- just to verify this argument.}
 
@@
 
-In Table~\ref{tab:races}, we can notice that \whoop is much faster than \corral. For example, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
+In Table~\ref{tab:results}, we can notice that \whoop is much faster than \corral. For example, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
+
+% We did not discover any data races with larger context switch bounds.
 

### 2015-04-26 intro.tex
update to text

@@
 \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in device drivers.
-The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
+The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race-checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
 The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
@@
 \item We discuss static pairwise lockset analysis,  a sound and scalable technique for automatically verifying the absence of data races in device drivers.
-\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
-\item We analyze \sizeOfBenchmarks Linux drivers to show that \whoop is efficient in finding data races.
+\item We show that we can achieve a sound partial-order reduction using our technique and significantly speedup \corral, a state-of-the-art bug-finder.
+\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient in race-checking and accelerating \corral.
 \end{itemize}

### 2015-04-26 whoop.tex
fix typo
update to text

@@
 
-The key idea behind our novel static analysis is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
+The key idea behind our novel static analysis is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race-checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
 The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
@@
 
-Next, \whoop instruments the program with global variables for lockset analysis and race checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:
+Next, \whoop instruments the program with global variables for lockset analysis and race-checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:
 
@@
 The arbitrary watched offset implies that every offset of each memory region is race-free.
-Watchdog race checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
+Watchdog race-checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
 
@@
 \whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialization}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by creating a checker function that calls the two instrumented entry points and then performs race checking assertions.
+This is achieved by creating a checker function that calls the two instrumented entry points and then performs race-checking assertions.
 This process repeats until all possible pairs of entry points have been sequentialized.
@@
 
-The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{whoop:method}).
+The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race-checking (see Section~\ref{whoop:method}).
 The key idea behind watchdog summarization is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
@@
 
-\subsubsection{Race Checking Instrumentation}
+\subsubsection{Race-Checking Instrumentation}
 \label{bf:racechecking}
@@
 
-Another optimization, is to soundly reduce the checked memory locations.
+Another optimization is to soundly reduce the checked memory locations.
 If a memory location is accessed only by one entry point in the pair, then it is safe to not check it.
@@
 Before this function is called, the kernel is not able to invoke any of the driver's entry points.
-Modeling this functionality can reduce false positives when race checking accesses before the driver's registration.
+Modeling this functionality can reduce false positives when race-checking accesses before the driver's registration.
 

### 2015-04-26 evaluation.tex
various updates

@@
 \centering
-\caption{Runtime comparison with different yield instrumentation granularities and context-switch bounds (csb).}
+\caption{Runtime comparison with different yield instrumentation granularities and context-switch bounds.}
 \label{tab:results}
@@
 
-%\begin{table*}[t]
-%\small
-%\centering
-%\caption{Program statistics, analysis and scalability results. Corral uses a context switch bound of 2.}
-%\label{tab:races}
-%\begin{tabular}{l rrr rrr rr r r}
-%\centering
-%\input{experiments/tables/granularity.tex}
-%\end{tabular}
-%\end{table*}
-%
-%\begin{table*}[t]
-%\small
-%\centering
-%\caption{Analysis and scalability results. Corral uses a context switch bound of 5 and 9, respectively.}
-%\label{tab:races2}
-%\begin{tabular}{l rr r r rr r r}
-%\centering
-%\input{experiments/tables/granularity2.tex}
-%\end{tabular}
-%\end{table*}
+We performed experiments to validate the usefulness of the \whoop approach (see Section~\ref{whoop}).
+We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution\footnote{https://www.kernel.org}.
+We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds, to show that \whoop can efficiently accelerate bug-finding with \corral.
 
-\subsection{Experimental Setup}
-\label{eval:setup}
+\noindent
+\textbf{Experimental Setup}\xspace\xspace All experiments were performed on a ... machine running ..., LLVM 3.5; SMACK 1.5; Z3 4.3.2; Boogie rev. 4192 and \corral rev. 534.
 
-All experiments were performed on a ... machine running ..., LLVM 3.5; SMACK 1.5; Z3 4.3.2; Boogie rev. 4192 and \corral rev. 534.
+\noindent
+\textbf{Benchmarks}\xspace\xspace We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
+To thoroughly test \whoop against the Linux driver APIs, we chose drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+We manually modeled the environment for all these drivers.
+This process required more than a month of work.
+We also focused on choosing non-trivial drivers (e.g.
+large number of entry points, deep nesting of entry point helper functions, complex use of locking, use of function pointers, etc).
 
-\subsection{Benchmarks}
-\label{eval:benchmarks}
+\noindent
+\textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs that \whoop identified (\#Racy Pairs); number of racy memory regions that \whoop reported (\#Racy MRs); and number of races that \corral discovered (\#Races Found) using a context-switch bound of 2.
 
-We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel. \PDComment{Say something more about the benchmarks}
+We can see that \whoop reports many more races than \corral.
+This is expected, though, as \whoop over-approximates the original driver for soundness and scalability.
+This means that if a race exists, \whoop will report it (assuming tool and model correctness).
+On the other hand, \corral performs bounded checking, which means that it will report a race only under pre-defined bounds.
+A common problem for \corral when analyzing drivers is loop coverage.
+To deal with this we enable the over-approximation technique described in~\cite{lal2014powering}.
+This can potentially lead \corral to report false positives, but we have not seen this in practice.
 
-Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs that \whoop identified (\#Racy Pairs); number of racy memory regions that \whoop reported (\#Racy MRs); and number of races that \corral discovered (\#Races Found).
+We found the following data race in the generic\_nvram driver: during the \texttt{llseek} file operation, the driver is accessing the file offset (\texttt{file->f\_pos}) without using a lock.
+When we investigated another similar driver that uses the same API, we saw that it was protecting the same offset access with a mutex.
+This made us suspicious that the generic\_nvram driver might have a real race.
+We filed a bug report, but did not manage to get a response yet.
 
-\subsection{Context-Switch Granularity of Yield Instrumentation}
-\label{eval:granularity}
+A common case of races that we discovered had to do with incrementing a counter without using a lock.
+This might happen for performance, but even benign races might lead to serious issues, especially if a compiler tries to perform optimizations assuming no races exist.
 
-We evaluate the scalability of our acceleration technique with respect to our benchmark sets (Table~\ref{tab:results}).
+It is worthwhile to mention that we did not discover any additional data races with larger context switch bounds.
+This might mean that most races in these drivers require only a small amount of context-switches to manifest, or that \corral run out of resources before discovering the deeper bug.
+
+\noindent
+\textbf{Granularity of Context-Switches}\xspace\xspace Table~\ref{tab:results} presents runtime results from running \whoop, \corral and \whoop + \corral.
 All reported times are in seconds and averages of three runs.
-\corral was configured with a time budget of 3000 seconds per tool invocation (one for each pair of entry points); \whoop did not use a timeout.
+\corral was configured with a time budget of 7200 seconds (T.O.
+denotes a timeout) and a context-switch bound (csb) of 2, 5 and 9.
+By default, \corral instruments context-switches (i.e.
+\texttt{yield}) in all visible operations (denoted by \yieldall in the table).
+\whoop + \corral, uses two context-switch instrumentation granularities (see Section~\ref{whoop:bugfinding}): \yieldcoarse and \yieldmr.
 
-Our experiments use the \corral context switch bounds \texttt{/k:2}, \texttt{/k:5} and \texttt{/k:9}.
-A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialized program.
-Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings.
+A higher context switch bound results into deeper interleavings being explored and thus a larger sequentialized program.
+Intuitively, this means that we see even greater speedups using information from \whoop, when exploring deeper interleavings.
+
+In Table~\ref{tab:results}, we can notice that \whoop is much faster than \corral. For example, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
 
@@
 Using \whoop + \corral, though, will inherently discard spurious data races (assuming a precise environmental model).
-
-In Table~\ref{tab:results}, we can notice that \whoop is much faster than \corral. For example, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
-
-% We did not discover any data races with larger context switch bounds.
 

### 2015-04-26 whoop.tex
various updates

@@
 Watchdog summarization uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
+The reason we need loop invariants is that we over-approximate loops using loop-cutting (also available inside Boogie).
 
@@
 
-\subsubsection{Sound Partial-Order Reduction}
-\label{bf:reduction}
-
-By default, \corral instruments a \texttt{yield} statement (which denotes a
-potential nondeterministic context-switch) before each \emph{visible operation}
-(i.e., synchronization operations and shared memory accesses) of every thread.
-It then leverages sequentialization to explore all possible thread interleavings
-up to a pre-defined bound.
+\noindent
+\textbf{Sound Partial-Order Reduction}\xspace\xspace By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before each \emph{visible operation} (i.e., synchronization operations and shared memory accesses) of every thread.
+It then leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
 
@@
 
-\subsubsection{Race-Checking Instrumentation}
-\label{bf:racechecking}
-
-We employ a simple, but effective, encoding of data-race checks within \corral.
+\noindent
+\textbf{Race-Checking Instrumentation}\xspace\xspace We employ a simple, but effective, encoding of data-race checks within \corral.
 For example, whenever there is a write access to a shared variable, such as

### 2015-04-26 evaluation.tex
various updates

@@
 \setlength{\tabcolsep}{0.3em}
-\caption{Program statistics and data race detection results from running \whoop and \corral on our benchmarks. Corral uses a context switch bound of 2.}
+\caption{Program statistics and race-checking results from running \whoop and \corral on our benchmarks.}
 \label{tab:stats}
@@
 \noindent
-\textbf{Granularity of Context-Switches}\xspace\xspace Table~\ref{tab:results} presents runtime results from running \whoop, \corral and \whoop + \corral.
+\textbf{Granularity of Context-Switches}\xspace\xspace Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks.
 All reported times are in seconds and averages of three runs.
-\corral was configured with a time budget of 7200 seconds (T.O.
-denotes a timeout) and a context-switch bound (csb) of 2, 5 and 9.
+\corral was given a time budget of 2 hours (T.O.
+denotes a tool timeout) and a context-switch bound (csb) of 2, 5 and 9.
 By default, \corral instruments context-switches (i.e.
 \texttt{yield}) in all visible operations (denoted by \yieldall in the table).
-\whoop + \corral, uses two context-switch instrumentation granularities (see Section~\ref{whoop:bugfinding}): \yieldcoarse and \yieldmr.
+\whoop + \corral, instead, uses two different context-switch instrumentation granularities (see Section~\ref{whoop:bugfinding}): \yieldcoarse and \yieldmr.
 

### 2015-04-26 evaluation.tex
various updates

@@
 
-We can see that \whoop reports many more races than \corral.
-This is expected, though, as \whoop over-approximates the original driver for soundness and scalability.
+We can see that, in general, \whoop reports (many) more races than \corral.
+This is expected, though, as \whoop over-approximates the original driver for soundness.
 This means that if a race exists, \whoop will report it (assuming tool and model correctness).
 On the other hand, \corral performs bounded checking, which means that it will report a race only under pre-defined bounds.
-A common problem for \corral when analyzing drivers is loop coverage.
-To deal with this we enable the over-approximation technique described in~\cite{lal2014powering}.
+It is worthwhile to note that a common problem for \corral when analyzing drivers is loop coverage.
+To deal with this we enable the over-approximation described in~\cite{lal2014powering}.
 This can potentially lead \corral to report false positives, but we have not seen this in practice.
 
-We found the following data race in the generic\_nvram driver: during the \texttt{llseek} file operation, the driver is accessing the file offset (\texttt{file->f\_pos}) without using a lock.
-When we investigated another similar driver that uses the same API, we saw that it was protecting the same offset access with a mutex.
-This made us suspicious that the generic\_nvram driver might have a real race.
+Most of the data races that \whoop and \corral found can be classified in two cases.
+The first case had to do with incrementing a global counter without using a lock.
+The reason behind this might be performance, but even benign races can lead to undefined behavior according to the C standard.
+The second case had to do with an entry point modifying a field of an object (either global or passed as parameter) without using a lock.
+This can easily lead to a data race if another concurrently running entry point accesses the same field of the same object.
+As an example of the second case, we discovered the following data race in the generic\_nvram char driver: during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} (\texttt{file} is passed as a parameter to the entry point) without first acquiring a lock.
+This can lead to a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
+When we investigated another char driver that uses the same API, we saw that it was protecting the offset access in the same entry point with a mutex.
+This made us suspicious that the race we found in generic\_nvram is real.
 We filed a bug report, but did not manage to get a response yet.
-
-A common case of races that we discovered had to do with incrementing a counter without using a lock.
-This might happen for performance, but even benign races might lead to serious issues, especially if a compiler tries to perform optimizations assuming no races exist.
 
@@
 
-In Table~\ref{tab:results}, we can notice that \whoop is much faster than \corral. For example, \whoop analyses the dtlk char driver in 6.51 seconds, whereas \corral requires 1287.05 seconds.
+We can notice that \whoop is significantly faster than \corral.
+This is expected as \whoop achieves scalability using over-approximation and summarization techniques.
+This allows \whoop to perform especially well in complex drivers such as the r8169 ethernet driver: although \corral ...
+and timeouts with a csb of 9, \whoop manages to analyze the driver in 228.1 seconds.
+We believe that the reason behind this is that r8169 has many loops and uses deeply-nested recursion in some entry points, which \corral might not be able to handle efficiently.
 

### 2015-04-26 evaluation.tex
various updates

@@
 To thoroughly test \whoop against the Linux driver APIs, we chose drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
-We manually modeled the environment for all these drivers.
-This process required more than a month of work.
-We also focused on choosing non-trivial drivers (e.g.
+Further, we focused on choosing non-trivial drivers (e.g.
 large number of entry points, deep nesting of entry point helper functions, complex use of locking, use of function pointers, etc).
+We manually modeled the environment for all these drivers, a process that required more than a month of work.
 
@@
 We can see that, in general, \whoop reports (many) more races than \corral.
-This is expected, though, as \whoop over-approximates the original driver for soundness.
-This means that if a race exists, \whoop will report it (assuming tool and model correctness).
-On the other hand, \corral performs bounded checking, which means that it will report a race only under pre-defined bounds.
-It is worthwhile to note that a common problem for \corral when analyzing drivers is loop coverage.
+This is expected because \whoop over-approximates the original driver for soundness: if a race exists, the tool will report it (assuming tool and model correctness), but \whoop can also report false races.
+On the other hand, \corral reports only real races, but does that under pre-defined bounds.
+It is worthwhile to note that a common problem for \corral (especially when analyzing drivers) is loop coverage.
 To deal with this we enable the over-approximation described in~\cite{lal2014powering}.

### 2015-04-26 whoop.tex
various updates

@@
 const {:lock} unique lock$0: int;
-
 call mutex_lock(lock$0);
@@
 foo.c: error: potential write-write race:
-
   write by entry point ep1, foo.c:36:2
     shared->resource = 4;
-
   write by entry point ep2, foo.c:52:2
@@
 procedure foo$ep1(dev: int);
-  ...
   ensures _b$ls$entrypoint1$0

### 2015-04-26 conclusion.tex
some fixes

@@
-In this paper, we presented \whoop, a novel infrastructure for (i) analyzing the concurrent behavior of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis.
-In the heart of our approach, we combine static lock set analysis with sound sequentialization and automated theorem proving.
-Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.
+In this paper, we present \whoop, a new fully automatic approach for (i) finding data races in device drivers and (ii) accelerating bug-finding using race-related information.
+In the heart of our approach, we combine static lockset analysis with sound sequentialization and automated theorem proving.
+Our lockset analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.
 

### 2015-04-26 conclusion.tex
fix

@@
-In this paper, we present \whoop, a new fully automatic approach for (i) finding data races in device drivers and (ii) accelerating bug-finding using race-related information.
+In this paper, we presented \whoop, a new fully automatic approach for (i) finding data races in device drivers and (ii) accelerating bug-finding using race-related information.
 In the heart of our approach, we combine static lockset analysis with sound sequentialization and automated theorem proving.

### 2015-04-26 whoop.tex
update

@@
 \noindent
-\textbf{Race-Checking Instrumentation}\xspace\xspace We employ a simple, but effective, encoding of data-race checks within \corral.
+\textbf{Race-Checking Instrumentation}\xspace\xspace In this paper, we treat races as bugs. To detect them, we employ a simple, but effective, encoding of data race checks within \corral.
 For example, whenever there is a write access to a shared variable, such as

### 2015-04-28 evaluation.tex
update

@@
 We can see that, in general, \whoop reports (many) more races than \corral.
-This is expected because \whoop over-approximates the original driver for soundness: if a race exists, the tool will report it (assuming tool and model correctness), but \whoop can also report false races.
+This is expected because \whoop over-approximates the original driver for soundness: if a race exists, the tool will report it (assuming tool and model correctness), but \whoop can, and does, report false races.
 On the other hand, \corral reports only real races, but does that under pre-defined bounds.
@@
 All reported times are in seconds and averages of three runs.
-\corral was given a time budget of 2 hours (T.O.
+\corral was given a time budget of 8 hours (T.O.
 denotes a tool timeout) and a context-switch bound (csb) of 2, 5 and 9.

### 2015-04-28 evaluation.tex
update in the text

@@
 This is expected because \whoop over-approximates the original driver for soundness: if a race exists, the tool will report it (assuming tool and model correctness), but \whoop can, and does, report false races.
-On the other hand, \corral reports only real races, but does that under pre-defined bounds.
-It is worthwhile to note that a common problem for \corral (especially when analyzing drivers) is loop coverage.
-To deal with this we enable the over-approximation described in~\cite{lal2014powering}.
-This can potentially lead \corral to report false positives, but we have not seen this in practice.
+On the contrary, \corral reports only real races, but does that under pre-defined bounds and thus can miss races.
 

### 2015-04-28 whoop.tex
update
update in the text

@@
 The race-related information generated by \whoop can then be used to accelerate the bug-finder.
-Currently, we support \corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers.
+Currently, we support \corral~\cite{lal2012corral, lal2014powering}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers.
 \corral accepts Boogie IVL and hence it was easy to integrate into our toolchain (see Figure~\ref{fig:whoop} -- C).
@@
 It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.
+\corral also employs sophisticated strategies for improving loop coverage, these are described in more detail in~\cite{lal2014powering}.
 
@@
 Note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
-We introduce the instrumentation as a precise feasibility check for the potential data races that \whoop reports.
+An advantage of this instrumentation is that it can be used as a precise feasibility check for the potential data races that \whoop reports.
 

### 2015-04-28 evaluation.tex
various updates in evaluation

@@
 \textbf{Benchmarks}\xspace\xspace We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
-To thoroughly test \whoop against the Linux driver APIs, we chose drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+We chose drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
 Further, we focused on choosing non-trivial drivers (e.g.
@@
 
-We can see that, in general, \whoop reports (many) more races than \corral.
+We can see that, in general, \whoop reports more races than \corral.
 This is expected because \whoop over-approximates the original driver for soundness: if a race exists, the tool will report it (assuming tool and model correctness), but \whoop can, and does, report false races.
-On the contrary, \corral reports only real races, but does that under pre-defined bounds and thus can miss races.
+On the contrary, \corral is precise, but can miss races as it explores the driver under pre-defined bounds.
+Further, a common problem for \corral is loop coverage.
+To tackle this we enable the build-in loop over-approximation described in~\cite{lal2014powering}.
+This can potentially lead \corral to report false positives, but we have not seen this in practice.
 

### 2015-05-04 background.tex
Changes to, and feedback on, the background.  Very first thoughts on how to restructure the Whoop section.

@@
 \label{bg:concurrency}
+
+\ADComment{I believe that \emph{responsiveness} is a key motivator for concurrency in device drivers: even if there is only a single core, we may wish for driver entry points of distinct processes to run concurrently so that no single process hogs the core.
+Can you please consider weaving this into what follows?}
 
@@
 \label{definition:datarace}
-A data race occurs when two distinct threads access a shared memory location, at least one of the accesses modifies the location, and there is no mechanism in place to prevent these accesses from being simultaneous.
+A data race occurs when two distinct threads access a shared memory location, at least one of the accesses modifies the location, at least one of the accesses is non-atomic, and there is no mechanism in place to prevent these accesses from being simultaneous.
 \end{definition}
 
-Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver.
+\ADComment{In the definition, I added the part about at least one of the accesses being non-atomic.
+I believe this is important in drivers, since there are cases where people use atomics.
+Could you somewhere briefly acknowledge, referring back to Definition~\ref{definition:datarace}, that we treat atomics soundly in whoop, in the sense that we do not regard them as providing any protection between threads?}
+
+\ADComment{It would be much more compelling, even if it takes more space, if you could find a small snippet from a real driver such that there is a data race say on a field of a struct.
+Could you replace the contrived example with such an example?}  Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver.
 The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}.
@@
 A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behavior.
+\ADComment{If you stick with this example, it would be worth pointing out that a write-write race is also possible.}
 
-\begin{lstlisting}[caption = Simple data race in a Linux driver, label = fig:data_race_example]
+\begin{lstlisting}[caption = Simple data race in a Linux driver \ADComment{I think the caption is misleading, since this is clearly not a driver}, label = fig:data_race_example]
 static void ep1 (struct shared *dev) {
@@
 Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}.
-Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency.
-On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread.
-Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, in this work we focus on locks.
+Coarse-grained locking can \ADComment{I removed the word ``severely''.
+My PhD supervisor always advised me to avoid using words that are subjective.} hurt performance, because it limits the opportunity for concurrency.
+On the other hand, fine-grained locking can easily lead to deadlocks.
+\ADComment{Removed description of deadlock, as it is well known.}
+\ADComment{Suggestion: expand the following a little to say what one of the lock-free synchronization methods in Linux is.
+Perhaps take this opportunity to refer back to the stuff about ``atomic'' I added in the data race definition.}  Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, in this work we focus on locks.
+\ADComment{Can we justify the focus on locks a bit more?
+E.g., are you able to survey, via grepping, what percentage of Linux drivers make use of locks (likely 100\%?) vs.\ what percentage also make use of lock-free mechanisms?}
 
@@
 \label{bg:happensbefore}
+
+\ADComment{I am unsure as to whether we need this background, but I'm undecided as I'm still getting on top of the rest of the material.}
 
@@
 An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behavior of concurrent programs.
-Lockset analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution.
+Lockset analysis was proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution.
 If that lockset ever becomes empty, the analysis reports a potential data race on that memory location.
-This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
+\ADComment{This could be better explained, I think.} This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
 
@@
 
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a thread of a concurrent program.
-Lockset analysis starts by creating one lockset for each thread $T$ in the program, called current lockset of $T$ or $CLS_T$, and one lockset for each shared variable $v$, called lockset of $v$ or $LS_v$.
-In this example we analyze only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$).
-Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\varnothing$.
-Likewise, $LS_A$ starts as the set of all existing locks in the program, which in this case is $\{M, N\}$.
-Lockset analysis then proceeds as follows:
+\ADComment{What is missing from the explanation, and from the example you give, is the idea that with a dynamic approach, you run the program and a thread operates in a truly concurrent environment.
+You thus only get to see what the thread did on the particular schedule that was followed.
+I also think you should use a two-threaded example, instead of a one-threaded example, to show that although thread $T$ might consistently use mutex $M$ to protect variable $A$, this is no good if thread $U$ accesses variable $A$ without holding $M$.}  \ADComment{If you use multiple letters in math mode, you should use ``mathit'' to get them typeset properly.
+E.g., you should write $\mathit{CLS}_T$, not $CLS_T$, and $\mathit{LS}_v$, not $LS_v$.
+Can you please address this throughout the paper?
+It can be cumbersome to write, making it worth thinking about adding macros.}
 
+Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $CLS_T$, for each thread $T$ of the program, and a lockset, $LS_v$, for each shared variable $v$ used in the program.
+Initially, $CLS_T$ is empty for every thread $T$, because the threads do not hold any locks on program start.
+The lockset $LS_v$ for each variable $v$ is initialised to the set of all locks manipulated by the program,
+because initially each access to $v$ has (vacuously) been protected by every lock.
+The program is executed as usual (with threads scheduled according to the OS scheduler), except that
+instrumentation is added to update locksets as follows:
+%
 \begin{enumerate}
-\item In each \emph{lock} and \emph{unlock} operation, update ($CLS$) with the currently held locks.
-\item In each shared variable access, update ($LS_A$) with the intersection of the latest ($CLS$) and the ($LS_A$) of the previous statement.
-\item If ($LS_A$) ever becomes empty, then report that the access to $A$ is unprotected.
+\item After each \emph{lock} and \emph{unlock} operation by thread $T$, $CLS_T$ is updated to reflect the locks currently held by $T$.
+\item When thread $T$ accesses shared variable $v$, any locks that are not common to $LS_v$ and $CLS_T$ are removed from $LS_v$.
+If $LS_v$ becomes empty as a result, a warning is issued that the access to $v$ may be unprotected.
 \end{enumerate}
 
-Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs.
+In the example of Figure~\ref{fig:locksets}, we analyze only one thread $T$ and one shared variable $A$ and thus there are two locksets: the current lockset for $T$ ($CLS_T$) and the lockset for $A$ ($LS_A$).
+
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a thread of a concurrent program.
+\ADComment{I folded some of the discussion of the example into the above more general description of locksets.
+Assuming you do produce a two-threaded example, please put in a brief explanation of it.}
+
+Lockset analysis avoids reasoning about arbitrary thread interleavings \ADComment{is this true of standard lockset analysis?
+With Eraser, a single interleaving would indeed provide a lot of information about races, but it would still be necessary to explore lots of interleavings to get high coverage, since the program behaviour may be schedule-dependent}, and thus has the potential to scale well in realistic concurrent programs.
 The technique, though, suffers from imprecision (i.e.
@@
 Furthermore, as a classic dynamic analyzer, Eraser's bug-finding ability is limited to the execution paths that the tool explores.
-To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.
+\ADComment{Perhaps re-phrase this to say that some existing techniques (cite them) have explored the idea of static lock set analysis, based on dataflow analysis, and that we also present a static lockset analysis method, that involves generating verification conditions to be discharged to a theorem prover} To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

### 2015-05-04 whoop.tex
various updates in evaluation
Changes to, and feedback on, the background.  Very first thoughts on how to restructure the Whoop section.

@@
+\ADComment{Please ignore my comments in this section for now.}
+
+We now present \emph{static pairwise lockset analysis}, a novel method for data race analysis in concurrent device drivers.
+In Section~\ref{ } we explain conceptually how the approach works.
+In Section~\ref{ } we describe our implementation of static pairwise lockset analysis in a practical tool,
+Whoop, that can be applied directly to driver source code, to check whether pairs of driver entry points are 
+free from data races.
+
+We demonstrate in Section~\ref{ } both that the Whoop technique and tool has value in its own right as a stand-alone static analyser, and that results from Whoop can be exploited to significantly boost performance of a more precise static analysis for concurrency offered by the CORRAL tool, based on context bounding and stratified inlining.
+
+\subsection{Static pairwise lockset analysis}
+
+\ADComment{Ally: first, let us try to explain really clearly in words how the technique works.}
+
+\ADComment{Second, let's introduce some notation to describe the thing more formally.}
+
+\ADComment{Finally, try to set up some kind of theorem giving the guarantee we think we have.}
+
+\subsection{Implementation in Whoop}
+
+The entry points for
+
 We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyze Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.
@@
 It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.
-\corral also employs sophisticated strategies for improving loop coverage, these are described in more detail in~\cite{lal2014powering}.
 

### 2015-05-05 background.tex
updates in the first part of the background section

@@
-\subsection{Concurrency in the Linux Kernel}
+\subsection{Concurrency in Device Drivers}
 \label{bg:concurrency}
 
-\ADComment{I believe that \emph{responsiveness} is a key motivator for concurrency in device drivers: even if there is only a single core, we may wish for driver entry points of distinct processes to run concurrently so that no single process hogs the core.
-Can you please consider weaving this into what follows?}
+Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput).
+In device drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core.
+Likewise, performance is also important: e.g.
+network drivers should be able to process incoming packages in a timely manner.
 
-Nowadays, multicore architectures have become mainstream, leading to an ever-increasing demand for concurrency in operating systems.
-Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
+Modern operating systems address the demand for responsiveness and performance by providing multiple sources of concurrency~\cite{corbet2005linux}: a kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
 All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:
@@
 
-\ADComment{In the definition, I added the part about at least one of the accesses being non-atomic.
-I believe this is important in drivers, since there are cases where people use atomics.
-Could you somewhere briefly acknowledge, referring back to Definition~\ref{definition:datarace}, that we treat atomics soundly in whoop, in the sense that we do not regard them as providing any protection between threads?}
+Example~\ref{fig:data_race_example} presents multiple data races that \whoop automatically found in the generic\_nvram Linux char driver (see Section~\ref{evaluation}).
+The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.
+However, this can lead to multiple possible data races because $T_1$ and $T_2$ can read/write access the \texttt{f\_pos} field of the \texttt{file} object in arbitrary order resulting into nondeterministic behavior.
 
-\ADComment{It would be much more compelling, even if it takes more space, if you could find a small snippet from a real driver such that there is a data race say on a field of a struct.
-Could you replace the contrived example with such an example?}  Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver.
-The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}.
-Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5.
-A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behavior.
-\ADComment{If you stick with this example, it would be worth pointing out that a write-write race is also possible.}
-
-\begin{lstlisting}[caption = Simple data race in a Linux driver \ADComment{I think the caption is misleading, since this is clearly not a driver}, label = fig:data_race_example]
-static void ep1 (struct shared *dev) {
-  dev->resource = 5;
+\begin{lstlisting}[caption = Data races in the generic\_nvram char driver that \whoop found., label = fig:data_race_example]
+static loff_t nvram_llseek(struct file *file,
+    loff_t offset, int origin) {
+  switch (origin) {
+    case 0:
+      break;
+    case 1:
+      offset += file->f_pos; // racy
+      break;
+    case 2:
+      offset += nvram_len;
+      break;
+    default:
+      offset = -1;
   }
-
-static void ep2 (struct shared *dev) {
-  if (dev->resource != 5)
-    dev->resource = 3;
+  if (offset < 0)
+    return -EINVAL;
+  file->f_pos = offset; // racy
+  return file->f_pos; // racy
 }
@@
 Example~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Example~\ref{fig:data_race_example}.
+The example is tricky, because the return statement can potentially race on the \texttt{f\_pos} field.
+To fix this issue, we store the result in a temporary variable, inside the critical section.
 
-\begin{lstlisting}[caption = Using locks to eliminate the data race, label = fig:lock_example]
-static void ep1 (struct shared *dev) {
-  mutex_lock(&dev->lock);
-  dev->resource = 5;
-  mutex_unlock(&dev->lock);
+\begin{lstlisting}[caption = Using a lock to eliminate the data races., label = fig:lock_example]
+static loff_t nvram_llseek(struct file *file,
+    loff_t offset, int origin) {
+  mutex_lock(&nvram_mutex); // lock
+  switch (origin) {
+    case 0:
+      break;
+    case 1:
+      offset += file->f_pos;
+      break;
+    case 2:
+      offset += nvram_len;
+      break;
+    default:
+      offset = -1;
   }
-
-static void ep2 (struct shared *dev) {
-  mutex_lock(&dev->lock);
-  if (dev->resource != 5)
-    dev->resource = 3;
-  mutex_unlock(&dev->lock);
+  if (offset < 0)
+    return -EINVAL;
+  file->f_pos = offset;
+  loff_t res = file->f_pos; // store result
+  mutex_unlock(&nvram_mutex); // unlock
+  return res;
 }
@@
 Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}.
-Coarse-grained locking can \ADComment{I removed the word ``severely''.
-My PhD supervisor always advised me to avoid using words that are subjective.} hurt performance, because it limits the opportunity for concurrency.
+Coarse-grained locking can hurt performance, because it limits the opportunity for concurrency.
 On the other hand, fine-grained locking can easily lead to deadlocks.
-\ADComment{Removed description of deadlock, as it is well known.}
-\ADComment{Suggestion: expand the following a little to say what one of the lock-free synchronization methods in Linux is.
-Perhaps take this opportunity to refer back to the stuff about ``atomic'' I added in the data race definition.}  Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, in this work we focus on locks.
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read and write operations, in this work we focus on locks.
+However, we still treat lock-free operations (e.g.
+atomics) soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads.
+This is sound but can potentially introduce false positives.
+
 \ADComment{Can we justify the focus on locks a bit more?

### 2015-05-05 whoop.tex
updates in the first part of the background section

@@
 \whoop inherits all limitations of lockset analysis (as discussed in Section~\ref{bg:lockset}): a violation of the locking discipline does not always correspond to a real error (e.g.
-\whoop does not reasons about lockfree data structures in a special way, it just over-approximates their read accesses); and we do not check for dynamically created locks (or for locks that are provided by external libraries).
+\whoop does not reason about lockfree data structures in a special way, it just over-approximates their read accesses); and we do not check for dynamically created locks (or for locks that are provided by external libraries).
 Because we perform two thread reduction, we also over-approximate the read accesses, which is another source of false positives.
@@
 
-\PDComment{Corral limitations ...}
+Finally, \whoop is unable to verify drivers where the programmer implicitly enforces serialization of entry points that would otherwise race.
+For example, in the intel\_scu\_wd driver (see Section~\ref{evaluation}) the programmer sets flag $A$ in entry point $ep1$ that that does not allow entry point $ep2$ ...
+(TODO).

### 2015-05-05 background.tex
updates in the lockset analysis part of the background section

@@
 The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.
-However, this can lead to multiple possible data races because $T_1$ and $T_2$ can read/write access the \texttt{f\_pos} field of the \texttt{file} object in arbitrary order resulting into nondeterministic behavior.
+However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can read/write access the \texttt{f\_pos} field of the \texttt{file} object in arbitrary order resulting into nondeterministic behavior.
 
@@
 You thus only get to see what the thread did on the particular schedule that was followed.
-I also think you should use a two-threaded example, instead of a one-threaded example, to show that although thread $T$ might consistently use mutex $M$ to protect variable $A$, this is no good if thread $U$ accesses variable $A$ without holding $M$.}  \ADComment{If you use multiple letters in math mode, you should use ``mathit'' to get them typeset properly.
-E.g., you should write $\mathit{CLS}_T$, not $CLS_T$, and $\mathit{LS}_v$, not $LS_v$.
-Can you please address this throughout the paper?
-It can be cumbersome to write, making it worth thinking about adding macros.}
+I also think you should use a two-threaded example, instead of a one-threaded example, to show that although thread $T$ might consistently use mutex $M$ to protect variable $A$, this is no good if thread $U$ accesses variable $A$ without holding $M$.}
 
-Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $CLS_T$, for each thread $T$ of the program, and a lockset, $LS_v$, for each shared variable $v$ used in the program.
-Initially, $CLS_T$ is empty for every thread $T$, because the threads do not hold any locks on program start.
-The lockset $LS_v$ for each variable $v$ is initialised to the set of all locks manipulated by the program,
-because initially each access to $v$ has (vacuously) been protected by every lock.
-The program is executed as usual (with threads scheduled according to the OS scheduler), except that
-instrumentation is added to update locksets as follows:
+Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_v$, for each shared variable $v$ used in the program.
+Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start.
+The lockset $\mathit{LS}_v$ for each variable $v$ is initialized to the set of all locks manipulated by the program, because initially each access to $v$ has (vacuously) been protected by every lock.
+The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows:
 %
 \begin{enumerate}
-\item After each \emph{lock} and \emph{unlock} operation by thread $T$, $CLS_T$ is updated to reflect the locks currently held by $T$.
-\item When thread $T$ accesses shared variable $v$, any locks that are not common to $LS_v$ and $CLS_T$ are removed from $LS_v$.
-If $LS_v$ becomes empty as a result, a warning is issued that the access to $v$ may be unprotected.
+\item After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
+\item When $T$ accesses shared variable $v$, any locks that are not common to $\mathit{LS}_v$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_v$.
+If $\mathit{LS}_v$ becomes empty as a result, a warning is issued that the access to $v$ may be unprotected.
 \end{enumerate}
 
-In the example of Figure~\ref{fig:locksets}, we analyze only one thread $T$ and one shared variable $A$ and thus there are two locksets: the current lockset for $T$ ($CLS_T$) and the lockset for $A$ ($LS_A$).
+In the example of Figure~\ref{fig:locksets}, we analyze only one thread $T$ and one shared variable $A$ and thus there are two locksets: the current lockset for $T$ ($\mathit{CLS}_T$) and the lockset for $A$ ($\mathit{LS}_A$).
 
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a thread of a concurrent program.
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing the shared variable $A$.
+Thus, we have to analyze three locksets: the current locksets $\mathit{CLS}_1$ and $\mathit{CLS}_2$ for $T_1$ and $T_2$ respectively; and the lockset $\mathit{LS}_A$ for $A$.
+
 \ADComment{I folded some of the discussion of the example into the above more general description of locksets.
@@
 
-Lockset analysis avoids reasoning about arbitrary thread interleavings \ADComment{is this true of standard lockset analysis?
-With Eraser, a single interleaving would indeed provide a lot of information about races, but it would still be necessary to explore lots of interleavings to get high coverage, since the program behaviour may be schedule-dependent}, and thus has the potential to scale well in realistic concurrent programs.
+In contrast to analyses based on happens-before, lockset analysis has the potential to scale well, because it does not need to reason about all possible thread interleavings to achieve good code coverage.
 The technique, though, suffers from imprecision (i.e.
-can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
-Furthermore, as a classic dynamic analyzer, Eraser's bug-finding ability is limited to the execution paths that the tool explores.
-\ADComment{Perhaps re-phrase this to say that some existing techniques (cite them) have explored the idea of static lock set analysis, based on dataflow analysis, and that we also present a static lockset analysis method, that involves generating verification conditions to be discharged to a theorem prover} To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.
+can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
+Furthermore, code coverage in dynamic lockset analysis tools, such as Eraser, is limited by the number of explored execution paths.
+To counter these limitations, some techniques~\cite{pratikakis2006locksmith, voung2007relay} have explored the idea of applying lockset analysis in a static context, using dataflow analysis.
+In this paper, we also present a static lockset analysis method that involves generating verification conditions to be discharged to a theorem prover.

### 2015-05-05 background.tex
more updates in the lockset analysis part of the background section

@@
 Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs.
-Furthermore, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}, a problem known as state space explosion.
+Moreover, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}, a problem known as state space explosion.
 Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.
@@
 
-\ADComment{What is missing from the explanation, and from the example you give, is the idea that with a dynamic approach, you run the program and a thread operates in a truly concurrent environment.
-You thus only get to see what the thread did on the particular schedule that was followed.
-I also think you should use a two-threaded example, instead of a one-threaded example, to show that although thread $T$ might consistently use mutex $M$ to protect variable $A$, this is no good if thread $U$ accesses variable $A$ without holding $M$.}
-
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_v$, for each shared variable $v$ used in the program.
@@
 
-\ADComment{I folded some of the discussion of the example into the above more general description of locksets.
-Assuming you do produce a two-threaded example, please put in a brief explanation of it.}
+\ADComment{I also think you should use a two-threaded example, instead of a one-threaded example, to show that although thread $T$ might consistently use mutex $M$ to protect variable $A$, this is no good if thread $U$ accesses variable $A$ without holding $M$.}
 
@@
 can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
-Furthermore, code coverage in dynamic lockset analysis tools, such as Eraser, is limited by the number of explored execution paths.
-To counter these limitations, some techniques~\cite{pratikakis2006locksmith, voung2007relay} have explored the idea of applying lockset analysis in a static context, using dataflow analysis.
-In this paper, we also present a static lockset analysis method that involves generating verification conditions to be discharged to a theorem prover.
+Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler.
+This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.
+
+To counter the above limitations, techniques such as Locksmith~\cite{pratikakis2006locksmith} and RELAY~\cite{voung2007relay} have explored the idea of applying lockset analysis in a static context, using dataflow analysis.
+In this paper, we present a novel static lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.

### 2015-05-05 background.tex
more updates in the background section

@@
 
-\ADComment{Can we justify the focus on locks a bit more?
-E.g., are you able to survey, via grepping, what percentage of Linux drivers make use of locks (likely 100\%?) vs.\ what percentage also make use of lock-free mechanisms?}
+We believe that our focus in locks is justified, because the majority of Linux drivers still use locks.
+As a sanity check, we used a simple grep command and found that 68\% of the drivers from the 4.0 Linux kernel distribution use locks, versus 13\% that use atomic operations.
+We also found that 31\% of the drivers use neither locking nor atomic primitives, which suggests that they are either simple (non-concurrent) drivers or use some other lock-free synchronization primitive.
 
@@
 
-In the example of Figure~\ref{fig:locksets}, we analyze only one thread $T$ and one shared variable $A$ and thus there are two locksets: the current lockset for $T$ ($\mathit{CLS}_T$) and the lockset for $A$ ($\mathit{LS}_A$).
-
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing the shared variable $A$.
-Thus, we have to analyze three locksets: the current locksets $\mathit{CLS}_1$ and $\mathit{CLS}_2$ for $T_1$ and $T_2$ respectively; and the lockset $\mathit{LS}_A$ for $A$.
-
-\ADComment{I also think you should use a two-threaded example, instead of a one-threaded example, to show that although thread $T$ might consistently use mutex $M$ to protect variable $A$, this is no good if thread $U$ accesses variable $A$ without holding $M$.}
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing the same shared variable $A$.
+Thus, we have to track three locksets: the current locksets $\mathit{CLS}_1$ and $\mathit{CLS}_2$ for $T_1$ and $T_2$ respectively; and the lockset $\mathit{LS}_A$ for $A$.
+Initially, $\mathit{LS}_A$ contains the two locks $M$ and $N$ that the two threads acquire during execution.
+When $T_1$ terminates, $\mathit{LS}_A$ contains only the lock $N$, because the thread write accessed $A$ without holding $N$.
+Next, during the execution of $T_2$, the thread write accesses $A$ without holding $M$.
+Thus, $\mathit{LS}_A$ becomes the empty set.
+This means that there is a potential data race between the two threads on $A$ as $T_2$ does not consistently protect $A$ with at least one common with $T_1$ lock.
 

### 2015-05-05 background.tex
more updates in the background section

@@
 
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing the same shared variable $A$.
-Thus, we have to track three locksets: the current locksets $\mathit{CLS}_1$ and $\mathit{CLS}_2$ for $T_1$ and $T_2$ respectively; and the lockset $\mathit{LS}_A$ for $A$.
-Initially, $\mathit{LS}_A$ contains the two locks $M$ and $N$ that the two threads acquire during execution.
-When $T_1$ terminates, $\mathit{LS}_A$ contains only the lock $N$, because the thread write accessed $A$ without holding $N$.
-Next, during the execution of $T_2$, the thread write accesses $A$ without holding $M$.
-Thus, $\mathit{LS}_A$ becomes the empty set.
-This means that there is a potential data race between the two threads on $A$ as $T_2$ does not consistently protect $A$ with at least one common with $T_1$ lock.
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$.
+Initially, $\mathit{LS}_A$, which is the lockset for A, contains all possible locks in the program: $M$ and $N$.
+During execution of $T_1$, the thread writes $A$ without holding $N$, and thus $N$ is removed from $\mathit{LS}_A$.
+Next, during execution of $T_2$, the thread writes $A$ without holding $M$, and thus $\mathit{LS}_A$ becomes empty.
+Finally, a warning is reported because the two threads do not consistently protect $A$.
 

### 2015-05-05 background.tex
small fix

@@
 
-\begin{lstlisting}[caption = Data races in the generic\_nvram char driver that \whoop found., label = fig:data_race_example]
+\begin{lstlisting}[caption = Data races in a Linux driver that we discovered using \whoop., label = fig:data_race_example]
 static loff_t nvram_llseek(struct file *file,

### 2015-05-07 whoop.tex
Wrote down simple language, and instrumentation.  Still a lot TODO.

@@
 \ADComment{Second, let's introduce some notation to describe the thing more formally.}
+
+We present our approach in a semi-formal manner, in the context of the following simple concurrent programming model.
+For ease of presentation only, the model does not features aggregate data types, pointers or dynamic memory allocation.
+These \emph{are} handled by our implementation, and in \ADComment{TODO} we discuss interesting practical issues arising from the handling of a full-blown language.
+\ADComment{I want to say that we do not treat dynamic thread creation because driver threads do not create further driver threads.
+However, I am not sure (a) whether this is true, nor (b) whether even if it is true it would be a problem for us.}
+
+\paragraph{Simple concurrent programming model}
+
+A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of threads, $T_1,\dots,T_n$ (for some $n > 0$).
+A thread, $T$, is described by a finite set of procedures, $\procedures{T}$, and a finite set of thread-private variables, $\variables{T}$.
+Thread execution commences from a designated procedure $\mainforthread{T} \in \procedures{T}$.
+A procedure $P \in \procedures{T}$ is represented by a control flow graph of basic blocks.
+Each basic block contains a sequence of statements, where the forms of statement are shown in Figure~\ref{TODO}.
+A basic block either has a single successor or a pair of successors.
+In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.
+
+\begin{figure}
+\begin{tabular}{lp{5.5cm}}
+\textbf{Statement form} & \textbf{Notes} \\\hline
+$x = e;$ & private assignment, where $x \in \variables{T}$ and $e$ is an expression over $\variables{T}$ \\\hline
+$x = f(\overline{e});$ & procedure call, where $x \in \variables{T}$, $\overline{e}$ is a sequence of expressions over $\variables{T}$, $f$ is the name of a procedure in $\procedures{T}$ \\\hline
+$s = e;$ & shared write, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$ \\\hline
+$x = s;$ & shared read,  where $x \in \variables{T}$ and $s \in \sharedvars$ \\\hline
+$\mutexlock{m};$   & mutex lock, where $m \in \mutexes$ \\\hline
+$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$
+\end{tabular}
+\caption{Forms of statements in our simple programming model}\label{fig:statements}
+\end{figure}
+
+\ADComment{Need to say something clear, but intuitive rather than precise, about the semantics.  Must say what a data race is.  Must address the weak memory issue.}
+
+\paragraph{Instrumentation}
+
+\begin{figure}
+\begin{tabular}{ll}
+\textbf{Original statement} & \textbf{Instrumented statement} \\\hline
+$x = e;$                & $\instrument{x}{i} = \instrument{e}{i}$;  \\\hline
+$x = f(\overline{e});$ & $\instrument{x}{i} = \instrument{f}{i}(\instrument{\overline{e}}{i});$ \\\hline
+$s = e;$ & $\hasbeenwritten{i} = \hasbeenwritten{i} \cup \{ s \};$ \\
+         & $\lockset{s}{i} = \lockset{s}{i} \cap \currentlockset{i};$ \\\hline
+$x = s;$ & $\hasbeenread{i} = \hasbeenread{i} \cup \{ s \};$ \\
+         & $\lockset{s}{i} = \lockset{s}{i} \cap \currentlockset{i};$ \\
+         & $\havoc{\instrument{x}{i}};$ \\\hline
+$\mutexlock{m};$   & $\currentlockset{i} = \currentlockset{i} \cup \{ m \};$ \\\hline
+$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\\hline
+\end{tabular}
+\caption{Instrumenting statements for static lockset analysis}\label{fig:instrumentation}
+\end{figure}
 

### 2015-05-07 evaluation.tex
updates to evaluation

@@
 \centering
-\caption{Runtime comparison with different yield instrumentation granularities and context-switch bounds.}
+\setlength{\tabcolsep}{0.5em}
+\caption{Comparison with different yield instrumentation granularities and context-switch bounds.}
 \label{tab:results}
-\begin{tabular}{l rrr rrr rr r r}
+\begin{tabular}{l r rrrr rrrr rrrr}
 \centering
@@
 \texttt{yield}) in all visible operations (denoted by \yieldall in the table).
-\whoop + \corral, instead, uses two different context-switch instrumentation granularities (see Section~\ref{whoop:bugfinding}): \yieldcoarse and \yieldmr.
+\whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{whoop:bugfinding}), and also uses all the optimisations discussed in Section~\ref{whoop:optimizations}.
 

### 2015-05-07 whoop.tex
Progress

@@
 
-\paragraph{Simple concurrent programming model}
+\paragraph{A simple concurrent programming model}
 
-A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of threads, $T_1,\dots,T_n$ (for some $n > 0$).
-A thread, $T$, is described by a finite set of procedures, $\procedures{T}$, and a finite set of thread-private variables, $\variables{T}$.
-Thread execution commences from a designated procedure $\mainforthread{T} \in \procedures{T}$.
-A procedure $P \in \procedures{T}$ is represented by a control flow graph of basic blocks.
-Each basic block contains a sequence of statements, where the forms of statement are shown in Figure~\ref{TODO}.
+We consider a concurrent programming model in which an unbounded number of threads execute a set of pre-defined \emph{thread templates}.
+In the context of device drivers, a thread template corresponds to a driver entry point.
+At any given point of execution a certain number of threads are active, each thread executing a particular template.
+As in the case of drivers, where multiple distinct entry points as well as multiple instances of the same entry point may run concurrently, two distinct threads in our model may execute according to the same template or according to different templates.
+Further threads may be launched at any point during execution, just as in the context of device drivers where the OS may invoke additional driver entry points.
+
+A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of thread templates, $T_1,\dots,T_n$ (for some $n > 0$).
+A thread template, $T$, is described by a finite set of procedures, $\procedures{T}$, and a finite set of thread-private variables, $\variables{T}$.
+A designated procedure, procedure $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of a template by a thread.
+A procedure $P \in \procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
 A basic block either has a single successor or a pair of successors.
 In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.
+
+The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
+In particular, shared variables may not appear in arbitrary expressions.
+This restriction simplifies our presentation of lockset instrumentation, and a program that does not satisfy this restriction can be trivially pre-processed into one that does through the introduction of additional private temporaries to record values read from the shared state.
+We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.
 
@@
 
-\ADComment{Need to say something clear, but intuitive rather than precise, about the semantics.  Must say what a data race is.  Must address the weak memory issue.}
+\ADComment{Say something clear, but intuitive rather than precise, about the semantics.}
 
-\paragraph{Instrumentation}
+\paragraph{Data races}
+
+\ADComment{Say what a data race is, and address the weak memory issue.}
+
+\paragraph{Lockset instrumentation}
 

### 2015-05-08 intro.tex
update

@@
 
-In this paper, we present \whoop, a new fully automatic approach that (i) statically analyses device drivers written in the C language for data races and (ii) exploits the race-related results from the static analysis to accelerate bug-finding.
-\whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in device drivers.
+In this paper, we present \whoop, a new fully automatic approach that (i) statically analyzes device drivers written in the C language for data races and (ii) exploits the race-related results from the symbolic analysis to accelerate bug-finding.
+\whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in device drivers.
 The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race-checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
@@
 
-We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
+We then show that by exploiting the race-freedom guarantees that our symbolic analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
@@
 \item We present \whoop, a new fully automatic approach for analyzing device drivers for data races and accelerating bug-finding.
-\item We discuss static pairwise lockset analysis,  a sound and scalable technique for automatically verifying the absence of data races in device drivers.
+\item We discuss symbolic pairwise lockset analysis,  a sound and scalable technique for automatically verifying the absence of data races in device drivers.
 \item We show that we can achieve a sound partial-order reduction using our technique and significantly speedup \corral, a state-of-the-art bug-finder.

### 2015-05-08 background.tex
update to the table

@@
 To counter the above limitations, techniques such as Locksmith~\cite{pratikakis2006locksmith} and RELAY~\cite{voung2007relay} have explored the idea of applying lockset analysis in a static context, using dataflow analysis.
-In this paper, we present a novel static lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.
+In this paper, we present a novel symbolic lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.

### 2015-05-08 whoop.tex
A bit of progress

@@
-\ADComment{Please ignore my comments in this section for now.}
+We now present \emph{symbolic pairwise lockset analysis}, a novel method
+for data race analysis in concurrent device drivers.  In
+Section~\ref{sec:symbolicpairwise} we describe how the approach works
+in a semi-formal manner, with respect to a simple concurrent
+programming model.  In Section~\ref{sec:whoopimplementation} we
+explain how we have implemented symbolic pairwise lockset analysis in
+a practical tool, Whoop, that can be applied directly to driver source
+code, to check whether pairs of driver entry points can race with one
+another.
 
-We now present \emph{static pairwise lockset analysis}, a novel method for data race analysis in concurrent device drivers.
-In Section~\ref{ } we explain conceptually how the approach works.
-In Section~\ref{ } we describe our implementation of static pairwise lockset analysis in a practical tool,
-Whoop, that can be applied directly to driver source code, to check whether pairs of driver entry points are 
-free from data races.
+Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the Whoop technique has value as a stand-alone analyser, and that results from Whoop can be exploited to significantly boost performance of a more precise symbolic analysis for concurrency offered by the CORRAL tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{whoop:bugfinding}.
 
-We demonstrate in Section~\ref{ } both that the Whoop technique and tool has value in its own right as a stand-alone static analyser, and that results from Whoop can be exploited to significantly boost performance of a more precise static analysis for concurrency offered by the CORRAL tool, based on context bounding and stratified inlining.
+\subsection{Symbolic pairwise lockset analysis}\label{sec:symbolicpairwise}
 
-\subsection{Static pairwise lockset analysis}
+In short, symbolic pairwise lockset analysis works by considering, for a given device driver, every pair of entry points that can potentially execute concurrently.
+For each such pair, we use symbolic verification to check whether it is possible for the pair to race, accounting for the effects of additional entry points by treating the driver shared state abstractly so that when an entry point reads from the shared state, a nondeterministic value is returned.
+Symbolic verification of an entry point pair works by instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and attempting to verify a sequential program that executes the instrumented entry points in sequence, and subsequently asserts that the lockset for each shared variable is non-empty.
+Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.
 
-\ADComment{Ally: first, let us try to explain really clearly in words how the technique works.}
-
-\ADComment{Second, let's introduce some notation to describe the thing more formally.}
-
-We present our approach in a semi-formal manner, in the context of the following simple concurrent programming model.
-For ease of presentation only, the model does not features aggregate data types, pointers or dynamic memory allocation.
-These \emph{are} handled by our implementation, and in \ADComment{TODO} we discuss interesting practical issues arising from the handling of a full-blown language.
-\ADComment{I want to say that we do not treat dynamic thread creation because driver threads do not create further driver threads.
-However, I am not sure (a) whether this is true, nor (b) whether even if it is true it would be a problem for us.}
+We now detail the approach in a semi-formal manner, in the context of a simple concurrent programming model.
 
@@
 Further threads may be launched at any point during execution, just as in the context of device drivers where the OS may invoke additional driver entry points.
+For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
+These \emph{are} handled by our implementation, and in Section~\ref{sec:whoopimplementation} we discuss interesting practical issues arising from the handling of a full-blown language.
 
@@
 
-\subsection{Implementation in Whoop}
+\subsection{Implementation in Whoop}\label{sec:whoopimplementation}
 

### 2015-05-08 whoop.tex
First draft of the simple programming language with informal semantics.

@@
 
-In short, symbolic pairwise lockset analysis works by considering, for a given device driver, every pair of entry points that can potentially execute concurrently.
+Symbolic pairwise lockset analysis works by considering, for a given device driver, every pair of entry points that can potentially execute concurrently.
 For each such pair, we use symbolic verification to check whether it is possible for the pair to race, accounting for the effects of additional entry points by treating the driver shared state abstractly so that when an entry point reads from the shared state, a nondeterministic value is returned.
 Symbolic verification of an entry point pair works by instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and attempting to verify a sequential program that executes the instrumented entry points in sequence, and subsequently asserts that the lockset for each shared variable is non-empty.
-Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.
+Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.
 
@@
 
-\paragraph{A simple concurrent programming model}
+\medskip\noindent\textbf{Concurrent programming model }
+%
+We consider a simple concurrent programming model in which an
+unbounded number of threads execute a set of pre-defined \emph{thread
+  templates}.  At any given point of execution a certain number of
+threads are active, each thread executing a particular template.  In
+the context of device drivers, a thread template corresponds to a
+driver entry point, and multiple instances of the same thread template
+may execute concurrently, just as multiple invocations of a single
+driver entry point may be concurrent.  Further threads may be launched
+at any point during execution; in the context of device drivers this
+corresponds to the OS invoking additional driver entry points.  For
+ease of presentation only, our model does not feature aggregate data
+types, pointers or dynamic memory allocation.  These \emph{are}
+handled by our implementation, and in
+Section~\ref{sec:whoopimplementation} we discuss interesting practical
+issues arising from the handling of a full-blown language.
 
-We consider a concurrent programming model in which an unbounded number of threads execute a set of pre-defined \emph{thread templates}.
-In the context of device drivers, a thread template corresponds to a driver entry point.
-At any given point of execution a certain number of threads are active, each thread executing a particular template.
-As in the case of drivers, where multiple distinct entry points as well as multiple instances of the same entry point may run concurrently, two distinct threads in our model may execute according to the same template or according to different templates.
-Further threads may be launched at any point during execution, just as in the context of device drivers where the OS may invoke additional driver entry points.
-For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
-These \emph{are} handled by our implementation, and in Section~\ref{sec:whoopimplementation} we discuss interesting practical issues arising from the handling of a full-blown language.
+A \emph{concurrent program} is described by a finite set of
+\emph{shared} variables, $\sharedvars$, a finite set of mutexes,
+$\mutexes$, and a finite set of \emph{thread templates}.  A thread
+template, $T$, consists of a finite set of procedures,
+$\procedures{T}$, and a finite set of private variables,
+$\variables{T}$.  A designated procedure, procedure $\mainforthread{T}
+\in \procedures{T}$, denotes the starting point for execution of a
+template by a thread.  A procedure $P \in \procedures{T}$ is
+represented by a control flow graph of basic blocks, where each block
+contains a sequence of statements.  A basic block either has a single
+successor or a pair of successors.  In the latter case, an \emph{exit
+  condition} over thread-private variables determines, at runtime, the
+successor to which control should flow on block exit.
 
-A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of thread templates, $T_1,\dots,T_n$ (for some $n > 0$).
-A thread template, $T$, is described by a finite set of procedures, $\procedures{T}$, and a finite set of thread-private variables, $\variables{T}$.
-A designated procedure, procedure $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of a template by a thread.
-A procedure $P \in \procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
-A basic block either has a single successor or a pair of successors.
-In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.
-
-The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
-In particular, shared variables may not appear in arbitrary expressions.
-This restriction simplifies our presentation of lockset instrumentation, and a program that does not satisfy this restriction can be trivially pre-processed into one that does through the introduction of additional private temporaries to record values read from the shared state.
-We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.
+The forms of statements are shown in Figure~\ref{fig:statements}, and
+include designated statements for reading from and writing to shared
+variables.  In particular, shared variables may not appear in
+arbitrary expressions.  This restriction simplifies our presentation
+of lockset instrumentation, below, and a program that does not satisfy
+this restriction can be trivially pre-processed into one that does
+through the introduction of additional private temporaries to record
+values read from the shared state.  We do not specify the form of
+expressions, nor the types of variables, assuming a standard set of
+data types and operations.
 
@@
 
-\ADComment{Say something clear, but intuitive rather than precise, about the semantics.}
+\medskip\noindent\textbf{Semantics} Let $\ids$ be an infinite set from
+which dynamic thread ids will be drawn.  The state of a running
+concurrent program consists of: a valuation of the shared variables,
+$\sharedvars$; a mapping that associates each mutex in $\mutexes$ with
+an id from $\ids$, recording which thread currently holds the mutex,
+or with a special value $\bot \notin \ids$ to indicate that the
+mutex is not held by any thread; a list of \emph{threads}.  Each
+thread consists of an id, drawn from $\ids$, a thread template, $T$,
+an index indicating the next statement of the template $T$ to be
+executed by the thread, and a valuation of the thread private
+variables, $\variables{T}$.  If multiple threads are instances of the
+same template $T$, then each thread carries a \emph{separate}
+valuation of the private variables for this template.
 
-\paragraph{Data races}
+At the start of execution the valuation of shared variables is
+arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$),
+and the list of threads is empty.
 
-\ADComment{Say what a data race is, and address the weak memory issue.}
+At any point of execution, a new thread may be added to the list of
+thread states.  This involves selecting an id $i \in \ids$ that has
+not been previously used during program execution and a thread
+template $T$, setting the point of execution for the new thread to be
+the first statement of $\mainforthread{T}$, and choosing an arbitrary
+valuation for the private variables $\variables{T}$.
 
-\paragraph{Lockset instrumentation}
+We consider a standard interleaving model of concurrency: at any
+execution point, a thread may execute its current statement, unless
+that statement has the form $\mutexlock(m)$ and mutex $m$ is already held
+by some thread.  Executing a statement causes the state of the thread,
+and the shared state, to be updated in a standard manner.  For
+example, if a thread following template $T$ executes $s = e$, where $s
+\in \sharedvars$ and $e$ is an expression over $\variables{T}$, the
+shared variable valuation is updated so that $s$ has the value
+determined by evaluating $e$ in the context of the thread's private
+variable valuation.  The thread's next statement is updated according
+to the control flow graph of the current procedure.
+
+A thread terminates if it reaches the end of $\mainforthread{T}$; in this
+case the thread is removed from the list of threads.  Because we are
+interested in the analysis of device drivers, which are
+\emph{reactive} concurrent programs, we do not consider the notion of
+global program termination.
+
+\medskip\noindent\textbf{Data races } \ADComment{We now need to define
+what a data race is.  This could be done using happens-before.  I am undecided as to whether
+we should refer back to the happens-before material in the background section, or whether
+we should move that material here, so that we explain what happens-before means, and then
+explain what a data races is.  My preference is for the latter: doing it all here.}
+
+\medskip\noindent\textbf{Lockset instrumentation} \ADComment{This is as far as I have gotten with the writing.
+I now want to explain formally and somewhat abstractly how our lockset instrumentation works.}
 

### 2015-05-08 experiments/figures/plot.py
added draft plots in evaluation section

@@ -1,0 +1,90 @@
+OUTPUT='yieldmr_vs_yieldall.pdf'
+# OUTPUT='yieldepp_vs_yieldall.pdf'
+XLABEL='Analysis time (in seconds) using Whoop+Corral (Yield-MR)'
+# XLABEL='Analysis time (in seconds) using Whoop+Corral (Yield-EPP)'
+YLABEL='Analysis time (in seconds) using Corral (Yield-ALL)'
+TIMEOUT=36000
+
+import numpy as np
+import matplotlib.pyplot as plt
+import math
+import sys
+
+from matplotlib.backends.backend_pdf import PdfPages
+
+plt.rcParams['xtick.minor.size'] = 0
+plt.rcParams['ytick.minor.size'] = 0
+
+x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              15.9]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           18.5]
+x2 = [49.3, 0, 23.9, 3.1,       19.3, 104.6,   107.6,    278.0,               24.3]
+
+z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        16.1]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       18.5]
+z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1,    1543.4,          24.5]
+
+y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            12.9]
+y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           15.2]
+y2 = [197.5, 36000, 78.4, 37.3   , 51.8, 793.0,   392.1,    1539.9,             21.4]
+
+# setup a figure
+pp = PdfPages(OUTPUT)
+fig, ax = plt.subplots(figsize=(8,8))
+
+# plot the data (csb = 2)
+ax.scatter(x0, y0, s=80, color='b', zorder = 10, marker='$2$', label='csb = 2')
+# plot the data (csb = 5)
+ax.scatter(x1, y1, s=80, color='g', zorder = 10, marker='$5$', label='csb = 5')
+# plot the data (csb = 9)
[…]

### 2015-05-08 experiments/figures/plot_corr.py

@@ -1,0 +1,86 @@
+OUTPUT='correlation.pdf'
+XLABEL='Analysis time (in seconds)'
+YLABEL='Number of instrumented yields'
+TIMEOUT=36000
+YIELDS=18000
+
+import numpy as np
+import matplotlib.pyplot as plt
+import math
+import sys
+
+from matplotlib.backends.backend_pdf import PdfPages
+
+plt.rcParams['xtick.minor.size'] = 0
+plt.rcParams['ytick.minor.size'] = 0
+
+x0 = [13.5, 41.3, 15.7, 3.1   , 14.5, 41.6,    99.2,     21.0,            15.9]
+x1 = [16.9, 79.4, 17.3, 3.1   , 16.1, 52.9,    102.4,    32.0,            18.5]
+x2 = [49.3, 0, 23.9, 3.1   , 19.3, 104.6,   107.6,    278.0,            24.3]
+
+z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0,    44.4,            16.1]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5,    89.6,            18.5]
+z2 = [132.1, 0, 55.2, 3.2,   33.0, 633.2,    107.1,    1543.4,            24.5]
+
+y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0,    39.8,            12.9]
+y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8,    85.0,            15.2]
+y2 = [197.5, 36000, 78.4, 37.3   , 51.8, 793.0,   392.1,    1539.9,            21.4]
+
+yx = [29, 167, 22, 0,    129, 286,     812,       601,      217]
+
+yz = [47, 500, 51, 0,    245, 664,     1143,      732,      227]
+
+yy = [92, 691, 104, 82,  513, 801,     2058,      732,      227]
+
+# setup a figure
+pp = PdfPages(OUTPUT)
+fig, ax = plt.subplots(figsize=(8,8))
+
+# plot the data (csb = 2)
+ax.scatter(x0, yx, s=80, color='b', zorder = 10, marker='$2$', label='csb = 2 (Yield-MR)')
+# plot the data (csb = 5)
+ax.scatter(x1, yx, s=80, color='b', zorder = 10, marker='$5$', label='csb = 5 (Yield-MR)')
+# plot the data (csb = 9)
[…]

### 2015-05-08 evaluation.tex
updates to evaluation
added draft plots in evaluation section
update in evaluation

@@
 \centering
-\setlength{\tabcolsep}{0.5em}
+\setlength{\tabcolsep}{0.45em}
 \caption{Comparison with different yield instrumentation granularities and context-switch bounds.}
@@
 \end{table*}
+
+\begin{figure}
+\centering
+\includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
+\caption{draft}
+\label{fig:plot1}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.99\linewidth]{experiments/figures/correlation.pdf}
+\caption{draft}
+\label{fig:plot2}
+\end{figure}
 
@@
 \textbf{Benchmarks}\xspace\xspace We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
-We chose drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
-Further, we focused on choosing non-trivial drivers (e.g.
-large number of entry points, deep nesting of entry point helper functions, complex use of locking, use of function pointers, etc).
-We manually modeled the environment for all these drivers, a process that required more than a month of work.
+We chose non-trivial drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+We had to manually model the environment for these drivers, a process that required two months of work.
 
@@
 \textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs that \whoop identified (\#Racy Pairs); number of racy memory regions that \whoop reported (\#Racy MRs); and number of races that \corral discovered (\#Races Found) using a context-switch bound of 2.
+We did not discover new races with larger context switch bounds.
+This might mean that these drivers only have races that manifest with at least one or two context-switches, or that \corral run hit its bounds before discovering a deeper race.
 
-We can see that, in general, \whoop reports more races than \corral.
-This is expected because \whoop over-approximates the original driver for soundness: if a race exists, the tool will report it (assuming tool and model correctness), but \whoop can, and does, report false races.
-On the contrary, \corral is precise, but can miss races as it explores the driver under pre-defined bounds.
-Further, a common problem for \corral is loop coverage.
-To tackle this we enable the build-in loop over-approximation described in~\cite{lal2014powering}.
+We can see that \whoop reports more races than \corral.
+This is expected, as our approach uses over-approximation (to find all races), and thus can introduce false bugs.
+On the other hand, \corral is precise, but can miss races as it explores programs under pre-defined bounds.
+Another issue with \corral is loop coverage due to unsound loop unrolling.
+To tackle this, we enable the built-in loop over-approximation described in~\cite{lal2014powering}.
 This can potentially lead \corral to report false positives, but we have not seen this in practice.
 
-Most of the data races that \whoop and \corral found can be classified in two cases.
-The first case had to do with incrementing a global counter without using a lock.
-The reason behind this might be performance, but even benign races can lead to undefined behavior according to the C standard.
-The second case had to do with an entry point modifying a field of an object (either global or passed as parameter) without using a lock.
-This can easily lead to a data race if another concurrently running entry point accesses the same field of the same object.
-As an example of the second case, we discovered the following data race in the generic\_nvram char driver: during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} (\texttt{file} is passed as a parameter to the entry point) without first acquiring a lock.
+Most of the races that \whoop and \corral discovered can be classified in two cases.
+The first case is accessing a global counter (or flag) from concurrent entry points, without a lock.
+This might be for performance, and indeed a lot of the races we found might be benign.
+Even benign races, though, can lead to undefined behavior according to the C standard, which can cause many issues (e.g.
+with compilers).
+The second case is an entry point accessing a field of an object (either global or passed as a parameter) without a lock.
+This can lead to a race if another entry point simultaneously accesses the same field of the same object.
+
+As an example of the second case, we discovered the following race in the generic\_nvram char driver: during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} without first acquiring a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This can lead to a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
-When we investigated another char driver that uses the same API, we saw that it was protecting the offset access in the same entry point with a mutex.
-This made us suspicious that the race we found in generic\_nvram is real.
+When we investigated another char driver that uses the same APIs, we saw that it was protecting the offset access in its \texttt{llseek} entry point with a mutex.
+This made us suspicious that the race we found in generic\_nvram could be real.
 We filed a bug report, but did not manage to get a response yet.
-
-It is worthwhile to mention that we did not discover any additional data races with larger context switch bounds.
-This might mean that most races in these drivers require only a small amount of context-switches to manifest, or that \corral run out of resources before discovering the deeper bug.
 

### 2015-05-08 evaluation.tex
update in evaluation

@@
 
-As an example of the second case, we discovered the following race in the generic\_nvram char driver: during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} without first acquiring a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+As an example of the second case, we found the following race in the generic\_nvram char driver (see Example~\ref{fig:data_race_example} for the source code): during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} without first acquiring a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This can lead to a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.

### 2015-05-09 experiments/figures/plot.py
update in measurements

@@ -15,17 +15,17 @@ from matplotlib.backends.backend_pdf import PdfPages
 plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
-x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              15.9]
-x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           18.5]
-x2 = [49.3, 0, 23.9, 3.1,       19.3, 104.6,   107.6,    278.0,               24.3]
+x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           0, 18.5]
+x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               0, 24.3]
 
-z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        16.1]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       18.5]
-z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1,    1543.4,          24.5]
+z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       0, 18.5]
+z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          0, 24.5]
 
-y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            12.9]
-y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           15.2]
-y2 = [197.5, 36000, 78.4, 37.3   , 51.8, 793.0,   392.1,    1539.9,             21.4]
[…]

### 2015-05-09 experiments/figures/plot_corr.py

@@ -14,23 +14,23 @@ from matplotlib.backends.backend_pdf import PdfPages
 plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
-x0 = [13.5, 41.3, 15.7, 3.1   , 14.5, 41.6,    99.2,     21.0,            15.9]
-x1 = [16.9, 79.4, 17.3, 3.1   , 16.1, 52.9,    102.4,    32.0,            18.5]
-x2 = [49.3, 0, 23.9, 3.1   , 19.3, 104.6,   107.6,    278.0,            24.3]
+x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           0, 18.5]
+x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               0, 24.3]
 
-z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0,    44.4,            16.1]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5,    89.6,            18.5]
-z2 = [132.1, 0, 55.2, 3.2,   33.0, 633.2,    107.1,    1543.4,            24.5]
+z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       0, 18.5]
+z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          0, 24.5]
 
-y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0,    39.8,            12.9]
-y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8,    85.0,            15.2]
-y2 = [197.5, 36000, 78.4, 37.3   , 51.8, 793.0,   392.1,    1539.9,            21.4]
+y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            1966.5, 12.9]
[…]

### 2015-05-09 experiments/figures/plot.py
update in evaluation

@@ -29,7 +29,7 @@ y2 = [197.5, 36000, 78.4, 37.3   , 51.8, 793.0,   392.1, 36000, 1539.9,         
 
 # setup a figure
 pp = PdfPages(OUTPUT)
-fig, ax = plt.subplots(figsize=(8,8))
+fig, ax = plt.subplots(figsize=(6,6))
 
 # plot the data (csb = 2)
 ax.scatter(x0, y0, s=80, color='b', zorder = 10, marker='$2$', label='csb = 2')

### 2015-05-10 experiments/figures/plot.py
update in measurements

@@ -16,12 +16,12 @@ plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
 x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9]
-x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           0, 18.5]
-x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               0, 24.3]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5]
+x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3]
 
 z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       0, 18.5]
-z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          0, 24.5]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5]
+z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5]
 
 y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            1966.5, 12.9]
 y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2]

### 2015-05-10 experiments/figures/plot_corr.py

@@ -15,12 +15,12 @@ plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
 x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9]
-x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           0, 18.5]
-x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               0, 24.3]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5]
+x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3]
 
 z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       0, 18.5]
-z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          0, 24.5]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5]
+z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5]
 
 y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            1966.5, 12.9]
 y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2]

### 2015-05-10 experiments/figures/plot.py
update in evaluation

@@ -32,11 +32,11 @@ pp = PdfPages(OUTPUT)
 fig, ax = plt.subplots(figsize=(6,6))
 
 # plot the data (csb = 2)
-ax.scatter(x0, y0, s=80, color='b', zorder = 10, marker='$2$', label='csb = 2')
+ax.scatter(x0, y0, s=60, color='b', zorder = 10, marker='+')
 # plot the data (csb = 5)
-ax.scatter(x1, y1, s=80, color='g', zorder = 10, marker='$5$', label='csb = 5')
+ax.scatter(x1, y1, s=60, color='g', zorder = 10, marker='o', facecolors='none')
 # plot the data (csb = 9)
-ax.scatter(x2, y2, s=80, color='r', zorder = 10, marker='$9$', label='csb = 9')
+ax.scatter(x2, y2, s=60, color='r', zorder = 10, marker='x', facecolors='none')
 
 # plot the data (csb = 2)
 # ax.scatter(z0, y0, s=80, color='b', zorder = 10, marker='$2$', label='csb = 2')

### 2015-05-10 evaluation.tex
update in evaluation
added experimental setup

@@
 \includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
-\caption{draft}
-\label{fig:plot1}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=.99\linewidth]{experiments/figures/correlation.pdf}
-\caption{draft}
-\label{fig:plot2}
+\caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
+The symbols $+$, $\circ$ and $\times$, represent a context-switch bound of 2, 5 and 9, respectively.}
+\label{fig:plot}
 \end{figure}
@@
 \noindent
-\textbf{Experimental Setup}\xspace\xspace All experiments were performed on a ... machine running ..., LLVM 3.5; SMACK 1.5; Z3 4.3.2; Boogie rev. 4192 and \corral rev. 534.
+\textbf{Experimental Setup}\xspace\xspace We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5, Z3 4.3.2, Boogie rev.
+4192 and \corral rev.
+534.
 

### 2015-05-10 experiments/figures/plot.py

@@ -15,17 +15,17 @@ from matplotlib.backends.backend_pdf import PdfPages
 plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
-x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9]
-x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5]
-x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3]
+x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5,        4805.4, 36000]
+x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3,        36000]
 
-z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5]
-z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5]
+z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5,        36000]
+z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
 
-y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            1966.5, 12.9]
-y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2]
[…]

### 2015-05-10 experiments/figures/plot.py
update in evaluation

@@ -20,12 +20,12 @@ x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36
 x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3,        36000]
 
 z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5,        36000]
+z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
 z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
 
 y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            1966.5, 12.9,        548.2, 36000]
-y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2,        36000]
-y2 = [197.5, 36000, 78.4, 37.3   , 51.8, 793.0,   392.1, 36000, 1539.9,         36000, 21.4,        36000]
+y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2,        36000,36000]
+y2 = [197.5, 27337.6, 78.4, 37.3   , 51.8, 793.0,   392.1, 36000, 1539.9,         36000, 21.4,        36000]
 
 # setup a figure
 pp = PdfPages(OUTPUT)
@@ -51,15 +51,19 @@ ax.plot(xyline, xyline, 'k--')
 
 x2sp = np.logspace(math.log(0.00425,10), math.log(20500,10))
 y2sp = np.logspace(math.log(0.0085,10), math.log(40000,10))
-ax.plot(x2sp, y2sp, 'c-.', label='2x speedup')
+ax.plot(x2sp, y2sp, 'm-.', label='2x speedup')
 
 x5sp = np.logspace(math.log(0.0017,10), math.log(8200,10))
[…]

### 2015-05-10 experiments/figures/plot.py
update in evaluation

@@ -16,7 +16,7 @@ plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
 x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
-x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5,        4805.4, 36000]
+x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5,        36000]
 x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3,        36000]
 
 z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
@@ -24,12 +24,12 @@ z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000,
 z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
 
 y0 = [33.5, 169.3, 38.1, 11.1   , 35.0, 182.6,   388.0, 271.0, 39.8,            1966.5, 12.9,        548.2, 36000]
-y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2,        36000,36000]
+y1 = [47.0, 595.1, 45.6, 13.8   , 40.2, 263.7,   390.8, 2746.6, 85.0,           36000, 15.2,        36000]
 y2 = [197.5, 27337.6, 78.4, 37.3   , 51.8, 793.0,   392.1, 36000, 1539.9,         36000, 21.4,        36000]
 
 # setup a figure
 pp = PdfPages(OUTPUT)
-fig, ax = plt.subplots(figsize=(6,6))
+fig, ax = plt.subplots(figsize=(6.5,6.5))
 
 # plot the data (csb = 2)
 ax.scatter(x0, y0, s=60, color='b', zorder = 10, marker='+')

### 2015-05-11 conclusion.tex
update
rewrote conclusion

--- 2015-04-26 conclusion.tex
@@
-In this paper, we present \whoop, a new fully automatic approach for (i) finding data races in device drivers and (ii) accelerating bug-finding using race-related information.
-In the heart of our approach, we combine static lockset analysis with sound sequentialization and automated theorem proving.
-Our lockset analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.
+In this paper we presented \whoop, a new fully automatic approach that can soundly detect all possible data races in device drivers.
+Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation, summarization and symbolic pairwise lockset analysis, and thus can scale well.
 
-Compared to traditional data race detection tools that typically attempt to explore as many thread interleavings as feasible, and thus face scalability issues in the presence of realistic concurrent programs, \textsc{Whoop} not only entirely avoids reasoning about thread interleavings, but also allows the reuse of existing successful sequential verification techniques.
-The main limitation of our approach is that it can potentially report many false positives, as we over-approximate the device driver shared state.
-To tackle this problem, we use the race-related information that \whoop generates to speed up \corral, a state-of-the-art precise bug-finder for concurrent programs.
+Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art precise bug-finder for concurrent programs.
+Using the combination of \whoop + \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our approach is efficient for race-checking and can achieve x-y speedups over vanilla \corral.

### 2015-05-11 conclusion.tex
some updates to include speedups

@@
 Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation, summarization and symbolic pairwise lockset analysis, and thus can scale well.
-
 Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art precise bug-finder for concurrent programs.
-Using the combination of \whoop + \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our approach is efficient for race-checking and can achieve x-y speedups over vanilla \corral.
+Using the combination of \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our approach can analyze programs 1.5--20$\times$ faster than standalone \corral.

### 2015-05-11 experiments/figures/plot.py

@@ -15,17 +15,17 @@ from matplotlib.backends.backend_pdf import PdfPages
 plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
-x0 = [13.5, 41.3, 15.7, 3.1,    14.5, 41.6,    99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
-x1 = [16.9, 79.4, 17.3, 3.1,    16.1, 52.9,    102.4, 1847.9, 32.0,           36000, 18.5,        36000]
-x2 = [49.3, 1358.9, 23.9, 3.1,  19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3,        36000]
+x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
+x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0,           36000, 18.5,        36000]
+x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3,        36000]
 
-z0 = [17.7, 79.4, 26.1, 3.1,   22.3, 91.2,     104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
-z1 = [24.2, 432.4, 31.2, 3.1,  25.7, 150.7,    105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
-z2 = [132.1, 0, 55.2, 3.2,     33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
+z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
+z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
+z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
 
[…]

### 2015-05-11 evaluation.tex
rewriten limitations

@@
 We performed experiments to validate the usefulness of the \whoop approach (see Section~\ref{whoop}).
-We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution\footnote{https://www.kernel.org}.
+We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution\footnote{\url{https://www.kernel.org}}.
 We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds, to show that \whoop can efficiently accelerate bug-finding with \corral.

### 2015-05-11 whoop.tex
rewriten limitations

@@
 Boogie and \corral).
-To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur\footnote{https://github.com/mc-imperial/chauffeur}, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.
+To achieve this translation, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools: Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}.
 
@@
 
-\whoop inherits all limitations of lockset analysis (as discussed in Section~\ref{bg:lockset}): a violation of the locking discipline does not always correspond to a real error (e.g.
-\whoop does not reason about lockfree data structures in a special way, it just over-approximates their read accesses); and we do not check for dynamically created locks (or for locks that are provided by external libraries).
-Because we perform two thread reduction, we also over-approximate the read accesses, which is another source of false positives.
+\whoop is based on lockset analysis and thus can report false bugs, because a violation of the locking discipline does not always correspond to a real data race (e.g.
+when lockfree synchronization is used instead of locking).
+\whoop also uses over-approximation (e.g.
+when reading from the shared state) and summarization, which can be sources of false positives.
+Further, the tool does not currently check for dynamically created locks or for locks that are provided by external libraries, although the later could be easily addressed by providing a mechanism for users to declare custom locks.
 
-Environmental modeling is another limitation of \whoop.
-As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.
+Another limitation of \whoop is that it is unable to verify cases where a driver is forced to be accessed by a single process at a time.
+This \emph{single-open device}~\cite{corbet2005linux} mode is enforced by using an atomic counter to indicate device availability: an entry point executes if the counter has value 1, else blocks.
+Because \whoop performs pair-wise analysis, it over-approximates this counter, and regard two entry points as running concurrently, even if this is not the case.
+However,~\cite{corbet2005linux} advises against using single-open device mode, as it hinders user ingenuity (e.g.
+a user might expect that a device can be accessed concurrently for performance).
+Thus, because using single-open device mode is considered bad practice, we burden the developer with disregarding any related false bugs.
+
+Finally, another limitation of our approach is the need for precise environmental modeling.
+Enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.
 We argue that further increasing the precision of our models is orthogonal to the contributions of this paper.
-Moreover, despite unavoidable imprecisions during the sound static data race analysis, \whoop can use the reported information to significantly speedup precise bug-finders for concurrent programs, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.
-
-Finally, \whoop is unable to verify drivers where the programmer implicitly enforces serialization of entry points that would otherwise race.
-For example, in the intel\_scu\_wd driver (see Section~\ref{evaluation}) the programmer sets flag $A$ in entry point $ep1$ that that does not allow entry point $ep2$ ...
-(TODO).
+Moreover, even if the symbolic lockset analysis can be imprecise, \whoop can still use any race-freedom guarantees to significantly speedup a precise bug-finders, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.

### 2015-05-11 experiments/figures/plot.py
update in evaluation

@@ -17,15 +17,15 @@ plt.rcParams['ytick.minor.size'] = 0
 
 x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
 x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0,           36000, 18.5,        36000]
-x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6,   107.6, 0, 278.0,               36000, 24.3,        36000]
+x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 0, 278.0,               36000, 24.3,        36000]
 
 z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
 z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
 z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
 
-y0 = [33.5, 169.3, 38.1, 11.1, 22.8, 35.0, 182.6, 77.5, 388.0, 271.0, 39.8,            1966.5, 12.9,        548.2, 36000]
-y1 = [47.0, 595.1, 45.6, 13.8, 130.3, 40.2, 263.7, 438.8, 390.8, 2746.6, 85.0,           36000, 15.2,        36000]
-y2 = [197.5, 27337.6, 78.4, 37.3, 1571.7, 51.8, 793.0,   392.1, 36000, 1539.9,         36000, 21.4,        36000]
+y0 = [33.5, 169.3, 38.1, 11.1, 22.8, 35.0, 182.6, 81.5, 388.0, 271.0, 39.8,            1966.5, 12.9,        548.2, 36000]
+y1 = [47.0, 595.1, 45.6, 13.8, 130.3, 40.2, 263.7, 419.4, 390.8, 2746.6, 85.0,           36000, 15.2,        36000]
[…]

### 2015-05-11 evaluation.tex
fix

@@
 We did not discover new races with larger context switch bounds.
-This might mean that these drivers only have races that manifest with at least one or two context-switches, or that \corral run hit its bounds before discovering a deeper race.
+This might mean that these drivers only have races that manifest with at least one or two context-switches, or that \corral hit its bounds before discovering a deeper race.
 

### 2015-05-11 experiments/figures/plot.py

@@ -51,19 +51,19 @@ ax.plot(xyline, xyline, 'k--')
 
 x2sp = np.logspace(math.log(0.00425,10), math.log(20500,10))
 y2sp = np.logspace(math.log(0.0085,10), math.log(40000,10))
-ax.plot(x2sp, y2sp, 'm-.', label='2x speedup')
+ax.plot(x2sp, y2sp, 'c-.', label='2x speedup')
 
-x5sp = np.logspace(math.log(0.0017,10), math.log(8200,10))
-y5sp = np.logspace(math.log(0.0085,10), math.log(41000,10))
-ax.plot(x5sp, y5sp, 'c-.', label='5x speedup')
+# x5sp = np.logspace(math.log(0.0017,10), math.log(8200,10))
+# y5sp = np.logspace(math.log(0.0085,10), math.log(41000,10))
+# ax.plot(x5sp, y5sp, 'g-.', label='5x speedup')
 
 x10sp = np.logspace(math.log(0.00085,10), math.log(4100,10))
 y10sp = np.logspace(math.log(0.0085,10), math.log(41000,10))
-ax.plot(x10sp, y10sp, 'g-.', label='10x speedup')
+ax.plot(x10sp, y10sp, 'm-.', label='10x speedup')
 
-x20sp = np.logspace(math.log(0.000425,10), math.log(2050,10))
-y20sp = np.logspace(math.log(0.0085,10), math.log(41000,10))
-ax.plot(x20sp, y20sp, 'y-.', label='20x speedup')
+# x20sp = np.logspace(math.log(0.000425,10), math.log(2050,10))
+# y20sp = np.logspace(math.log(0.0085,10), math.log(41000,10))
+# ax.plot(x20sp, y20sp, 'y-.', label='20x speedup')
 
 # setup log scales
 ax.set_xscale('log')

### 2015-05-11 whoop.tex
fix

@@
 Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronization bugs that arise from careless fine-grained locking.
-The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
+Note that the actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
 
@@
 
-Another limitation of \whoop is that it is unable to verify cases where a driver is forced to be accessed by a single process at a time.
-This \emph{single-open device}~\cite{corbet2005linux} mode is enforced by using an atomic counter to indicate device availability: an entry point executes if the counter has value 1, else blocks.
-Because \whoop performs pair-wise analysis, it over-approximates this counter, and regard two entry points as running concurrently, even if this is not the case.
-However,~\cite{corbet2005linux} advises against using single-open device mode, as it hinders user ingenuity (e.g.
+Another limitation of \whoop is that it is unable to verify drivers that are designed to be accessed by a single process at a time.
+This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of each entry point) a flag that indicates device availability: if the flag is set to true, then the respective entry point executes, else it blocks.
+Because \whoop performs pair-wise analysis, it over-approximates this flag, and regards a pair of entry points as running concurrently, even if this is not the case.
+However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.
 a user might expect that a device can be accessed concurrently for performance).
-Thus, because using single-open device mode is considered bad practice, we burden the developer with disregarding any related false bugs.
+Because using single-open device mode is considered bad practice, we thus burden the developer with disregarding any related false bug reports.
 

### 2015-05-11 intro.tex
update in intro

@@
 
-Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system.
-According to the C language standard, data races also lead to undefined behavior, which can be the source of various compilation and optimization issues \PDComment{which is the best cite here?}.
-Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races.
-Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyze the program behavior of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs.
-The few of these techniques that attempt to deal with data races are usually either \emph{unsound} (i.e.
+Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of complex operating systems.
+Further, according to the C language standard, races result in undefined behavior, which can be the source of tricky compilation and optimization issues (as compilers typically assume race-freedom).
+Realistic concurrent programs also have an exponentially large state-space, which presents a serious challenge for implementing efficient race-checking techniques~\cite{musuvathi2008finding}.
+
+Techniques such as~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been successfully used to analyze device drivers, but mostly focus on generic sequential program properties and finding protocol bugs.
+Well-regarded Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races.
+Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.
 can miss real bugs) or \emph{imprecise} (i.e.
-can report false bugs).
-Furthermore, most previous techniques that attempt to reason about thread interleavings face significant scalability issues, because of the exponentially large state-space of realistic concurrent programs~\cite{musuvathi2008finding}.
+can report false bugs), and typically sacrifice precision for scalability.
+Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.
 
@@
 
-We then show that by exploiting the race-freedom guarantees that our symbolic analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
+We then show that by exploiting the race-freedom guarantees that our symbolic analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.

### 2015-05-11 related.tex
some fixes
update in intro

@@
-Using static analysis to detect data races is a promising alternative to dynamic concurrency bug finding, which is inherently limited by scalability and execution path coverage issues.
+Statically analyzing concurrent programs to detect data races is a promising alternative to dynamic techniques, which commonly face scalability and code coverage issues as they typically rely on a (controlled) scheduler to explore execution paths.
 Notable previous work includes the static analyzers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}.
@@
 
-Most related to our work are the static lock set analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
+Most related to our work are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
 Both tools, though, have significant limitations.

### 2015-05-11 related.tex
update in related work

@@
-Statically analyzing concurrent programs to detect data races is a promising alternative to dynamic techniques, which commonly face scalability and code coverage issues as they typically rely on a (controlled) scheduler to explore execution paths.
-Notable previous work includes the static analyzers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}.
-However, all these three tools heavily rely on user annotations, and thus require a lot of manual effort that limits their applicability.
-\textsc{Whoop}, on the other hand, does not require any source code modifications, and thus can be applied with zero effort.
+Statically analyzing concurrent programs to detect races is a promising alternative to dynamic analyzers, which commonly face code coverage issues as they rely on a (controlled) scheduler for exploring execution paths~\cite{musuvathi2008finding}.
+Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable examples of static race analyzers.
+However, both tools have limited applicability, as they heavily rely on user annotations.
+\textsc{Whoop}, on the other hand, does not require any source code modifications, and thus can be applied with minimal effort.
 
@@
 The authors also reported that Locksmith was unable to run on several large programs, showcasing its limited scalability.
+In contrast, \whoop is based on novel over-approximation techniques and uses modern SMT solvers to accelerate state-of-the-art concurrency bug-finders, such as \corral, for precise \emph{and} scalable analysis.

### 2015-05-11 related.tex
update in related work

@@
 
-Most related to our work are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
+Most related to our static lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
 Both tools, though, have significant limitations.
@@
 In contrast, \whoop is based on novel over-approximation techniques and uses modern SMT solvers to accelerate state-of-the-art concurrency bug-finders, such as \corral, for precise \emph{and} scalable analysis.
+
+Our work is not the first to exploit static analysis for achieving efficient race-checking.
+Examples include: \cite{choi2002efficient}, which combined static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combined a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combined inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.

### 2015-05-11 evaluation.tex
fix

@@
 \setlength{\tabcolsep}{0.3em}
-\caption{Program statistics and race-checking results from running \whoop and \corral on our benchmarks.}
+\caption{Program statistics and race-checking results from applying \whoop and \corral on our benchmarks.}
 \label{tab:stats}
@@
 All reported times are in seconds and averages of three runs.
-\corral was given a time budget of 8 hours (T.O.
+\corral was given a time budget of 10 hours (T.O.
 denotes a tool timeout) and a context-switch bound (csb) of 2, 5 and 9.

### 2015-05-11 evaluation.tex
fix

@@
 \textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs that \whoop identified (\#Racy Pairs); number of racy memory regions that \whoop reported (\#Racy MRs); and number of races that \corral discovered (\#Races Found) using a context-switch bound of 2.
-We did not discover new races with larger context switch bounds.
+We did not discover new races with larger context-switch bounds.
 This might mean that these drivers only have races that manifest with at least one or two context-switches, or that \corral hit its bounds before discovering a deeper race.
@@
 By default, \corral instruments context-switches (i.e.
-\texttt{yield}) in all visible operations (denoted by \yieldall in the table).
+\texttt{yield} statements) in all visible operations (denoted by \yieldall).
 \whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{whoop:bugfinding}), and also uses all the optimisations discussed in Section~\ref{whoop:optimizations}.
 
-A higher context switch bound results into deeper interleavings being explored and thus a larger sequentialized program.
+A higher context-switch bound results into deeper interleavings being explored and thus a larger sequentialized program.
 Intuitively, this means that we see even greater speedups using information from \whoop, when exploring deeper interleavings.

### 2015-05-11 whoop.tex
fix

@@
 
-Because \corral performs bounded verification both in terms of the number of allowed context switches and loop/recursion unrollings, it is inherently unsound; in other words, it can miss real bugs.
+Because \corral performs bounded verification both in terms of the number of allowed context-switches and loop/recursion unrollings, it is inherently unsound; in other words, it can miss real bugs.
 On the other hand, \corral is precise, and assuming a precise model of the program environment it will only report true bugs.
@@
 all bugs that are triggered by investigating a larger than $r$ recursion bound
-and larger than $k$ context switch bound. Let \corral then run the same $n$
+and larger than $k$ context-switch bound. Let \corral then run the same $n$
 amount of pairs, but this time with all \texttt{yield} statements in non-racy

### 2015-05-11 experiments/figures/plot.py
update in evaluation

@@ -21,7 +21,7 @@ x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 0, 278.0,     
 
 z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
 z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
-z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2,    107.1, 0, 1543.4,          36000, 24.5,        36000]
+z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 0, 1543.4,          36000, 24.5,        36000]
 
 y0 = [33.5, 169.3, 38.1, 11.1, 22.8, 35.0, 182.6, 81.5, 388.0, 271.0, 39.8,            1966.5, 12.9,        548.2, 36000]
 y1 = [47.0, 595.1, 45.6, 13.8, 130.3, 40.2, 263.7, 419.4, 390.8, 2746.6, 85.0,           36000, 15.2,        36000]

### 2015-05-12 experiments/figures/plot.py
update in evaluation

@@ -17,11 +17,11 @@ plt.rcParams['ytick.minor.size'] = 0
 
 x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
 x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0,           36000, 18.5,        36000]
-x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 0, 278.0,               36000, 24.3,        36000]
+x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0,               36000, 24.3,        36000]
 
 z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
 z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
-z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 0, 1543.4,          36000, 24.5,        36000]
+z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4,          36000, 24.5,        36000]
 
 y0 = [33.5, 169.3, 38.1, 11.1, 22.8, 35.0, 182.6, 81.5, 388.0, 271.0, 39.8,            1966.5, 12.9,        548.2, 36000]
 y1 = [47.0, 595.1, 45.6, 13.8, 130.3, 40.2, 263.7, 419.4, 390.8, 2746.6, 85.0,           36000, 15.2,        36000]

### 2015-05-12 whoop.tex
update

@@
 
-We also optimize the use of function pointers.
+%We also optimize the use of function pointers.
 LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site.

### 2015-05-12 experiments/figures/plot.py
update in evaluation

@@ -15,17 +15,17 @@ from matplotlib.backends.backend_pdf import PdfPages
 plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
-x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0,              906.0, 15.9,        359.8, 36000]
-x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0,           36000, 18.5,        36000]
-x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0,               36000, 24.3,        36000]
+x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0,              906.0, 15.9, 359.8, 36000]
+x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0,           36000, 18.5, 5664.5, 36000]
+x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0,               36000, 24.3, 36000, 36000]
 
-z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1,        474.7, 36000]
-z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5,        2953.5, 36000]
-z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4,          36000, 24.5,        36000]
+z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1, 474.7, 36000]
+z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5, 12677.3, 36000]
[…]

### 2015-05-12 whoop.tex
update in optimisations

@@
 
-A common practice for statically analyzing software is to ``close'' the environment by abstracting away the low level implementation details.
-Any abstract environmental model, though, can and will ultimately result in decreased precision and thus false positives.
-This is one of the main limitations of statically analyzing real life software that use complex external APIs.
-In our case, we developed a simple environmental model for the Linux kernel that consists of stub header files.
-Because we only focus on lockset-based properties during the race-checking phase of \whoop, we can get away with abstracting a lot of the underlying functionality.
-Our model, though, can still result in a lot of false positives.
+To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel interacts with its drivers.
+This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then exploit to make \whoop more precise.
 
-To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various drivers.
-This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then use for increased precision.
+In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding kernel-imposed serialization.
+The Linux kernel can serialize calls to entry points, thus forcing them to not run concurrently with each other.
+An an example, a large number of networking and ethernet entry points are mutually serialized with RTNL, a network-specific kernel lock.
+\whoop exploits this knowledge and does not create pairs for entry points that are serialized.
 
-In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding kernel imposed locks.
-The Linux kernel is able to serialize a number of entry point calls, which thus can never run concurrently with each other.
-An an example, a large number of networking and ethernet entry points are mutually serialized with RTNL, a network-specific kernel lock.
-\whoop will not create pairs for entry points that cannot run concurrently.
-
-Another optimization is to soundly reduce the checked memory locations.
-If a memory location is accessed only by one entry point in the pair, then it is safe to not check it.
-This can potentially decrease the amount of generated candidate invariants and speed up the verification process.
-
-Modeling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux.
+Modeling kernel methods is another opportunity to increase precision, but requires Linux kernel expertise.
 For example, the \texttt{register\_netdev()} function registers a network driver with the kernel.
@@
 Modeling this functionality can reduce false positives when race-checking accesses before the driver's registration.
+
+Another optimization \whoop uses is to soundly reduce the checked memory locations.
+If a memory location is accessed only by one entry point in a pair, then it is safe to not check this location for this pair.
+This reduces the burden to the SMT solver and speed up the verification process.
 
@@
 
-Finally, we assume that the formal parameters of an entry point do not alias, and thus do not race.
+Finally, we assume that the formal parameters of an entry point do not alias, and thus cannot race.
 This is a potentially unsound feature that can be turned off using a command line option.
@@
 
-Finally, another limitation of our approach is the need for precise environmental modeling.
-Enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel.
-We argue that further increasing the precision of our models is orthogonal to the contributions of this paper.
-Moreover, even if the symbolic lockset analysis can be imprecise, \whoop can still use any race-freedom guarantees to significantly speedup a precise bug-finders, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.
+Statically analyzing system-level software, such as drivers, requires to ``close'' the environment by abstracting away the low level implementation details.
+In this work, we developed a simple environmental model for the Linux kernel that consists of (nondeterministic) stub functions.
+A limitation of our model is that it can, and will, ultimately result in decreased precision and, thus, false positives.
+However, because we currently only focus on finding data races, we can get away with over-approximating a lot of the underlying Linux kernel functionality, without losing too much precision.
+Enriching our model with domain-specific knowledge to make it more precise is an ongoing manual effort, but requires Linux expertise.
+We argue that further increasing the precision is orthogonal to the contributions of this paper.
+Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.

### 2015-05-12 related.tex
fix

--- 2015-05-11 related.tex
@@
 In contrast, \whoop is based on novel over-approximation techniques and uses modern SMT solvers to accelerate state-of-the-art concurrency bug-finders, such as \corral, for precise \emph{and} scalable analysis.
+
+Our work is not the first to exploit static analysis for achieving efficient race-checking.
+Prior work includes: \cite{choi2002efficient}, which combined static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combined a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combined inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.

### 2015-05-12 whoop.tex
Progress

@@
 \in \procedures{T}$, denotes the starting point for execution of a
-template by a thread.  A procedure $P \in \procedures{T}$ is
+template by a thread.  A procedure in $\procedures{T}$ is
 represented by a control flow graph of basic blocks, where each block
@@
 execution point, a thread may execute its current statement, unless
-that statement has the form $\mutexlock(m)$ and mutex $m$ is already held
+that statement has the form $\mutexlock{m}$ and mutex $m$ is already held
 by some thread.  Executing a statement causes the state of the thread,
@@
 
-\medskip\noindent\textbf{Lockset instrumentation} \ADComment{This is as far as I have gotten with the writing.
-I now want to explain formally and somewhat abstractly how our lockset instrumentation works.}
+\medskip\noindent\textbf{Lockset instrumentation }
+Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibly that $S=T$.
+We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is
+an instance of $T$.
+
+We achieve this in the following manner.  Given an arbitrary symbol, $i$, we define the \emph{instrumentation} of a
+template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to instrumentation, \emph{renaming} and \emph{lockset instrumentation}.
+
+Renaming is simple: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+The purpose of renaming is to ensure that when we analyse a pair of templates, $S$ and $T$, we ensure that $S$ and $T$ execute distinct procedures and operate on distinct private data.
+Shared variables and mutexes are \emph{not} renamed, because these are indeed shared across templates.
+
+\ADComment{Next up: describe lockset instrumentation with respect to Figure~\ref{ }.
+This involves explaining what the sets $W$, $R$, $\mathit{LS}$ and $\mathit{CLS}$ are, and explaining how the instrumentation works.}
+
+\ADComment{Finally, explain the main program that we check, which initializes the sets and runs the mains of each template.}
+
+\ADComment{Last of all, pay attention to termination.}
 

### 2015-05-12 whoop.tex
Getting longer and longer

@@
 We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is
-an instance of $T$.
+an instance of $T$, in the presence of arbitrarily many further threads.
 
-We achieve this in the following manner.  Given an arbitrary symbol, $i$, we define the \emph{instrumentation} of a
+The first phase in achieving this is to \emph{instrument} a thread template to enable lockset analysis.  Given an arbitrary symbol, $i$, we define the instrumentation of a
 template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to instrumentation, \emph{renaming} and \emph{lockset instrumentation}.
@@
 
-\ADComment{Next up: describe lockset instrumentation with respect to Figure~\ref{ }.
-This involves explaining what the sets $W$, $R$, $\mathit{LS}$ and $\mathit{CLS}$ are, and explaining how the instrumentation works.}
-
-\ADComment{Finally, explain the main program that we check, which initializes the sets and runs the mains of each template.}
-
-\ADComment{Last of all, pay attention to termination.}
-
 \begin{figure}
@@
 \textbf{Original statement} & \textbf{Instrumented statement} \\\hline
-$x = e;$                & $\instrument{x}{i} = \instrument{e}{i}$;  \\\hline
-$x = f(\overline{e});$ & $\instrument{x}{i} = \instrument{f}{i}(\instrument{\overline{e}}{i});$ \\\hline
 $s = e;$ & $\hasbeenwritten{i} = \hasbeenwritten{i} \cup \{ s \};$ \\
@@
 
-\ADComment{Finally, try to set up some kind of theorem giving the guarantee we think we have.}
+Lockset instrumentation first requires the introduction of sets to track those variables that have been read from and written to, the mutexes that have been consistently used to protect each variable, and the set of current locks held by a thread.
+To this end, when instrumenting template $T$ with respect to symbol $i$, we add a sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track those shared variables that have been read from and written to, respectively, by the thread executing $T$.
+We add a current lock set, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread, and for each shared variable $s \in \sharedvars$ we add a lock set, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
+
+The statements of each procedure in $\instrument{T}{i}$ that manipulate shared variables and mutexes are then instrumented to take account of lock sets.
+This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.
+A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lock set for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).
+Notice that in the instrumented form, the shared variable $s$ is no longer assigned to.
+We discuss the reasons for this when we describe \emph{shared state abstraction} below.
+A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command applied to the receiving private variable $x$; again this relates to shared state abstraction and is discussed below.
+Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lock set, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
+
+\medskip\noindent\textbf{Shared state abstraction }
+\ADComment{Explain why the shared variables are gone.}
+
+\medskip\noindent\textbf{Sequentialisation }
+\ADComment{Show what we actually check.}
+
+\ADComment{Now state the theoretical result, paying attention to termination.}
 

### 2015-05-12 background.tex
Edits to code example.

@@
 
-\begin{lstlisting}[caption = Data races in a Linux driver that we discovered using \whoop., label = fig:data_race_example]
+\begin{figure}[t]
+\begin{lstlisting}
 static loff_t nvram_llseek(struct file *file,
@@
   switch (origin) {
-    case 0:
-      break;
-    case 1:
-      offset += file->f_pos; // racy
-      break;
-    case 2:
-      offset += nvram_len;
-      break;
-    default:
-      offset = -1;
+    case 0: break;
+    case 1: offset += file->f_pos; break; // racy
+    case 2: offset += nvram_len; break;
+    default: offset = -1;
   }
@@
 \end{lstlisting}
+\caption{Data races in a Linux driver that we discovered using \whoop}\label{fig:data_race_example}
+\end{figure}
 
@@
 source code that cannot be executed by more than one thread simultaneously.
-Example~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Example~\ref{fig:data_race_example}.
+Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}.
 The example is tricky, because the return statement can potentially race on the \texttt{f\_pos} field.
 To fix this issue, we store the result in a temporary variable, inside the critical section.
-
-\begin{lstlisting}[caption = Using a lock to eliminate the data races., label = fig:lock_example]
-static loff_t nvram_llseek(struct file *file,
-    loff_t offset, int origin) {
-  mutex_lock(&nvram_mutex); // lock
-  switch (origin) {
-    case 0:
-      break;
-    case 1:
-      offset += file->f_pos;
-      break;
-    case 2:
-      offset += nvram_len;
-      break;
-    default:
-      offset = -1;
-  }
-  if (offset < 0)
-    return -EINVAL;
-  file->f_pos = offset;
-  loff_t res = file->f_pos; // store result
-  mutex_unlock(&nvram_mutex); // unlock
-  return res;
-}
-\end{lstlisting}
 
@@
 We also found that 31\% of the drivers use neither locking nor atomic primitives, which suggests that they are either simple (non-concurrent) drivers or use some other lock-free synchronization primitive.
+
+\begin{figure}[t]
+\begin{lstlisting}
+static loff_t nvram_llseek(struct file *file,
+    loff_t offset, int origin) {
+  mutex_lock(&nvram_mutex); // lock
+  switch (origin) {
+    case 0: break;
+    case 1: offset += file->f_pos; break;
+    case 2: offset += nvram_len; break;
+    default: offset = -1;
+  }
+  if (offset < 0)
+    return -EINVAL;
+  file->f_pos = offset;
+  loff_t res = file->f_pos; // store result
+  mutex_unlock(&nvram_mutex); // unlock
+  return res;
+}
+\end{lstlisting}
+\caption{Using a lock to eliminate the data races}\label{fig:lock_example}
+\end{figure}
 

### 2015-05-12 background.tex
Worked with Pantazis on the theory part of the paper and the new proposed structure of the implementation section.

@@
 
-\subsection{Reasoning about Thread Interleavings}
-\label{bg:happensbefore}
-
-\ADComment{I am unsure as to whether we need this background, but I'm undecided as I'm still getting on top of the rest of the material.}
-
-A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during the execution of a concurrent program.
-Happens-before can be defined as follows:
-
-\begin{definition}
-\label{definition:datarace}
-Let $E_1$ and $E_2$ be two events during a program's concurrent execution.
-If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronization mechanism (e.g.
-locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
-\end{definition}
-
-Happens-before is transitive, i.e.
-for any three events $E_1$, $E_2$ and $E_3$, if $E_1 \rightarrow E_2$ and $E_2 \rightarrow E_3$, then $E_1 \rightarrow E_3$.
-If two events $E_1$ and $E_2$ occur in different threads, and neither $E_1 \rightarrow E_2$, nor $E_2 \rightarrow E_1$ holds, then $E_1$ and $E_2$ happen concurrently, which indicates a data race if the two events access the same shared resource and at least one of the accesses is a modifier.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=1\linewidth]{img/happens_before.pdf}
-\caption{Applying happens-before in two concurrently running threads}
-\label{fig:happens_before}
-\end{figure}
-
-Figure~\ref{fig:happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points.
-Each of the three statements $\{A, B, C\}$ in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation.
-The same applies for each of the four statements $\{D, E, F, G\}$ in thread 2.
-Statement $C$ in thread 1 releases the lock, and subsequently allows $D$ in thread 2 to acquire the same lock.
-Thus, $C \rightarrow D$.
-This means that for this specific thread interleaving the set of statements $\{A, B, C\}$ happens before the set of statements $\{D, E, F, G\}$.
-
-The above example also showcases the main limitation of happens-before.
-If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements $B$ and $G$, because $B \rightarrow G$ in this specific schedule.
-Thus, a tool based on happens-before requires a scheduler that produces as many different thread interleavings as possible to increase execution path coverage~\cite{savage1997eraser}.
-Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs.
-Moreover, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}, a problem known as state space explosion.
-Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.
-
-\subsection{Reasoning about Locksets}
+\subsection{Lockset analysis}
 \label{bg:lockset}
 
-An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behavior of concurrent programs.
-Lockset analysis was proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution.
+Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.
+The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution.
 If that lockset ever becomes empty, the analysis reports a potential data race on that memory location.
-\ADComment{This could be better explained, I think.} This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
+This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
 
@@
 
-In contrast to analyses based on happens-before, lockset analysis has the potential to scale well, because it does not need to reason about all possible thread interleavings to achieve good code coverage.
-The technique, though, suffers from imprecision (i.e.
-can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
+In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{TODO} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.
+The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
 Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler.

### 2015-05-12 whoop.tex
Worked with Pantazis on the theory part of the paper and the new proposed structure of the implementation section.

@@
 \subsection{Symbolic pairwise lockset analysis}\label{sec:symbolicpairwise}
+
+\ADComment{Top level question about this section: is the motivation super-clear?  Is it obvious why we go for the pairwise approach?}
 
@@
 
-\medskip\noindent\textbf{Data races } \ADComment{We now need to define
-what a data race is.  This could be done using happens-before.  I am undecided as to whether
-we should refer back to the happens-before material in the background section, or whether
-we should move that material here, so that we explain what happens-before means, and then
-explain what a data races is.  My preference is for the latter: doing it all here.}
-
 \medskip\noindent\textbf{Lockset instrumentation }
@@
 \medskip\noindent\textbf{Shared state abstraction }
-\ADComment{Explain why the shared variables are gone.}
+Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account
+for possible side-effects due to other threads that are running concurrently.  The instrumentation
+of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$,
+a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command,
+which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in
+this fashion, it is possible to completely dispense with the shared variables after instrumentation
+has been performed.  As a result, when instrumenting a shared variable write, the effect of the write
+is not explicitly modelled.
 
 \medskip\noindent\textbf{Sequentialisation }
-\ADComment{Show what we actually check.}
+The psuedocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
+analyse in order to prove race-freedom for a pair of thread templates $T$ and $U$.
+Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding
+$\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
+First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are
+initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and
+$\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures
+of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and
+$\mainforthread{\instrument{U}{j}}$ are then executed in turn.  Finally, an assertion checks
+for consistent use of mutexes: if a shared variables is written during execution of $\instrument{T}{i}$
+and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$
+and $\lockset{s}{j}$ must contain a common mutex.
 
-\ADComment{Now state the theoretical result, paying attention to termination.}
+\begin{figure}
+\begin{tabular}{l}
+
+$\currentlockset{i} = \emptyset;$ $\hasbeenread{i} = \emptyset;$ $\hasbeenwritten{i} = \emptyset;$ \\
+$\currentlockset{j} = \emptyset;$ $\hasbeenread{j} = \emptyset;$ $\hasbeenwritten{j} = \emptyset;$ \\
+\texttt{for} $s \in \sharedvars$ \texttt{do} $\lockset{s}{i} = \mutexes;$ $\lockset{s}{j} = \mutexes;$ \medskip
+\\
+
+$\mainforthread{\instrument{T}{i}}();$ \\
+$\mainforthread{\instrument{U}{j}}();$ \medskip\\
+
+\texttt{assert} $\forall s \in \sharedvars \; .$ \\
+
+$\quad s \in \hasbeenwritten{i} \cap (\hasbeenread{j} \cup \hasbeenwritten{j}) \vee s \in \hasbeenwritten{j} \cap (\hasbeenread{i} \cup \hasbeenwritten{i}) \implies$ \\
+
+$\quad\quad \lockset{s}{i} \cap \lockset{s}{j} \neq \emptyset;$ \\
+
+\end{tabular}
+\caption{The sequential program to be analysed in order to prove race-freedom for a pair of thread templates}
+\label{fig:sequentialization}
+\end{figure}
+
+\medskip\noindent\textbf{Soundness }
+We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct,
+it is impossible for a thread executing template $T$ to race with a thread executing template $U$,
+under the assumption that the threads are guaranteed to terminate.
+Let us assume that the program of Figure~\ref{fig:sequentialization} is correct,
+and suppose (by way of contradition) that a thread executing $T$ can in fact race with a thread
+executing $U$, on some shared variable $s$.  By our hypothesis that the program
+of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate,
+the assertion checked at the end of the program guarantees at least one mutex, $m$ say,
+belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset
+(and according to the manner in which shared accesses are instrumented in Figure~\ref{ }),
+this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
+As a result, $m$ must be unlocked and locked between the two accesses, which
+contradicts that the pair of accesses is racing.
+
+We require the threads to terminate becauase in the presence of non-termination the assertion
+at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis
+problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
+In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyse
+in our experimental evaluation (Section~\ref{evaluation}) are terminating.
 
 \subsection{Implementation in Whoop}\label{sec:whoopimplementation}
+
+\ADComment{Acknowledge very clearly that the simple concurrent programming model of Section~\ref{...} is deliberately idealistic.
+In practice, we do not know upfront what the entry points for a driver are, we don't have named set of locks, and rather than a given set of named shared variables, we have arbitrary memory accesses via pointers.
+We now explain how we have taken the conceptual ideas from Section~\ref{...} and used them to build a practical tool.
+
+\ADComment{Explain briefly the architecture, and specifically introduce Chauffer, SMACK and Boogie.}
+
+\ADComment{Explain how we identify entry points (Chauffer)}
+
+\ADComment{Explain how pointers get handled (SMACK) -- be brief, talk about memory regions, don't show Boogie code.}
+
+\ADComment{Explain how we identify mutexes (inside our Boogie-based analysis)}
+
+\ADComment{Explain how we avoid quantifiers in performing race analysis}
+
+\ADComment{Briefly talk about how we tried inlining and do summarisation instead}
+
+\ADComment{Discuss optimizations and assumptions.  Bring these to before the Corral stuff.}
+
+\ADComment{Note, unrelated to structure: put mention of interrupts in limitations}
 

### 2015-05-12 background.tex
update in background

@@
 Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput).
-In device drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core.
+In drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core.
 Likewise, performance is also important: e.g.
@@
 \end{lstlisting}
-\caption{Data races in a Linux driver that we discovered using \whoop}\label{fig:data_race_example}
+\caption{Example of data races in the generic\_nvram Linux char driver}\label{fig:data_race_example}
 \end{figure}
@@
 We believe that our focus in locks is justified, because the majority of Linux drivers still use locks.
-As a sanity check, we used a simple grep command and found that 68\% of the drivers from the 4.0 Linux kernel distribution use locks, versus 13\% that use atomic operations.
+To check this, we used a simple grep command and found that 68\% of the drivers from the 4.0 Linux kernel distribution use locks, versus 13\% that use atomic operations.
 We also found that 31\% of the drivers use neither locking nor atomic primitives, which suggests that they are either simple (non-concurrent) drivers or use some other lock-free synchronization primitive.
@@
 \end{lstlisting}
-\caption{Using a lock to eliminate the data races}\label{fig:lock_example}
+\caption{Using a lock to eliminate the data races in the previous example}\label{fig:lock_example}
 \end{figure}
 
-\subsection{Lockset analysis}
+\subsection{Lockset Analysis}
 \label{bg:lockset}
@@
 
-In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{TODO} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.
+In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.
 The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.

### 2015-05-12 whoop.tex
update in background
updates

@@
-We now present \emph{symbolic pairwise lockset analysis}, a novel method
-for data race analysis in concurrent device drivers.  In
-Section~\ref{sec:symbolicpairwise} we describe how the approach works
-in a semi-formal manner, with respect to a simple concurrent
-programming model.  In Section~\ref{sec:whoopimplementation} we
-explain how we have implemented symbolic pairwise lockset analysis in
-a practical tool, Whoop, that can be applied directly to driver source
-code, to check whether pairs of driver entry points can race with one
-another.
+We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.
+In Section~\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.
+In Section~\ref{sec:whoopimplementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code, to check whether pairs of driver entry points can race with one another.
 
-Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the Whoop technique has value as a stand-alone analyser, and that results from Whoop can be exploited to significantly boost performance of a more precise symbolic analysis for concurrency offered by the CORRAL tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{whoop:bugfinding}.
+Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency offered by the \corral tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{whoop:bugfinding}.
 
-\subsection{Symbolic pairwise lockset analysis}\label{sec:symbolicpairwise}
+\subsection{Symbolic Pairwise Lockset Analysis}\label{sec:symbolicpairwise}
 
@@
 
-\medskip\noindent\textbf{Concurrent programming model }
+\medskip\noindent\textbf{Concurrent Programming Model }
 %
@@
 
-\subsection{Implementation in Whoop}\label{sec:whoopimplementation}
+\subsection{Implementation in \whoop}\label{sec:whoopimplementation}
 
@@
 In practice, we do not know upfront what the entry points for a driver are, we don't have named set of locks, and rather than a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from Section~\ref{...} and used them to build a practical tool.
+We now explain how we have taken the conceptual ideas from Section~\ref{...} and used them to build a practical tool.}
 
@@
 
-\subsection{Accelerating Bug-Finding}
-\label{whoop:bugfinding}
-
-\whoop is a sound but potentially imprecise static data race analyzer.
-While developers do appreciate soundness, it is more important to ensure they are not overwhelmed by countless reports of false bugs.
-Hence, to improve precision, we enable an external precise bug-finder for concurrent programs to be plugged into \whoop.
-The race-related information generated by \whoop can then be used to accelerate the bug-finder.
-Currently, we support \corral~\cite{lal2012corral, lal2014powering}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers.
-\corral accepts Boogie IVL and hence it was easy to integrate into our toolchain (see Figure~\ref{fig:whoop} -- C).
-Our technique, though, is general and capable of accelerating any arbitrary bug-finder for concurrent programs.
-
-\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors.
-It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found.
-\corral performs bounded exploration of a concurrent program in two steps.
-First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.
-Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program.
-It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.
-
-Because \corral performs bounded verification both in terms of the number of allowed context-switches and loop/recursion unrollings, it is inherently unsound; in other words, it can miss real bugs.
-On the other hand, \corral is precise, and assuming a precise model of the program environment it will only report true bugs.
-\whoop takes advantage of this precision to report only feasible data races.
-
-\noindent
-\textbf{Sound Partial-Order Reduction}\xspace\xspace By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before each \emph{visible operation} (i.e., synchronization operations and shared memory accesses) of every thread.
-It then leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
-
-Our approach to accelerating \corral is simple and yet effective: we simplify
-the sequentialization that \corral performs by instrumenting \texttt{yield}
-statements only at shared memory accesses that \whoop reported as potentially
-racy. This effectively acts as a partial-order reduction that can significantly
-reduce the number of interleavings that \corral needs to explore, and thus
-increase performance and scalability.
-
-\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
-about entry point pairs, or could we prove it just for any pair?}\PDComment{Lets also see what Ally thinks about this part.}
-We now show that our partial-order reduction is sound: it does not miss any more
-bugs than what \corral would miss even without taking advantage of the \whoop
-race-related information. Let \corral run an $n$ amount of entry point pairs
-with yields in all visible operations. Let this configuration be known as $C$.
-Each entry point pair is running in a separate \corral process. \corral will
-miss all bugs that are triggered by more than 3 bugs running concurrently and
-all bugs that are triggered by investigating a larger than $r$ recursion bound
-and larger than $k$ context-switch bound. Let \corral then run the same $n$
-amount of pairs, but this time with all \texttt{yield} statements in non-racy
-shared memory accesses that are removed according to information from \whoop. Let this configuration be
-known as $W+C$. Because these shared memory accesses cannot race, performing a
-\texttt{yield} will not introduce any new program behaviors. Thus, \corral will
-find the same bugs using the configuration $W+C$, as using the configuration
-$C$.
-
-We have implemented two different levels of context-switch instrumentation granularity:
-
-\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
-finds that a pair of entry points cannot race, then it will only instrument
-context-switches in synchronization points of this pair; else if \whoop finds
-that a pair of entry points can race, then it will instrument context-switches
-in all visible operations of this pair.
-
-\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
-in all synchronization points, but only instruments context-switches in memory
-regions that can potentially race (regardless if a pair has not been fully
-verified as race-free). In our experiments (see Section~\ref{evaluation}),
-\yieldmr performs significantly faster than \yieldcoarse.
-
-\noindent
-\textbf{Race-Checking Instrumentation}\xspace\xspace In this paper, we treat races as bugs. To detect them, we employ a simple, but effective, encoding of data race checks within \corral.
-For example, whenever there is a write access to a shared variable, such as
-\texttt{\$M.13 := 0}, we instrument the Boogie IVL code to check for write-write
-races as follows:
-%
-\begin{boogie}
-$M.13 := 0;        // original write
-yield;             // allow for a context-switch
-assert $M.13 == 0; // check written value
-\end{boogie}
-%
-Likewise, for a read access to a shared variable, such as \texttt{\$p0 :=
-\$M.13}, we instrument the Boogie IVL code to check for read-write races as
-follows:
-%
-\begin{boogie}
-$p0 := $M.13;        // original read
-yield;               // allow for a context-switch
-assert $p0 == $M.13; // check read value
-\end{boogie}
-
-Note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
-An advantage of this instrumentation is that it can be used as a precise feasibility check for the potential data races that \whoop reports.
-
 \subsection{Optimizations and Assumptions}

### 2015-05-12 corral.tex
updates

\whoop is a sound but potentially imprecise static data race analyzer.  While developers do appreciate soundness, it is more important to ensure they are not overwhelmed by countless reports of false bugs.  Hence, to improve precision, we enable an external precise bug-finder for concurrent programs to be plugged into \whoop. The race-related information generated by \whoop can then be used to accelerate the bug-finder.  Currently, we support \corral~\cite{lal2012corral, lal2014powering}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers. \corral accepts Boogie IVL and hence it was easy to integrate into our toolchain (see Figure~\ref{fig:whoop} -- C). Our technique, though, is general and capable of accelerating any arbitrary bug-finder for concurrent programs.

\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors. It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found. \corral performs bounded exploration of a concurrent program in two steps. First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}. Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program. It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.

Because \corral performs bounded verification both in terms of the number of allowed context-switches and loop/recursion unrollings, it is inherently unsound; in other words, it can miss real bugs. On the other hand, \corral is precise, and assuming a precise model of the program environment it will only report true bugs. \whoop takes advantage of this precision to report only feasible data races.

\noindent
\textbf{Sound Partial-Order Reduction}\xspace\xspace By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before each \emph{visible operation} (i.e., synchronization operations and shared memory accesses) of every thread. It then leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.

Our approach to accelerating \corral is simple and yet effective: we simplify
the sequentialization that \corral performs by instrumenting \texttt{yield}
statements only at shared memory accesses that \whoop reported as potentially
racy. This effectively acts as a partial-order reduction that can significantly
reduce the number of interleavings that \corral needs to explore, and thus
increase performance and scalability.

\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
about entry point pairs, or could we prove it just for any pair?}\PDComment{Lets also see what Ally thinks about this part.}
We now show that our partial-order reduction is sound: it does not miss any more
bugs than what \corral would miss even without taking advantage of the \whoop
race-related information. Let \corral run an $n$ amount of entry point pairs
with yields in all visible operations. Let this configuration be known as $C$.
Each entry point pair is running in a separate \corral process. \corral will
miss all bugs that are triggered by more than 3 bugs running concurrently and
all bugs that are triggered by investigating a larger than $r$ recursion bound
and larger than $k$ context-switch bound. Let \corral then run the same $n$
amount of pairs, but this time with all \texttt{yield} statements in non-racy
shared memory accesses that are removed according to information from \whoop. Let this configuration be
known as $W+C$. Because these shared memory accesses cannot race, performing a
\texttt{yield} will not introduce any new program behaviors. Thus, \corral will
find the same bugs using the configuration $W+C$, as using the configuration
$C$.

We have implemented two different levels of context-switch instrumentation granularity:

\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
finds that a pair of entry points cannot race, then it will only instrument
context-switches in synchronization points of this pair; else if \whoop finds
that a pair of entry points can race, then it will instrument context-switches
in all visible operations of this pair.

\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
in all synchronization points, but only instruments context-switches in memory
regions that can potentially race (regardless if a pair has not been fully
verified as race-free). In our experiments (see Section~\ref{evaluation}),
\yieldmr performs significantly faster than \yieldcoarse.

\noindent
\textbf{Race-Checking Instrumentation}\xspace\xspace In this paper, we treat races as bugs. To detect them, we employ a simple, but effective, encoding of data race checks within \corral.
For example, whenever there is a write access to a shared variable, such as
\texttt{\$M.13 := 0}, we instrument the Boogie IVL code to check for write-write
races as follows:
%
\begin{boogie}
$M.13 := 0;        // original write
yield;             // allow for a context-switch
assert $M.13 == 0; // check written value
\end{boogie}
%
Likewise, for a read access to a shared variable, such as \texttt{\$p0 :=
\$M.13}, we instrument the Boogie IVL code to check for read-write races as
follows:
%
\begin{boogie}
$p0 := $M.13;        // original read
yield;               // allow for a context-switch
assert $p0 == $M.13; // check read value
\end{boogie}

Note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.  An advantage of this instrumentation is that it can be used as a precise feasibility check for the potential data races that \whoop reports.

### 2015-05-12 experiments/figures/plot.py
update in evaluation

@@ -21,7 +21,7 @@ x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0, 
 
 z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1, 474.7, 36000]
 z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5, 12677.3, 36000]
-z2 = [132.1, 0, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4,          36000, 24.5, 36000, 36000]
+z2 = [132.1, 22514.1, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4,          36000, 24.5, 36000, 36000]
 
 y0 = [33.5, 169.3, 38.1, 11.1, 22.8, 35.0, 182.6, 81.5, 388.0, 271.0, 39.8,            1966.5, 12.9, 548.2, 36000]
 y1 = [47.0, 595.1, 45.6, 13.8, 130.3, 40.2, 263.7, 419.4, 390.8, 2746.6, 85.0,           36000, 15.2, 14469.0, 36000]

### 2015-05-12 background.tex
fix

@@
 Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput).
-In drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core.
-Likewise, performance is also important: e.g.
-network drivers should be able to process incoming packages in a timely manner.
+In drivers, responsiveness is important (even if there is only a single core), because entry points of distinct processes should run concurrently so that no single process hogs the core.
+Likewise, performance is also important: e.g.\ network drivers should be able to process incoming packages in a timely manner.
 
@@
 The most common technique for eliminating data races is using a \emph{locking} mechanism.
-Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e.
-source code that cannot be executed by more than one thread simultaneously.
+Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e.\ source code that cannot be executed by more than one thread simultaneously.
 Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}.
@@
 Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read and write operations, in this work we focus on locks.
-However, we still treat lock-free operations (e.g.
-atomics) soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads.
+However, we still treat lock-free operations (e.g.\ atomics) soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads.
 This is sound but can potentially introduce false positives.

### 2015-05-12 intro.tex
fix

@@
 Well-regarded Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races.
-Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.
-can miss real bugs) or \emph{imprecise} (i.e.
-can report false bugs), and typically sacrifice precision for scalability.
+Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability.
 Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.
@@
 
-\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g.
-Windows) as well as other concurrent programs with similar structure (e.g.
-filesystems).
+\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g.\ Windows) as well as other concurrent programs with similar structure (e.g.\ filesystems).
 

### 2015-05-12 whoop.tex
fix

@@
 
-Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency offered by the \corral tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{whoop:bugfinding}.
+Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency offered by the \corral tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{corral}.
 
@@
 The summarization that we perform to achieve scalability is presented in Section~\ref{whoop:summarization}.
-After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{whoop:bugfinding})
+After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{corral})
 
@@
 This function acts as the main entry point to the driver after the latter has been registered with the kernel.
-This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{whoop:method} and~\ref{whoop:bugfinding}).
+This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{whoop:method} and~\ref{corral}).
 The output is a file that contains information regarding the entry points and a file with the rewritten program.
@@
 We argue that further increasing the precision is orthogonal to the contributions of this paper.
-Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.
+Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{corral} and~\ref{evaluation}.

### 2015-05-12 background.tex
fix

@@
 Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput).
-In drivers, responsiveness is important (even if there is only a single core), because entry points of distinct processes should run concurrently so that no single process hogs the core.
-Likewise, performance is also important: e.g.\ network drivers should be able to process incoming packages in a timely manner.
+In drivers, responsiveness is important (even if there is only a single core available), because entry points of distinct processes should run concurrently so that no single process hogs the core.
+Likewise, performance is important, as drivers should be able to timely process incoming requests (e.g.\ network packages).
 
 Modern operating systems address the demand for responsiveness and performance by providing multiple sources of concurrency~\cite{corbet2005linux}: a kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
-All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:
+All these possibilities can easily lead to \emph{data races}, which can be defined as follows:
 

### 2015-05-12 whoop.tex
fix

@@
 
-Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency offered by the \corral tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{corral}.
+Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in Section~\ref{corral}.
 

### 2015-05-12 whoop.tex
update in main section

@@
 
-Symbolic pairwise lockset analysis works by considering, for a given device driver, every pair of entry points that can potentially execute concurrently.
-For each such pair, we use symbolic verification to check whether it is possible for the pair to race, accounting for the effects of additional entry points by treating the driver shared state abstractly so that when an entry point reads from the shared state, a nondeterministic value is returned.
-Symbolic verification of an entry point pair works by instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and attempting to verify a sequential program that executes the instrumented entry points in sequence, and subsequently asserts that the lockset for each shared variable is non-empty.
+Symbolic pairwise lockset analysis works by considering, for a given driver, every pair of entry points that can potentially execute concurrently.
+For each such pair, we use symbolic verification to check whether it is possible for the pair to race, modeling the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
+Symbolic verification of an entry point pair works by (i) instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts that the lockset for each shared variable is non-empty.
 Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.
@@
 when reading from the shared state) and summarization, which can be sources of false positives.
-Further, the tool does not currently check for dynamically created locks or for locks that are provided by external libraries, although the later could be easily addressed by providing a mechanism for users to declare custom locks.
+Furthermore, the tool does not currently check for dynamically created locks or for locks that are provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.
 

### 2015-05-12 whoop.tex
update in main section

@@
 
-Symbolic pairwise lockset analysis works by considering, for a given driver, every pair of entry points that can potentially execute concurrently.
-For each such pair, we use symbolic verification to check whether it is possible for the pair to race, modeling the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Symbolic verification of an entry point pair works by (i) instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts that the lockset for each shared variable is non-empty.
+Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.
+For each such pair, we use symbolic verification to check whether it is possible for the pair to race in a shared variable.
+We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
+
+Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts that the lockset for each shared variable is non-empty.
 Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.

### 2015-05-12 whoop.tex
update in main section

@@
 %
-We consider a simple concurrent programming model in which an
-unbounded number of threads execute a set of pre-defined \emph{thread
-  templates}.  At any given point of execution a certain number of
-threads are active, each thread executing a particular template.  In
-the context of device drivers, a thread template corresponds to a
-driver entry point, and multiple instances of the same thread template
-may execute concurrently, just as multiple invocations of a single
-driver entry point may be concurrent.  Further threads may be launched
-at any point during execution; in the context of device drivers this
-corresponds to the OS invoking additional driver entry points.  For
-ease of presentation only, our model does not feature aggregate data
-types, pointers or dynamic memory allocation.  These \emph{are}
-handled by our implementation, and in
-Section~\ref{sec:whoopimplementation} we discuss interesting practical
-issues arising from the handling of a full-blown language.
+We consider a simple concurrent programming model in which an unbounded number of threads execute a set of pre-defined \emph{thread templates}.
+At any given point of execution a certain number of threads are active, each thread executing a particular template.
+In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
+Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.
+For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
+These \emph{are} handled by our implementation, and in Section~\ref{sec:whoopimplementation} we discuss interesting practical issues arising from the handling of a full-blown language.
 
@@
 \begin{tabular}{lp{5.5cm}}
-\textbf{Statement form} & \textbf{Notes} \\\hline
-$x = e;$ & private assignment, where $x \in \variables{T}$ and $e$ is an expression over $\variables{T}$ \\\hline
-$x = f(\overline{e});$ & procedure call, where $x \in \variables{T}$, $\overline{e}$ is a sequence of expressions over $\variables{T}$, $f$ is the name of a procedure in $\procedures{T}$ \\\hline
-$s = e;$ & shared write, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$ \\\hline
-$x = s;$ & shared read,  where $x \in \variables{T}$ and $s \in \sharedvars$ \\\hline
-$\mutexlock{m};$   & mutex lock, where $m \in \mutexes$ \\\hline
-$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$
+\textbf{Statement Form} & \textbf{Notes} \\
+\toprule
+
+$x = e;$ & private assignment, where $x \in \variables{T}$ and $e$ is an expression over $\variables{T}$ \\
+\midrule
+
+$x = f(\overline{e});$ & procedure call, where $x \in \variables{T}$, $\overline{e}$ is a sequence of expressions over $\variables{T}$, $f$ is the name of a procedure in $\procedures{T}$ \\
+\midrule
+
+$s = e;$ & shared write, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$ \\
+\midrule
+
+$x = s;$ & shared read,  where $x \in \variables{T}$ and $s \in \sharedvars$ \\
+\midrule
+
+$\mutexlock{m};$   & mutex lock, where $m \in \mutexes$ \\
+\midrule
+
+$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$\\
+\bottomrule
 \end{tabular}
@@
 \begin{tabular}{ll}
-\textbf{Original statement} & \textbf{Instrumented statement} \\\hline
+\textbf{Original Statement} & \textbf{Instrumented Statement} \\
+\toprule
+
 $s = e;$ & $\hasbeenwritten{i} = \hasbeenwritten{i} \cup \{ s \};$ \\
-         & $\lockset{s}{i} = \lockset{s}{i} \cap \currentlockset{i};$ \\\hline
+         & $\lockset{s}{i} = \lockset{s}{i} \cap \currentlockset{i};$ \\
+\midrule
+
 $x = s;$ & $\hasbeenread{i} = \hasbeenread{i} \cup \{ s \};$ \\
          & $\lockset{s}{i} = \lockset{s}{i} \cap \currentlockset{i};$ \\
-         & $\havoc{\instrument{x}{i}};$ \\\hline
-$\mutexlock{m};$   & $\currentlockset{i} = \currentlockset{i} \cup \{ m \};$ \\\hline
-$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\\hline
+         & $\havoc{\instrument{x}{i}};$ \\
+\midrule
+
+$\mutexlock{m};$   & $\currentlockset{i} = \currentlockset{i} \cup \{ m \};$ \\
+\midrule
+
+$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
+\bottomrule
 \end{tabular}

### 2015-05-12 whoop.tex
fix

@@
 
-A \emph{concurrent program} is described by a finite set of
-\emph{shared} variables, $\sharedvars$, a finite set of mutexes,
-$\mutexes$, and a finite set of \emph{thread templates}.  A thread
-template, $T$, consists of a finite set of procedures,
-$\procedures{T}$, and a finite set of private variables,
-$\variables{T}$.  A designated procedure, procedure $\mainforthread{T}
-\in \procedures{T}$, denotes the starting point for execution of a
-template by a thread.  A procedure in $\procedures{T}$ is
-represented by a control flow graph of basic blocks, where each block
-contains a sequence of statements.  A basic block either has a single
-successor or a pair of successors.  In the latter case, an \emph{exit
-  condition} over thread-private variables determines, at runtime, the
-successor to which control should flow on block exit.
+A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of \emph{thread templates}.
+A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.
+A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of a template by a thread.
+A procedure in $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
+A basic block either has a single successor or a pair of successors.
+In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.
 
-The forms of statements are shown in Figure~\ref{fig:statements}, and
-include designated statements for reading from and writing to shared
-variables.  In particular, shared variables may not appear in
-arbitrary expressions.  This restriction simplifies our presentation
-of lockset instrumentation, below, and a program that does not satisfy
-this restriction can be trivially pre-processed into one that does
-through the introduction of additional private temporaries to record
-values read from the shared state.  We do not specify the form of
-expressions, nor the types of variables, assuming a standard set of
-data types and operations.
+The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
+In particular, shared variables may not appear in arbitrary expressions.
+This restriction simplifies our presentation of lockset instrumentation, below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.
+We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.
 

### 2015-05-12 whoop.tex
updates in main section

@@
 
-\medskip\noindent\textbf{Semantics} Let $\ids$ be an infinite set from
-which dynamic thread ids will be drawn.  The state of a running
-concurrent program consists of: a valuation of the shared variables,
-$\sharedvars$; a mapping that associates each mutex in $\mutexes$ with
-an id from $\ids$, recording which thread currently holds the mutex,
-or with a special value $\bot \notin \ids$ to indicate that the
-mutex is not held by any thread; a list of \emph{threads}.  Each
-thread consists of an id, drawn from $\ids$, a thread template, $T$,
-an index indicating the next statement of the template $T$ to be
-executed by the thread, and a valuation of the thread private
-variables, $\variables{T}$.  If multiple threads are instances of the
-same template $T$, then each thread carries a \emph{separate}
-valuation of the private variables for this template.
+\medskip\noindent\textbf{Semantics }
+%
+Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.
+The state of a running concurrent program consists of: a valuation of the shared variables, $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
+Each thread consists of an id, drawn from $\ids$, a thread template, $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.
+If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.
 
-At the start of execution the valuation of shared variables is
-arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$),
-and the list of threads is empty.
+At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
 
-At any point of execution, a new thread may be added to the list of
-thread states.  This involves selecting an id $i \in \ids$ that has
-not been previously used during program execution and a thread
-template $T$, setting the point of execution for the new thread to be
-the first statement of $\mainforthread{T}$, and choosing an arbitrary
-valuation for the private variables $\variables{T}$.
+At any point of execution, a new thread may be added to the list of thread states.
+This involves selecting an id $i \in \ids$ that has not been previously used during program execution and a thread template $T$, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
 
-We consider a standard interleaving model of concurrency: at any
-execution point, a thread may execute its current statement, unless
-that statement has the form $\mutexlock{m}$ and mutex $m$ is already held
-by some thread.  Executing a statement causes the state of the thread,
-and the shared state, to be updated in a standard manner.  For
-example, if a thread following template $T$ executes $s = e$, where $s
-\in \sharedvars$ and $e$ is an expression over $\variables{T}$, the
-shared variable valuation is updated so that $s$ has the value
-determined by evaluating $e$ in the context of the thread's private
-variable valuation.  The thread's next statement is updated according
-to the control flow graph of the current procedure.
+We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.
+Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.
+For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.
+The thread's next statement is updated \PDComment{Do you mean execute?} according to the control flow graph of the current procedure.
 
-A thread terminates if it reaches the end of $\mainforthread{T}$; in this
-case the thread is removed from the list of threads.  Because we are
-interested in the analysis of device drivers, which are
-\emph{reactive} concurrent programs, we do not consider the notion of
-global program termination.
+A thread terminates if it reaches the end of $\mainforthread{T}$; in this case the thread is removed from the list of threads.
+Because we are interested in the analysis of device drivers, which are \emph{reactive} concurrent programs, we do not consider the notion of global program termination.
 
 \medskip\noindent\textbf{Lockset instrumentation }
+%
 Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibly that $S=T$.

### 2015-05-12 whoop.tex
fixes

@@
 %
-Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibly that $S=T$.
-We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is
-an instance of $T$, in the presence of arbitrarily many further threads.
+Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibility that $S$ and $T$ are equal.
+We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is an instance of $T$, in the presence of arbitrarily many further threads that may execute concurrently.
 
-The first phase in achieving this is to \emph{instrument} a thread template to enable lockset analysis.  Given an arbitrary symbol, $i$, we define the instrumentation of a
-template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to instrumentation, \emph{renaming} and \emph{lockset instrumentation}.
+To achieve this, we first \emph{instrument} a thread template for lockset analysis.
+Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.
+There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.
 
-Renaming is simple: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
-The purpose of renaming is to ensure that when we analyse a pair of templates, $S$ and $T$, we ensure that $S$ and $T$ execute distinct procedures and operate on distinct private data.
-Shared variables and mutexes are \emph{not} renamed, because these are indeed shared across templates.
+Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.
+Shared variables and mutexes are \emph{not} renamed, because these are shared across templates.
 
@@
 \end{tabular}
-\caption{Instrumenting statements for static lockset analysis}\label{fig:instrumentation}
+\caption{Instrumenting statements for lockset analysis}\label{fig:instrumentation}
 \end{figure}
@@
 Lockset instrumentation first requires the introduction of sets to track those variables that have been read from and written to, the mutexes that have been consistently used to protect each variable, and the set of current locks held by a thread.
-To this end, when instrumenting template $T$ with respect to symbol $i$, we add a sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track those shared variables that have been read from and written to, respectively, by the thread executing $T$.
+To this end, when instrumenting template $T$ with respect to symbol $i$, we add sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track those shared variables that have been read from and written to, respectively, by the thread executing $T$.
 We add a current lock set, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread, and for each shared variable $s \in \sharedvars$ we add a lock set, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
@@
 \medskip\noindent\textbf{Shared state abstraction }
+%
 Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account
@@
 has been performed.  As a result, when instrumenting a shared variable write, the effect of the write
-is not explicitly modelled.
+is not explicitly modeled.
 
 \medskip\noindent\textbf{Sequentialisation }
-The psuedocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
-analyse in order to prove race-freedom for a pair of thread templates $T$ and $U$.
+%
+The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
+analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
 Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding
@@
 \medskip\noindent\textbf{Soundness }
+%
 We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct,

### 2015-05-12 whoop.tex
fixes

@@
 
-To achieve this, we first \emph{instrument} a thread template for lockset analysis.
+To achieve this, we first \emph{instrument} a thread template for lockset analysis (see Section~\ref{bg:lockset}).
 Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.
@@
 
-Lockset instrumentation first requires the introduction of sets to track those variables that have been read from and written to, the mutexes that have been consistently used to protect each variable, and the set of current locks held by a thread.
-To this end, when instrumenting template $T$ with respect to symbol $i$, we add sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track those shared variables that have been read from and written to, respectively, by the thread executing $T$.
-We add a current lock set, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread, and for each shared variable $s \in \sharedvars$ we add a lock set, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
+Initially, lockset instrumentation introduces sets that track: the shared variables that have been read from and written to; the mutexes that have been consistently used to protect each variable; and the mutexes that are currently held by a thread.
+To this end, when instrumenting template $T$ with respect to symbol $i$, we add the sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$.
+For each shared variable $s \in \sharedvars$ we add a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
+Finally, we add a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread.
 
-The statements of each procedure in $\instrument{T}{i}$ that manipulate shared variables and mutexes are then instrumented to take account of lock sets.
+The statements of each procedure in $\instrument{T}{i}$ that manipulate shared variables and mutexes are then instrumented to take account of the above sets.
 This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.
-A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lock set for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).
+A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).
 Notice that in the instrumented form, the shared variable $s$ is no longer assigned to.
@@
 A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command applied to the receiving private variable $x$; again this relates to shared state abstraction and is discussed below.
-Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lock set, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
+Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
 
@@
 %
-Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account
-for possible side-effects due to other threads that are running concurrently.  The instrumentation
-of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$,
-a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command,
-which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in
-this fashion, it is possible to completely dispense with the shared variables after instrumentation
-has been performed.  As a result, when instrumenting a shared variable write, the effect of the write
-is not explicitly modeled.
+Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account for possible side-effects due to other threads that are running concurrently.
+The instrumentation of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.
+This is reflected by the use of a $\havockeyword$ command, which sets its argument to an arbitrary value.
+Because all shared state accesses are abstracted in this fashion, it is possible to completely dispense with the shared variables after the lockset instrumentation has been performed.
+As a result, when instrumenting a shared variable write, the effect of the write is not explicitly modeled.
 

### 2015-05-12 whoop.tex
fixes

@@
 %
-The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
-analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
-Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding
-$\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
-First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are
-initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and
-$\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures
-of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and
-$\mainforthread{\instrument{U}{j}}$ are then executed in turn.  Finally, an assertion checks
-for consistent use of mutexes: if a shared variables is written during execution of $\instrument{T}{i}$
-and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$
-and $\lockset{s}{j}$ must contain a common mutex.
+The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
+Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
+First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.
+The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter, due to renaming).
+Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.
 

### 2015-05-12 whoop.tex
more fixes

@@
 \end{tabular}
-\caption{Forms of statements in our simple programming model}\label{fig:statements}
+\caption{Forms of statements in our simple programming model}
+\label{fig:statements}
 \end{figure}
@@
 \end{tabular}
-\caption{Instrumenting statements for lockset analysis}\label{fig:instrumentation}
+\caption{Instrumenting statements for lockset analysis}
+\label{fig:instrumentation}
 \end{figure}
@@
 \end{tabular}
-\caption{The sequential program to be analysed in order to prove race-freedom for a pair of thread templates}
+\caption{The sequential program to be analyzed in order to prove race-freedom for a pair of thread templates}
 \label{fig:sequentialization}
@@
 %
-We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct,
-it is impossible for a thread executing template $T$ to race with a thread executing template $U$,
-under the assumption that the threads are guaranteed to terminate.
-Let us assume that the program of Figure~\ref{fig:sequentialization} is correct,
-and suppose (by way of contradition) that a thread executing $T$ can in fact race with a thread
-executing $U$, on some shared variable $s$.  By our hypothesis that the program
-of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate,
-the assertion checked at the end of the program guarantees at least one mutex, $m$ say,
-belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset
-(and according to the manner in which shared accesses are instrumented in Figure~\ref{ }),
-this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
-As a result, $m$ must be unlocked and locked between the two accesses, which
-contradicts that the pair of accesses is racing.
+We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct, it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
+Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.
+By our hypothesis that the program of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate, the assertion checked at the end of the program guarantees at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.
+By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
+As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
 
-We require the threads to terminate becauase in the presence of non-termination the assertion
-at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis
-problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
-In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyse
-in our experimental evaluation (Section~\ref{evaluation}) are terminating.
+We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.
+The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
+In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see Section~\ref{evaluation}) are indeed terminating.
 

### 2015-05-12 intro.tex
added intro to the implementation

@@
 
-\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g.\ Windows) as well as other concurrent programs with similar structure (e.g.\ filesystems).
+\whoop currently supports only Linux drivers, but the approach is generic, and thus we believe that it can be used to analyze drivers from other operating systems (e.g.\ Windows) as well as other concurrent programs that use a similar programming model (e.g.\ filesystems).
 

### 2015-05-12 whoop.tex
added intro to the implementation

@@
 In Section~\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.
-In Section~\ref{sec:whoopimplementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code, to check whether pairs of driver entry points can race with one another.
+In Section~\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code, to check whether pairs of driver entry points can race with one another.
 
@@
 
-\subsection{Symbolic Pairwise Lockset Analysis}\label{sec:symbolicpairwise}
+\subsection{Symbolic Pairwise Lockset Analysis}
+\label{sec:symbolicpairwise}
 
@@
 For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
-These \emph{are} handled by our implementation, and in Section~\ref{sec:whoopimplementation} we discuss interesting practical issues arising from the handling of a full-blown language.
+These \emph{are} handled by our implementation, and in Section~\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
 
@@
 
-\subsection{Implementation in \whoop}\label{sec:whoopimplementation}
+\subsection{Implementation in \whoop}
+\label{sec:implementation}
 
-\ADComment{Acknowledge very clearly that the simple concurrent programming model of Section~\ref{...} is deliberately idealistic.
-In practice, we do not know upfront what the entry points for a driver are, we don't have named set of locks, and rather than a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from Section~\ref{...} and used them to build a practical tool.}
+The simple concurrent programming model discussed in Section~\ref{sec:symbolicpairwise} is deliberately idealistic to describe our symbolic verification technique.
+In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
+We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build a practical tool for soundly detecting data races in drivers.
 

### 2015-05-12 whoop.tex
added architecture to the implementation

@@
 In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build a practical tool for soundly detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build \whoop, a practical tool for soundly detecting data races in drivers.
 
-\ADComment{Explain briefly the architecture, and specifically introduce Chauffer, SMACK and Boogie.}
+\begin{figure*}
+\centering
+\includegraphics[width=.99\linewidth]{img/whoop.pdf}
+\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
+\label{fig:whoop}
+\end{figure*}
+
+\medskip\noindent\textbf{Architecture }
+%
+Figure~\ref{fig:whoop} depicts the \whoop toolchain.
+Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{Stub header files that model the Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
+Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and Section~\ref{sec:symbolicpairwise}) using the Boogie verification engine.
+After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and Section~\ref{corral}).
+
+We decided to reuse industrial-strength tools in this work, because (i) we do not want to reinvent the wheel and (ii) these tools are robust and already battle-proven.
 
@@
 
-\ADComment{Note, unrelated to structure: put mention of interrupts in limitations}
-
-The entry points for
-
-We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyze Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.
-
-\begin{figure*}
-\centering
-\includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\caption{The \whoop infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
-\label{fig:whoop}
-\end{figure*}
-
-The key idea behind our novel static analysis is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race-checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.
-
-Figure~\ref{fig:whoop} depicts the \whoop toolchain.
-Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analyzed.
-Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
-Next, \whoop performs \emph{static pairwise lockset analysis} to detect all potential data races in the abstract program.
-The proof sketch behind our static lockset analysis is given in Section~\ref{whoop:proof}.
-The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}.
-The summarization that we perform to achieve scalability is presented in Section~\ref{whoop:summarization}.
-After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{corral})
-
-Finally, we discuss optimizations and assumptions in Section~\ref{whoop:optimizations}, and limitations of our technique in Section~\ref{whoop:limitations}.
-
-\subsection{Translation to Boogie IVL}
-\label{whoop:translation}
-
+\medskip\noindent\textbf{Translation to Boogie IVL }
+%
 The aim of this phase (see Figure~\ref{fig:whoop} -- A) is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
@@
 \end{boogie}
-
-\subsection{Static Lockset Analysis Proof Sketch}
-\label{whoop:proof}
-
-We now formalize our approach to verifying data race freedom in a driver by statically analyzing its locksets.
-Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver.
-Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one).
-Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronization bugs that arise from careless fine-grained locking.
-Note that the actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.
-
-For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$.
-Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution.
-In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location.
-
-\begin{theorem}
-\label{theorem:locksets}
-For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
-\end{theorem}
-
-A sketch of how the above theorem can be proved is as follows.
-Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$.
-By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$.
-By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$.
-As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
 
@@
 
+\PDComment{briefly mention interrupts}
+
 Another limitation of \whoop is that it is unable to verify drivers that are designed to be accessed by a single process at a time.

### 2015-05-12 experiments/figures/plot.py
updated measurements

@@ -15,16 +15,16 @@ from matplotlib.backends.backend_pdf import PdfPages
 plt.rcParams['xtick.minor.size'] = 0
 plt.rcParams['ytick.minor.size'] = 0
 
-x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0,              906.0, 15.9, 359.8, 36000]
-x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0,           36000, 18.5, 5664.5, 36000]
+x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0, 55.8, 906.0, 15.9, 359.8, 36000]
+x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0, 1698.6, 36000, 18.5, 5664.5, 36000]
 x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0,               36000, 24.3, 36000, 36000]
 
-z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4,        1924.4, 16.1, 474.7, 36000]
-z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6,       36000, 18.5, 12677.3, 36000]
+z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4, 96.2, 1924.4, 16.1, 474.7, 36000]
+z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6, 2249.0, 36000, 18.5, 12677.3, 36000]
 z2 = [132.1, 22514.1, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4,          36000, 24.5, 36000, 36000]
 
[…]

### 2015-05-12 evaluation.tex
Some comments on the experimental section.  Though I realised I commented on the section you told me was not yet ready - oops.  Jeroen would hate that.

@@
 
-We performed experiments to validate the usefulness of the \whoop approach (see Section~\ref{whoop}).
-We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution\footnote{\url{https://www.kernel.org}}.
-We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds, to show that \whoop can efficiently accelerate bug-finding with \corral.
+We performed experiments to validate the usefulness of the \whoop approach (Section~\ref{whoop}) and its combination with \corral (Section~\ref{corral}).
+We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
+Our results demonstrate that \whoop can efficiently accelerate bug-finding with \corral.
 
@@
 534.
+\ADComment{Perhaps also state that Boogie runs inside Mono, and give the Mono version?}
 
@@
 \textbf{Benchmarks}\xspace\xspace We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
-We chose non-trivial drivers from many different domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
-We had to manually model the environment for these drivers, a process that required two months of work.
+We chose non-trivial drivers from several domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+We had to manually model the environment for these drivers, a process that required approximately two months of work.
+\ADComment{Question from me: are block and char domains, in the same way that ethernet is a domain?}
+
+\ADComment{At this point, is it absolutely clear what we do with \corral?  I.e., that we consider each entry point pair separately, and do not employ any abstraction to model additional threads?}
 
 \noindent
-\textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs that \whoop identified (\#Racy Pairs); number of racy memory regions that \whoop reported (\#Racy MRs); and number of races that \corral discovered (\#Races Found) using a context-switch bound of 2.
+\textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of potentially racy pairs identified by \whoop (\#Racy Pairs); number of potentially racy memory regions reported by \whoop (\#Racy MRs); and number of races that discovered by \corral using a context-switch bound of 2 (\#Races Found).
 We did not discover new races with larger context-switch bounds.
@@
 We can see that \whoop reports more races than \corral.
-This is expected, as our approach uses over-approximation (to find all races), and thus can introduce false bugs.
-On the other hand, \corral is precise, but can miss races as it explores programs under pre-defined bounds.
+This is expected, since \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analysing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false alarm race reports.
+On the other hand, \corral is precise, but can miss races because only a limited number of context switches are considered.
 Another issue with \corral is loop coverage due to unsound loop unrolling.
@@
 This can potentially lead \corral to report false positives, but we have not seen this in practice.
+\ADComment{Would this be a good point to reiterate the fact that when we apply \corral to an entry point pair, we are just checking the specific pair of threads, and are not accounting for actions by other threads?}
 
@@
 This might be for performance, and indeed a lot of the races we found might be benign.
-Even benign races, though, can lead to undefined behavior according to the C standard, which can cause many issues (e.g.
-with compilers).
+Even benign races, though, lead to undefined behavior according to the C standard, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
 The second case is an entry point accessing a field of an object (either global or passed as a parameter) without a lock.
@@
 \textbf{Granularity of Context-Switches}\xspace\xspace Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks.
-All reported times are in seconds and averages of three runs.
-\corral was given a time budget of 10 hours (T.O.
-denotes a tool timeout) and a context-switch bound (csb) of 2, 5 and 9.
+All reported times, in seconds, are averages over three runs.
+\ADComment{Can we say something about variance?
+Perhaps ask Nathan what best to say?} \corral was given a time budget of 10 hours (T.O.
+denotes a tool timeout), and we considered context-switch bounds (csb) of 2, 5 and 9.
 By default, \corral instruments context-switches (i.e.
-\texttt{yield} statements) in all visible operations (denoted by \yieldall).
+\texttt{yield} statements) in \ADComment{instead of ``in'' can we say either ``before'' or ``after'' (not sure which is accurate)} all visible operations (denoted \yieldall).
 \whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{whoop:bugfinding}), and also uses all the optimisations discussed in Section~\ref{whoop:optimizations}.

### 2015-05-12 background.tex
added chauffeur to the implementation

@@
 \end{lstlisting}
-\caption{Example of data races in the generic\_nvram Linux char driver}\label{fig:data_race_example}
+\caption{Example of data races in the generic\_nvram Linux char driver}
+\label{fig:data_race_example}
 \end{figure}
@@
 \end{lstlisting}
-\caption{Using a lock to eliminate the data races in the previous example}\label{fig:lock_example}
+\caption{Using a lock to eliminate the data races in the previous example}
+\label{fig:lock_example}
 \end{figure}

### 2015-05-12 whoop.tex
added chauffeur to the implementation

@@
 
-\medskip\noindent\textbf{Lockset instrumentation }
+\medskip\noindent\textbf{Lockset Instrumentation }
 %
@@
 
-\medskip\noindent\textbf{Shared state abstraction }
+\medskip\noindent\textbf{Shared State Abstraction }
 %
@@
 \begin{tabular}{l}
-
 $\currentlockset{i} = \emptyset;$ $\hasbeenread{i} = \emptyset;$ $\hasbeenwritten{i} = \emptyset;$ \\
@@
 Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{Stub header files that model the Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
-Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined verification-focused semantics that is used as the input to a number of cutting-edge verifiers (e.g.
+Boogie and \corral).
+
 Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and Section~\ref{sec:symbolicpairwise}) using the Boogie verification engine.
@@
 
-We decided to reuse industrial-strength tools in this work, because (i) we do not want to reinvent the wheel and (ii) these tools are robust and already battle-proven.
+In this work, we decided to reuse industrial-strength tools, because (i) we did not want to ``reinvent the wheel'' and (ii) these tools are already robust and battle-proven.
 
-\ADComment{Explain how we identify entry points (Chauffer)}
+\medskip\noindent\textbf{Extracting Entry Point Information }
+%
+We have developed Chauffeur, a Clang-frontend that performs two important tasks.
+The first is to traverse (using a Clang visitor pass) the abstract syntax tree (AST) of the original driver and extract all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
+Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is subsequently parsed by \whoop.
+
+\begin{figure}[t]
+\begin{lstlisting}
+const struct file_operations nvram_fops = {
+  .llseek = nvram_llseek,
+  .read = read_nvram,
+  .write = write_nvram,
+  .unlocked_ioctl = nvram_unlocked_ioctl,
+};
+\end{lstlisting}
+\caption{Entry points definitions in the generic\_nvram driver}
+\label{fig:entrypoints}
+\end{figure}
+
+The second task is to instrument the original driver with a function that calls all entry points, one after the other, passing them parameters of arbitrary value (to over-approximate kernel behavior for soundness).
+This function effectively acts as the \emph{main} function of the driver and is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions).
+This simple rewriting does not alter the semantics of the original driver, as \whoop only uses this function to create checker functions and instrument them with calls to entry points (see Figure~\ref{fig:sequentialization} for the pseudocode of such checker function).
+Chauffeur outputs the rewritten driver in a C file.
 
 \ADComment{Explain how pointers get handled (SMACK) -- be brief, talk about memory regions, don't show Boogie code.}
+
+\medskip\noindent\textbf{Translation to Boogie IVL }
+%
+Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
+SMACK then translates the LLVM-IR to Boogie IVL.
+SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL.
+To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.
+
+The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource.
+Notice that function calls (e.g.
+to lock and unlock a mutex) are preserved in the translation.
 
@@
 
-\ADComment{Briefly talk about how we tried inlining and do summarisation instead}
-
-\ADComment{Discuss optimizations and assumptions.  Bring these to before the Corral stuff.}
-
-\medskip\noindent\textbf{Translation to Boogie IVL }
+\medskip\noindent\textbf{Instrumentation and Analysis }
 %
-The aim of this phase (see Figure~\ref{fig:whoop} -- A) is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g.
-Boogie and \corral).
-To achieve this translation, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools: Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}.
-
-First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given driver, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points.
-This function acts as the main entry point to the driver after the latter has been registered with the kernel.
-This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{whoop:method} and~\ref{corral}).
-The output is a file that contains information regarding the entry points and a file with the rewritten program.
-
-Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
-SMACK then translates the LLVM-IR to Boogie IVL.
-SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL.
-To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.
-
-\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
-static void ep1(struct net_device *dev) {
-  struct shared *tp = netdev_priv(dev);
-  mutex_lock(&tp->wk.mutex);
-  tp->resource = 4;
-  mutex_unlock(&tp->wk.mutex);
-}
-\end{lstlisting}
-
-Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK.
-The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource.
-Notice that function calls (e.g.
-to lock and unlock a mutex) are preserved in the translation.
-
-\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
-procedure ep1(dev: int) modifies $M.0; {
-  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
-  $bb0:
-    $p0 := dev;
-    $p1 := $pa($p0, 32, 1);
-    $p2 := $pa($pa($p1, 0, 24), 12, 1);
-    $p3 := $pa($pa($p2, 0, 12), 0, 1);
-    call mutex_lock($p3);
-    $p4 := $pa($pa($p1, 0, 24), 4, 1);
-    $M.0[$p4] := 4;
-    $p5 := $pa($pa($p1, 0, 24), 12, 1);
-    $p6 := $pa($pa($p5, 0, 12), 0, 1);
-    call mutex_unlock($p6);
-    return;
-}
-\end{boogie}
-
-\subsection{Instrumentation and Analysis}
-\label{whoop:method}
-
 The core \whoop tool (see Figure~\ref{fig:whoop} -- B) is responsible for parsing, instrumenting and sequentializing the abstract Boogie program.
@@
 
-\subsection{Watchdog Summarization}
-\label{whoop:summarization}
+\ADComment{Briefly talk about how we tried inlining and do summarization instead}
 
+\medskip\noindent\textbf{Watchdog Summarization }
+%
 Early on during the development of \whoop, we realized that scalability would be a serious issue.
@@
 
-\subsection{Optimizations and Assumptions}
-\label{whoop:optimizations}
-
+\medskip\noindent\textbf{Optimizations and Assumptions }
+%
 To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel interacts with its drivers.
@@
 
-\subsection{Limitations}
-\label{whoop:limitations}
-
+\medskip\noindent\textbf{Limitations }
+%
 \whoop is based on lockset analysis and thus can report false bugs, because a violation of the locking discipline does not always correspond to a real data race (e.g.

### 2015-05-12 evaluation.tex
some fixes

@@
 We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
-Our results demonstrate that \whoop can efficiently accelerate bug-finding with \corral.
+Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.
 
@@
 \noindent
-\textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of potentially racy pairs identified by \whoop (\#Racy Pairs); number of potentially racy memory regions reported by \whoop (\#Racy MRs); and number of races that discovered by \corral using a context-switch bound of 2 (\#Races Found).
+\textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of potentially racy pairs identified by \whoop (\#Racy Pairs); number of potentially racy memory regions reported by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound of 2 (\#Races Found).
 We did not discover new races with larger context-switch bounds.
@@
 
-As an example of the second case, we found the following race in the generic\_nvram char driver (see Example~\ref{fig:data_race_example} for the source code): during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} without first acquiring a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+As an example of the second case, we found the following race in the generic\_nvram char driver (see Figure~\ref{fig:data_race_example} for the source code): during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} without first acquiring a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This can lead to a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.

### 2015-05-12 whoop.tex
updates to SMACK text

@@
 
-\ADComment{Explain how pointers get handled (SMACK) -- be brief, talk about memory regions, don't show Boogie code.}
+\medskip\noindent\textbf{Translation for Verification }
+%
+The rewritten, by Chauffeur, driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
+SMACK then translates the LLVM-IR to Boogie IVL.
+Function calls (e.g.
+to lock and unlock a mutex) are preserved in this translation, which is useful for the lockset instrumentation performed by \whoop.
 
-\medskip\noindent\textbf{Translation to Boogie IVL }
-%
-Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
-SMACK then translates the LLVM-IR to Boogie IVL.
-SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL.
+SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
 To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.
+This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap.
 
-The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource.
-Notice that function calls (e.g.
-to lock and unlock a mutex) are preserved in the translation.
+A benefit of using this SMACK model is that memory regions denote disjoint sections of the heap (i.e.
+pointers that cannot alias).
+We leverage this knowledge inside \whoop to guide and optimize our instrumentation and analysis.
+In reality, this memory model is unsound ...
+(e.g.
+integers, overflow?), but for our work we assume it is sound.
 

### 2015-05-12 whoop.tex
updates to SMACK text

@@
 %
-The rewritten, by Chauffeur, driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
-SMACK then translates the LLVM-IR to Boogie IVL.
+The rewritten driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
 Function calls (e.g.
-to lock and unlock a mutex) are preserved in this translation, which is useful for the lockset instrumentation performed by \whoop.
+for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
 
-SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
+Next, SMACK translates the LLVM-IR to Boogie IVL.
+An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
+This means that \whoop does not need to directly deal with pointers and alias analysis, allowing us to reuse existing proven techniques.
+
 To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.

### 2015-05-12 whoop.tex
updates to SMACK text

@@
 
-Next, SMACK translates the LLVM-IR to Boogie IVL.
+SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop.
 An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
-This means that \whoop does not need to directly deal with pointers and alias analysis, allowing us to reuse existing proven techniques.
+This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse proven existing techniques and focus instead on verification efforts.
 
@@
 This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap.
-
-A benefit of using this SMACK model is that memory regions denote disjoint sections of the heap (i.e.
+A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.
 pointers that cannot alias).

### 2015-05-12 whoop.tex
updates to SMACK text

@@
 pointers that cannot alias).
-We leverage this knowledge inside \whoop to guide and optimize our instrumentation and analysis.
-In reality, this memory model is unsound ...
-(e.g.
-integers, overflow?), but for our work we assume it is sound.
+We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in Section~\ref{corral}.
+
+In reality, SMACK is unsound ... (e.g. integers), but for our work we assume it is sound.
 

### 2015-05-12 whoop.tex
created draft paragraph for identifying locks

@@
 
-\ADComment{Explain how we identify mutexes (inside our Boogie-based analysis)}
-
-\ADComment{Explain how we avoid quantifiers in performing race analysis}
-
-\medskip\noindent\textbf{Instrumentation and Analysis }
+\medskip\noindent\textbf{Identifying Set of Locks }
 %
-The core \whoop tool (see Figure~\ref{fig:whoop} -- B) is responsible for parsing, instrumenting and sequentializing the abstract Boogie program.
-During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g.
-the following call to \texttt{mutex\_lock()}:
+\whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g. the following call to \texttt{mutex\_lock()}:
 
@@
 
-When the lock rewriting finishes, \whoop traverses the Boogie IVL code, uses the information extracted by Chauffeur and separates each independent entry point call-graph from the rest of the program.
-To achieve this, it duplicates helper functions and renames them accordingly: e.g.
-for entry point \texttt{ep1}, a helper function \texttt{foo} will be renamed to \texttt{foo\$ep1}.
-This allows \whoop to perform entry point sensitive instrumentation and analysis.
-Although the program can potentially become much larger, this does not affect the analysis as discussed later.
+\ADComment{Explain how we avoid quantifiers in performing race analysis}
 
+\medskip\noindent\textbf{Instrumentation and Analysis }
+%
 Next, \whoop instruments the program with global variables for lockset analysis and race-checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:

### 2015-05-12 whoop.tex
added identifying locks to the implementation

@@
 Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{Stub header files that model the Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
-Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined verification-focused semantics that is used as the input to a number of cutting-edge verifiers (e.g.
-Boogie and \corral).
+Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined verification-focused semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
 
@@
 The rewritten driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
-Function calls (e.g.
-for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
+Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
 
@@
 This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap.
-A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.
-pointers that cannot alias).
+A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.\ pointers that cannot alias).
 We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in Section~\ref{corral}.
 
-In reality, SMACK is unsound ... (e.g. integers), but for our work we assume it is sound.
+In reality, SMACK is unsound ... (e.g.\ integers), but for our work we assume it is sound.
 
@@
 %
-\whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g. the following call to \texttt{mutex\_lock()}:
+When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
+The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
 
-\begin{boogie}
-call mutex_lock($p3);
-\end{boogie}
-
-Will be rewritten to:
-
-\begin{boogie}
-const {:lock} unique lock$0: int;
-call mutex_lock(lock$0);
-\end{boogie}
-
-If \whoop cannot infer a lock it will exit with a warning.
-A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks).
-It is relatively straightforward to enhance \whoop, though, with new locking primitives.
-\whoop will also exit with a warning if it detects ``improper'' use of locks (e.g.
-dynamic lock creation).
-The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
+Currently, \whoop only supports mutexes and spinlocks from the Linux kernel APIs.
+However, it is relatively easy to enhance our tool with knowledge of other locking primitives.
+If \whoop cannot infer a lock, e.g.
+because it was created dynamically or was indexed from an array of locks, it will exit with a warning.
+The reason behind this is twofold: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
 
@@
 The key idea behind watchdog summarization is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
-This reduces the false positives from reporting errors if only e.g.
-a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not.
+This reduces the false positives from reporting errors if only e.g.\ a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not.
 An example of watchdog-based candidate post-conditions is the following:
@@
 %
-\whoop is based on lockset analysis and thus can report false bugs, because a violation of the locking discipline does not always correspond to a real data race (e.g.
-when lockfree synchronization is used instead of locking).
-\whoop also uses over-approximation (e.g.
-when reading from the shared state) and summarization, which can be sources of false positives.
+\whoop is based on lockset analysis and thus can report false bugs, because a violation of the locking discipline does not always correspond to a real data race (e.g.\ when lockfree synchronization is used instead of locking).
+\whoop also uses over-approximation (e.g.\ when reading from the shared state) and summarization, which can be sources of false positives.
 Furthermore, the tool does not currently check for dynamically created locks or for locks that are provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.
@@
 Because \whoop performs pair-wise analysis, it over-approximates this flag, and regards a pair of entry points as running concurrently, even if this is not the case.
-However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.
-a user might expect that a device can be accessed concurrently for performance).
+However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
 Because using single-open device mode is considered bad practice, we thus burden the developer with disregarding any related false bug reports.

### 2015-05-13 whoop.tex
draft sections for watchdog-based techniques

@@
 \ADComment{Top level question about this section: is the motivation super-clear?  Is it obvious why we go for the pairwise approach?}
+
+\PDComment{Mention this somewhere (it was in my previous implementation text)?
+Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.}
 
@@
 
-\medskip\noindent\textbf{Identifying Set of Locks }
+\medskip\noindent\textbf{Identifying Set of All Locks }
 %
@@
 
-\medskip\noindent\textbf{Instrumentation and Analysis }
+\medskip\noindent\textbf{Watchdog Race-Checking }
 %
@@
 Watchdog race-checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
-
-\whoop proceeds by instrumenting each entry point call-graph for lockset analysis by replacing each call to a locking or unlocking function with a \whoop-specific lock and unlock functions.
-This means that Example~\ref{fig:smack_translation} will become:
-
-\begin{boogie}
-procedure ep1(dev: int) modifies $M.0; {
-  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
-  $bb0:
-    $p0 := dev;
-    $p1 := $pa($p0, 32, 1);
-    $p2 := $pa($pa($p1, 0, 24), 12, 1);
-    $p3 := $pa($pa($p2, 0, 12), 0, 1);
-    call _UPDATE_CLS_$ep1(lock$0, true);
-    $p4 := $pa($pa($p1, 0, 24), 4, 1);
-    $M.0[$p4] := 4;
-    $p5 := $pa($pa($p1, 0, 24), 12, 1);
-    $p6 := $pa($pa($p5, 0, 12), 0, 1);
-    call _UPDATE_CLS_$ep1(lock$0, false);
-    return;
-}
-\end{boogie}
-
-The \texttt{\_UPDATE\_CLS\_\$ep1()} is a special function that updates the CLS global variable to either true or false, if the lock is held or released by \texttt{ep1} accordingly:
-
-\begin{boogie}
-procedure {:inline 1} _UPDATE_CLS_$ep1(lock: int,
-    isLocked: bool);
-  modifies lock$0_in_CLS_$ep1;
-
-implementation {:inline 1} _UPDATE_CLS_$ep1(
-    lock: int, isLocked: bool) {
-  _UPDATE:
-    lock$0_in_CLS_$ep1 := (if lock == lock$0 then
-      isLocked else lock$0_in_CLS_$ep1);
-  return;
-}
-\end{boogie}
-
-Next, \whoop replaces every write and read access in each entry point call-graph with a \whoop-specific function:
-
-\begin{boogie}
-procedure ep1(dev: int) modifies $M.0; {
-  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
-  $bb0:
-    $p0 := dev;
-    $p1 := $pa($p0, 32, 1);
-    $p2 := $pa($pa($p1, 0, 24), 12, 1);
-    $p3 := $pa($pa($p2, 0, 12), 0, 1);
-    call _UPDATE_CLS_$ep1(lock$0, true);
-    $p4 := $pa($pa($p1, 0, 24), 4, 1);
-    call _WRITE_LS_$M.0_$ep1($p4);
-    $p5 := $pa($pa($p1, 0, 24), 12, 1);
-    $p6 := $pa($pa($p5, 0, 12), 0, 1);
-    call _UPDATE_CLS_$ep1(lock$0, false);
-    return;
-}
-\end{boogie}
-
-The \texttt{\_WRITE\_LS\_\$M.0\_\$ep1()} is a special function that computes the intersection of the CLS and the lockset of \texttt{\$M.0} (see Section~\ref{bg:lockset}).
-It also updates the write-set of \texttt{\$M.0} to true, because there was a write access by \texttt{ep1}:
-
-\begin{boogie}
-procedure {:inline 1} _WRITE_LS_$M.0_$ep1(
-    ptr: int);
-  modifies lock$0_in_LS_$M.0_$ep1,
-    WRITTEN_$M.0_$ep1;
-
-implementation {:inline 1} _WRITE_LS_$M.0_$ep1(
-    ptr: int) {
-  _WRITE:
-    goto anon1_Then, anon1_Else;
-
-  anon1_Then:
-    assume {:partition} WATCHED_ACCESS_$M.0 ==
-      ptr && DEVICE_IS_REGISTERED_$ep;
-    lock$0_in_LS_$M.0_$ep := lock$0_in_CLS_$ep && lock$0_in_LS_$M.0_$ep;
-    WRITTEN_$M.0_$ep := true;
-    return;
-  anon1_Else:
-    assume {:partition} !(WATCHED_ACCESS_$M.0 ==
-      ptr && DEVICE_IS_REGISTERED_$ep);
-    return;
-}
-\end{boogie}
-
-The \texttt{\_READ\_LS\_\$M.0\_\$ep1()} updates the lockset of \texttt{\$M.0} and the read-set accordingly for a read access.
-
-\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialization}, which combines the two arbitrary threads in a single sequential pair.
-This is achieved by creating a checker function that calls the two instrumented entry points and then performs race-checking assertions.
-This process repeats until all possible pairs of entry points have been sequentialized.
-To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}).
-This over-approximates any effects from all the unmodeled threads on the shared state of the pair.
-Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.
-
-After each checker function calls a pair of entry points, it asserts that each shared memory location is protected by at least one common lock (see Section~\ref{bg:lockset}).
-The assertion check is triggered iff the memory location was accessed by both entry points and at least one of the accesses was a write.
-We do not check for read-read races, which are inherently benign.
-The following is an example of a checker function for entry points \texttt{ep1} and \texttt{ep2}:
-
-\begin{boogie}
-procedure check$ep1$ep2(dev: int);
-  requires !lock$0_in_CLS_$ep1;
-  requires !lock$0_in_CLS_$ep2;
-  requires lock$0_in_LS_$M.0_$ep1;
-  requires lock$0_in_LS_$M.0_$ep2;
-  requires !WRITTEN_$M.0_$ep1;
-  requires !WRITTEN_$M.0_$ep2;
-  requires !READ_$M.0_$ep1;
-  requires !READ_$M.0_$ep2;
-  modifies lock$0_in_CLS_$ep1, lock$0_in_CLS_$ep2,
-    lock$0_in_LS_$M.0_$ep1, lock$0_in_LS_$M.0_$ep2,
-    WRITTEN_ $M.0_$ep1,WRITTEN_$M.0_$ep2,
-    READ_$M.0_$ep1, READ_$M.0_$ep2;
-
-implementation check$ep1$ep2(dev: int) {
-  _CHECK:
-    call ep1(dev);
-    call ep2(dev);
-    assert (WRITTEN_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
-      || (WRITTEN_$M.0_$ep1 && READ_$M.0_$ep2)
-      || (READ_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
-      ==> lock$0_in_LS_$M.0_$ep1 &&
-      lock$0_in_LS_$M.0_$ep2;
-    return;
-}
-\end{boogie}
 

### 2015-05-13 whoop.tex
added watchdog race checking to the implementation

@@
 
-\ADComment{Explain how we avoid quantifiers in performing race analysis}
+\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
+%
+Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see Section~\ref{sec:symbolicpairwise}).
+Such sets can be typically modeled in Boogie using maps.
+However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness.
+To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog} race-checking instrumentation, which works as follows:
 
-\medskip\noindent\textbf{Watchdog Race-Checking }
-%
-Next, \whoop instruments the program with global variables for lockset analysis and race-checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:
-
-\begin{boogie}
-var {:current_lockset} lock$0_in_CLS_$ep1: bool;
-var {:lockset} lock$0_in_LS_$M.0_$ep1: bool;
-var {:access_checking} WRITTEN_$M.0_$ep1: bool;
-var {:access_checking} READ_$M.0_$ep1: bool;
-\end{boogie}
-
-\whoop also instruments a global watchdog variable per memory region that is common to all entry points:
-
-\begin{boogie}
-const {:watchdog} WATCHED_ACCESS_$M.0: int;
-\end{boogie}
-
-The above instrumented variables represent the follows: \texttt{lock\$0\_in\_CLS\_\$ep1} is the current lockset (CLS) of \texttt{ep1}; \texttt{WATCHED\_ACCESS\_\$M.0} is an \emph{unconstrained} constant representing the offset to memory region \texttt{\$M.0} that should be checked for races; \texttt{lock\$0\_in\_LS\_\$M.0\_\$ep1} is the lockset for \texttt{\$M.0}; \texttt{WRITTEN\_\$M.0\_\$ep1} is the write-set for \texttt{\$M.0}; and \texttt{READ\_\$M.0\_\$ep1} is the read-set for \texttt{\$M.0}.
+We introduce a global watchdog variable for each SMACK memory region that is common to both entry points of a pair.
+This variable is an \emph{unconstrained} constant representing the offset to memory region that should be checked for races.
 Verification involves proving that two entry points cannot race at the watched offset of every memory region.
@@
 Watchdog race-checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
-
-Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to the Z3~\cite{de2008z3} theorem prover.
-Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
-To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
-The following is an example of a reported error:
-
-\begin{lstlisting}
-foo.c: error: potential write-write race:
-  write by entry point ep1, foo.c:36:2
-    shared->resource = 4;
-  write by entry point ep2, foo.c:52:2
-    shared->resource = 5;
-\end{lstlisting}
 
@@
 
+\medskip\noindent\textbf{Verification and Error Reporting }
+%
+Finally, the instrumented and summarized abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to the Z3~\cite{de2008z3} theorem prover.
+Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
+To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
+The following is an example of a reported error:
+
+\begin{lstlisting}
+foo.c: error: potential write-write race:
+  write by entry point ep1, foo.c:36:2
+    shared->resource = 4;
+  write by entry point ep2, foo.c:52:2
+    shared->resource = 5;
+\end{lstlisting}
+
 \medskip\noindent\textbf{Optimizations and Assumptions }

### 2015-05-13 whoop.tex
split optimisations and assumptions
updates

@@
 However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness.
-To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog} race-checking instrumentation, which works as follows:
+To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, which works as follows:
 
@@
 Early on during the development of \whoop, we realized that scalability would be a serious issue.
-When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory.
+When we tried to apply \whoop on the r8169 ethernet driver ($>$7000 lines of C code) with all functions fully inlined, we run out of memory.
 One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory.
 Another reason was that some entry points recursively call helper functions.
-Inherently, recursion does not work with inlining.
-To tackle these issues we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze Linux drivers in a scalable fashion, but without destroying precision.
+Inherently, recursion does not work well with inlining.
+To tackle these issues, we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze drivers in a scalable fashion and without destroying precision.
 
@@
 
-The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race-checking (see Section~\ref{whoop:method}).
-The key idea behind watchdog summarization is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses.
-This reduces the false positives from reporting errors if only e.g.\ a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not.
-An example of watchdog-based candidate post-conditions is the following:
+The candidate invariants are based on the aforementioned global watchdog variables that \whoop instruments for watchdog race-checking.
+The key idea behind watchdog summarization is to automatically generate candidate invariants for each memory region in the granularity of watched accesses.
+This reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing.
+
+To generate the candidate invariants, \whoop performs an inter-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
+To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
+For example the previous \texttt{\$M.0} candidates would now be:
 
@@
     ==> WATCHED_ACCESS_$M.0 == dev + 8
-    ==> !WRITTEN_$M.0_$entrypoint1;
-  ...
-\end{boogie}
-
-To generate the candidate invariants, \whoop performs an intra-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
-To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
-For example the previous \texttt{\$M.0} candidates would now be:
-
-\begin{boogie}
-procedure foo$ep1(dev: int);
-  ensures _b$ls$entrypoint1$0
-    ==> lock$0_in_LS_$M.0_$entrypoint1;
-  ensures _b$ls$entrypoint1$1
-    ==> lock$0_in_LS_$M.0_$entrypoint1;
-  ensures _b$ac$entrypoint1$0
-    ==> !WRITTEN_$M.0_$entrypoint1;
-  ensures _b$ac$entrypoint1$1
     ==> !WRITTEN_$M.0_$entrypoint1;
@@
 
-\medskip\noindent\textbf{Optimizations and Assumptions }
+\medskip\noindent\textbf{Optimizations }
 %
@@
 
-%We also optimize the use of function pointers.
-LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site.
-To increase precision, Chauffeur analyses the AST of the original program and attempts to infer context-sensitive function pointer use information.
-This information is the used by \whoop to increase precision.
-To be sound, this optimization is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.
-
-Finally, we assume that the formal parameters of an entry point do not alias, and thus cannot race.
+\medskip\noindent\textbf{Assumptions }
+%
+We assume that the formal parameters of an entry point do not alias, and thus cannot race.
 This is a potentially unsound feature that can be turned off using a command line option.
@@
 We also assume that our environmental model and domain-specific knowledge are correct and do not cause any bugs to be missed.
+Finally, we assume that the external tools we use as part of the \whoop toolchain are sound and bug-free.
 

### 2015-05-13 whoop.tex
updates to optimisations

@@
 
-\ADComment{Briefly talk about how we tried inlining and do summarization instead}
-
 \medskip\noindent\textbf{Watchdog Summarization }
@@
 To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
-For example the previous \texttt{\$M.0} candidates would now be:
-
-\begin{boogie}
-procedure foo$ep1(dev: int);
-  ensures _b$ls$entrypoint1$0
-    ==> WATCHED_ACCESS_$M.0 == dev + 4
-    ==> lock$0_in_LS_$M.0_$entrypoint1;
-  ensures _b$ls$entrypoint1$1
-    ==> WATCHED_ACCESS_$M.0 == dev + 8
-    ==> lock$0_in_LS_$M.0_$entrypoint1;
-  ensures _b$ac$entrypoint1$0
-    ==> WATCHED_ACCESS_$M.0 == dev + 4
-    ==> !WRITTEN_$M.0_$entrypoint1;
-  ensures _b$ac$entrypoint1$1
-    ==> WATCHED_ACCESS_$M.0 == dev + 8
-    ==> !WRITTEN_$M.0_$entrypoint1;
-  ...
-\end{boogie}
 
@@
 This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then exploit to make \whoop more precise.
-
 In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding kernel-imposed serialization.
-The Linux kernel can serialize calls to entry points, thus forcing them to not run concurrently with each other.
-An an example, a large number of networking and ethernet entry points are mutually serialized with RTNL, a network-specific kernel lock.
+The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other.
+As an example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock.
 \whoop exploits this knowledge and does not create pairs for entry points that are serialized.
 
-Modeling kernel methods is another opportunity to increase precision, but requires Linux kernel expertise.
+%Modeling kernel methods is another opportunity to increase precision, but requires Linux kernel expertise.
 For example, the \texttt{register\_netdev()} function registers a network driver with the kernel.
@@
 
-Another optimization \whoop uses is to soundly reduce the checked memory locations.
-If a memory location is accessed only by one entry point in a pair, then it is safe to not check this location for this pair.
-This reduces the burden to the SMT solver and speed up the verification process.
+A performance optimization that \whoop uses is to soundly reduce the checked shared variables.
+If a shared variable $s$ is accessed only by one entry point in a pair, then it is safe to not check $s$ for this pair.
+This reduces the burden of the SMT solver and speeds up the verification process.
 

### 2015-05-13 whoop.tex
updates to optimisations

@@
 %
-To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel interacts with its drivers.
-This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then exploit to make \whoop more precise.
-In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding kernel-imposed serialization.
+We have implemented various optimizations to increase the precision and performance of \whoop.
+
+In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding \emph{kernel-imposed serialization}.
 The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other.
@@
 \whoop exploits this knowledge and does not create pairs for entry points that are serialized.
+This is an ongoing manual effort: the more drivers we study, the more such properties we discover, which we can then exploit to make \whoop more precise.
 

### 2015-05-13 experiments/figures/plot.py
final results

@@ -17,15 +17,15 @@ plt.rcParams['ytick.minor.size'] = 0
 
 x0 = [13.5, 41.3, 15.7, 3.1, 13.5, 14.5, 41.6, 40.7, 99.2, 95.5, 21.0, 55.8, 906.0, 15.9, 359.8, 36000]
 x1 = [16.9, 79.4, 17.3, 3.1, 66.1, 16.1, 52.9, 295.3, 102.4, 1847.9, 32.0, 1698.6, 36000, 18.5, 5664.5, 36000]
-x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0,               36000, 24.3, 36000, 36000]
+x2 = [49.3, 1358.9, 23.9, 3.1, 689.3, 19.3, 104.6, 8883.0, 107.6, 36000, 278.0, 36000, 36000, 24.3, 36000, 36000]
 
 z0 = [17.7, 79.4, 26.1, 3.1, 14.7, 22.3, 91.2, 62.6, 104.0, 164.9, 44.4, 96.2, 1924.4, 16.1, 474.7, 36000]
 z1 = [24.2, 432.4, 31.2, 3.1, 70.0, 25.7, 150.7, 405.7, 105.5, 2309.8, 89.6, 2249.0, 36000, 18.5, 12677.3, 36000]
-z2 = [132.1, 22514.1, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4,          36000, 24.5, 36000, 36000]
+z2 = [132.1, 22514.1, 55.2, 3.2, 748.0, 33.0, 633.2, 33468.4, 107.1, 36000, 1543.4, 36000, 36000, 24.5, 36000, 36000]
 
 y0 = [33.5, 169.3, 38.1, 11.1, 22.8, 35.0, 182.6, 81.5, 388.0, 271.0, 39.8, 99.6, 1966.5, 12.9, 548.2, 36000]
 y1 = [47.0, 595.1, 45.6, 13.8, 130.3, 40.2, 263.7, 419.4, 390.8, 2746.6, 85.0, 2376.8, 36000, 15.2, 14469.0, 36000]
-y2 = [197.5, 27337.6, 78.4, 37.3, 1571.7, 51.8, 793.0, 36000, 392.1, 36000, 1539.9,         36000, 21.4, 36000, 36000]
[…]

### 2015-05-13 whoop.tex
update

@@
 
-In this work, we decided to reuse industrial-strength tools, because (i) we did not want to ``reinvent the wheel'' and (ii) these tools are already robust and battle-proven.
+Besides the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}), we decided to reuse other industrial-strength tools, because (i) we did not want to ``reinvent the wheel'' and (ii) these tools are robust and battle-proven in real complex software.
 

### 2015-05-13 whoop.tex
update

@@
 In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build \whoop, a practical tool for soundly detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.
 
@@
 %
-We have implemented various optimizations to increase the precision and performance of \whoop.
+We have implemented various optimizations to increase the precision and performance of \whoop. In our experience, the following two are the most effective optimizations:
 
-In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding \emph{kernel-imposed serialization}.
+We enrich \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision.
 The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other.
@@
 
-A performance optimization that \whoop uses is to soundly reduce the checked shared variables.
-If a shared variable $s$ is accessed only by one entry point in a pair, then it is safe to not check $s$ for this pair.
-This reduces the burden of the SMT solver and speeds up the verification process.
+We soundly reduce the checked shared variables to increase the performance of \whoop.
+If a shared variable $s$ is accessed by only one entry point in a pair, then it is safe to not check $s$ for this pair.
+This can reduce the burden of the SMT solver and speedup the verification process.
 

### 2015-05-13 whoop.tex
update

@@
 
-Currently, \whoop only supports mutexes and spinlocks from the Linux kernel APIs.
+Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs.
 However, it is relatively easy to enhance our tool with knowledge of other locking primitives.
@@
 because it was created dynamically or was indexed from an array of locks, it will exit with a warning.
-The reason behind this is twofold: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
+The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
 

### 2015-05-13 whoop.tex
update

@@
 This is a potentially unsound feature that can be turned off using a command line option.
-In our experience so far, though, we have not missed any data races because of this assumption.
+In our experience so far, though, we have not missed any races because of this assumption.
 We also assume that our environmental model and domain-specific knowledge are correct and do not cause any bugs to be missed.
-Finally, we assume that the external tools we use as part of the \whoop toolchain are sound and bug-free.
+Finally, we assume that the external tools that we use as part of the \whoop toolchain are sound and bug-free.
 
@@
 Furthermore, the tool does not currently check for dynamically created locks or for locks that are provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.
-
-\PDComment{briefly mention interrupts}
+We also do not currently handle interrupt handlers in a special way; to be sound, we assume that they can execute concurrently with entry points at any time.
+We could address this in the future by modeling interrupt-specific kernel functions (e.g.
+for enabling/disabling interrupts).
 

### 2015-05-13 corral.tex
footnote about weak memory

@@
 statements only at shared memory accesses that \whoop reported as potentially
-racy. This effectively acts as a partial-order reduction that can significantly
+racy.\footnote{In the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} This effectively acts as a partial-order reduction that can significantly
 reduce the number of interleavings that \corral needs to explore, and thus

### 2015-05-13 corral.tex
update

@@
 
-\noindent
-\textbf{Sound Partial-Order Reduction}\xspace\xspace By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before each \emph{visible operation} (i.e., synchronization operations and shared memory accesses) of every thread.
+\medskip\noindent\textbf{Sound Partial-Order Reduction }
+%
+By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before each \emph{visible operation} (i.e., synchronization operations and shared memory accesses) of every thread.
 It then leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
@@
 
-\noindent
-\textbf{Race-Checking Instrumentation}\xspace\xspace In this paper, we treat races as bugs. To detect them, we employ a simple, but effective, encoding of data race checks within \corral.
-For example, whenever there is a write access to a shared variable, such as
-\texttt{\$M.13 := 0}, we instrument the Boogie IVL code to check for write-write
-races as follows:
+\medskip\noindent\textbf{Race-Checking Instrumentation }
+%
+In this paper, we treat races as bugs.
+To detect them, we employ a simple, but effective, encoding of data race checks within \corral.
+For example, whenever there is a write access to a shared variable $s$, we instrument our abstract Boogie program to check for write-write races as follows:
 %
 \begin{boogie}
-$M.13 := 0;        // original write
+s = e;         // original write
 yield;         // allow for a context-switch
-assert $M.13 == 0; // check written value
+assert s == e; // check written value
 \end{boogie}
 %
-Likewise, for a read access to a shared variable, such as \texttt{\$p0 :=
-\$M.13}, we instrument the Boogie IVL code to check for read-write races as
-follows:
+Likewise, for a read access to a shared variable $s$, we instrument the abstract Boogie program to check for read-write races as follows:
 %
 \begin{boogie}
-$p0 := $M.13;        // original read
+x = s;         // original read
 yield;         // allow for a context-switch
-assert $p0 == $M.13; // check read value
+assert x == s; // check read value
 \end{boogie}

### 2015-05-13 corral.tex
update in corral section to talk about pairs

@@
 
+In this work, we use \corral to analyze separate pairs of entry points (as outputted by \whoop), instead of the whole Boogie driver (as outputted by SMACK).
+We also do not employ any abstraction to model additional threads, as we want \corral to report only true races.
+Because we only analyze pairs, though, \corral can miss races that require more than two threads to manifest.
+We argue that this limitation is orthogonal to this work: as long as we compare the same pairs with standalone \corral and with \corral that takes advantage of \whoop, then our comparison is sound.
+In reality, we could allow \corral to analyze the whole driver, and then instrument \texttt{yield} statements in a coarser granularity (taking into account all threads).
+
 \medskip\noindent\textbf{Race-Checking Instrumentation }

### 2015-05-13 evaluation.tex
update in corral section to talk about pairs

@@
 \noindent
-\textbf{Experimental Setup}\xspace\xspace We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5, Z3 4.3.2, Boogie rev.
+\textbf{Experimental Setup}\xspace\xspace We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5, Z3~4.3.2, Boogie rev.
 4192 and \corral rev.
 534.
-\ADComment{Perhaps also state that Boogie runs inside Mono, and give the Mono version?}
+We also used Mono~4.1.0 to run Boogie and \corral.
 
@@
 \textbf{Benchmarks}\xspace\xspace We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
-We chose non-trivial drivers from several domains: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+We chose non-trivial drivers from several categories: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
 We had to manually model the environment for these drivers, a process that required approximately two months of work.
-\ADComment{Question from me: are block and char domains, in the same way that ethernet is a domain?}
-
-\ADComment{At this point, is it absolutely clear what we do with \corral?  I.e., that we consider each entry point pair separately, and do not employ any abstraction to model additional threads?}
 
@@
 This can potentially lead \corral to report false positives, but we have not seen this in practice.
+
 \ADComment{Would this be a good point to reiterate the fact that when we apply \corral to an entry point pair, we are just checking the specific pair of threads, and are not accounting for actions by other threads?}
+
+\ADComment{At this point, is it absolutely clear what we do with \corral?  I.e., that we consider each entry point pair separately, and do not employ any abstraction to model additional threads?}
 

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 This can potentially lead \corral to report false positives, but we have not seen this in practice.
-
-\ADComment{Would this be a good point to reiterate the fact that when we apply \corral to an entry point pair, we are just checking the specific pair of threads, and are not accounting for actions by other threads?}
-
-\ADComment{At this point, is it absolutely clear what we do with \corral?  I.e., that we consider each entry point pair separately, and do not employ any abstraction to model additional threads?}
+Finally, when we apply \corral to a pair of entry points, we are just checking the specific pair and are not accounting for the effect of other threads (see Section~\ref{corral}); this can also cause \corral to miss some races.
 

### 2015-05-13 corral.tex
update

@@
 %
-By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before each \emph{visible operation} (i.e., synchronization operations and shared memory accesses) of every thread.
+By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before every shared memory access of each thread.
 It then leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
-
-Our approach to accelerating \corral is simple and yet effective: we simplify
-the sequentialization that \corral performs by instrumenting \texttt{yield}
-statements only at shared memory accesses that \whoop reported as potentially
-racy.\footnote{In the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} This effectively acts as a partial-order reduction that can significantly
-reduce the number of interleavings that \corral needs to explore, and thus
-increase performance and scalability.
+Our approach to accelerating \corral is simple and yet effective: we simplify the sequentialization that \corral performs by instrumenting \texttt{yield} statements only at shared memory accesses that \whoop reported as potentially racy.\footnote{In the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} This effectively acts as a partial-order reduction that can significantly reduce the number of interleavings that \corral needs to explore, and thus increase performance and scalability.
 
 \ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
-about entry point pairs, or could we prove it just for any pair?}\PDComment{Lets also see what Ally thinks about this part.}
-We now show that our partial-order reduction is sound: it does not miss any more
-bugs than what \corral would miss even without taking advantage of the \whoop
-race-related information. Let \corral run an $n$ amount of entry point pairs
-with yields in all visible operations. Let this configuration be known as $C$.
-Each entry point pair is running in a separate \corral process. \corral will
-miss all bugs that are triggered by more than 3 bugs running concurrently and
-all bugs that are triggered by investigating a larger than $r$ recursion bound
-and larger than $k$ context-switch bound. Let \corral then run the same $n$
-amount of pairs, but this time with all \texttt{yield} statements in non-racy
-shared memory accesses that are removed according to information from \whoop. Let this configuration be
-known as $W+C$. Because these shared memory accesses cannot race, performing a
-\texttt{yield} will not introduce any new program behaviors. Thus, \corral will
-find the same bugs using the configuration $W+C$, as using the configuration
-$C$.
+about entry point pairs, or could we prove it just for any pair?}
+
+We now show that our partial-order reduction is sound: it does not miss any more bugs than what \corral would miss even without taking advantage of the \whoop race-related information.
+Let \corral run an $n$ amount of entry point pairs with yields in all \emph{visible} operations (synchronization points and shared memory accesses).
+Let this configuration be known as $C$.
+Each entry point pair is running in a separate \corral process.
+\corral will miss all bugs that are triggered by more than 3 threads running concurrently and all bugs that are triggered by investigating a larger than $r$ recursion bound and larger than $k$ context-switch bound.
+Let \corral then run the same $n$ amount of pairs, but this time \whoop removes the \texttt{yield} statements in all non-racy shared memory accesses.
+Let this configuration be known as $W+C$.
+Because these shared memory accesses cannot race, performing a \texttt{yield} will not introduce any new program behaviors.
+Thus, \corral will find the same bugs using the configuration $W+C$, as using the configuration $C$.
 
@@
 finds that a pair of entry points cannot race, then it will only instrument
-context-switches in synchronization points of this pair; else if \whoop finds
+context-switches in synchronization points (i.e.\ locks and unlocks) of this pair; else if \whoop finds
 that a pair of entry points can race, then it will instrument context-switches

### 2015-05-13 evaluation.tex
update

@@
 
-\noindent
-\textbf{Experimental Setup}\xspace\xspace We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5, Z3~4.3.2, Boogie rev.
+\medskip\noindent\textbf{Experimental Setup }
+%
+We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5, Z3~4.3.2, Boogie rev.
 4192 and \corral rev.
@@
 
-\noindent
-\textbf{Benchmarks}\xspace\xspace We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
+\medskip\noindent\textbf{Benchmarks }
+%
+We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
 We chose non-trivial drivers from several categories: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
@@
 
-\noindent
-\textbf{Race-Checking}\xspace\xspace Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of potentially racy pairs identified by \whoop (\#Racy Pairs); number of potentially racy memory regions reported by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound of 2 (\#Races Found).
+\medskip\noindent\textbf{Race-Checking }
+%
+Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of potentially racy pairs identified by \whoop (\#Racy Pairs); number of potentially racy memory regions reported by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound of 2 (\#Races Found).
 We did not discover new races with larger context-switch bounds.
@@
 We can see that \whoop reports more races than \corral.
-This is expected, since \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analysing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false alarm race reports.
+This is expected, since \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false alarm race reports.
 On the other hand, \corral is precise, but can miss races because only a limited number of context switches are considered.
@@
 
-\noindent
-\textbf{Granularity of Context-Switches}\xspace\xspace Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks.
+\medskip\noindent\textbf{Granularity of Context-Switches }
+%
+Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Figure~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
 All reported times, in seconds, are averages over three runs.
-\ADComment{Can we say something about variance?
-Perhaps ask Nathan what best to say?} \corral was given a time budget of 10 hours (T.O.
-denotes a tool timeout), and we considered context-switch bounds (csb) of 2, 5 and 9.
+\corral was given a time budget of 10 hours (T.O.
+denotes timeout), and we considered context-switch bounds (csb) of 2, 5 and 9, respectively.
 By default, \corral instruments context-switches (i.e.
 \texttt{yield} statements) in \ADComment{instead of ``in'' can we say either ``before'' or ``after'' (not sure which is accurate)} all visible operations (denoted \yieldall).
-\whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{whoop:bugfinding}), and also uses all the optimisations discussed in Section~\ref{whoop:optimizations}.
+\whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{corral}).
 
@@
 
-%When running \corral on its own, it can potentially report a spurious race (that \whoop does not report), because \corral is not enhanced with domain-specific information that \whoop uses to discard spurious races.
-Using \whoop + \corral, though, will inherently discard spurious data races (assuming a precise environmental model).
-
 %When \corral on its own tries to analyze the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyze a large number of memory regions.

### 2015-05-13 corral.tex
update

@@
 %
-By default, \corral instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) before every shared memory access of each thread.
-It then leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
-Our approach to accelerating \corral is simple and yet effective: we simplify the sequentialization that \corral performs by instrumenting \texttt{yield} statements only at shared memory accesses that \whoop reported as potentially racy.\footnote{In the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} This effectively acts as a partial-order reduction that can significantly reduce the number of interleavings that \corral needs to explore, and thus increase performance and scalability.
+By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) after each shared memory access of each entry point.\footnote{In the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
+Our approach to accelerating \corral is simple and yet effective: we simplify the sequentialization that \corral performs by instrumenting \texttt{yield} statements only at shared memory accesses that \whoop reported as potentially racy.
+This effectively acts as a partial-order reduction that can significantly reduce the number of interleavings that \corral needs to explore, and thus increase performance and scalability.
 
@@
 We now show that our partial-order reduction is sound: it does not miss any more bugs than what \corral would miss even without taking advantage of the \whoop race-related information.
-Let \corral run an $n$ amount of entry point pairs with yields in all \emph{visible} operations (synchronization points and shared memory accesses).
+Let \corral run an $n$ amount of entry point pairs with yields after all \emph{visible} operations (synchronization points and shared memory accesses).
 Let this configuration be known as $C$.
@@
 \corral will miss all bugs that are triggered by more than 3 threads running concurrently and all bugs that are triggered by investigating a larger than $r$ recursion bound and larger than $k$ context-switch bound.
-Let \corral then run the same $n$ amount of pairs, but this time \whoop removes the \texttt{yield} statements in all non-racy shared memory accesses.
+Let \corral then run the same $n$ amount of pairs, but this time \whoop removes the \texttt{yield} statements after all non-racy shared memory accesses.
 Let this configuration be known as $W+C$.

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 denotes timeout), and we considered context-switch bounds (csb) of 2, 5 and 9, respectively.
-By default, \corral instruments context-switches (i.e.
-\texttt{yield} statements) in \ADComment{instead of ``in'' can we say either ``before'' or ``after'' (not sure which is accurate)} all visible operations (denoted \yieldall).
+When we run standalone \corral we enable the \yieldall instrumentation, which instruments context-switches (i.e.
+\texttt{yield} statements) after all visible operations.
 \whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{corral}).

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 
+We can notice that \whoop is significantly faster than \corral.
+This is expected as \whoop achieves scalability using over-approximation.
+For example, \corral timeouts in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze r8169 in 144.5 seconds.
+We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might get \corral stuck.
+
 A higher context-switch bound results into deeper interleavings being explored and thus a larger sequentialized program.
 Intuitively, this means that we see even greater speedups using information from \whoop, when exploring deeper interleavings.
-
-We can notice that \whoop is significantly faster than \corral.
-This is expected as \whoop achieves scalability using over-approximation and summarization techniques.
-This allows \whoop to perform especially well in complex drivers such as the r8169 ethernet driver: although \corral ...
-and timeouts with a csb of 9, \whoop manages to analyze the driver in 228.1 seconds.
-We believe that the reason behind this is that r8169 has many loops and uses deeply-nested recursion in some entry points, which \corral might not be able to handle efficiently.
 

### 2015-05-13 related.tex
Addressed comments at the start of III

@@
 Prior work includes: \cite{choi2002efficient}, which combined static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combined a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combined inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+
+Our pairwise approach to analysing driver entry points, employing abstraction to model additional threads, was inspired by the \emph{two thread reduction} used by the GPUVerify tool in the analysis of data-parallel OpenCL and CUDA kernels~\cite{gpuverify,bardsley2014engineering}.
+The idea of pairwise analysis of components in a concurrent system has been broadly applied, notably in model checking of cache coherence protocols~\cite{mcmillan1999verification}.

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 All reported times, in seconds, are averages over three runs.
-\corral was given a time budget of 10 hours (T.O.
-denotes timeout), and we considered context-switch bounds (csb) of 2, 5 and 9, respectively.
-When we run standalone \corral we enable the \yieldall instrumentation, which instruments context-switches (i.e.
+\corral was configured with a time budget of 10 hours (T.O.
+denotes timeout) and a context-switch bound (csb) of 2, 5 and 9, respectively.
+For standalone \corral we enable the \yieldall instrumentation, which instruments context-switches (i.e.
 \texttt{yield} statements) after all visible operations.
 \whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{corral}).
+The table also shows the number of context-switches per instrumentation (\#Y).
 
-We can notice that \whoop is significantly faster than \corral.
-This is expected as \whoop achieves scalability using over-approximation.
-For example, \corral timeouts in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze r8169 in 144.5 seconds.
-We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might get \corral stuck.
+\whoop uses over-approximation to achieve scalability and, as expected, executes significantly faster than \corral in all our benchmarks.
+For example, \corral timeouts in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
+We believe that the reason behind this is that r8169 has deeply-nested recursion in some entry points, which might get \corral stuck.
+This is not an issue for \whoop, which uses procedure summarization.
+
+Using the race freedom guarantees that \whoop provides, we managed to significantly accelerate \corral in the majority of the cases. We noticed speedups between 1.5--20x.
 

### 2015-05-13 whoop.tex

@@
 In Section~\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.
-In Section~\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code, to check whether pairs of driver entry points can race with one another.
+In Section~\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
 
@@
 
-\ADComment{Top level question about this section: is the motivation super-clear?  Is it obvious why we go for the pairwise approach?}
-
-\PDComment{Mention this somewhere (it was in my previous implementation text)?
-Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.}
-
 Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.
@@
 We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
+Restricting to pairs of entry points, rather than analysing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.
+The trade-off is that a quadratic number of entry point pairs must be checked.
+In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.
 

### 2015-05-13 whoop.tex
Addressed Pantazis's comment by deleting the associated sentence :-)

@@
 Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.
-For each such pair, we use symbolic verification to check whether it is possible for the pair to race in a shared variable.
+For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location.
 We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
@@
 
-Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts that the lockset for each shared variable is non-empty.
+Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.
 Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.
@@
 %
-We consider a simple concurrent programming model in which an unbounded number of threads execute a set of pre-defined \emph{thread templates}.
+We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.
 At any given point of execution a certain number of threads are active, each thread executing a particular template.
@@
 A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.
-A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of a template by a thread.
-A procedure in $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
+A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of $T$ by a thread.
+Each procedure of $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
 A basic block either has a single successor or a pair of successors.
-In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.
+In the latter case, an \emph{exit condition} over thread-private variables determines the successor to which control should flow on block exit.
 
@@
 At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
-
-At any point of execution, a new thread may be added to the list of thread states.
-This involves selecting an id $i \in \ids$ that has not been previously used during program execution and a thread template $T$, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
+%
+At any point of execution, a new thread may be added to the list of threads.
+This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
 
@@
 For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.
-The thread's next statement is updated \PDComment{Do you mean execute?} according to the control flow graph of the current procedure.
 
@@
 
-Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+\ADComment{Renaming must talk about basic block exit conditions} Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
 The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.

### 2015-05-13 background.tex
update to change sections to S symbol

@@
 
-Example~\ref{fig:data_race_example} presents multiple data races that \whoop automatically found in the generic\_nvram Linux char driver (see Section~\ref{evaluation}).
+Example~\ref{fig:data_race_example} presents multiple data races that \whoop automatically found in the generic\_nvram Linux char driver (see \S\ref{evaluation}).
 The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.

### 2015-05-13 evaluation.tex
update to change sections to S symbol

@@
 
-We performed experiments to validate the usefulness of the \whoop approach (Section~\ref{whoop}) and its combination with \corral (Section~\ref{corral}).
+We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}).
 We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
@@
 This can potentially lead \corral to report false positives, but we have not seen this in practice.
-Finally, when we apply \corral to a pair of entry points, we are just checking the specific pair and are not accounting for the effect of other threads (see Section~\ref{corral}); this can also cause \corral to miss some races.
+Finally, when we apply \corral to a pair of entry points, we are just checking the specific pair and are not accounting for the effect of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
 
@@
 \texttt{yield} statements) after all visible operations.
-\whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see Section~\ref{corral}).
+\whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see \S\ref{corral}).
 The table also shows the number of context-switches per instrumentation (\#Y).

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 
-Using the race freedom guarantees that \whoop provides, we managed to significantly accelerate \corral in the majority of the cases. We noticed speedups between 1.5--20x.
-
-A higher context-switch bound results into deeper interleavings being explored and thus a larger sequentialized program.
-Intuitively, this means that we see even greater speedups using information from \whoop, when exploring deeper interleavings.
+Using the race freedom guarantees that \whoop provides, we managed to significantly accelerate \corral in most of our benchmarks; the best performance was achieved using the \yieldmr instrumentation (see Table~\ref{tab:results}).
+Figure~\ref{fig:plot} shows that most of our speedups using \yieldmr are in the range of 1.5--10$\times$; whereas in two drivers, ssu100 and pc8736x\_gpio, we achieved $>$10$\times$ and $>$20$\times$ speedups, respectively (with csb of 9).
+We notice that a higher csb can result into greater speedups, when exploiting \whoop.
+This is expected, as a higher csb directly translates into a larger sequentialized program.
 

### 2015-05-13 whoop.tex

@@
 For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.
+Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.
 
@@
 %
-Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibility that $S$ and $T$ are equal.
-We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is an instance of $T$, in the presence of arbitrarily many further threads that may execute concurrently.
+For thread templates $S$ and $T$ (including the possibility that $S$ and $T$ are equal) we wish to check whether it is possible for a thread executing $S$ to race with a thread executing $T$, in the presence of arbitrarily many further concurrently executing threads.
 
@@
 
-\ADComment{Renaming must talk about basic block exit conditions} Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
 The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.
-Shared variables and mutexes are \emph{not} renamed, because these are shared across templates.
+This is vital in the case where $S$ and $T$ are the same.
 
@@
 
-Initially, lockset instrumentation introduces sets that track: the shared variables that have been read from and written to; the mutexes that have been consistently used to protect each variable; and the mutexes that are currently held by a thread.
-To this end, when instrumenting template $T$ with respect to symbol $i$, we add the sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$.
-For each shared variable $s \in \sharedvars$ we add a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
-Finally, we add a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread.
+Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
 
-The statements of each procedure in $\instrument{T}{i}$ that manipulate shared variables and mutexes are then instrumented to take account of the above sets.
+The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.
 This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.
 A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).
-Notice that in the instrumented form, the shared variable $s$ is no longer assigned to.
-We discuss the reasons for this when we describe \emph{shared state abstraction} below.
-A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command applied to the receiving private variable $x$; again this relates to shared state abstraction and is discussed below.
+A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.
 Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.

### 2015-05-13 whoop.tex
update to change sections to S symbol

@@
 We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.
-In Section~\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.
-In Section~\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
+In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.
+In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
 
-Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in Section~\ref{corral}.
+Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.
 
@@
 For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
-These \emph{are} handled by our implementation, and in Section~\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
+These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
 
@@
 
-To achieve this, we first \emph{instrument} a thread template for lockset analysis (see Section~\ref{bg:lockset}).
+To achieve this, we first \emph{instrument} a thread template for lockset analysis (see \S\ref{bg:lockset}).
 Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.
@@
 The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
-In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see Section~\ref{evaluation}) are indeed terminating.
+In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are indeed terminating.
 
@@
 
-The simple concurrent programming model discussed in Section~\ref{sec:symbolicpairwise} is deliberately idealistic to describe our symbolic verification technique.
+The simple concurrent programming model discussed in \S\ref{sec:symbolicpairwise} is deliberately idealistic to describe our symbolic verification technique.
 In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from Section~\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.
 
@@
 
-Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and Section~\ref{sec:symbolicpairwise}) using the Boogie verification engine.
-After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and Section~\ref{corral}).
+Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.
+After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).
 
@@
 A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.\ pointers that cannot alias).
-We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in Section~\ref{corral}.
+We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in \S\ref{corral}.
 
@@
 %
-Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see Section~\ref{sec:symbolicpairwise}).
+Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise}).
 Such sets can be typically modeled in Boogie using maps.
@@
 We argue that further increasing the precision is orthogonal to the contributions of this paper.
-Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{corral} and~\ref{evaluation}.
+Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 Using the race freedom guarantees that \whoop provides, we managed to significantly accelerate \corral in most of our benchmarks; the best performance was achieved using the \yieldmr instrumentation (see Table~\ref{tab:results}).
-Figure~\ref{fig:plot} shows that most of our speedups using \yieldmr are in the range of 1.5--10$\times$; whereas in two drivers, ssu100 and pc8736x\_gpio, we achieved $>$10$\times$ and $>$20$\times$ speedups, respectively (with csb of 9).
-We notice that a higher csb can result into greater speedups, when exploiting \whoop.
-This is expected, as a higher csb directly translates into a larger sequentialized program.
+Figure~\ref{fig:plot} shows that most of our speedups using \yieldmr are in the range of 1.5--10$\times$; whereas in ssu100, and pc8736x\_gpio respectively, we achieved a speedup greater than 10$\times$, and 20$\times$ respectively, using a csb of 9.
+We notice that a higher csb can potentially result into greater speedups, when exploiting \whoop.
+This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
 

### 2015-05-13 evaluation.tex
update in evaluation text

@@
 We can see that \whoop reports more races than \corral.
-This is expected, since \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false alarm race reports.
-On the other hand, \corral is precise, but can miss races because only a limited number of context switches are considered.
+This is expected, since \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.
+On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.
 Another issue with \corral is loop coverage due to unsound loop unrolling.
 To tackle this, we enable the built-in loop over-approximation described in~\cite{lal2014powering}.
-This can potentially lead \corral to report false positives, but we have not seen this in practice.
-Finally, when we apply \corral to a pair of entry points, we are just checking the specific pair and are not accounting for the effect of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
+This can potentially lead \corral to report false bugs, but we have not seen this in practice.
+Finally, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
 
@@
 \whoop uses over-approximation to achieve scalability and, as expected, executes significantly faster than \corral in all our benchmarks.
-For example, \corral timeouts in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
+For example, \corral times out in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
 We believe that the reason behind this is that r8169 has deeply-nested recursion in some entry points, which might get \corral stuck.
 This is not an issue for \whoop, which uses procedure summarization.
+This shows that \whoop has value as a stand-alone analyzer.
 
@@
 
-%When \corral on its own tries to analyze the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyze a large number of memory regions.
-For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyze.
-The same entry point pair is found race-free by \whoop.
-Using this information in \corral with the \yieldmr instrumentation, the pair takes less than a second to analyze.
-The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.
+However, if \whoop fails to verify any race-related properties of a driver, it might actually slow down \corral.
+This can be noticed in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead.
+Even if \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr.
+This is probably because \corral manages to verify the only 2 pairs of sx8 quite fast (21.4 seconds with csb of 9), and thus the overhead of using \whoop overwhelms any benefits from our fine-grained instrumentation.

### 2015-05-13 intro.tex
update in intro

@@
 We then show that by exploiting the race-freedom guarantees that our symbolic analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
-Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
+Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution, and show that we can achieve speedups in the range of 1.5--10$\times$ in most of our benchmarks, while we managed to achieve a speedup greater than 10$\times$ and 20$\times$ in two cases.
 

### 2015-05-13 whoop.tex
A few tweaks, and wrote something about SMACK's unsoundness issues

@@
 
-The simple concurrent programming model discussed in \S\ref{sec:symbolicpairwise} is deliberately idealistic to describe our symbolic verification technique.
-In practice, the Linux drivers we analyze are written in the C language, we do not know upfront which are the entry points for a driver, we do not have a named set of locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
+The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique.
+In practice, Linux drivers are written in the C, we do not know up-front which are the entry points for a driver, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
 We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.
@@
 Figure~\ref{fig:whoop} depicts the \whoop toolchain.
-Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{Stub header files that model the Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
-Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined verification-focused semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
+Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
+Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
 
@@
 
-Besides the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}), we decided to reuse other industrial-strength tools, because (i) we did not want to ``reinvent the wheel'' and (ii) these tools are robust and battle-proven in real complex software.
+We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).
+For the remaining components, we were able to re-use industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
 
@@
 %
-We have developed Chauffeur, a Clang-frontend that performs two important tasks.
+Chauffeur is a Clang-based frontend that performs two important tasks.
 The first is to traverse (using a Clang visitor pass) the abstract syntax tree (AST) of the original driver and extract all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
@@
 
-The second task is to instrument the original driver with a function that calls all entry points, one after the other, passing them parameters of arbitrary value (to over-approximate kernel behavior for soundness).
-This function effectively acts as the \emph{main} function of the driver and is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions).
-This simple rewriting does not alter the semantics of the original driver, as \whoop only uses this function to create checker functions and instrument them with calls to entry points (see Figure~\ref{fig:sequentialization} for the pseudocode of such checker function).
+\ADComment{This is not clear to me.
+Is the purpose of the main function simply to make sure that the functions are not compiled away?
+I.e., do we then just ignore the main function?
+If that is the case then why does it matter what we pass to the entry points, i.e.
+why should this be some arbitrary value?}  The second task is to instrument the original driver with a function that calls all entry points, one after the other, passing them parameters of arbitrary value (to over-approximate kernel behavior for soundness).
+This function effectively acts as the \emph{main} function of the driver and is required so that Clang does not compile away any entry points that are not otherwise invoked (because we closed the environment using stub functions).
+\ADComment{It is not clear to me why an entry point would be invoked anywhere in the code other than in the main function we add.
+Is it the case that sometimes entry points call each other?
+Or does the kernel sometimes make direct calls to named entry points (rather than calls via function pointers)?} This simple rewriting does not alter the semantics of the original driver, as \whoop only uses this function to create checker functions and instrument them with calls to entry points (see Figure~\ref{fig:sequentialization} for the pseudocode of such checker function).
 Chauffeur outputs the rewritten driver in a C file.
@@
 An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
-This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse proven existing techniques and focus instead on verification efforts.
+This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to re-use robust existing techniques and focus instead on verification efforts.
 
-To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.
+To achieve scalability, SMACK uses a \emph{split} memory model that exploits the case where
+alias analysis asserts that a given memory location cannot be aliased by multiple types.
+This has been shown to lead to more tractable verification compared with a \emph{monolithic} model where the heap is considered to be an array of bytes~\cite{rakamaric2009scalable}.
 This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap.
-A benefit of using this model is that memory regions denote disjoint sections of the heap (i.e.\ pointers that cannot alias).
+A benefit of using this model is that distinct memory regions denote disjoint sections of the heap.
 We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in \S\ref{corral}.
 
-In reality, SMACK is unsound ... (e.g.\ integers), but for our work we assume it is sound.
+We acknowledge that the SMACK translator is subject to sources of
+unsoundness (for example, SMACK models integers as an infinite set,
+rather than as a finite set of bitvectors), and that the combination
+of Clang and SMACK commits our approach to specific choices related to
+undefined and implementation-defined aspects of the C language when
+translating to Boogie.  However, \whoop makes no fundamental
+assumptions related to these translation decisions, meaning that a
+more accurate C-to-Boogie translation would automatically lead to a
+more accurate analysis with \whoop.
 

### 2015-05-13 whoop.tex
Worked on watchdog race checking.

@@
 %
-For thread templates $S$ and $T$ (including the possibility that $S$ and $T$ are equal) we wish to check whether it is possible for a thread executing $S$ to race with a thread executing $T$, in the presence of arbitrarily many further concurrently executing threads.
+For thread templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we wish to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
 
@@
 Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
-The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.
-This is vital in the case where $S$ and $T$ are the same.
+The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.
+This is vital in the case where $T$ and $U$ are the same.
 
@@
 
-\medskip\noindent\textbf{Identifying Set of All Locks }
+\medskip\noindent\textbf{Identifying Locks }
 %
@@
 The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
-
+%
 Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs.
@@
 The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
+\ADComment{Can you re-phrase (ii) to say that using a small, fixed number of locks is advocated by Linux experts as good practice, and cite the drivers book?}
 
@@
 %
-Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise}).
-Such sets can be typically modeled in Boogie using maps.
-However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness.
-To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, which works as follows:
+Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}).
+These sets can be modeled directly in Boogie as characteristic functions, using maps.
+However, this requires the use of quantifiers to express properties related to set contents.
+For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.
+It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.
 
-We introduce a global watchdog variable for each SMACK memory region that is common to both entry points of a pair.
-This variable is an \emph{unconstrained} constant representing the offset to memory region that should be checked for races.
-Verification involves proving that two entry points cannot race at the watched offset of every memory region.
-The arbitrary watched offset implies that every offset of each memory region is race-free.
-Watchdog race-checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
+To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.
+Suppose we are analysing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $Q$.
+When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_Q$, representing some unspecified index into $Q$; we call this the \emph{watched offset} for $Q$.
+We then attempt to prove that it is impossible for $T$ and $U$ to race on $Q$ at index $\mathit{watched}_Q$.
+If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $Q$, since the watched offset was arbitrary.
+This technique of choosing an arbitrary index to analyse for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $Q$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $Q$.
 

### 2015-05-13 whoop.tex
Feedback on watchdog summarization

@@
 To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.
-Suppose we are analysing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $Q$.
-When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_Q$, representing some unspecified index into $Q$; we call this the \emph{watched offset} for $Q$.
-We then attempt to prove that it is impossible for $T$ and $U$ to race on $Q$ at index $\mathit{watched}_Q$.
-If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $Q$, since the watched offset was arbitrary.
-This technique of choosing an arbitrary index to analyse for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $Q$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $Q$.
+Suppose we are analysing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $\mathit{MR}$.
+When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_{\mathit{MR}}$, representing some unspecified index into $\mathit{MR}$; we call this the \emph{watched offset} for $\mathit{MR}$.
+We then attempt to prove that it is impossible for $T$ and $U$ to race on $\mathit{MR}$ at index $\mathit{watched}_{\mathit{MR}}$.
+If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $\mathit{MR}$, since the watched offset was arbitrary.
+This technique of choosing an arbitrary index to analyse for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $\mathit{MR}$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $\mathit{MR}$.
 
@@
 %
-Early on during the development of \whoop, we realized that scalability would be a serious issue.
-When we tried to apply \whoop on the r8169 ethernet driver ($>$7000 lines of C code) with all functions fully inlined, we run out of memory.
-One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory.
-Another reason was that some entry points recursively call helper functions.
-Inherently, recursion does not work well with inlining.
-To tackle these issues, we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze drivers in a scalable fashion and without destroying precision.
+Early in the development of \whoop we experimented with reasoning about recursion-free drivers by resorting to full inlining.
+We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver ($>$7000 lines of C code) with all functions fully inlined, we quickly exhausted the memory limits of our experimental platform.
+\ADComment{Is it thus the case that r8169 did not exhibit any recursion?}  The C-to-Boogie translation produced, after inlining, hundreds of thousands of lines of Boogie IVL code, consuming a lot of memory and leading to gigantic verification conditions that the Z3 theorem prover could not handle effectively.
+\ADComment{I rephrased the previous sentence, because if I remember correctly we did get to the point of invoking Z3.
+But maybe I imagined that - perhaps we could not even generate the VC?}  Furthermore, some entry points invoke recursive helper functions, which cannot be fully inlined.
 
+\ADComment{I think the following is the weakest part of this (in general very good) section at the moment.
+My concerns are that (a) we have not really invented anything new, we just applied Houdini having guessed candidate preconditions, postconditions and loop invariants based on heuristics, and (b) we have not given much intuition for what these candidates look like.
+Could you give an example candidate, using mathematical notation?
+For instance, if entry point $T$ has a procedure $f$ that releases mutex $m$, then when instrumenting $T$ with symbol $i$ (using the notation from \S\ref{sec:symbolicpairwise}) we might guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$.
+It would be ideal if you could be as specific as possible about the candidates we generate.}
+
+To make our analysis scalable, and to cater for recursion, we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze drivers in a scalable fashion and without destroying precision.
 Watchdog summarization uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-The reason we need loop invariants is that we over-approximate loops using loop-cutting (also available inside Boogie).
+The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process; Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
 

### 2015-05-13 whoop.tex
A few more thoughts

@@
 
-\medskip\noindent\textbf{Watchdog Summarization }
+\medskip\noindent\textbf{Watchdog Summarization }  \ADComment{Due to my thoughts below, I'm unsure about calling this ``Watchdog'' summarization, as I don't think we're doing anything new here.
+I think perhaps ``Generatino of Loop and Procedure Summaries'' would be better.}
 %
@@
 To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
+
+\ADComment{In the above, can you make it super-clear that it is fine for the candidates to be wrong: Houdini will just kick them out?
+You might want to look at the comment on this in the GPUVerify 2012 OOPSLA paper; I remember writing something nice along these lines, feel free to re-use it.
+If you search for Houdini you'll find it.}
 

### 2015-05-13 whoop.tex
fix

--- 2015-05-13 whoop.tex
@@
 %
-For thread templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we wish to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
+For thread templates $S$ and $T$ (including the possibility that $S$ and $T$ are equal) we wish to check whether it is possible for a thread executing $S$ to race with a thread executing $T$, in the presence of arbitrarily many further concurrently executing threads.
 
@@
 Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
-The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.
-This is vital in the case where $T$ and $U$ are the same.
+The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.
+This is vital in the case where $S$ and $T$ are the same.
 
@@
 The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique.
-In practice, Linux drivers are written in the C, we do not know up-front which are the entry points for a driver, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
+In practice, Linux drivers are written in C, we do not know up-front which are the entry points for a driver, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
 We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.
@@
 A benefit of using this model is that distinct memory regions denote disjoint sections of the heap.
-We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained granularity of context-switch instrumentation as discussed in \S\ref{corral}.
+We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained context-switch instrumentation as discussed in \S\ref{corral}.
 
@@
 
-\medskip\noindent\textbf{Identifying Locks }
+\medskip\noindent\textbf{Identifying Set of All Locks }
 %
@@
 The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
-%
+
 Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs.
@@
 The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
-\ADComment{Can you re-phrase (ii) to say that using a small, fixed number of locks is advocated by Linux experts as good practice, and cite the drivers book?}
 
@@
 %
-Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}).
-These sets can be modeled directly in Boogie as characteristic functions, using maps.
-However, this requires the use of quantifiers to express properties related to set contents.
-For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.
-It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.
+Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise}).
+Such sets can be typically modeled in Boogie using maps.
+However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness.
+To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, which works as follows:
 
-To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.
-Suppose we are analysing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $Q$.
-When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_Q$, representing some unspecified index into $Q$; we call this the \emph{watched offset} for $Q$.
-We then attempt to prove that it is impossible for $T$ and $U$ to race on $Q$ at index $\mathit{watched}_Q$.
-If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $Q$, since the watched offset was arbitrary.
-This technique of choosing an arbitrary index to analyse for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $Q$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $Q$.
+We introduce a global watchdog variable for each SMACK memory region that is common to both entry points of a pair.
+This variable is an \emph{unconstrained} constant representing the offset to memory region that should be checked for races.
+Verification involves proving that two entry points cannot race at the watched offset of every memory region.
+The arbitrary watched offset implies that every offset of each memory region is race-free.
+Watchdog race-checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.
 

### 2015-05-13 whoop.tex
Comments on error reporting

@@
 %
-Finally, the instrumented and summarized abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to the Z3~\cite{de2008z3} theorem prover.
-Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
+For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, sent to the Boogie verification engine.
+For each procedure in the program, Boogie generates a verification condition~\cite{barnett2005weakest} and discharges these to the Z3~\cite{de2008z3} theorem prover.
+In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
+Successful verification implies that the entry point pair is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
 To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
-The following is an example of a reported error:
+The following is an example of a reported error: \ADComment{Can you replace this with an error Whoop would give for the example device driver bug used earlier in the paper, and refer back to that example?
+That would look more convincing.}
 
@@
 \end{lstlisting}
+
+Because a driver is processed by Clang and SMACK during the translation from C to Boogie IVL, significant engineering effort was required to map Boogie-level errors back to original source locations.
 

### 2015-05-13 whoop.tex
addressed comment regarding locks

--- 2015-05-13 whoop.tex
@@
 because it was created dynamically or was indexed from an array of locks, it will exit with a warning.
-The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.
-\ADComment{Can you re-phrase (ii) to say that using a small, fixed number of locks is advocated by Linux experts as good practice, and cite the drivers book?}
+The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite{corbet2005linux}.
 

### 2015-05-13 whoop.tex
Soundiness

@@
 
-We acknowledge that the SMACK translator is subject to sources of
-unsoundness (for example, SMACK models integers as an infinite set,
-rather than as a finite set of bitvectors), and that the combination
-of Clang and SMACK commits our approach to specific choices related to
-undefined and implementation-defined aspects of the C language when
-translating to Boogie.  However, \whoop makes no fundamental
-assumptions related to these translation decisions, meaning that a
-more accurate C-to-Boogie translation would automatically lead to a
-more accurate analysis with \whoop.
-
 \medskip\noindent\textbf{Identifying Locks }
@@
 %
-We have implemented various optimizations to increase the precision and performance of \whoop. In our experience, the following two are the most effective optimizations:
+We have implemented various optimizations to increase the precision and performance of \whoop.  We comment on the two most effective optimizations.
 
-We enrich \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision.
+First, we enriched \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision.
 The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other.
@@
 This is an ongoing manual effort: the more drivers we study, the more such properties we discover, which we can then exploit to make \whoop more precise.
+\ADComment{Can you make it clearer that this information is currently provided manually to Whoop?
+Can you explain how this is achieved?
+And can you comment briefly on how we figured this out?}
 
@@
 
-We soundly reduce the checked shared variables to increase the performance of \whoop.
-If a shared variable $s$ is accessed by only one entry point in a pair, then it is safe to not check $s$ for this pair.
-This can reduce the burden of the SMT solver and speedup the verification process.
+Second, we soundly reduce the number of memory regions that are analyzed for races.
+If memory region $\mathit{MR}$ is accessed by only one entry point in a pair then, trivially, the pair cannot race on $\mathit{MR}$.
+We thus disable lockset analysis for $\mathit{MR}$.
+This can reduce the complexity of verification conditions that need to be solved by the theorem prover, speeding up the verification process.
 
-\medskip\noindent\textbf{Assumptions }
+\medskip\noindent\textbf{Practical Assumptions Related to Soundness}
 %
+\whoop is ``soundy''\footnote{http://soundiness.org/}~\cite{soundiness}: the tool aims in principle to perform a sound analysis
+that can prove absence of races, but suffers from some known sources of unsoundness
+which we now comment on.
+
 We assume that the formal parameters of an entry point do not alias, and thus cannot race.
-This is a potentially unsound feature that can be turned off using a command line option.
-In our experience so far, though, we have not missed any races because of this assumption.
-We also assume that our environmental model and domain-specific knowledge are correct and do not cause any bugs to be missed.
-Finally, we assume that the external tools that we use as part of the \whoop toolchain are sound and bug-free.
+This is a potentially unsound feature that can be turned off using a command line option, but without this assumption the false alarm rate of the tool is very high.
+In our experience so far we have not missed any races by assuming non-overlapping parameters.
+We also realy on the soundness of our best-effort environmental model, and on the domain-specific knowledge exploit related to entry point serialization.
+
+As well as inheriting soundness issues arising from currently unknown bugs in 
+\whoop and in the external tools that \whoop relies on,
+we acknowledge that the SMACK translator is subject to sources of
+unsoundness (for example, SMACK models integers as an infinite set,
+rather than as a finite set of bitvectors), and that the combination
+of Clang and SMACK commits our approach to specific choices related to
+undefined and implementation-defined aspects of the C language when
+translating to Boogie.  However, \whoop makes no fundamental
+assumptions related to these translation decisions, meaning that a
+more accurate C-to-Boogie translation would automatically lead to a
+more accurate analysis with \whoop.
 
 \medskip\noindent\textbf{Limitations }
+%
+\ADComment{This section is good, but could be shorter.  I suggest you do a ``squeezing'' pass.}
 %

### 2015-05-13 background.tex
addressed some comments

@@
 
+\PDComment{Do we really need the following? I am not even sure how entirely accurate it is as I just used some simple greping ...}
 We believe that our focus in locks is justified, because the majority of Linux drivers still use locks.

### 2015-05-13 whoop.tex
addressed some comments

@@
 %
-Chauffeur is a Clang-based frontend that performs two important tasks.
-The first is to traverse (using a Clang visitor pass) the abstract syntax tree (AST) of the original driver and extract all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
-Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is subsequently parsed by \whoop.
+Chauffeur is a Clang-based frontend that traverses the abstract syntax tree (AST) of the original driver (using a Clang visitor pass) and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
+Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is subsequently parsed by \whoop and used during the instrumentation.
 
@@
 
-\ADComment{This is not clear to me.
-Is the purpose of the main function simply to make sure that the functions are not compiled away?
-I.e., do we then just ignore the main function?
-If that is the case then why does it matter what we pass to the entry points, i.e.
-why should this be some arbitrary value?}  The second task is to instrument the original driver with a function that calls all entry points, one after the other, passing them parameters of arbitrary value (to over-approximate kernel behavior for soundness).
-This function effectively acts as the \emph{main} function of the driver and is required so that Clang does not compile away any entry points that are not otherwise invoked (because we closed the environment using stub functions).
-\ADComment{It is not clear to me why an entry point would be invoked anywhere in the code other than in the main function we add.
-Is it the case that sometimes entry points call each other?
-Or does the kernel sometimes make direct calls to named entry points (rather than calls via function pointers)?} This simple rewriting does not alter the semantics of the original driver, as \whoop only uses this function to create checker functions and instrument them with calls to entry points (see Figure~\ref{fig:sequentialization} for the pseudocode of such checker function).
-Chauffeur outputs the rewritten driver in a C file.
-
 \medskip\noindent\textbf{Translation for Verification }
 %
-The rewritten driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
+Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
 Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
@@
 To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
-The following is an example of a reported error: \ADComment{Can you replace this with an error Whoop would give for the example device driver bug used earlier in the paper, and refer back to that example?
+The following is an example of a reported error: \ADComment{Can you replace this with an error \whoop would give for the example device driver bug used earlier in the paper, and refer back to that example?
 That would look more convincing.}
@@
 This is an ongoing manual effort: the more drivers we study, the more such properties we discover, which we can then exploit to make \whoop more precise.
-\ADComment{Can you make it clearer that this information is currently provided manually to Whoop?
+\ADComment{Can you make it clearer that this information is currently provided manually to \whoop?
 Can you explain how this is achieved?

### 2015-05-13 whoop.tex
fix

@@
 In practice, Linux drivers are written in C, we do not know up-front which are the entry points for a driver, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for soundly detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for detecting data races in drivers.
 
@@
 %
-\whoop is ``soundy''\footnote{http://soundiness.org/}~\cite{soundiness}: the tool aims in principle to perform a sound analysis
-that can prove absence of races, but suffers from some known sources of unsoundness
-which we now comment on.
+\whoop is ``soundy''\footnote{http://soundiness.org/}~\cite{soundiness}: the tool aims in principle to perform a sound analysis that can prove absence of races, but suffers from some known sources of unsoundness which we now comment on.
 
@@
 
-As well as inheriting soundness issues arising from currently unknown bugs in 
-\whoop and in the external tools that \whoop relies on,
-we acknowledge that the SMACK translator is subject to sources of
-unsoundness (for example, SMACK models integers as an infinite set,
-rather than as a finite set of bitvectors), and that the combination
-of Clang and SMACK commits our approach to specific choices related to
-undefined and implementation-defined aspects of the C language when
-translating to Boogie.  However, \whoop makes no fundamental
-assumptions related to these translation decisions, meaning that a
-more accurate C-to-Boogie translation would automatically lead to a
-more accurate analysis with \whoop.
+As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that the SMACK translator is subject to sources of unsoundness (for example, SMACK models integers as an infinite set, rather than as a finite set of bitvectors), and that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
+However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.
 

### 2015-05-13 whoop.tex
addressed some comments

@@
 %
-For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, sent to the Boogie verification engine.
-For each procedure in the program, Boogie generates a verification condition~\cite{barnett2005weakest} and discharges these to the Z3~\cite{de2008z3} theorem prover.
+For each pair of entry points, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
+Boogie generates a verification condition~\cite{barnett2005weakest}, for each procedure in the program, and discharges it to the Z3~\cite{de2008z3} theorem prover.
 In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
-Successful verification implies that the entry point pair is free of data races, while an error denotes a \emph{potential} data race and is reported to the user.
+Successful verification implies that the entry point pair is free of data races, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user.
 To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
-The following is an example of a reported error: \ADComment{Can you replace this with an error \whoop would give for the example device driver bug used earlier in the paper, and refer back to that example?
-That would look more convincing.}
+The following is a race that \whoop found and reported for the example of Figure~\ref{fig:data_race_example}:
 
-\begin{lstlisting}
-foo.c: error: potential write-write race:
-  write by entry point ep1, foo.c:36:2
-    shared->resource = 4;
-  write by entry point ep2, foo.c:52:2
-    shared->resource = 5;
+\begin{lstlisting}[keywordstyle=\ttfamily]
+generic_nvram.c: error: potential read-write race:
+  read by entry point nvram_llseek, generic_nvram.c:54:2
+    return file->f_pos;
+  write by entry point nvram_llseek, generic_nvram.c:53:2
+    file->f_pos = offset;
 \end{lstlisting}
 
-Because a driver is processed by Clang and SMACK during the translation from C to Boogie IVL, significant engineering effort was required to map Boogie-level errors back to original source locations.
+Because a driver is processed by Clang and SMACK during the translation from C to Boogie IVL, significant engineering effort was required to map Boogie-level errors back to original source code locations.
 

### 2015-05-13 whoop.tex
addressed some comments

@@
 Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.
-Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.
+Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged to an automated theorem prover.
 
@@
 For each pair of entry points, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
-Boogie generates a verification condition~\cite{barnett2005weakest}, for each procedure in the program, and discharges it to the Z3~\cite{de2008z3} theorem prover.
+Boogie generates a VC, for each procedure in the program, and discharges it to the Z3~\cite{de2008z3} theorem prover.
 In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
@@
 As an example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock.
-\whoop exploits this knowledge and does not create pairs for entry points that are serialized.
+We discovered this when we first analyzed the r8169 driver (see \S\ref{evaluation}) and \whoop reported many races between a number of networking entry points; when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL.
+\whoop exploits this knowledge and does not create pairs for entry points that are serialized by the kernel.
 This is an ongoing manual effort: the more drivers we study, the more such properties we discover, which we can then exploit to make \whoop more precise.
-\ADComment{Can you make it clearer that this information is currently provided manually to \whoop?
-Can you explain how this is achieved?
-And can you comment briefly on how we figured this out?}
-
-%Modeling kernel methods is another opportunity to increase precision, but requires Linux kernel expertise.
-For example, the \texttt{register\_netdev()} function registers a network driver with the kernel.
-Before this function is called, the kernel is not able to invoke any of the driver's entry points.
-Modeling this functionality can reduce false positives when race-checking accesses before the driver's registration.
+Currently, this information is hard-coded inside \whoop.
+In the future, we plan to provide a configuration file where the user can denote such kernel specific information.
 
@@
 We thus disable lockset analysis for $\mathit{MR}$.
-This can reduce the complexity of verification conditions that need to be solved by the theorem prover, speeding up the verification process.
+This can reduce the complexity of VCs that need to be solved by the theorem prover, speeding up the verification process.
 

### 2015-05-13 whoop.tex
addressed some comments

@@
 %
-\whoop is ``soundy''\footnote{http://soundiness.org/}~\cite{soundiness}: the tool aims in principle to perform a sound analysis that can prove absence of races, but suffers from some known sources of unsoundness which we now comment on.
+\whoop is ``soundy''\footnote{http://soundiness.org/}~\cite{soundiness}: it aims in principle to perform a sound analysis that can prove absence of races, but suffers from some known sources of unsoundness, which we now comment on.
 
 We assume that the formal parameters of an entry point do not alias, and thus cannot race.
-This is a potentially unsound feature that can be turned off using a command line option, but without this assumption the false alarm rate of the tool is very high.
-In our experience so far we have not missed any races by assuming non-overlapping parameters.
-We also realy on the soundness of our best-effort environmental model, and on the domain-specific knowledge exploit related to entry point serialization.
+This is a potentially unsound feature that can be turned off using a command line option, but without this assumption the false alarm rate of \whoop is very high.
+In our experience so far, we have not missed any races by assuming non-overlapping parameters.
+We also rely on the soundness of our best-effort environmental model, and on exploiting domain-specific knowledge related to entry point serialization by the Linux kernel.
 
-As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that the SMACK translator is subject to sources of unsoundness (for example, SMACK models integers as an infinite set, rather than as a finite set of bitvectors), and that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
+As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that SMACK is subject to sources of unsoundness (e.g.\ SMACK models integers as an infinite set, rather than as a finite set of bit-vectors) and that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
 However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.
@@
 \medskip\noindent\textbf{Limitations }
-%
-\ADComment{This section is good, but could be shorter.  I suggest you do a ``squeezing'' pass.}
 %

### 2015-05-13 whoop.tex
trimming down limitations

@@
 %
-\whoop is based on lockset analysis and thus can report false bugs, because a violation of the locking discipline does not always correspond to a real data race (e.g.\ when lockfree synchronization is used instead of locking).
-\whoop also uses over-approximation (e.g.\ when reading from the shared state) and summarization, which can be sources of false positives.
-Furthermore, the tool does not currently check for dynamically created locks or for locks that are provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.
-We also do not currently handle interrupt handlers in a special way; to be sound, we assume that they can execute concurrently with entry points at any time.
-We could address this in the future by modeling interrupt-specific kernel functions (e.g.
+As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lockfree synchronization is used).
+\whoop also over-approximates drivers, which can be another source of imprecision.
+Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.
+We also do not currently treat interrupt handlers in a special way; we just assume that they can execute concurrently with any entry point.
+One way to address this is to model interrupt-specific kernel functions (e.g.
 for enabling/disabling interrupts).

### 2015-05-13 whoop.tex
trimming down limitations

@@
 
-Another limitation of \whoop is that it is unable to verify drivers that are designed to be accessed by a single process at a time.
-This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of each entry point) a flag that indicates device availability: if the flag is set to true, then the respective entry point executes, else it blocks.
-Because \whoop performs pair-wise analysis, it over-approximates this flag, and regards a pair of entry points as running concurrently, even if this is not the case.
+Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time.
+This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks.
+Because \whoop performs pair-wise analysis, it over-approximates this flag, and thus can falsely report a pair as racy.
 However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
-Because using single-open device mode is considered bad practice, we thus burden the developer with disregarding any related false bug reports.
+Because this serialization is considered bad practice, we burden the developer with disregarding any related false bug reports.
 
-Statically analyzing system-level software, such as drivers, requires to ``close'' the environment by abstracting away the low level implementation details.
-In this work, we developed a simple environmental model for the Linux kernel that consists of (nondeterministic) stub functions.
-A limitation of our model is that it can, and will, ultimately result in decreased precision and, thus, false positives.
-However, because we currently only focus on finding data races, we can get away with over-approximating a lot of the underlying Linux kernel functionality, without losing too much precision.
-Enriching our model with domain-specific knowledge to make it more precise is an ongoing manual effort, but requires Linux expertise.
-We argue that further increasing the precision is orthogonal to the contributions of this paper.
-Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.
+Statically analyzing drivers requires to ``close'' the environment by abstracting away the low-level implementation details.
+To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions.
+A limitation of our model is that it can, and will, ultimately result in false positives.
+However, we currently only focus on finding data races, and thus can get away with over-approximating a lot of the underlying kernel functionality, without losing too much precision.
+Making our model more precise is an ongoing manual effort, but requires Linux expertise.
+We argue that further work on the model is orthogonal to the contributions of this paper.
+Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.

### 2015-05-13 whoop.tex
fix

@@
 
-\medskip\noindent\textbf{Watchdog Summarization }  \ADComment{Due to my thoughts below, I'm unsure about calling this ``Watchdog'' summarization, as I don't think we're doing anything new here.
-I think perhaps ``Generatino of Loop and Procedure Summaries'' would be better.}
+\medskip\noindent\textbf{Generating Loop and Procedure Summaries }
 %
 Early in the development of \whoop we experimented with reasoning about recursion-free drivers by resorting to full inlining.
-We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver ($>$7000 lines of C code) with all functions fully inlined, we quickly exhausted the memory limits of our experimental platform.
+We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver ($>$7000 lines of C code and use of recursion) with all functions fully inlined, we quickly exhausted the memory limits of our experimental platform.
 \ADComment{Is it thus the case that r8169 did not exhibit any recursion?}  The C-to-Boogie translation produced, after inlining, hundreds of thousands of lines of Boogie IVL code, consuming a lot of memory and leading to gigantic verification conditions that the Z3 theorem prover could not handle effectively.

### 2015-05-13 whoop.tex
update in summarisation

@@
 Early in the development of \whoop we experimented with reasoning about recursion-free drivers by resorting to full inlining.
-We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver ($>$7000 lines of C code and use of recursion) with all functions fully inlined, we quickly exhausted the memory limits of our experimental platform.
-\ADComment{Is it thus the case that r8169 did not exhibit any recursion?}  The C-to-Boogie translation produced, after inlining, hundreds of thousands of lines of Boogie IVL code, consuming a lot of memory and leading to gigantic verification conditions that the Z3 theorem prover could not handle effectively.
-\ADComment{I rephrased the previous sentence, because if I remember correctly we did get to the point of invoking Z3.
-But maybe I imagined that - perhaps we could not even generate the VC?}  Furthermore, some entry points invoke recursive helper functions, which cannot be fully inlined.
+We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined and blows up.
 
-\ADComment{I think the following is the weakest part of this (in general very good) section at the moment.
-My concerns are that (a) we have not really invented anything new, we just applied Houdini having guessed candidate preconditions, postconditions and loop invariants based on heuristics, and (b) we have not given much intuition for what these candidates look like.
+\ADComment{We have not given much intuition for what these candidates look like.
 Could you give an example candidate, using mathematical notation?
@@
 
-To make our analysis scalable, and to cater for recursion, we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze drivers in a scalable fashion and without destroying precision.
-Watchdog summarization uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process; Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
+To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
+The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process.
+Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
 
-The candidate invariants are based on the aforementioned global watchdog variables that \whoop instruments for watchdog race-checking.
-The key idea behind watchdog summarization is to automatically generate candidate invariants for each memory region in the granularity of watched accesses.
+The candidate invariants are based on the aforementioned watchdog constants that \whoop instruments for watchdog race-checking.
+Our candidate generation rules are purely heuristic.
+The key idea is to automatically generate candidates for each memory region in the granularity of watched accesses.
 This reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing.
@@
 
-\ADComment{In the above, can you make it super-clear that it is fine for the candidates to be wrong: Houdini will just kick them out?
-You might want to look at the comment on this in the GPUVerify 2012 OOPSLA paper; I remember writing something nice along these lines, feel free to re-use it.
-If you search for Houdini you'll find it.}
+We emphasize that the candidate invariants generated by \whoop are just that: candidates.
+The tool is free to speculatively generate candidates that later turn out to be incorrect: these are simply discarded by Houdini.
+A consequence is that incorrect or unintended candidates generated due to bugs in \whoop cannot compromise the soundness of verification.
 

### 2015-05-13 whoop.tex
updates

@@
 In practice, Linux drivers are written in C, we do not know up-front which are the entry points for a driver, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, fully automatic tool for detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop\footnote{\url{https://github.com/pdeligia/whoop}}, a practical, fully automatic tool for detecting data races in drivers.
 
@@
 
+To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
+The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process.
+Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
+
 \ADComment{We have not given much intuition for what these candidates look like.
@@
 It would be ideal if you could be as specific as possible about the candidates we generate.}
-
-To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process.
-Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
 
@@
 %
-\whoop is ``soundy''\footnote{http://soundiness.org/}~\cite{soundiness}: it aims in principle to perform a sound analysis that can prove absence of races, but suffers from some known sources of unsoundness, which we now comment on.
+\whoop is ``soundy''\footnote{\url{http://soundiness.org/}}~\cite{soundiness}: it aims in principle to perform a sound analysis that can prove absence of races, but suffers from some known sources of unsoundness, which we now comment on.
 

### 2015-05-13 whoop.tex
update in summarisation

@@
 We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analysing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.
 The trade-off is that a quadratic number of entry point pairs must be checked.
@@
 To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.
-Suppose we are analysing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $\mathit{MR}$.
+Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $\mathit{MR}$.
 When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_{\mathit{MR}}$, representing some unspecified index into $\mathit{MR}$; we call this the \emph{watched offset} for $\mathit{MR}$.
@@
 If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $\mathit{MR}$, since the watched offset was arbitrary.
-This technique of choosing an arbitrary index to analyse for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $\mathit{MR}$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $\mathit{MR}$.
+This technique of choosing an arbitrary index to analyze for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $\mathit{MR}$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $\mathit{MR}$.
 
@@
 
-\ADComment{We have not given much intuition for what these candidates look like.
-Could you give an example candidate, using mathematical notation?
-For instance, if entry point $T$ has a procedure $f$ that releases mutex $m$, then when instrumenting $T$ with symbol $i$ (using the notation from \S\ref{sec:symbolicpairwise}) we might guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$.
-It would be ideal if you could be as specific as possible about the candidates we generate.}
-
-The candidate invariants are based on the aforementioned watchdog constants that \whoop instruments for watchdog race-checking.
+The candidate invariants are based on the watchdog constants that \whoop instruments for watchdog race-checking.
 Our candidate generation rules are purely heuristic.
 The key idea is to automatically generate candidates for each memory region in the granularity of watched accesses.
-This reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing.
+For example, if entry point $T$ has a procedure $f$ that only writes (and not reads) the offset $z$ of memory region $\mathit{MR}$, which is associated with the write set $\hasbeenwritten{\mathit{MR}, i}$ and read set $\hasbeenread{\mathit{MR}, i}$, then when instrumenting $T$ with symbol $i$ (using the notation from \S\ref{sec:symbolicpairwise}), we might guess $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \in \hasbeenwritten{\mathit{MR}, i}$ and $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \notin \hasbeenread{\mathit{MR}, i}$, as postconditions for $\instrument{f}{i}$.
+Likewise, if $f$ releases mutex $m$, then we might guess $m \notin \currentlockset{i}$ as a postcondition.
 
+Using watchdog-based candidates reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing.
 To generate the candidate invariants, \whoop performs an inter-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
-To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region.
+To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region, e.g.\  $z \in \hasbeenwritten{\mathit{MR}, i}$ and $z \notin \hasbeenread{\mathit{MR}, i}$ for the above example.
 

### 2015-05-13 related.tex
update

@@
-Statically analyzing concurrent programs to detect races is a promising alternative to dynamic analyzers, which commonly face code coverage issues as they rely on a (controlled) scheduler for exploring execution paths~\cite{musuvathi2008finding}.
+Statically analyzing concurrent programs for races is a promising alternative to dynamic analyzers, which typically rely on a (controlled) scheduler for exploring execution paths, and thus can face code coverage issues~\cite{musuvathi2008finding}.
 Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable examples of static race analyzers.
@@
 
-Most related to our static lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
-Both tools, though, have significant limitations.
-The authors of RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races.
-To limit these false positives, RELAY employs post-analysis unsound filters, but these can also filter out true races.
-Although Locksmith successfully detected data races in several small Pthreads applications and 7 medium-sized Linux device drivers, it also reported a significant number of false positives.
-The authors also reported that Locksmith was unable to run on several large programs, showcasing its limited scalability.
-In contrast, \whoop is based on novel over-approximation techniques and uses modern SMT solvers to accelerate state-of-the-art concurrency bug-finders, such as \corral, for precise \emph{and} scalable analysis.
+Most related to our lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
+Both tools, though, have several limitations.
+RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races.
+To tackle this issue, RELAY employs post-analysis unsound filters, but these can also filter out real bugs.
+Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, showcasing its limited scalability.
+\whoop aims to achieve scalability \emph{and} precision: the first, via novel symbolic pairwise lockset analysis; and the second, by accelerating \corral, an industrial-strength precise bug-finder.
 

### 2015-05-13 conclusion.tex
update

@@
-In this paper we presented \whoop, a new fully automatic approach that can soundly detect all possible data races in device drivers.
-Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation, summarization and symbolic pairwise lockset analysis, and thus can scale well.
-Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art precise bug-finder for concurrent programs.
-Using the combination of \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our approach can analyze programs 1.5--20$\times$ faster than standalone \corral.
+In this paper we presented \whoop, a new fully automatic approach for detecting all possible data races in device drivers.
+Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which has the potential to scale well.
+Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrent bug-finder.
+Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our technique can analyze programs 1.5--20$\times$ faster than standalone \corral.

### 2015-05-13 evaluation.tex
fix
update

@@
 For example, \corral times out in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
-We believe that the reason behind this is that r8169 has deeply-nested recursion in some entry points, which might get \corral stuck.
+We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might get \corral stuck.
 This is not an issue for \whoop, which uses procedure summarization.
@@
 This is probably because \corral manages to verify the only 2 pairs of sx8 quite fast (21.4 seconds with csb of 9), and thus the overhead of using \whoop overwhelms any benefits from our fine-grained instrumentation.
+
+\medskip\noindent\textbf{Other Tools }
+%
+We tried to compare \whoop with other similar approaches (see \S\ref{related}).
+However, we found this hard to achieve, as some tools are not freely available, while others are outdated for several years and are not working on the 4.0 Linux drivers.

### 2015-05-13 evaluation.tex
update

@@
 
-Using the race freedom guarantees that \whoop provides, we managed to significantly accelerate \corral in most of our benchmarks; the best performance was achieved using the \yieldmr instrumentation (see Table~\ref{tab:results}).
-Figure~\ref{fig:plot} shows that most of our speedups using \yieldmr are in the range of 1.5--10$\times$; whereas in ssu100, and pc8736x\_gpio respectively, we achieved a speedup greater than 10$\times$, and 20$\times$ respectively, using a csb of 9.
-We notice that a higher csb can potentially result into greater speedups, when exploiting \whoop.
+Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr.
+Figure~\ref{fig:plot} shows that most speedups using \yieldmr are in the range of 1.5--10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup greater than 10$\times$, and 20$\times$ respectively, using a csb of 9.
+We notice that a higher csb typically results into greater speedups, when exploiting \whoop.
 This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
@@
 However, if \whoop fails to verify any race-related properties of a driver, it might actually slow down \corral.
-This can be noticed in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead.
-Even if \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr.
-This is probably because \corral manages to verify the only 2 pairs of sx8 quite fast (21.4 seconds with csb of 9), and thus the overhead of using \whoop overwhelms any benefits from our fine-grained instrumentation.
+We noticed this in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead.
+Although \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr.
+This might be because \corral verifies the only 2 pairs of sx8 quite fast (21.4 seconds with csb of 9), and thus the overhead of using \whoop overwhelms any benefits from our fine-grained instrumentation.
 

### 2015-05-13 whoop.tex
Added a blurn about the soudness of SMACK's emory model.

@@
 An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
-This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to re-use robust existing techniques and focus instead on verification efforts.
+This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse robust existing techniques and focus instead on verification efforts.
 
-To achieve scalability, SMACK uses a \emph{split} memory model that exploits the case where
-alias analysis asserts that a given memory location cannot be aliased by multiple types.
+To achieve scalability, SMACK uses a \emph{split} memory model that exploits an alias analysis to soundly partition memory locations into non-overlapping equivalence classes that do not alias.
 This has been shown to lead to more tractable verification compared with a \emph{monolithic} model where the heap is considered to be an array of bytes~\cite{rakamaric2009scalable}.
-This split memory model is based on \emph{memory regions}, which are maps of integers that model the heap.
+The split memory model is based on \emph{memory regions}, which are maps of integers that model the heap.
 A benefit of using this model is that distinct memory regions denote disjoint sections of the heap.
 We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained context-switch instrumentation as discussed in \S\ref{corral}.
+Note that the SMACK's memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses.
+The latest release of SMACK addresses this issue, and as future work we plan to integrate the latest SMACK with \whoop to further improve soundness.
 

### 2015-05-14 background.tex
cutting space in bg

@@
 
-Modern operating systems address the demand for responsiveness and performance by providing multiple sources of concurrency~\cite{corbet2005linux}: a kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging).
+Modern operating systems address the demand for responsiveness and performance by providing multiple sources of concurrency~\cite{corbet2005linux}, such as: a kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.
 All these possibilities can easily lead to \emph{data races}, which can be defined as follows:

### 2015-05-14 whoop.tex
cutting space in bg

@@
 We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).
-For the remaining components, we were able to re-use industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
+For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
 

### 2015-05-14 background.tex
cutting space in bg

@@
 The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.
-However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can read/write access the \texttt{f\_pos} field of the \texttt{file} object in arbitrary order resulting into nondeterministic behavior.
+However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of the \texttt{file} object in arbitrary order.
 

### 2015-05-14 background.tex
improvements in bg

@@
 
-Example~\ref{fig:data_race_example} presents multiple data races that \whoop automatically found in the generic\_nvram Linux char driver (see \S\ref{evaluation}).
+Example~\ref{fig:data_race_example} shows a racy entry point that \whoop (see \S\ref{whoop}) automatically found in the generic\_nvram Linux char driver (see \S\ref{evaluation}).
 The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.
@@
 
-The most common technique for eliminating data races is using a \emph{locking} mechanism.
-Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e.\ source code that cannot be executed by more than one thread simultaneously.
+The most common technique for eliminating races is using a \emph{locking} mechanism.
+Locking a set of program statements that access a shared resource with a lock $l$ creates a \emph{critical section}, i.e.\ source code that cannot execute concurrently by threads that hold $l$.
 Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}.

### 2015-05-14 background.tex
improvements in bg

@@
 
-The most common technique for eliminating races is using a \emph{locking} mechanism.
-Locking a set of program statements that access a shared resource with a lock $l$ creates a \emph{critical section}, i.e.\ source code that cannot execute concurrently by threads that hold $l$.
+The most common technique for eliminating races is using a lock $l$.
+Locking a set of program statements that access a shared resource with $l$ creates a \emph{critical section}, i.e.\ source code that can only execute by the thread that holds $l$.
 Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}.

### 2015-05-14 background.tex
fix
improvements in bg

@@
 Locking a set of program statements that access a shared resource with $l$ creates a \emph{critical section}, i.e.\ source code that can only execute by the thread that holds $l$.
-Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}.
+Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}.
 The example is tricky, because the return statement can potentially race on the \texttt{f\_pos} field.
@@
 
-Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}.
-Coarse-grained locking can hurt performance, because it limits the opportunity for concurrency.
+Lock-based synchronization is a powerful tool for preventing races in drivers, but careless use of locks has many pitfalls~\cite{sutter2005software}.
+Coarse-grained locking can hurt performance, as it limits the opportunity for concurrency.
 On the other hand, fine-grained locking can easily lead to deadlocks.
 Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read and write operations, in this work we focus on locks.
-However, we still treat lock-free operations (e.g.\ atomics) soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads.
-This is sound but can potentially introduce false positives.
+However, we still treat lock-free operations soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads.
+This is sound, but can potentially introduce false positives.
 

### 2015-05-14 background.tex
fixes for consistency

@@
 The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution.
-If that lockset ever becomes empty, the analysis reports a potential data race on that memory location.
+If that lockset ever becomes empty, the analysis reports a potential race on that memory location.
 This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
@@
 
-Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_v$, for each shared variable $v$ used in the program.
+Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_s$, for each shared variable $s$ used in the program.
 Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start.
-The lockset $\mathit{LS}_v$ for each variable $v$ is initialized to the set of all locks manipulated by the program, because initially each access to $v$ has (vacuously) been protected by every lock.
+The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock.
 The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows:
@@
 \item After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
-\item When $T$ accesses shared variable $v$, any locks that are not common to $\mathit{LS}_v$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_v$.
-If $\mathit{LS}_v$ becomes empty as a result, a warning is issued that the access to $v$ may be unprotected.
+\item When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
+If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.
 \end{enumerate}

### 2015-05-14 whoop.tex

@@
 To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process.
-Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
+Given a pool of candidate invariants, Houdini repeatedly tries to verify each procedure with respect to its current candidate summary.
+If verification fails due to an incorrect candidate, this candidate is discarded.
+The process repeats until a fixpoint is reached, whereby the largest conjunction of the candidate invariants that can be proven to hold reamins.
+The remaining candidates, which are indeed invariants, can then be assumed during further analysis of the program.
 
-The candidate invariants are based on the watchdog constants that \whoop instruments for watchdog race-checking.
-Our candidate generation rules are purely heuristic.
-The key idea is to automatically generate candidates for each memory region in the granularity of watched accesses.
-For example, if entry point $T$ has a procedure $f$ that only writes (and not reads) the offset $z$ of memory region $\mathit{MR}$, which is associated with the write set $\hasbeenwritten{\mathit{MR}, i}$ and read set $\hasbeenread{\mathit{MR}, i}$, then when instrumenting $T$ with symbol $i$ (using the notation from \S\ref{sec:symbolicpairwise}), we might guess $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \in \hasbeenwritten{\mathit{MR}, i}$ and $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \notin \hasbeenread{\mathit{MR}, i}$, as postconditions for $\instrument{f}{i}$.
-Likewise, if $f$ releases mutex $m$, then we might guess $m \notin \currentlockset{i}$ as a postcondition.
+The Houdini algorithm does not generate the initial pool of candidate invariants; these are given to the algorithm as a starting point.
+\whoop generates an initial pool of candidates using a set of heuristics.
+The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedureal pass over the code for an entry point.
+We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.
+If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$.
+Then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.
+These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.
+Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
+We stress that guessing of incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.
+The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.
 
-Using watchdog-based candidates reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing.
-To generate the candidate invariants, \whoop performs an inter-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
-To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region, e.g.\  $z \in \hasbeenwritten{\mathit{MR}, i}$ and $z \notin \hasbeenread{\mathit{MR}, i}$ for the above example.
-
-We emphasize that the candidate invariants generated by \whoop are just that: candidates.
-The tool is free to speculatively generate candidates that later turn out to be incorrect: these are simply discarded by Houdini.
-A consequence is that incorrect or unintended candidates generated due to bugs in \whoop cannot compromise the soundness of verification.
+In practice, the candidates are generated at the Boogie level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race instrumentation.
 

### 2015-05-14 corral.tex
update to change sections to S symbol
update
Worked on Whoop+Corral section, though some of it I'd like to discuss.

@@
-\whoop is a sound but potentially imprecise static data race analyzer.
-While developers do appreciate soundness, it is more important to ensure they are not overwhelmed by countless reports of false bugs.
-Hence, to improve precision, we enable an external precise bug-finder for concurrent programs to be plugged into \whoop.
-The race-related information generated by \whoop can then be used to accelerate the bug-finder.
-Currently, we support \corral~\cite{lal2012corral, lal2014powering}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers.
-\corral accepts Boogie IVL and hence it was easy to integrate into our toolchain (see Figure~\ref{fig:whoop} -- C).
-Our technique, though, is general and capable of accelerating any arbitrary bug-finder for concurrent programs.
+\ADComment{Top-level questions on this section: do we make it clear how \corral identifies a race?
+And do we make it clear enough that although in our experiments we consider accelerating race detection, we can more generally accelerate assertion checking?}
 
-\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors.
+\whoop is a sound but imprecise static data race analyzer.
+For developers for whom false alarms are not acceptable, we consider a method for leveraging full or partial
+race-freedom guarantees provided by \whoop in order to accelerate the perforance of 
+\corral~\cite{lal2012corral}, a precise bug-finding tool.
+\corral is used by Microsoft for analysis of Windows device drivers~\cite{lal2014powering},
+and has capabilities for analysing concurrent programs.  Because \corral operates on Boogie programs,
+it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C).
+Our technique, though, is general and capable in principle of accelerating any concurrency bug-finder that operates
+by interleaving threads at shared memory operations.
+
+\corral is a symbolic bounded verifier for Boogie that uses the Z3 SMT solver to statically reason about program behaviors.
 It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found.
@@
 Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program.
-It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.
+It uses two key techniques to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.
 
-Because \corral performs bounded verification both in terms of the number of allowed context-switches and loop/recursion unrollings, it is inherently unsound; in other words, it can miss real bugs.
+Because \corral performs bounded verification both in terms of the number of allowed context-switches and loop/recursion unrollings, it is inherently unsound: it can miss real bugs.
 On the other hand, \corral is precise, and assuming a precise model of the program environment it will only report true bugs.
@@
 %
-By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) after each shared memory access of each entry point.\footnote{In the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
+By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) after each shared memory access of each entry point.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \ADComment{We should say here that lock/unlock operations also have yields.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
 Our approach to accelerating \corral is simple and yet effective: we simplify the sequentialization that \corral performs by instrumenting \texttt{yield} statements only at shared memory accesses that \whoop reported as potentially racy.
@@
 
-\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
-about entry point pairs, or could we prove it just for any pair?}
-
-We now show that our partial-order reduction is sound: it does not miss any more bugs than what \corral would miss even without taking advantage of the \whoop race-related information.
+We now argue that our partial-order reduction is sound: it does not cause \corral to miss any additional bugs.
+\ADComment{Let's go over this together.}
 Let \corral run an $n$ amount of entry point pairs with yields after all \emph{visible} operations (synchronization points and shared memory accesses).
@@
 regions that can potentially race (regardless if a pair has not been fully
-verified as race-free). In our experiments (see Section~\ref{evaluation}),
+verified as race-free). In our experiments (see \S\ref{evaluation}),
 \yieldmr performs significantly faster than \yieldcoarse.
 
-In this work, we use \corral to analyze separate pairs of entry points (as outputted by \whoop), instead of the whole Boogie driver (as outputted by SMACK).
-We also do not employ any abstraction to model additional threads, as we want \corral to report only true races.
-Because we only analyze pairs, though, \corral can miss races that require more than two threads to manifest.
-We argue that this limitation is orthogonal to this work: as long as we compare the same pairs with standalone \corral and with \corral that takes advantage of \whoop, then our comparison is sound.
-In reality, we could allow \corral to analyze the whole driver, and then instrument \texttt{yield} statements in a coarser granularity (taking into account all threads).
+In this work, we use \corral to analyze individual pairs of entry points.
+We do not employ any abstraction to model additional threads, because we want \corral to report only true races.
+Because we only analyze pairs, though, \corral will miss races that require more than two threads to manifest.
+We could extend our setup so that more than two threads are considered by \corral, though because the number of threads that may be launched by an OS kernel is unknown in general, we are inevitably limited by some fixed maximum thread count.
 

### 2015-05-14 intro.tex
Space
Cuts to intro

--- 2015-05-12 intro.tex
@@
 This has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
-
+%
 Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
@@
 For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.
-All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (such as locks) are in place to control concurrent access to shared resources.
+All these scenarios can easily cause \emph{data races} if proper synchronization mechanisms (such as locks) are not in place to control concurrent access to shared resources.
 
-Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of complex operating systems.
-Further, according to the C language standard, races result in undefined behavior, which can be the source of tricky compilation and optimization issues (as compilers typically assume race-freedom).
-Realistic concurrent programs also have an exponentially large state-space, which presents a serious challenge for implementing efficient race-checking techniques~\cite{musuvathi2008finding}.
-
-Techniques such as~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been successfully used to analyze device drivers, but mostly focus on generic sequential program properties and finding protocol bugs.
-Well-regarded Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races.
+Data races are a source of undefined behavior in C, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
+%
+Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs.
+Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races.
 Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability.
@@
 In this paper, we present \whoop, a new fully automatic approach that (i) statically analyzes device drivers written in the C language for data races and (ii) exploits the race-related results from the symbolic analysis to accelerate bug-finding.
-\whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in device drivers.
-The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race-checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock.
-The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.
+\whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for verifying the absence of data races in device drivers.
+We attempt to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it to record \emph{locksets}, and (iii) using the locksets to assert that all accesses to the same shared resource are protected by at least one common lock.
+Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
 
-We then show that by exploiting the race-freedom guarantees that our symbolic analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs.
+We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art bug-finder used by Microsoft to analyze device drivers, as well as other concurrent programs.
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
+We find that by combining \whoop and \corral we achieve concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
+For two drivers, we observe even greater speedups of \ADComment{first speedup} and \ADComment{second speedup}.
 
-\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g.\ Windows) as well as other concurrent programs with similar structure (e.g.\ filesystems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic, thus could be applied to analyze drivers for other operating systems (e.g.\ Windows) as well as other concurrent systems that use a similar programming model (e.g.\ filesystems).
 
@@
 \begin{itemize}
-\item We present \whoop, a new fully automatic approach for analyzing device drivers for data races and accelerating bug-finding.
-\item We discuss symbolic pairwise lockset analysis,  a sound and scalable technique for automatically verifying the absence of data races in device drivers.
-\item We show that we can achieve a sound partial-order reduction using our technique and significantly speedup \corral, a state-of-the-art bug-finder.
-\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient in race-checking and accelerating \corral.
+\item We propose symbolic pairwise lockset analysis, a sound and scalable technique for automatically verifying the absence of data races in device drivers.
+\item We present \whoop, a tool that leverages our approach for analyzing data races in device drivers.
+\item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, a state-of-the-art bug-finder.
+\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race checking and accelerating \corral.
 \end{itemize}

### 2015-05-14 background.tex
Space

@@
 
-Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput).
-In drivers, responsiveness is important (even if there is only a single core available), because entry points of distinct processes should run concurrently so that no single process hogs the core.
-Likewise, performance is important, as drivers should be able to timely process incoming requests (e.g.\ network packages).
-
-Modern operating systems address the demand for responsiveness and performance by providing multiple sources of concurrency~\cite{corbet2005linux}, such as: a kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.
-All these possibilities can easily lead to \emph{data races}, which can be defined as follows:
+Modern operating systems address the demand for responsiveness and performance in device drivers by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.
+These forms of concurrent execution are prone to \emph{data races}.
 
@@
 \label{definition:datarace}
-A data race occurs when two distinct threads access a shared memory location, at least one of the accesses modifies the location, at least one of the accesses is non-atomic, and there is no mechanism in place to prevent these accesses from being simultaneous.
+A \emph{data race} occurs when two distinct threads access a shared memory location, at least one of the accesses modifies the location, at least one of the accesses is non-atomic, and there is no mechanism in place to prevent these accesses from being simultaneous.
 \end{definition}
 
-Example~\ref{fig:data_race_example} shows a racy entry point that \whoop (see \S\ref{whoop}) automatically found in the generic\_nvram Linux char driver (see \S\ref{evaluation}).
+Example~\ref{fig:data_race_example} shows a racy entry point that in the generic\_nvram Linux char driver;
+our \whoop (see \S\ref{whoop}) was able to find this race automatically 
+(see \S\ref{evaluation}).
 The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.
@@
 
-The most common technique for eliminating races is using a lock $l$.
-Locking a set of program statements that access a shared resource with $l$ creates a \emph{critical section}, i.e.\ source code that can only execute by the thread that holds $l$.
+The most common method for avoiding races is to protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.  
 Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}.
-The example is tricky, because the return statement can potentially race on the \texttt{f\_pos} field.
-To fix this issue, we store the result in a temporary variable, inside the critical section.
+Because the return statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
 
@@
 On the other hand, fine-grained locking can easily lead to deadlocks.
-Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read and write operations, in this work we focus on locks.
-However, we still treat lock-free operations soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads.
-This is sound, but can potentially introduce false positives.
-
-\PDComment{Do we really need the following? I am not even sure how entirely accurate it is as I just used some simple greping ...}
-We believe that our focus in locks is justified, because the majority of Linux drivers still use locks.
-To check this, we used a simple grep command and found that 68\% of the drivers from the 4.0 Linux kernel distribution use locks, versus 13\% that use atomic operations.
-We also found that 31\% of the drivers use neither locking nor atomic primitives, which suggests that they are either simple (non-concurrent) drivers or use some other lock-free synchronization primitive.
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations by not regarding them as providing any protection between threads.
+This is sound, but can lead to false alarm race reports.}
 

### 2015-05-14 evaluation.tex
fix in evaluation

@@
 Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr.
-Figure~\ref{fig:plot} shows that most speedups using \yieldmr are in the range of 1.5--10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup greater than 10$\times$, and 20$\times$ respectively, using a csb of 9.
+Figure~\ref{fig:plot} shows that most speedups using \yieldmr are in the range of 1.5--10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9.
 We notice that a higher csb typically results into greater speedups, when exploiting \whoop.

### 2015-05-14 corral.tex
POR

@@
-\ADComment{Top-level questions on this section: do we make it clear how \corral identifies a race?
-And do we make it clear enough that although in our experiments we consider accelerating race detection, we can more generally accelerate assertion checking?}
+\ADComment{Suggestions: discuss race instrumentation for Corral immediately after explaining how corral works.
+Don't say ``we regard races as bugs''.
+After POR part, emphasise that this idea can be used to accelerate Corral for arbitrary bug-finding in concurrent programs.
+Then make it clear that in the remainder of the paper we regard races as bugs and use corral solely to look for races.}
 
@@
 %
-By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) after each shared memory access of each entry point.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \ADComment{We should say here that lock/unlock operations also have yields.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
-Our approach to accelerating \corral is simple and yet effective: we simplify the sequentialization that \corral performs by instrumenting \texttt{yield} statements only at shared memory accesses that \whoop reported as potentially racy.
-This effectively acts as a partial-order reduction that can significantly reduce the number of interleavings that \corral needs to explore, and thus increase performance and scalability.
-
-We now argue that our partial-order reduction is sound: it does not cause \corral to miss any additional bugs.
-\ADComment{Let's go over this together.}
-Let \corral run an $n$ amount of entry point pairs with yields after all \emph{visible} operations (synchronization points and shared memory accesses).
-Let this configuration be known as $C$.
-Each entry point pair is running in a separate \corral process.
-\corral will miss all bugs that are triggered by more than 3 threads running concurrently and all bugs that are triggered by investigating a larger than $r$ recursion bound and larger than $k$ context-switch bound.
-Let \corral then run the same $n$ amount of pairs, but this time \whoop removes the \texttt{yield} statements after all non-racy shared memory accesses.
-Let this configuration be known as $W+C$.
-Because these shared memory accesses cannot race, performing a \texttt{yield} will not introduce any new program behaviors.
-Thus, \corral will find the same bugs using the configuration $W+C$, as using the configuration $C$.
+By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
+Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, we do not instrument a \texttt{yield} after this statement.
+This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.
+This is because each access for which a \texttt{yield} is suppressed is guaranteed to be protected by some lock (a consequence of lockset analysis).
+If the access is a write, its effects are not visible by the other entry point until the lock is released.
+If the access is a read, the value of the shared location cannot change until the lock is released.
+The fact that a \texttt{yield} is placed after each unlock operation suffices to take account of communication between entry points via the shared memory location.
 

### 2015-05-14 whoop.tex
Tweaks

@@
 In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
-Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.
-For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
+Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
 These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
@@
 The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.
-This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.
+This transformation is described in Figure~\ref{fig:instrumentation}.
 A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).

### 2015-05-14 intro.tex
POR
hacks in introduction

@@
 
-A major reason behind these concurrency bugs is that device drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}.
-For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation.
-All these scenarios can easily cause \emph{data races} if proper synchronization mechanisms (such as locks) are not in place to control concurrent access to shared resources.
+A major reason behind these concurrency bugs is that drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}.
+For example the Linux kernel can invoke its drivers from multiple concurrent threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may decide to remove a device (hot-plugging) while some operation is still running.
+All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (e.g.\ locks) are in place to control concurrent access to shared resources.
 
@@
 
-In this paper, we present \whoop, a new fully automatic approach that (i) statically analyzes device drivers written in the C language for data races and (ii) exploits the race-related results from the symbolic analysis to accelerate bug-finding.
+In this paper, we present \whoop, a new fully automatic approach that statically analyzes device drivers written in C for data races.
 \whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for verifying the absence of data races in device drivers.
-We attempt to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it to record \emph{locksets}, and (iii) using the locksets to assert that all accesses to the same shared resource are protected by at least one common lock.
+We attempt to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are \emph{consistently} protected by at least one common lock.
 Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
 
-We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art bug-finder used by Microsoft to analyze device drivers, as well as other concurrent programs.
-Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.
+We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers, as well as other concurrent programs.
+Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.
 We find that by combining \whoop and \corral we achieve concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
-For two drivers, we observe even greater speedups of \ADComment{first speedup} and \ADComment{second speedup}.
+For two drivers, we observe even greater speedups of 12$\times$ and 20$\times$.
 
@@
 \item We present \whoop, a tool that leverages our approach for analyzing data races in device drivers.
-\item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, a state-of-the-art bug-finder.
-\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race checking and accelerating \corral.
+\item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, an industrial-strength bug-finder.
+\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race-checking and accelerating \corral.
 \end{itemize}

### 2015-05-14 background.tex
hacks in bg

@@
 
-Example~\ref{fig:data_race_example} shows a racy entry point that in the generic\_nvram Linux char driver;
-our \whoop (see \S\ref{whoop}) was able to find this race automatically 
-(see \S\ref{evaluation}).
-The driver can invoke the \texttt{nvram\_llseek} entry point concurrently from two threads $T_1$ and $T_2$, while passing the same \texttt{file} object as a parameter.
-However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of the \texttt{file} object in arbitrary order.
+Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux driver.
+The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} object as a parameter.
+However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order.
+Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).
 
@@
 \end{lstlisting}
-\caption{Example of data races in the generic\_nvram Linux char driver}
+\caption{Racy entry point in the generic\_nvram Linux driver}
 \label{fig:data_race_example}
@@
 On the other hand, fine-grained locking can easily lead to deadlocks.
-Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations by not regarding them as providing any protection between threads.
-This is sound, but can lead to false alarm race reports.}
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
 

### 2015-05-14 background.tex
lockset figure hack
hacks in bg

@@
 The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution.
-If that lockset ever becomes empty, the analysis reports a potential race on that memory location.
+If that lockset ever becomes empty, the analysis reports a \emph{potential} race on that memory location.
 This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
@@
 Next, during execution of $T_2$, the thread writes $A$ without holding $M$, and thus $\mathit{LS}_A$ becomes empty.
-Finally, a warning is reported because the two threads do not consistently protect $A$.
+As a result, a warning is reported because the two threads do not consistently protect $A$.
 

### 2015-05-14 whoop.tex
hacks with paragraphs
reduced the font of some figures
merged SMACK unsoundness text

@@
 
-\medskip\noindent\textbf{Concurrent Programming Model }
+\noindent\textbf{Concurrent Programming Model }
 %
@@
 \begin{figure}
+\footnotesize
+\center
 \begin{tabular}{lp{5.5cm}}
@@
 
-\medskip\noindent\textbf{Semantics }
+\noindent\textbf{Semantics }
 %
@@
 
-\medskip\noindent\textbf{Lockset Instrumentation }
+\noindent\textbf{Lockset Instrumentation }
 %
@@
 \begin{figure}
+\footnotesize
+\center
 \begin{tabular}{ll}
@@
 
-\medskip\noindent\textbf{Shared State Abstraction }
+\noindent\textbf{Shared State Abstraction }
 %
@@
 
-\medskip\noindent\textbf{Sequentialisation }
+\noindent\textbf{Sequentialisation }
 %
@@
 \begin{figure}
+\footnotesize
 \begin{tabular}{l}
@@
 
-\medskip\noindent\textbf{Soundness }
+\noindent\textbf{Soundness }
 %
@@
 
-\medskip\noindent\textbf{Architecture }
+\noindent\textbf{Architecture }
 %
@@
 
-\medskip\noindent\textbf{Extracting Entry Point Information }
+\noindent\textbf{Extracting Entry Point Information }
 %
@@
 
-\medskip\noindent\textbf{Translation for Verification }
+\noindent\textbf{Translation for Verification }
 %
@@
 We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained context-switch instrumentation as discussed in \S\ref{corral}.
-Note that the SMACK's memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses.
-The latest release of SMACK addresses this issue, and as future work we plan to integrate the latest SMACK with \whoop to further improve soundness.
 
-\medskip\noindent\textbf{Identifying Locks }
+\noindent\textbf{Identifying Locks }
 %
@@
 
-\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
+\noindent\textbf{Watchdog Race-Checking Instrumentation }
 %
@@
 
-\medskip\noindent\textbf{Generating Loop and Procedure Summaries }
+\noindent\textbf{Generating Loop and Procedure Summaries }
 %
@@
 
-\medskip\noindent\textbf{Verification and Error Reporting }
+\noindent\textbf{Verification and Error Reporting }
 %
@@
 
-\medskip\noindent\textbf{Optimizations }
+\noindent\textbf{Optimizations }
 %
@@
 
-\medskip\noindent\textbf{Practical Assumptions Related to Soundness}
+\noindent\textbf{Practical Assumptions Related to Soundness}
 %
@@
 
-As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that SMACK is subject to sources of unsoundness (e.g.\ SMACK models integers as an infinite set, rather than as a finite set of bit-vectors) and that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
+As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
 However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.
 
-\medskip\noindent\textbf{Limitations }
+\noindent\textbf{Limitations }
 %

### 2015-05-14 whoop.tex
some space hack

@@
 %
-Early in the development of \whoop we experimented with reasoning about recursion-free drivers by resorting to full inlining.
-We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined and blows up.
+Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.
+We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined.
 
 To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-Given a pool of candidate invariants, Houdini repeatedly tries to verify each procedure with respect to its current candidate summary.
+Given a pool of candidates, Houdini repeatedly attempts to verify each procedure with respect to its current candidate summary.
 If verification fails due to an incorrect candidate, this candidate is discarded.
-The process repeats until a fixpoint is reached, whereby the largest conjunction of the candidate invariants that can be proven to hold reamins.
+The process repeats until a fixpoint is reached, whereby the largest conjunction of the candidate invariants that can be proven to hold remains.
 The remaining candidates, which are indeed invariants, can then be assumed during further analysis of the program.
 
-The Houdini algorithm does not generate the initial pool of candidate invariants; these are given to the algorithm as a starting point.
-\whoop generates an initial pool of candidates using a set of heuristics.
-The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedureal pass over the code for an entry point.
+Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.
+The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.
 We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.
-If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$.
-Then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.
+If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.
 These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.
 Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
-We stress that guessing of incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.
+We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.
 The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.
 
-In practice, the candidates are generated at the Boogie level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race instrumentation.
+In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.
 

### 2015-05-14 corral.tex
hacks with paragraphs
some fixes

@@
-\ADComment{Suggestions: discuss race instrumentation for Corral immediately after explaining how corral works.
-Don't say ``we regard races as bugs''.
-After POR part, emphasise that this idea can be used to accelerate Corral for arbitrary bug-finding in concurrent programs.
-Then make it clear that in the remainder of the paper we regard races as bugs and use corral solely to look for races.}
+\whoop is a sound but imprecise static data race analyzer.
+For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate the performance of \corral~\cite{lal2012corral}, a precise bug-finder.
+\corral, which is used by Microsoft to analyze Windows drivers~\cite{lal2014powering}, has capabilities for analyzing concurrent programs.
+Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C).
+Our technique, though, is general and capable in principle of accelerating any concurrency bug-finder that operates by interleaving threads at shared memory operations.
 
-\whoop is a sound but imprecise static data race analyzer.
-For developers for whom false alarms are not acceptable, we consider a method for leveraging full or partial
-race-freedom guarantees provided by \whoop in order to accelerate the perforance of 
-\corral~\cite{lal2012corral}, a precise bug-finding tool.
-\corral is used by Microsoft for analysis of Windows device drivers~\cite{lal2014powering},
-and has capabilities for analysing concurrent programs.  Because \corral operates on Boogie programs,
-it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C).
-Our technique, though, is general and capable in principle of accelerating any concurrency bug-finder that operates
-by interleaving threads at shared memory operations.
-
-\corral is a symbolic bounded verifier for Boogie that uses the Z3 SMT solver to statically reason about program behaviors.
+\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors.
 It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found.
@@
 
-Because \corral performs bounded verification both in terms of the number of allowed context-switches and loop/recursion unrollings, it is inherently unsound: it can miss real bugs.
-On the other hand, \corral is precise, and assuming a precise model of the program environment it will only report true bugs.
+\noindent\textbf{Race-Checking Instrumentation }
+%
+To detect data races with \corral, \whoop outputs a Boogie program instrumented with a simple, but effective encoding of race checks.
+Whenever there is a write access to a shared variable $s$, \whoop instruments the program as follows:
+%
+\begin{boogie}
+s = e;         // original write
+yield;         // allow for a context-switch
+assert s == e; // check written value
+\end{boogie}
+%
+Likewise, whenever there is a read access to $s$, \whoop instruments the program as follows:
+%
+\begin{boogie}
+x = s;         // original read
+yield;         // allow for a context-switch
+assert x == s; // check read value
+\end{boogie}
+
+\corral is inherently unsound (i.e.\ can miss real races), because it performs bounded verification.
+However, \corral is precise and, assuming a precise environmental model, it will only report true races.
 \whoop takes advantage of this precision to report only feasible data races.
 
-\medskip\noindent\textbf{Sound Partial-Order Reduction }
+Note also that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
+
+\noindent\textbf{Sound Partial-Order Reduction }
 %
@@
 
+\ADComment{Don't say ``we regard races as bugs''.
+After POR part, emphasize that this idea can be used to accelerate Corral for arbitrary bug-finding in concurrent programs.
+Then make it clear that in the remainder of the paper we regard races as bugs and use corral solely to look for races.}
+
+\noindent\textbf{Granularity of Context-Switches }
+%
 We have implemented two different levels of context-switch instrumentation granularity:
@@
 We could extend our setup so that more than two threads are considered by \corral, though because the number of threads that may be launched by an OS kernel is unknown in general, we are inevitably limited by some fixed maximum thread count.
-
-\medskip\noindent\textbf{Race-Checking Instrumentation }
-%
-In this paper, we treat races as bugs.
-To detect them, we employ a simple, but effective, encoding of data race checks within \corral.
-For example, whenever there is a write access to a shared variable $s$, we instrument our abstract Boogie program to check for write-write races as follows:
-%
-\begin{boogie}
-s = e;         // original write
-yield;         // allow for a context-switch
-assert s == e; // check written value
-\end{boogie}
-%
-Likewise, for a read access to a shared variable $s$, we instrument the abstract Boogie program to check for read-write races as follows:
-%
-\begin{boogie}
-x = s;         // original read
-yield;         // allow for a context-switch
-assert x == s; // check read value
-\end{boogie}
-
-Note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
-An advantage of this instrumentation is that it can be used as a precise feasibility check for the potential data races that \whoop reports.

### 2015-05-14 evaluation.tex
hacks with paragraphs
some fixes

@@
 
-\medskip\noindent\textbf{Experimental Setup }
+\noindent\textbf{Experimental Setup }
 %
@@
 
-\medskip\noindent\textbf{Benchmarks }
+\noindent\textbf{Benchmarks }
 %
@@
 
-\medskip\noindent\textbf{Race-Checking }
+\noindent\textbf{Race-Checking }
 %
-Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of potentially racy pairs identified by \whoop (\#Racy Pairs); number of potentially racy memory regions reported by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound of 2 (\#Races Found).
-We did not discover new races with larger context-switch bounds.
-This might mean that these drivers only have races that manifest with at least one or two context-switches, or that \corral hit its bounds before discovering a deeper race.
+Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs reported by \whoop (\#Racy Pairs); number of racy memory regions identified by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
+Note that we did not discover any new races with a higher csb; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug.
+Furthermore, \corral did not discover any races that \whoop did not already report, giving us extra confidence in the implementation of \whoop.
 
@@
 
-\medskip\noindent\textbf{Granularity of Context-Switches }
+\noindent\textbf{Granularity of Context-Switches }
 %
@@
 \corral was configured with a time budget of 10 hours (T.O.
-denotes timeout) and a context-switch bound (csb) of 2, 5 and 9, respectively.
+denotes timeout) and a csb of 2, 5 and 9, respectively.
 For standalone \corral we enable the \yieldall instrumentation, which instruments context-switches (i.e.
@@
 
-\medskip\noindent\textbf{Other Tools }
+\noindent\textbf{Other Tools }
 %

### 2015-05-14 corral.tex
fix

@@
 \whoop takes advantage of this precision to report only feasible data races.
-
-Note also that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
+Also note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
 

### 2015-05-14 corral.tex
final updates in corral section

@@
 \end{boogie}
+%
+A \texttt{yield} statement denotes a nondeterministic context-switch, and is used by \corral to guide the sequentialization.
 
@@
 
+In this work, we use \corral to analyze individual pairs of entry points.
+We do not use any abstraction to model additional threads, as we want \corral to report only true races.
+Because we only analyze pairs, though, \corral will miss races that require more than two threads to manifest.
+We could extend our setup so that more than two threads are considered by \corral, but because the number of threads that an OS kernel might launch is unknown in general, we are inevitably limited by some fixed maximum thread count.
+
 \noindent\textbf{Sound Partial-Order Reduction }
 %
-By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} statement (which denotes a potential nondeterministic context-switch) after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
+By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
 Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, we do not instrument a \texttt{yield} after this statement.
 This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.
-This is because each access for which a \texttt{yield} is suppressed is guaranteed to be protected by some lock (a consequence of lockset analysis).
-If the access is a write, its effects are not visible by the other entry point until the lock is released.
+This is because each access, for which a \texttt{yield} is suppressed, is guaranteed to be protected by some lock (a consequence of lockset analysis).
+If the access is a write, its effects are not visible by the other entry point in the pair until the lock is released.
 If the access is a read, the value of the shared location cannot change until the lock is released.
@@
 
-\ADComment{Don't say ``we regard races as bugs''.
-After POR part, emphasize that this idea can be used to accelerate Corral for arbitrary bug-finding in concurrent programs.
-Then make it clear that in the remainder of the paper we regard races as bugs and use corral solely to look for races.}
-
-\noindent\textbf{Granularity of Context-Switches }
-%
 We have implemented two different levels of context-switch instrumentation granularity:
@@
 
-In this work, we use \corral to analyze individual pairs of entry points.
-We do not employ any abstraction to model additional threads, because we want \corral to report only true races.
-Because we only analyze pairs, though, \corral will miss races that require more than two threads to manifest.
-We could extend our setup so that more than two threads are considered by \corral, though because the number of threads that may be launched by an OS kernel is unknown in general, we are inevitably limited by some fixed maximum thread count.
+Our partial-order reduction can be used to accelerate \corral for arbitrary bug-finding in concurrent programs.
+In the remainder of this paper, though, we regard races as bugs and use \corral solely to look for races.

### 2015-05-14 evaluation.tex
A couple of changes

@@
 Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs reported by \whoop (\#Racy Pairs); number of racy memory regions identified by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
-Note that we did not discover any new races with a higher csb; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug.
-Furthermore, \corral did not discover any races that \whoop did not already report, giving us extra confidence in the implementation of \whoop.
+Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug.
+\corral did not discover any races that \whoop did not already report.
+This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.
 

### 2015-05-14 related.tex
A couple of changes

@@
-Statically analyzing concurrent programs for races is a promising alternative to dynamic analyzers, which typically rely on a (controlled) scheduler for exploring execution paths, and thus can face code coverage issues~\cite{musuvathi2008finding}.
-Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable examples of static race analyzers.
-However, both tools have limited applicability, as they heavily rely on user annotations.
-\textsc{Whoop}, on the other hand, does not require any source code modifications, and thus can be applied with minimal effort.
+Statically race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}.
+Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable static race analyzers.
+In comparison to \whoop, these tools rely heavily on user annotations.
 
@@
 
-Our work is not the first to exploit static analysis for achieving efficient race-checking.
-Prior work includes: \cite{choi2002efficient}, which combined static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combined a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combined inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+Prior works on static race checking include: \cite{choi2002efficient}, which combined static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combined a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combined inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
 

### 2015-05-14 related.tex
Tweaks.

@@
 
-Prior works on static race checking include: \cite{choi2002efficient}, which combined static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combined a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combined inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+Prior works on static race checking include: \cite{choi2002efficient}, which combines static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combines a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combines inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
 

### 2015-05-14 background.tex
Space hacking.  Resorting to a bit of negative spacing.

@@
 \end{lstlisting}
+\vspace{-2mm}
 \caption{Racy entry point in the generic\_nvram Linux driver}
@@
 Because the return statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
-
-Lock-based synchronization is a powerful tool for preventing races in drivers, but careless use of locks has many pitfalls~\cite{sutter2005software}.
-Coarse-grained locking can hurt performance, as it limits the opportunity for concurrency.
-On the other hand, fine-grained locking can easily lead to deadlocks.
+%
+Careless use of locks has many pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks.
 Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
@@
 \end{lstlisting}
+\vspace{-2mm}
 \caption{Using a lock to eliminate the data races in the previous example}
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=1\linewidth]{img/lockset.pdf}
-\caption{Applying lockset analysis on a thread of a concurrent program}
-\label{fig:locksets}
-\end{figure}
-
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_s$, for each shared variable $s$ used in the program.
@@
 The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock.
-The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows:
+The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
 %
-\begin{enumerate}
-\item After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
-\item When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
+After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
+%
+When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
 If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.
-\end{enumerate}
 
@@
 In this paper, we present a novel symbolic lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1\linewidth]{img/lockset.pdf}
+\caption{Applying lockset analysis on a thread of a concurrent program}
+\label{fig:locksets}
+\end{figure}

### 2015-05-14 corral.tex
Space hacking.  Resorting to a bit of negative spacing.

@@
 
-We have implemented two different levels of context-switch instrumentation granularity:
-
-\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
+We have implemented two different levels of context-switch instrumentation granularity.
+%
+\textbf{\yieldcoarse} instruments \texttt{yield} statement in a binary fashion: if \whoop
 finds that a pair of entry points cannot race, then it will only instrument
@@
 in all visible operations of this pair.
-
-\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
+%
+\textbf{\yieldmr} is a finer-grained instrumentation: \whoop instruments context-switches
 in all synchronization points, but only instruments context-switches in memory
@@
 verified as race-free). In our experiments (see \S\ref{evaluation}),
-\yieldmr performs significantly faster than \yieldcoarse.
+\yieldmr significantly outperforms \yieldcoarse.
 

### 2015-05-14 evaluation.tex
Space hacking.  Resorting to a bit of negative spacing.

@@
 This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
-
-However, if \whoop fails to verify any race-related properties of a driver, it might actually slow down \corral.
+%
+However, if \whoop fails to verify any race-related properties of a driver, it can slow down \corral.
 We noticed this in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead.
 Although \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr.
-This might be because \corral verifies the only 2 pairs of sx8 quite fast (21.4 seconds with csb of 9), and thus the overhead of using \whoop overwhelms any benefits from our fine-grained instrumentation.
+\corral checks the only two pairs associated with sx8 quite quickly (21.4 seconds with csb of 9), so that the overhead of running \whoop outweighs the benefits of fine-grained instrumentation.
 

### 2015-05-14 intro.tex
Space hacking.  Resorting to a bit of negative spacing.

@@
 
-A major reason behind these concurrency bugs is that drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}.
-For example the Linux kernel can invoke its drivers from multiple concurrent threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may decide to remove a device (hot-plugging) while some operation is still running.
+A major cause of concurrency bugs is the complicated and hostile environment in which drivers operate~\cite{corbet2005linux}.
+The Linux kernel can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may decide to remove a device (hot-plugging) while some operation is still running.
 All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (e.g.\ locks) are in place to control concurrent access to shared resources.
-
+%
 Data races are a source of undefined behavior in C, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
@@
 Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
-
+%
 We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers, as well as other concurrent programs.
@@
 
-\whoop currently supports Linux drivers, but our approach is conceptually generic, thus could be applied to analyze drivers for other operating systems (e.g.\ Windows) as well as other concurrent systems that use a similar programming model (e.g.\ filesystems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ filesystems).
 

### 2015-05-14 whoop.tex
Space hacking.  Resorting to a bit of negative spacing.

@@
 In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
-
+%
 Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.
@@
 This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
-
+%
 We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.
@@
 Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
-
-The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.
-This transformation is described in Figure~\ref{fig:instrumentation}.
-A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).
-A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.
+%
+The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.
+For a shared variable assignment $s = e$ we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update the lockset $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).
+A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.
 Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
@@
 
-We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.
-The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
-In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are indeed terminating.
+In the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.
+The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.
 

### 2015-05-14 evaluation.tex
space hack

@@
 %
-Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs reported by \whoop (\#Racy Pairs); number of racy memory regions identified by \whoop (\#Racy MRs); and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
+Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
 Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug.

### 2015-05-14 evaluation.tex
fix

@@
 
-We can see that \whoop reports more races than \corral.
-This is expected, since \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.
+We can see that \whoop reports more races than \corral does.
+This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.
 On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.

### 2015-05-14 background.tex
fixes

@@
 
-The most common method for avoiding races is to protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.  
+The most common method for avoiding races is by protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.
 Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}.
@@
 %
-Careless use of locks has many pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks.
-Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
+Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency; while fine-grained locking can easily lead to deadlocks.
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
 

### 2015-05-14 related.tex
fixes

@@
-Statically race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}.
+Static race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}.
 Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable static race analyzers.

### 2015-05-14 corral.tex
fixes

@@
 
-We have implemented two different levels of context-switch instrumentation granularity.
+We have implemented two different levels of context-switch instrumentation granularity: \yieldcoarse and \yieldmr.
 %
-\textbf{\yieldcoarse} instruments \texttt{yield} statement in a binary fashion: if \whoop
+The first instruments \texttt{yield} statement in a binary fashion: if \whoop
 finds that a pair of entry points cannot race, then it will only instrument

### 2015-05-14 evaluation.tex
fixes

@@
 
-\noindent\textbf{Granularity of Context-Switches }
+\noindent\textbf{Accelerating \corral }
 %
 Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Figure~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
-All reported times, in seconds, are averages over three runs.
+All reported times are in seconds and averages over three runs.
 \corral was configured with a time budget of 10 hours (T.O.
 denotes timeout) and a csb of 2, 5 and 9, respectively.
-For standalone \corral we enable the \yieldall instrumentation, which instruments context-switches (i.e.
-\texttt{yield} statements) after all visible operations.
-\whoop + \corral, instead, uses two different context-switch instrumentation granularities, \yieldcoarse and \yieldmr (see \S\ref{corral}).
+Standalone \corral uses \yieldall, which instruments context-switches (i.e.
+\texttt{yield} statements) after all visible operations, while \whoop + \corral uses \yieldcoarse and \yieldmr, which instrument context-switches in a more fine-grained fashion (see \S\ref{corral}).
 The table also shows the number of context-switches per instrumentation (\#Y).

### 2015-05-14 corral.tex
fixes

--- 2015-05-14 corral.tex
@@
 
-We have implemented two different levels of context-switch instrumentation granularity:
-
-\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
-finds that a pair of entry points cannot race, then it will only instrument
-context-switches in synchronization points (i.e.\ locks and unlocks) of this pair; else if \whoop finds
-that a pair of entry points can race, then it will instrument context-switches
-in all visible operations of this pair.
-
-\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
-in all synchronization points, but only instruments context-switches in memory
-regions that can potentially race (regardless if a pair has not been fully
-verified as race-free). In our experiments (see \S\ref{evaluation}),
-\yieldmr performs significantly faster than \yieldcoarse.
+We have implemented two different \texttt{yield} instrumentations in \whoop: \yieldcoarse and \yieldmr.
+%
+\yieldcoarse instruments \texttt{yield} statements in a binary fashion: if \whoop finds that a pair of entry points cannot race, then it will instrument a \texttt{yield} only after each lock and unlock statement of this pair; else if \whoop finds that a pair of entry points might race, then it will instrument context-switches after all visible operations of this pair.
+%
+\yieldmr performs a finer-grained instrumentation: it instruments context-switches only after each memory region that can potentially race (regardless if a pair has not been fully verified as race-free) and after each lock and unlock.
+In our experiments (see \S\ref{evaluation}), \yieldmr significantly outperforms \yieldcoarse.
 

### 2015-05-14 corral.tex
fixes

@@
 %
-\yieldcoarse instruments \texttt{yield} statements in a binary fashion: if \whoop finds that a pair of entry points cannot race, then it will instrument a \texttt{yield} only after each lock and unlock statement of this pair; else if \whoop finds that a pair of entry points might race, then it will instrument context-switches after all visible operations of this pair.
+The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair as race free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
 %
-\yieldmr performs a finer-grained instrumentation: it instruments context-switches only after each memory region that can potentially race (regardless if a pair has not been fully verified as race-free) and after each lock and unlock.
+\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proved as race-free), and after each lock and unlock.
 In our experiments (see \S\ref{evaluation}), \yieldmr significantly outperforms \yieldcoarse.

### 2015-05-14 evaluation.tex
fix

@@
 Although \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr.
-\corral checks the only two pairs associated with sx8 quite quickly (21.4 seconds with csb of 9), so that the overhead of running \whoop outweighs the benefits of fine-grained instrumentation.
+\corral checks the only two pairs associated with sx8 quite quickly (21.4 seconds with csb of 9), so the overhead of running \whoop outweighs the benefits of fine-grained instrumentation.
 

### 2015-05-14 evaluation.tex
fixes

@@
 Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr.
-Figure~\ref{fig:plot} shows that most speedups using \yieldmr are in the range of 1.5--10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9.
-We notice that a higher csb typically results into greater speedups, when exploiting \whoop.
+Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9.
+We noticed that a higher csb typically results into greater speedups, when exploiting \whoop.
 This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
@@
 We noticed this in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead.
-Although \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr.
+Although \whoop verified 46 out of 47 memory regions of sx8, it still failed to speedup \corral using \yieldmr.
 \corral checks the only two pairs associated with sx8 quite quickly (21.4 seconds with csb of 9), so the overhead of running \whoop outweighs the benefits of fine-grained instrumentation.

### 2015-05-14 intro.tex
fixes

@@
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.
-We find that by combining \whoop and \corral we achieve concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
-For two drivers, we observe even greater speedups of 12$\times$ and 20$\times$.
+We found that by combining \whoop and \corral we achieve can concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
+For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
 

### 2015-05-14 whoop.tex
space hack

@@
 
-A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of \emph{thread templates}.
-A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.
-A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of $T$ by a thread.
+A \emph{concurrent program} is described by a finite set of \emph{shared} variables $\sharedvars$, a finite set of mutexes $\mutexes$, and a finite set of \emph{thread templates}.
+A thread template $T$ consists of a finite set of procedures $\procedures{T}$ and a finite set of private variables $\variables{T}$.
+A designated procedure $\mainforthread{T} \in \procedures{T}$ denotes the starting point for execution of $T$ by a thread.
 Each procedure of $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
@@
 Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.
-The state of a running concurrent program consists of: a valuation of the shared variables, $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
-Each thread consists of an id, drawn from $\ids$, a thread template, $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.
+The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
+Each thread consists of an id, drawn from $\ids$, a thread template $T$ an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.
 If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.
@@
 %
-For thread templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we wish to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
-
-To achieve this, we first \emph{instrument} a thread template for lockset analysis (see \S\ref{bg:lockset}).
-Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.
+For templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
+%
+To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{bg:lockset}).
+Given an arbitrary symbol $i$, we define the instrumentation of $T$ with respect to $i$, denoted $\instrument{T}{i}$.
 There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.
 
-Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+Renaming is straightforward: all occurrences of each private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
 The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.

### 2015-05-14 whoop.tex
space hack

@@
 
-Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
+Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset $\currentlockset{i} \subseteq \powerset{\mutexes}$ to record the mutexes currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset $\lockset{s}{i}$ to record the mutexes that are consistently held when the thread accesses $s$.
 %
 The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.
-For a shared variable assignment $s = e$ we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update the lockset $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).
-A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.
-Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
+For a shared variable assignment $s = e$, we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).
+A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss below.
+Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating $\currentlockset{i}$ to add and remove mutex $m$, respectively.
 

### 2015-05-14 conclusion.tex
fix

@@
 Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrent bug-finder.
-Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our technique can analyze programs 1.5--20$\times$ faster than standalone \corral.
+Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.

### 2015-05-14 evaluation.tex
update in evaluation

@@
 We tried to compare \whoop with other similar approaches (see \S\ref{related}).
-However, we found this hard to achieve, as some tools are not freely available, while others are outdated for several years and are not working on the 4.0 Linux drivers.
+However, we found this hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007 and might miss some of the latest features); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.

### 2015-05-14 intro.tex
fixes

@@
 A major cause of concurrency bugs is the complicated and hostile environment in which drivers operate~\cite{corbet2005linux}.
-The Linux kernel can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may decide to remove a device (hot-plugging) while some operation is still running.
+The Linux kernel can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, or the user may decide to remove a device (hot-plugging) while some operation is still running.
 All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (e.g.\ locks) are in place to control concurrent access to shared resources.

### 2015-05-14 background.tex
fixes
fix

@@
 If that lockset ever becomes empty, the analysis reports a \emph{potential} race on that memory location.
-This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
+This is because an empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
 
@@
 \includegraphics[width=1\linewidth]{img/lockset.pdf}
-\caption{Applying lockset analysis on a thread of a concurrent program}
+\caption{Applying lockset analysis on a concurrent program}
 \label{fig:locksets}

### 2015-05-14 whoop.tex
fixes

@@
 In particular, shared variables may not appear in arbitrary expressions.
-This restriction simplifies our presentation of lockset instrumentation, below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.
+This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.
 We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.
@@
 %
-Chauffeur is a Clang-based frontend that traverses the abstract syntax tree (AST) of the original driver (using a Clang visitor pass) and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
-Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is subsequently parsed by \whoop and used during the instrumentation.
+Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
+Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.
 

### 2015-05-14 whoop.tex
fixes

@@
 When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
-The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
+The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
 %
@@
 For each pair of entry points, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
-Boogie generates a VC, for each procedure in the program, and discharges it to the Z3~\cite{de2008z3} theorem prover.
+Boogie generates a VC, for each procedure in the program, and discharges it to the Z3 theorem prover.
 In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.

### 2015-05-14 whoop.tex
fix

@@
 We assume that the formal parameters of an entry point do not alias, and thus cannot race.
-This is a potentially unsound feature that can be turned off using a command line option, but without this assumption the false alarm rate of \whoop is very high.
+This is a potentially unsound feature that can be turned off using a command line option, but without this assumption the false alarm rate of \whoop is high.
 In our experience so far, we have not missed any races by assuming non-overlapping parameters.

### 2015-05-14 corral.tex
fixes

@@
 %
-\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proved as race-free), and after each lock and unlock.
+\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock.
 In our experiments (see \S\ref{evaluation}), \yieldmr significantly outperforms \yieldcoarse.

### 2015-05-14 whoop.tex
cutting space
fixes

--- 2015-05-14 whoop.tex
@@
 
-Because a driver is processed by Clang and SMACK during the translation from C to Boogie IVL, significant engineering effort was required to map Boogie-level errors back to original source code locations.
-
 \noindent\textbf{Optimizations }
@@
 As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lockfree synchronization is used).
-\whoop also over-approximates drivers, which can be another source of imprecision.
+\whoop also uses over-approximation, which can be another source of imprecision.
 Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.

### 2015-05-14 evaluation.tex
fix

@@
 %
-We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel.
-We chose non-trivial drivers from several categories: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
-We had to manually model the environment for these drivers, a process that required approximately two months of work.
+We evaluate our methodology against \sizeOfBenchmarks drivers from the 4.0 Linux kernel.
+We chose non-trivial drivers of several types: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+For all these drivers, we had to understand their environment and manually model it; this required approximately two months of work.
 

### 2015-05-14 whoop.tex
some text about pairwise analysis

@@
 We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.
-The trade-off is that a quadratic number of entry point pairs must be checked.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity than the level of pairs.
+However, a pair-wise analysis can be trivially parallelized for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
 In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

### 2015-05-14 whoop.tex
fix

@@
   .write = write_nvram,
-  .unlocked_ioctl = nvram_unlocked_ioctl,
+  .unlocked_ioctl = nvram_unlocked_ioctl
 };

### 2015-05-14 whoop.tex
space hack

@@
 Figure~\ref{fig:whoop} depicts the \whoop toolchain.
-Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
-Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
-
+The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
+Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
 Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.

### 2015-05-15 whoop.tex
update
space cutting

@@
 Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
-
+%
 SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop.
@@
 %
-When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
+When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
 The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.

### 2015-05-15 whoop.tex
update

@@
 %
-For each pair of entry points, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
-Boogie generates a VC, for each procedure in the program, and discharges it to the Z3 theorem prover.
+For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
+For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.
 In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
-Successful verification implies that the entry point pair is free of data races, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user.
+Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user.
 To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.

### 2015-05-15 corral.tex
fix

@@
 Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, we do not instrument a \texttt{yield} after this statement.
-This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.
+This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.
 This is because each access, for which a \texttt{yield} is suppressed, is guaranteed to be protected by some lock (a consequence of lockset analysis).

### 2015-05-15 whoop.tex
fix

@@
 The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other.
-As an example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock.
-We discovered this when we first analyzed the r8169 driver (see \S\ref{evaluation}) and \whoop reported many races between a number of networking entry points; when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL.
-\whoop exploits this knowledge and does not create pairs for entry points that are serialized by the kernel.
-This is an ongoing manual effort: the more drivers we study, the more such properties we discover, which we can then exploit to make \whoop more precise.
+For example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock.
+We discovered this when \whoop reported many races between a number of networking entry points of the r8169 driver (see \S\ref{evaluation}); when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL.
+\whoop exploits this knowledge and does not create pairs for entry points that are mutually serialized by the kernel.
+This is an ongoing manual effort: the more drivers we study, the more such properties we discover to make \whoop more precise.
 Currently, this information is hard-coded inside \whoop.
-In the future, we plan to provide a configuration file where the user can denote such kernel specific information.
+In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.
 

### 2015-05-15 whoop.tex
Minor fix.

--- 2015-05-15 whoop.tex
@@
 In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
-%
+
 Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.
@@
 We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity than the level of pairs.
-However, a pair-wise analysis can be trivially parallelized for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.
+The trade-off is that a quadratic number of entry point pairs must be checked.
 In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.
@@
 Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.
-Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged to an automated theorem prover.
+Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged by an automated theorem prover.
 
@@
 
-\noindent\textbf{Concurrent Programming Model }
+\medskip\noindent\textbf{Concurrent Programming Model }
 %
@@
 In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
-Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
+Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.
+For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
 These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
 
-A \emph{concurrent program} is described by a finite set of \emph{shared} variables $\sharedvars$, a finite set of mutexes $\mutexes$, and a finite set of \emph{thread templates}.
-A thread template $T$ consists of a finite set of procedures $\procedures{T}$ and a finite set of private variables $\variables{T}$.
-A designated procedure $\mainforthread{T} \in \procedures{T}$ denotes the starting point for execution of $T$ by a thread.
+A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of \emph{thread templates}.
+A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.
+A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of $T$ by a thread.
 Each procedure of $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.
@@
 In particular, shared variables may not appear in arbitrary expressions.
-This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.
+This restriction simplifies our presentation of lockset instrumentation, below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.
 We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.
@@
 \begin{figure}
-\footnotesize
-\center
 \begin{tabular}{lp{5.5cm}}
@@
 
-\noindent\textbf{Semantics }
+\medskip\noindent\textbf{Semantics }
 %
 Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.
-The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
-Each thread consists of an id, drawn from $\ids$, a thread template $T$ an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.
+The state of a running concurrent program consists of: a valuation of the shared variables, $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
+Each thread consists of an id, drawn from $\ids$, a thread template, $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.
 If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.
@@
 This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
-%
+
 We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.
@@
 
-\noindent\textbf{Lockset Instrumentation }
+\medskip\noindent\textbf{Lockset Instrumentation }
 %
-For templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
-%
-To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{bg:lockset}).
-Given an arbitrary symbol $i$, we define the instrumentation of $T$ with respect to $i$, denoted $\instrument{T}{i}$.
+For thread templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we wish to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
+
+To achieve this, we first \emph{instrument} a thread template for lockset analysis (see \S\ref{bg:lockset}).
+Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.
 There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.
 
-Renaming is straightforward: all occurrences of each private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
 The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.
@@
 \begin{figure}
-\footnotesize
-\center
 \begin{tabular}{ll}
@@
 
-Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset $\currentlockset{i} \subseteq \powerset{\mutexes}$ to record the mutexes currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset $\lockset{s}{i}$ to record the mutexes that are consistently held when the thread accesses $s$.
-%
-The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.
-For a shared variable assignment $s = e$, we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).
-A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss below.
-Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating $\currentlockset{i}$ to add and remove mutex $m$, respectively.
+Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
 
-\noindent\textbf{Shared State Abstraction }
+The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.
+This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.
+A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).
+A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.
+Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
+
+\medskip\noindent\textbf{Shared State Abstraction }
 %
@@
 
-\noindent\textbf{Sequentialisation }
+\medskip\noindent\textbf{Sequentialisation }
 %
@@
 \begin{figure}
-\footnotesize
 \begin{tabular}{l}
@@
 
-\noindent\textbf{Soundness }
+\medskip\noindent\textbf{Soundness }
 %
@@
 
-In the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.
-The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.
+We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.
+The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
+In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are indeed terminating.
 
@@
 
-\noindent\textbf{Architecture }
+\medskip\noindent\textbf{Architecture }
 %
 Figure~\ref{fig:whoop} depicts the \whoop toolchain.
-The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
-Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
+Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races.
+Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
+
 Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.
@@
 We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).
-For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
+For the remaining components, we were able to re-use industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
 
-\noindent\textbf{Extracting Entry Point Information }
+\medskip\noindent\textbf{Extracting Entry Point Information }
 %
-Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
-Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.
+Chauffeur is a Clang-based frontend that traverses the abstract syntax tree (AST) of the original driver (using a Clang visitor pass) and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
+Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is subsequently parsed by \whoop and used during the instrumentation.
 
@@
   .write = write_nvram,
-  .unlocked_ioctl = nvram_unlocked_ioctl
+  .unlocked_ioctl = nvram_unlocked_ioctl,
 };
@@
 
-\noindent\textbf{Translation for Verification }
+\medskip\noindent\textbf{Translation for Verification }
 %
@@
 Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
-%
+
 SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop.
@@
 We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained context-switch instrumentation as discussed in \S\ref{corral}.
+Note that the SMACK's memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses.
+The latest release of SMACK addresses this issue, and as future work we plan to integrate the latest SMACK with \whoop to further improve soundness.
 
-\noindent\textbf{Identifying Locks }
+\medskip\noindent\textbf{Identifying Locks }
 %
-When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
-The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
+When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
+The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
 %
@@
 
-\noindent\textbf{Watchdog Race-Checking Instrumentation }
+\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
 %
@@
 
-\noindent\textbf{Generating Loop and Procedure Summaries }
+\medskip\noindent\textbf{Generating Loop and Procedure Summaries }
 %
-Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.
-We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined.
+Early in the development of \whoop we experimented with reasoning about recursion-free drivers by resorting to full inlining.
+We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined and blows up.
 
 To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-Given a pool of candidates, Houdini repeatedly attempts to verify each procedure with respect to its current candidate summary.
-If verification fails due to an incorrect candidate, this candidate is discarded.
-The process repeats until a fixpoint is reached, whereby the largest conjunction of the candidate invariants that can be proven to hold remains.
-The remaining candidates, which are indeed invariants, can then be assumed during further analysis of the program.
+The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process.
+Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.
 
-Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.
-The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.
-We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.
-If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.
-These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.
-Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
-We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.
-The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.
+The candidate invariants are based on the watchdog constants that \whoop instruments for watchdog race-checking.
+Our candidate generation rules are purely heuristic.
+The key idea is to automatically generate candidates for each memory region in the granularity of watched accesses.
+For example, if entry point $T$ has a procedure $f$ that only writes (and not reads) the offset $z$ of memory region $\mathit{MR}$, which is associated with the write set $\hasbeenwritten{\mathit{MR}, i}$ and read set $\hasbeenread{\mathit{MR}, i}$, then when instrumenting $T$ with symbol $i$ (using the notation from \S\ref{sec:symbolicpairwise}), we might guess $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \in \hasbeenwritten{\mathit{MR}, i}$ and $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \notin \hasbeenread{\mathit{MR}, i}$, as postconditions for $\instrument{f}{i}$.
+Likewise, if $f$ releases mutex $m$, then we might guess $m \notin \currentlockset{i}$ as a postcondition.
 
-In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.
+Using watchdog-based candidates reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing.
+To generate the candidate invariants, \whoop performs an inter-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations.
+To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region, e.g.\  $z \in \hasbeenwritten{\mathit{MR}, i}$ and $z \notin \hasbeenread{\mathit{MR}, i}$ for the above example.
 
-\noindent\textbf{Verification and Error Reporting }
+We emphasize that the candidate invariants generated by \whoop are just that: candidates.
+The tool is free to speculatively generate candidates that later turn out to be incorrect: these are simply discarded by Houdini.
+A consequence is that incorrect or unintended candidates generated due to bugs in \whoop cannot compromise the soundness of verification.
+
+\medskip\noindent\textbf{Verification and Error Reporting }
 %
 For each pair of entry points, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
-Boogie generates a VC, for each procedure in the program, and discharges it to the Z3 theorem prover.
+Boogie generates a VC, for each procedure in the program, and discharges it to the Z3~\cite{de2008z3} theorem prover.
 In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
@@
 
-\noindent\textbf{Optimizations }
+Because a driver is processed by Clang and SMACK during the translation from C to Boogie IVL, significant engineering effort was required to map Boogie-level errors back to original source code locations.
+
+\medskip\noindent\textbf{Optimizations }
 %
@@
 
-\noindent\textbf{Practical Assumptions Related to Soundness}
+\medskip\noindent\textbf{Practical Assumptions Related to Soundness}
 %
@@
 
-As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
+As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that SMACK is subject to sources of unsoundness (e.g.\ SMACK models integers as an infinite set, rather than as a finite set of bit-vectors) and that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
 However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.
 
-\noindent\textbf{Limitations }
+\medskip\noindent\textbf{Limitations }
 %
 As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lockfree synchronization is used).
-\whoop also uses over-approximation, which can be another source of imprecision.
+\whoop also over-approximates drivers, which can be another source of imprecision.
 Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.

### 2015-05-15 evaluation.tex

@@
 Even benign races, though, lead to undefined behavior according to the C standard, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
-The second case is an entry point accessing a field of an object (either global or passed as a parameter) without a lock.
-This can lead to a race if another entry point simultaneously accesses the same field of the same object.
+The second case is an entry point modifying a field of a struct (either global or passed as a parameter) without a lock.
+This can lead to a race if another entry point simultaneously accesses the same field of the same struct.
 

### 2015-05-15 related.tex

@@
 
-Prior works on static race checking include: \cite{choi2002efficient}, which combines static analysis, optimized instrumentation and runtime access caching to speedup dynamic data race detection; \cite{kahlon2007fast}, which combines a divide-and-conquer strategy that partitions all pointers of a program that cannot alias in disjoint sets and function summarization to achieve a scalable context-sensitive alias analysis for data races; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false data race warnings; and \cite{das2015section}, which combines inter-procedural pointer analysis with verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+Prior works on static race checking include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speedup dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve a scalable race analysis; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
 

### 2015-05-15 evaluation.tex
Can you be specific about to whom you filed the bug report?  Ideally, provide a link to a bug tracker.

@@
 
-As an example of the second case, we found the following race in the generic\_nvram char driver (see Figure~\ref{fig:data_race_example} for the source code): during the \texttt{llseek} entry point, the driver is accessing the file offset \texttt{file->f\_pos} without first acquiring a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
-This can lead to a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
-When we investigated another char driver that uses the same APIs, we saw that it was protecting the offset access in its \texttt{llseek} entry point with a mutex.
-This made us suspicious that the race we found in generic\_nvram could be real.
-We filed a bug report, but did not manage to get a response yet.
+As an example of the second case, we found the following race in the generic\_nvram char driver (see Figure~\ref{fig:data_race_example}): in the \texttt{llseek} entry point the driver accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
+We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.
+We have filed a bug report \ADComment{to whom?} and are awaiting a response.
 

### 2015-05-15 intro.tex
Introduction now on one page.  Minor negative spacing used.

@@
-Device drivers exist at the thin boundary between hardware and software.
-They are complex pieces of system-level software responsible for the interaction between the operating system and any hardware devices that are attached to a computer.
-Drivers are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}.
+Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software
+to provide an interface between the operating system and hardware devices that are attached to a computer.
+Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}.
 This has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
@@
 
-A major cause of concurrency bugs is the complicated and hostile environment in which drivers operate~\cite{corbet2005linux}.
-The Linux kernel can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, or the user may decide to remove a device (hot-plugging) while some operation is still running.
-All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (e.g.\ locks) are in place to control concurrent access to shared resources.
+Concurrency bugs are exaccerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}.
+The OS can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may remove a device (hot-plugging) while some operation is still running.
+These scenarios can cause \emph{data races} if insufficient synchronization mechanisms are in place to control concurrent access to shared resources.
 %
@@
 
-In this paper, we present \whoop, a new fully automatic approach that statically analyzes device drivers written in C for data races.
-\whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for verifying the absence of data races in device drivers.
-We attempt to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are \emph{consistently} protected by at least one common lock.
+We present \whoop, an automated approach for static data race analysis in device drivers.
+\whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock.
 Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
 %
-We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers, as well as other concurrent programs.
+We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers and other concurrent programs.
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.
-We found that by combining \whoop and \corral we achieve can concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
+By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
 For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
-
+%
 \whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ filesystems).
@@
 To summarize, our contributions are as follows:
+\vspace{-0.5mm}
 \begin{itemize}
 \item We propose symbolic pairwise lockset analysis, a sound and scalable technique for automatically verifying the absence of data races in device drivers.
+\vspace{-0.2mm}
 \item We present \whoop, a tool that leverages our approach for analyzing data races in device drivers.
+\vspace{-0.2mm}
 \item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, an industrial-strength bug-finder.
+\vspace{-0.2mm}
 \item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race-checking and accelerating \corral.

### 2015-05-15 whoop.tex
Cut to the chase in the abstract.
Removed subsections from background and replaced with paragraph headings.  A bit reluctant to do this, but I think it makes sense to save cutting too much text.

@@
 %
-To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{bg:lockset}).
+To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{sec:background}).
 Given an arbitrary symbol $i$, we define the instrumentation of $T$ with respect to $i$, denoted $\instrument{T}{i}$.
@@
 $\currentlockset{j} = \emptyset;$ $\hasbeenread{j} = \emptyset;$ $\hasbeenwritten{j} = \emptyset;$ \\
-\texttt{for} $s \in \sharedvars$ \texttt{do} $\lockset{s}{i} = \mutexes;$ $\lockset{s}{j} = \mutexes;$ \medskip
+\texttt{for} $s \in \sharedvars$ \texttt{do} $\lockset{s}{i} = \mutexes;$ $\lockset{s}{j} = \mutexes;$ \smallskip
 \\
@@
 $\mainforthread{\instrument{T}{i}}();$ \\
-$\mainforthread{\instrument{U}{j}}();$ \medskip\\
+$\mainforthread{\instrument{U}{j}}();$ \smallskip\\
 
@@
 \label{fig:sequentialization}
+\vspace{-2mm}
 \end{figure}
@@
 \includegraphics[width=.99\linewidth]{img/whoop.pdf}
+\vspace{-2mm}
 \caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
+\vspace{-2mm}
 \label{fig:whoop}
@@
 \end{lstlisting}
-\caption{Entry points definitions in the generic\_nvram driver}
+\vspace{-2mm}
+\caption{Entry point definitions in the generic\_nvram driver}
 \label{fig:entrypoints}
+\vspace{-2mm}
 \end{figure}
@@
 \end{lstlisting}
+\vspace{-2mm}
 

### 2015-05-15 whoop.tex
Done it.

@@
 \label{fig:statements}
+\vspace{-2mm}
 \end{figure}
@@
 
-A thread terminates if it reaches the end of $\mainforthread{T}$; in this case the thread is removed from the list of threads.
-Because we are interested in the analysis of device drivers, which are \emph{reactive} concurrent programs, we do not consider the notion of global program termination.
+A thread terminates if it reaches the end of $\mainforthread{T}$, and is removed from the list of threads.
+Since our interest is in analysis of device drivers, which are reactive, we do not consider the notion of global program termination.
 
@@
 There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.
-
+%
 Renaming is straightforward: all occurrences of each private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
@@
 
-To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.
-Given a pool of candidates, Houdini repeatedly attempts to verify each procedure with respect to its current candidate summary.
+To make our analysis scale while maintaining precision, and to support recursion, we use the Houdini algorithm~\cite{flanagan2001houdini} to automatically compute summaries (pre- and post-conditions and loop invariants).
+Given a pool of \emph{candidate} pre-conditions, post-conditions, and loop invariants, Houdini repeatedly attempts to verify each procedure.
+Initially, the entire candidate pool is considered.
 If verification fails due to an incorrect candidate, this candidate is discarded.
-The process repeats until a fixpoint is reached, whereby the largest conjunction of the candidate invariants that can be proven to hold remains.
-The remaining candidates, which are indeed invariants, can then be assumed during further analysis of the program.
+The process repeats until a fixpoint is reached.
+The (possibly empty) set of remaining candidates have been proven to hold, and can used to summarize calls and loops during further program analysis.
 

### 2015-05-15 whoop.tex
Addressed comment from Akis about false alarm rate.

@@
 We assume that the formal parameters of an entry point do not alias, and thus cannot race.
-This is a potentially unsound feature that can be turned off using a command line option, but without this assumption the false alarm rate of \whoop is very high.
-In our experience so far, we have not missed any races by assuming non-overlapping parameters.
+This is a potentially unsound feature that can be turned off using a command line option.
+Without this assumption we have observed \whoop reporting false alarms, and in our experience so far we have not missed any races by assuming non-overlapping parameters.
 We also rely on the soundness of our best-effort environmental model, and on exploiting domain-specific knowledge related to entry point serialization by the Linux kernel.

### 2015-05-15 evaluation.tex
Done it.
URL and paper
address a comment

@@
 \end{tabular}
+\vspace{-3mm}
 \end{table}
@@
 \end{tabular}
+\vspace{-3mm}
 \end{table*}
@@
 \includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
+\vspace{-2mm}
 \caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
@@
 \label{fig:plot}
+\vspace{-2mm}
 \end{figure}
@@
 
-As an example of the second case, we found the following race in the generic\_nvram char driver (see Figure~\ref{fig:data_race_example}): in the \texttt{llseek} entry point the driver accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+As an example of the second case, we found the following race in the generic\_nvram driver (see Figure~\ref{fig:data_race_example}): in the \texttt{llseek} entry point the driver accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
 We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.
-We have filed a bug report \ADComment{to whom?} and are awaiting a response.
+We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.
 

### 2015-05-15 intro.tex

@@
 
-Concurrency bugs are exaccerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}.
+Concurrency bugs are exacerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}.
 The OS can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may remove a device (hot-plugging) while some operation is still running.
@@
 %
-Data races are a source of undefined behavior in C, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
+Data races are a source of undefined behavior in C~\cite[p.\ 38]{iso/iec11}, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
 %
@@
 %
-\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ filesystems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ file systems).
 

### 2015-05-15 whoop.tex
Removed sentence

@@
 
-\noindent\textbf{Sequentialisation }
+\noindent\textbf{Sequentialization }
 %
@@
 We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.
-If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.
+If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.
 These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.
-Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
+Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
 We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.
@@
 %
-As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lockfree synchronization is used).
+As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lock-free synchronization is used).
 \whoop also uses over-approximation, which can be another source of imprecision.
@@
 However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
-Because this serialization is considered bad practice, we burden the developer with disregarding any related false bug reports.
 

### 2015-05-15 intro.tex
minor fixes in intro

@@
-Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software
-to provide an interface between the operating system and hardware devices that are attached to a computer.
+Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer.
 Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}.
@@
 %
-We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers and other concurrent programs.
+We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze drivers and other concurrent programs.
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.
@@
 %
-\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ file systems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as to concurrent systems that use a similar programming model (e.g.\ file systems).
 

### 2015-05-15 background.tex
Removed subsections from background and replaced with paragraph headings.  A bit reluctant to do this, but I think it makes sense to save cutting too much text.
Pulled background back a page.
Done it.
fix

@@
-\subsection{Concurrency in Device Drivers}
-\label{bg:concurrency}
-
+\noindent\textbf{Concurrency in Device Drivers }
+%
 Modern operating systems address the demand for responsiveness and performance in device drivers by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.
@@
 Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux driver.
-The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} object as a parameter.
+The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter.
 However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order.
@@
 \label{fig:data_race_example}
+\vspace{-2mm}
 \end{figure}
 
-The most common method for avoiding races is by protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.
-Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}.
-Because the return statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
+The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.
+Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}.
+Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
 %
 Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency; while fine-grained locking can easily lead to deadlocks.
-Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
 
@@
 \label{fig:lock_example}
+\vspace{-2mm}
 \end{figure}
 
-\subsection{Lockset Analysis}
-\label{bg:lockset}
-
+\noindent\textbf{Lockset Analysis }
+%
 Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.
@@
 \includegraphics[width=1\linewidth]{img/lockset.pdf}
+\vspace{-2mm}
 \caption{Applying lockset analysis on a concurrent program}
 \label{fig:locksets}
+\vspace{-2mm}
 \end{figure}

### 2015-05-15 whoop.tex
fix
added page in book

@@
 The process repeats until a fixpoint is reached.
-The (possibly empty) set of remaining candidates have been proven to hold, and can used to summarize calls and loops during further program analysis.
+The (possibly empty) set of remaining candidates has been proven to hold, and can used to summarize calls and loops during further program analysis.
 
@@
 Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time.
-This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks.
+This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks.
 Because \whoop performs pair-wise analysis, it over-approximates this flag, and thus can falsely report a pair as racy.

### 2015-05-15 whoop.tex
added page in book and slight rewording

@@
 Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time.
-This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks.
-Because \whoop performs pair-wise analysis, it over-approximates this flag, and thus can falsely report a pair as racy.
-However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
+This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks.
+Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy.
+However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
 

### 2015-05-15 whoop.tex
added page in book and slight rewording

@@
 When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
-The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
+The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
 %
@@
 because it was created dynamically or was indexed from an array of locks, it will exit with a warning.
-The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite{corbet2005linux}.
+The are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
 

### 2015-05-15 intro.tex
added racerx cite

@@
 Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races.
-Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability.
+Techniques that can detect races in drivers~\cite{dawson2003racerx, qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability.
 Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

### 2015-05-15 intro.tex
fix

@@
 %
-Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
+Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
 Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance.

### 2015-05-15 evaluation.tex
added url in tool link
some fixes

@@
 \centering
-\setlength{\tabcolsep}{0.45em}
+\setlength{\tabcolsep}{0.4em}
 \caption{Comparison with different yield instrumentation granularities and context-switch bounds.}
@@
 Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.
+
+For reproducibility, we make all of our tools (including source) and benchmarks available online:
+
+\begin{scriptsize}
+\textbf{\texttt{\url{http://multicore.doc.ic.ac.uk/tools/Whoop/ASE2015}}}
+\end{scriptsize}
 
@@
 Most of the races that \whoop and \corral discovered can be classified in two cases.
-The first case is accessing a global counter (or flag) from concurrent entry points, without a lock.
+The first case is accessing a global counter (or flag) from concurrent entry points, without holding a lock.
 This might be for performance, and indeed a lot of the races we found might be benign.
 Even benign races, though, lead to undefined behavior according to the C standard, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
-The second case is an entry point modifying a field of a struct (either global or passed as a parameter) without a lock.
+The second case is an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock.
 This can lead to a race if another entry point simultaneously accesses the same field of the same struct.

### 2015-05-15 evaluation.tex
some fixes

@@
 Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Figure~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
-All reported times are in seconds and averages over three runs.
+All reported times are in seconds and averaged over three runs.
 \corral was configured with a time budget of 10 hours (T.O.
-denotes timeout) and a csb of 2, 5 and 9, respectively.
+denotes timeout) and a csb of 2, 5 and 9.
 Standalone \corral uses \yieldall, which instruments context-switches (i.e.
@@
 Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9.
-We noticed that a higher csb typically results into greater speedups, when exploiting \whoop.
+We noticed that a higher csb typically results into greater speedups when exploiting \whoop.
 This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
@@
 We tried to compare \whoop with other similar approaches (see \S\ref{related}).
-However, we found this hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007 and might miss some of the latest features); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.
+However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007 and might be missing some of the latest features); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.

### 2015-05-15 related.tex
Done it.
some fixes

@@
 In comparison to \whoop, these tools rely heavily on user annotations.
-
+%
 Most related to our lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}.
@@
 RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races.
-To tackle this issue, RELAY employs post-analysis unsound filters, but these can also filter out real bugs.
+To tackle this issue, RELAY employs unsound post-analysis filters, but these can also filter out real bugs.
 Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, showcasing its limited scalability.
@@
 
-Prior works on static race checking include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speedup dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve a scalable race analysis; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speedup dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve a scalable race analysis; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
 

### 2015-05-15 intro.tex
fix

@@
 %
-\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as to concurrent systems that use a similar programming model (e.g.\ file systems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as to analyze concurrent systems that use a similar programming model (e.g.\ file systems).
 

### 2015-05-15 intro.tex
fix

--- 2015-05-15 intro.tex
@@
 %
-\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as to concurrent systems that use a similar programming model (e.g.\ file systems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as concurrent systems that use a similar programming model (e.g.\ file systems).
 

### 2015-05-15 background.tex
fix

@@
 Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}.
-Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
+Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
 %

### 2015-05-15 background.tex
fix

@@
 
-Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_s$, for each shared variable $s$ used in the program.
+Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program.
 Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start.

### 2015-05-15 whoop.tex
fix

@@
 Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.
-Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged by an automated theorem prover.
+Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged to an automated theorem prover.
 

### 2015-05-15 corral.tex
Cut to the chase in the abstract.
fix

@@
 assert s == e; // check written value
-\end{boogie}
+\end{boogie}\vspace{-2mm}
 %
@@
 assert x == s; // check read value
-\end{boogie}
+\end{boogie}\vspace{-2mm}
 %
@@
 By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
-Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, we do not instrument a \texttt{yield} after this statement.
+Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement.
 This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.

### 2015-05-15 evaluation.tex
fix

@@
 \whoop uses over-approximation to achieve scalability and, as expected, executes significantly faster than \corral in all our benchmarks.
-For example, \corral times out in all configurations (without and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
+For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
 We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might get \corral stuck.

### 2015-05-15 evaluation.tex
rephrase

@@
 We noticed that a higher csb typically results into greater speedups when exploiting \whoop.
-This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
+This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
 %
-However, if \whoop fails to verify any race-related properties of a driver, it can slow down \corral.
-We noticed this in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead.
-Although \whoop verified 46 out of 47 memory regions of sx8, it still failed to speedup \corral using \yieldmr.
-\corral checks the only two pairs associated with sx8 quite quickly (21.4 seconds with csb of 9), so the overhead of running \whoop outweighs the benefits of fine-grained instrumentation.
+However, if \whoop fails to verify any race-related properties of a driver, it might slow down \corral.
+We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but it did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver quite quickly (21.4 seconds with csb of 9).
+We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.
 

### 2015-05-15 whoop.tex
fix

@@
 The process repeats until a fixpoint is reached.
-The (possibly empty) set of remaining candidates has been proven to hold, and can used to summarize calls and loops during further program analysis.
+The (possibly empty) set of remaining candidates has been proven to hold, and can be used to summarize calls and loops during further program analysis.
 
@@
 These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.
-Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
+Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
 We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.

### 2015-05-15 whoop.tex
fix

@@
 Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
-We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.
+We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded by Houdini.
 The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

### 2015-05-15 whoop.tex
fix

@@
 Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity than the level of pairs.
-However, a pair-wise analysis can be trivially parallelized for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
+However, a pairwise analysis can be trivially parallelized for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
 In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

### 2015-05-15 conclusion.tex
fixes

@@
 Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrent bug-finder.
-Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.
+Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.

### 2015-05-15 evaluation.tex
fixes

@@
 We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}).
-We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel distribution.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
+We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
 Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.
@@
 %
-We evaluate our methodology against \sizeOfBenchmarks drivers from the 4.0 Linux kernel.
+We evaluate our methodology against \sizeOfBenchmarks drivers from the Linux 4.0 kernel.
 We chose non-trivial drivers of several types: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.

### 2015-05-15 intro.tex
fixes
fix

@@
 Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}.
-This has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
+This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
 %
-Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
+Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
 Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance.
@@
 We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze drivers and other concurrent programs.
-Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.
+Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel.
 By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.

### 2015-05-15 background.tex
fixes

@@
 %
-Modern operating systems address the demand for responsiveness and performance in device drivers by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.
+Modern operating systems address the demand for responsiveness and performance in device drivers by providing multiple sources of concurrency~\cite{corbet2005linux}: an arbitrary number of entry points from the same driver can be invoked concurrently; a running driver process can block, causing the driver to switch execution to another process; and hardware interrupts can be handled concurrently.
 These forms of concurrent execution are prone to \emph{data races}.

### 2015-05-15 intro.tex
fixes

@@
 %
-Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
+Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
 Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance.
@@
 %
-\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as concurrent systems that use a similar programming model (e.g.\ file systems).
+\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as concurrent systems that use a similar programming model (e.g.\ file systems).
 

### 2015-05-15 background.tex
fix in figure 2

@@
     case 1: offset += file->f_pos; break; // racy
-    case 2: offset += nvram_len; break;
     default: offset = -1;
   }
-  if (offset < 0)
-    return -EINVAL;
+  if (offset < 0) return -EINVAL;
   file->f_pos = offset; // racy
@@
     case 1: offset += file->f_pos; break;
-    case 2: offset += nvram_len; break;
     default: offset = -1;
   }
-  if (offset < 0)
+  if (offset < 0) {
+    mutex_unlock(&nvram_mutex); // unlock
     return -EINVAL;
+  }
   file->f_pos = offset;
@@
 
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$.
+Figure~\ref{fig:locksets} shows an example of applying lockset analysis to a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$.
 Initially, $\mathit{LS}_A$, which is the lockset for A, contains all possible locks in the program: $M$ and $N$.

### 2015-05-15 whoop.tex
reworded a footnote

@@
 We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity than the level of pairs.
-However, a pairwise analysis can be trivially parallelized for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
+However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
 In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

### 2015-05-15 whoop.tex
reworded a footnote

@@
 Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
-However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
+However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
 In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

### 2015-05-15 corral.tex
fixes

@@
 \whoop is a sound but imprecise static data race analyzer.
-For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate the performance of \corral~\cite{lal2012corral}, a precise bug-finder.
-\corral, which is used by Microsoft to analyze Windows drivers~\cite{lal2014powering}, has capabilities for analyzing concurrent programs.
+For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate the performance of \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze Windows drivers~\cite{lal2014powering}.
 Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C).

### 2015-05-15 evaluation.tex
fixes

@@
 We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}).
-We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
+We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral using different yield instrumentations and context-switch bounds.
 Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.

### 2015-05-15 whoop.tex
fixes

@@
 In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.
-In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
+In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.
 %
-Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.
+Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.
 
@@
 
-The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
+The allowed statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
 In particular, shared variables may not appear in arbitrary expressions.
@@
 \center
-\begin{tabular}{lp{5.5cm}}
-\textbf{Statement Form} & \textbf{Notes} \\
+\setlength{\tabcolsep}{0.45em}
+\begin{tabular}{lp{6.8cm}}
+\textbf{Statement} & \textbf{Notes} \\
 \toprule
@@
 \end{tabular}
-\caption{Forms of statements in our simple programming model}
+\caption{The allowed statements in our simple programming model}
 \label{fig:statements}
@@
 
-At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
+Initially, the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
 %
@@
 %
-We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.
+We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is held by some thread.
 Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.
@@
 \label{fig:instrumentation}
+\vspace{-2mm}
 \end{figure}
@@
 %
-We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct, it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
+We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, which is --- described above --- at the end of the program holds), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
 Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.
-By our hypothesis that the program of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate, the assertion checked at the end of the program guarantees at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.
+By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.
 By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.

### 2015-05-15 whoop.tex
moved toolchain figure earlier
fix

@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=.99\linewidth]{img/whoop.pdf}
+\vspace{-2mm}
+\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
+\vspace{-3mm}
+\label{fig:whoop}
+\end{figure*}
+
 \noindent\textbf{Soundness }
@@
 The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique.
-In practice, Linux drivers are written in C, we do not know up-front which are the entry points for a driver, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
+In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
 We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop\footnote{\url{https://github.com/pdeligia/whoop}}, a practical, fully automatic tool for detecting data races in drivers.
-
-\begin{figure*}
-\centering
-\includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\vspace{-2mm}
-\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
-\vspace{-2mm}
-\label{fig:whoop}
-\end{figure*}
 

### 2015-05-15 whoop.tex
fixes

@@
 Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
-Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
+Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
 %
@@
 An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
-This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse robust existing techniques and focus instead on verification efforts.
+Thus, \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts.
 
@@
 because it was created dynamically or was indexed from an array of locks, it will exit with a warning.
-The are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
+There are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
 
@@
 %
-Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}).
+Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}).
 These sets can be modeled directly in Boogie as characteristic functions, using maps.
@@
 Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
-We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded by Houdini.
+We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculate candidates that are later deemed to be incorrect, and thus discarded by Houdini.
 The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.
@@
 First, we enriched \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision.
-The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other.
+The Linux kernel can serialize calls to entry points, thus forcing them to run in sequence instead of an interleaved manner.
 For example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock.
@@
 Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy.
-However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
+However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it ``inhibits user ingenuity'' (e.g.\ a user might expect that a driver can be accessed concurrently for performance).
 

### 2015-05-15 corral.tex
fixes

@@
-\whoop is a sound but imprecise static data race analyzer.
-For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate the performance of \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze Windows drivers~\cite{lal2014powering}.
+\whoop is a sound but imprecise static race analyzer.
+For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze Windows drivers~\cite{lal2014powering}.
 Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C).

### 2015-05-15 whoop.tex
fixes

@@
 
-Statically analyzing drivers requires to ``close'' the environment by abstracting away the low-level implementation details.
+Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details.
 To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions.
 A limitation of our model is that it can, and will, ultimately result in false positives.
-However, we currently only focus on finding data races, and thus can get away with over-approximating a lot of the underlying kernel functionality, without losing too much precision.
+However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the underlying kernel functionality, without losing too much precision.
 Making our model more precise is an ongoing manual effort, but requires Linux expertise.

### 2015-05-15 evaluation.tex
fixes

@@
 
-Most of the races that \whoop and \corral discovered can be classified in two cases.
-The first case is accessing a global counter (or flag) from concurrent entry points, without holding a lock.
+Most of the races that \whoop and \corral discovered fall into two categories.
+The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock.
 This might be for performance, and indeed a lot of the races we found might be benign.
 Even benign races, though, lead to undefined behavior according to the C standard, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
-The second case is an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock.
+The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock.
 This can lead to a race if another entry point simultaneously accesses the same field of the same struct.
 
-As an example of the second case, we found the following race in the generic\_nvram driver (see Figure~\ref{fig:data_race_example}): in the \texttt{llseek} entry point the driver accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+As an example of the second category, we found the following race in the generic\_nvram driver (see Figure~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.

### 2015-05-15 related.tex
fix

@@
 RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races.
-To tackle this issue, RELAY employs unsound post-analysis filters, but these can also filter out real bugs.
+To tackle this issue, RELAY employs unsound post-analysis filters and, hence, can also filter out real bugs.
 Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, showcasing its limited scalability.
@@
 Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speedup dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve a scalable race analysis; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+Our tool, \whoop, uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then uses \corral to filter out false races.
 

### 2015-07-29 background.tex
fixes
update towards camera ready and also included the reviews

@@
 
-Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux driver.
+Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver.
 The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter.
@@
   switch (origin) {
-    case 0: break;
+    ...
     case 1: offset += file->f_pos; break; // racy
-    default: offset = -1;
+    ...
   }
@@
 \vspace{-2mm}
-\caption{Racy entry point in the generic\_nvram Linux driver}
+\caption{Racy entry point in a Linux 4.0 driver (drivers/char/generic\_nvram.c)}
 \label{fig:data_race_example}
@@
   switch (origin) {
-    case 0: break;
+    ...
     case 1: offset += file->f_pos; break;
-    default: offset = -1;
+    ...
   }
@@
 \includegraphics[width=1\linewidth]{img/lockset.pdf}
-\vspace{-2mm}
+\vspace{-5mm}
 \caption{Applying lockset analysis on a concurrent program}
 \label{fig:locksets}
-\vspace{-2mm}
+\vspace{-3mm}
 \end{figure}

### 2015-07-29 corral.tex
fixes
update towards camera ready and also included the reviews

@@
 This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.
-This is because each access, for which a \texttt{yield} is suppressed, is guaranteed to be protected by some lock (a consequence of lockset analysis).
+This is because each access for which a \texttt{yield} is suppressed is guaranteed to be protected by some lock (a consequence of lockset analysis).
 If the access is a write, its effects are not visible by the other entry point in the pair until the lock is released.
@@
 Our partial-order reduction can be used to accelerate \corral for arbitrary bug-finding in concurrent programs.
-In the remainder of this paper, though, we regard races as bugs and use \corral solely to look for races.
+In the remainder of this paper, though, we only regard races as bugs and use \corral solely to look for races.

### 2015-07-29 evaluation.tex
update towards camera ready and also included the reviews

@@
 %
-Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
+Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
 Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug.
@@
 
-\whoop uses over-approximation to achieve scalability and, as expected, executes significantly faster than \corral in all our benchmarks.
+\whoop uses over-approximation to scale and, as expected, executes significantly faster than \corral in all our benchmarks.
 For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
-We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might get \corral stuck.
+We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might cause \corral to get stuck.
 This is not an issue for \whoop, which uses procedure summarization.
@@
 Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr.
-Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9.
+Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9.
 We noticed that a higher csb typically results into greater speedups when exploiting \whoop.
-This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
+This is expected, as a higher csb translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
 %
-However, if \whoop fails to verify any race-related properties of a driver, it might slow down \corral.
-We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but it did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver quite quickly (21.4 seconds with csb of 9).
+However, there are some cases where \whoop might slow down \corral.
+We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9).
 We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.
@@
 We tried to compare \whoop with other similar approaches (see \S\ref{related}).
-However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007 and might be missing some of the latest features); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.
+However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.

### 2015-07-29 related.tex
update towards camera ready and also included the reviews

@@
 To tackle this issue, RELAY employs unsound post-analysis filters and, hence, can also filter out real bugs.
-Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, showcasing its limited scalability.
+Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, hinting at its limited scalability.
 \whoop aims to achieve scalability \emph{and} precision: the first, via novel symbolic pairwise lockset analysis; and the second, by accelerating \corral, an industrial-strength precise bug-finder.
 
-Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speedup dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve a scalable race analysis; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speed up dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
 Our tool, \whoop, uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then uses \corral to filter out false races.

### 2015-07-29 whoop.tex
update towards camera ready and also included the reviews

@@
 For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location.
-We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
-However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.
+We soundly model the effects of additional threads by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.
+\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
+However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}
+The trade-off is that a quadratic number of entry point pairs must be checked.
 In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.
@@
 %
-We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, which is --- described above --- at the end of the program holds), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
+We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
 Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.
@@
 %
-Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions.
-Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.
+Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel functions.
+Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for \texttt{file\_operations}); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.
 
@@
 Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs.
-However, it is relatively easy to enhance our tool with knowledge of other locking primitives.
 If \whoop cannot infer a lock, e.g.
@@
 A limitation of our model is that it can, and will, ultimately result in false positives.
-However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the underlying kernel functionality, without losing too much precision.
-Making our model more precise is an ongoing manual effort, but requires Linux expertise.
+However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision.
+Making our model more precise is an ongoing manual effort, but requires Linux kernel expertise.
 We argue that further work on the model is orthogonal to the contributions of this paper.

### 2015-08-04 background.tex
pass over the paper addressing Zvonimir's feedback

@@
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program.
-Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start.
+Initially, $\mathit{CLS}_T$ is empty for every $T$, because the threads do not hold any locks on program start.
 The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock.
@@
 %
-When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
+When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
+%
 If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

### 2015-08-04 corral.tex
pass over the paper addressing Zvonimir's feedback

@@
 \yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock.
-In our experiments (see \S\ref{evaluation}), \yieldmr significantly outperforms \yieldcoarse.
+In our experiments (see \S\ref{evaluation}), \yieldmr is significantly faster than \yieldcoarse.
 
-Our partial-order reduction can be used to accelerate \corral for arbitrary bug-finding in concurrent programs.
-In the remainder of this paper, though, we only regard races as bugs and use \corral solely to look for races.
+Our partial-order reduction is able in principle to accelerate \corral for \emph{arbitrary} bug-finding in concurrent programs.
+Although we did preliminary explorations in this direction, identifying useful safety properties to check was challenging since there are typically no assertions in drivers.
+Thus, in the remainder of this paper, we only regard data races as bugs and use \corral solely to look for data races.

### 2015-08-04 evaluation.tex
pass over the paper addressing Zvonimir's feedback

@@
 %
-Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).
-Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug.
-\corral did not discover any races that \whoop did not already report.
-This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.
+Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral, because \whoop might find that a memory region is racy, but the same memory region might race in many points in the program.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of $\leq$ 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
 
@@
 Another issue with \corral is loop coverage due to unsound loop unrolling.
-To tackle this, we enable the built-in loop over-approximation described in~\cite{lal2014powering}.
+To tackle this, we enable the built-in loop over-approximation described in previous work~\cite{lal2014powering}.
 This can potentially lead \corral to report false bugs, but we have not seen this in practice.
-Finally, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
+Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
+Standalone \corral did not discover any data races that \whoop did not already report.
+This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.
 

### 2015-08-04 intro.tex
pass over the paper addressing Zvonimir's feedback

@@
 Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races.
-Techniques that can detect races in drivers~\cite{dawson2003racerx, qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability.
-Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.
+Techniques that can detect races in drivers~\cite{dawson2003racerx, qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either overly \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability.
+Thus, there is a clear need for new tools that can detect races efficiently and precisely.
 

### 2015-08-04 whoop.tex
pass over the paper addressing Zvonimir's feedback

@@
 We soundly model the effects of additional threads by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.
-\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
 However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}
+%
 The trade-off is that a quadratic number of entry point pairs must be checked.
@@
 In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop\footnote{\url{https://github.com/pdeligia/whoop}}, a practical, fully automatic tool for detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, fully automatic tool for detecting data races in drivers.
 
@@
 The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
-Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
+Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
 Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.
@@
 %
-When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable.
-The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
+When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant.
+Representing all locks statically, instead of their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
 %
-Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs.
-If \whoop cannot infer a lock, e.g.
-because it was created dynamically or was indexed from an array of locks, it will exit with a warning.
-There are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
+Currently, \whoop supports Linux kernel mutex and spinlock operations.  In addition, it is easy to enhance it with knowledge of other locking primitives.
+%
+If \whoop cannot reliably infer a unique lock instance (e.g., because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning.
+However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
 

### 2015-08-05 background.tex
Removed vspace hacks for now.
Done up to Section 3

@@
 Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver.
-The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter.
-However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order.
+The driver can invoke the entry point concurrently from two threads, with the same \texttt{file} struct as a parameter.
+This can lead to multiple possible data races because the threads can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order.
 Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).
@@
 \end{lstlisting}
-\vspace{-2mm}
-\caption{Racy entry point in a Linux 4.0 driver (drivers/char/generic\_nvram.c)}
+%\vspace{-2mm}
+\caption{Racy entry point in the generic\_nvram Linux 4.0 driver}
 \label{fig:data_race_example}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 %
-Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency; while fine-grained locking can easily lead to deadlocks.
-Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}
+Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks.
+Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by regarding them as not providing any protection between threads; this can lead to false alarms.}
 
@@
 \end{lstlisting}
-\vspace{-2mm}
-\caption{Using a lock to eliminate the data races in the previous example}
+%\vspace{-2mm}
+\caption{Introducing a lock to eliminate the data races in the previous example}
 \label{fig:lock_example}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.
-The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution.
-If that lockset ever becomes empty, the analysis reports a \emph{potential} race on that memory location.
-This is because an empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
+The idea is to track the set of locks (i.e.\ \emph{lockset}) that are \emph{consistently} used to protect a memory location during program execution.
+An empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.
+Consequently, the analysis reports a \emph{potential} race on that memory location.
 
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program.
-Initially, $\mathit{CLS}_T$ is empty for every $T$, because the threads do not hold any locks on program start.
-The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock.
+Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start.
+In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock.
 The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
@@
 In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.
-The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
+The technique, though, suffers from imprecision (i.e.\ can report false bugs) since a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
 Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler.
-This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.
+Since the tool runs the program in a truly concurrent environment, it only gets to see what the thread did on the particular schedule that was followed.
 
-To counter the above limitations, techniques such as Locksmith~\cite{pratikakis2006locksmith} and RELAY~\cite{voung2007relay} have explored the idea of applying lockset analysis in a static context, using dataflow analysis.
+To counter the above limitations, techniques such as Locksmith~\cite{pratikakis2006locksmith} and RELAY~\cite{voung2007relay} have explored the idea of applying lockset analysis statically, using dataflow analysis.
 In this paper, we present a novel symbolic lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.
@@
 \includegraphics[width=1\linewidth]{img/lockset.pdf}
-\vspace{-5mm}
+%\vspace{-5mm}
 \caption{Applying lockset analysis on a concurrent program}
 \label{fig:locksets}
-\vspace{-3mm}
+%\vspace{-3mm}
 \end{figure}

### 2015-08-05 intro.tex
Removed vspace hacks for now.
Done up to Section 3

@@
 Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}.
-This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
+This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification, and validation schedules~\cite{yavatkar2012era}.
 %
 Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers.
-Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance.
-The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.
+Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} established that they account for 19\% of the total bugs in Linux drivers, showcasing their significance.
+The majority of these concurrency bugs were \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.
 
@@
 Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel.
-By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
+Our combination achieves analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.
 For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
@@
 To summarize, our contributions are as follows:
-\vspace{-0.5mm}
+%\vspace{-0.5mm}
 \begin{itemize}
 \item We propose symbolic pairwise lockset analysis, a sound and scalable technique for automatically verifying the absence of data races in device drivers.
-\vspace{-0.2mm}
+%\vspace{-0.2mm}
 \item We present \whoop, a tool that leverages our approach for analyzing data races in device drivers.
-\vspace{-0.2mm}
+%\vspace{-0.2mm}
 \item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, an industrial-strength bug-finder.
-\vspace{-0.2mm}
+%\vspace{-0.2mm}
 \item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race-checking and accelerating \corral.

### 2015-08-05 whoop.tex
Removed vspace hacks for now.
Done with 3.A.

@@
 We soundly model the effects of additional threads by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.
-Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself.
+Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity, such as by considering all entry points one after the other, taking into account that an entry point can race with itself.
 However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}
@@
 In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
-Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.
+Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers, or dynamic memory allocation.
 These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
@@
 \label{fig:statements}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
-Each thread consists of an id, drawn from $\ids$, a thread template $T$ an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.
+Each thread consists of an id (drawn from $\ids$), a thread template $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.
 If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.
@@
 %
-For templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
+For templates $T$ and $U$, which can be equal, we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrary many further concurrently executing threads.
 %
@@
 \label{fig:instrumentation}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
-First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.
-The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter, due to renaming).
+First, the read, write, and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes $\mutexes$.
+The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter due to renaming).
 Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.
@@
 \label{fig:sequentialization}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 \includegraphics[width=.99\linewidth]{img/whoop.pdf}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
-\vspace{-3mm}
+%\vspace{-3mm}
 \label{fig:whoop}
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Entry point definitions in the generic\_nvram driver}
 \label{fig:entrypoints}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 

### 2015-08-05 background.tex
Changed Figure into Fig.

@@
 
-Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver.
+Fig.~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver.
 The driver can invoke the entry point concurrently from two threads, with the same \texttt{file} struct as a parameter.
@@
 The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.
-Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}.
+Fig.~\ref{fig:lock_example} shows how to use locking to eliminate the races in Fig.~\ref{fig:data_race_example}.
 Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
@@
 
-Figure~\ref{fig:locksets} shows an example of applying lockset analysis to a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$.
+Fig.~\ref{fig:locksets} shows an example of applying lockset analysis to a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$.
 Initially, $\mathit{LS}_A$, which is the lockset for A, contains all possible locks in the program: $M$ and $N$.

### 2015-08-05 evaluation.tex
Removed vspace hacks for now.
Changed Figure into Fig.

@@
 \end{tabular}
-\vspace{-3mm}
+%\vspace{-3mm}
 \end{table}
@@
 \end{tabular}
-\vspace{-3mm}
+%\vspace{-3mm}
 \end{table*}
@@
 \includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
@@
 \label{fig:plot}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 
-As an example of the second category, we found the following race in the generic\_nvram driver (see Figure~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
@@
 %
-Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Figure~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
+Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
 All reported times are in seconds and averaged over three runs.
@@
 Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr.
-Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9.
+Fig.~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9.
 We noticed that a higher csb typically results into greater speedups when exploiting \whoop.

### 2015-08-05 whoop.tex
Changed Figure into Fig.

@@
 
-The allowed statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
+The allowed statements are shown in Fig.~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.
 In particular, shared variables may not appear in arbitrary expressions.
@@
 %
-The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.
+The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Fig.~\ref{fig:instrumentation}.
 For a shared variable assignment $s = e$, we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).
@@
 Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account for possible side-effects due to other threads that are running concurrently.
-The instrumentation of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.
+The instrumentation of Fig.~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.
 This is reflected by the use of a $\havockeyword$ command, which sets its argument to an arbitrary value.
@@
 %
-The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
+The pseudocode of Fig.~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
 Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
@@
 %
-We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
-Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.
+We sketch an argument that if the program of Fig.~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
+Let us assume that the program of Fig.~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.
 By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.
-By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
+By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Fig.~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
 As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
 
-In the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.
+In the presence of non-termination the assertion at the end of Fig.~\ref{fig:sequentialization} may not be reached.
 The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.
@@
 In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, fully automatic tool for detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, automated tool for detecting data races in drivers.
 
@@
 %
-Figure~\ref{fig:whoop} depicts the \whoop toolchain.
+Fig.~\ref{fig:whoop} depicts the \whoop toolchain.
 The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
 Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
-Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.
-After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).
+Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.
+After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Fig.~\ref{fig:whoop}--C and \S\ref{corral}).
 
-We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).
+We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Fig.~\ref{fig:whoop}).
 For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
@@
 Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel functions.
-Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for \texttt{file\_operations}); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.
+Linux drivers define entry points in a standard way (see Fig.~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for \texttt{file\_operations}); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.
 
@@
 %
-Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}).
+Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Fig.~\ref{fig:instrumentation}).
 These sets can be modeled directly in Boogie as characteristic functions, using maps.
@@
 For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.
-In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
+In particular, the verification for the root-level procedure, encoding the sequential program sketched in Fig.~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
 Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user.
 To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code.
-The following is a race that \whoop found and reported for the example of Figure~\ref{fig:data_race_example}:
+The following is a race that \whoop found and reported for the example of Fig.~\ref{fig:data_race_example}:
 

### 2015-08-05 whoop.tex
Minor tweaks to save a line here and there

@@
 We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is held by some thread.
-Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.
+Executing a statement updates the thread-private and shared state in a standard manner.
 For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.
@@
 %
-The pseudocode of Fig.~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
+Fig.~\ref{fig:sequentialization} shows the sequential program that we analyze to prove race-freedom for a pair of thread templates $T$ and $U$.
 Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
@@
 \end{tabular}
-\caption{The sequential program to be analyzed in order to prove race-freedom for a pair of thread templates}
+\caption{The sequential program to be analyzed to prove race-freedom for a pair of thread templates}
 \label{fig:sequentialization}

### 2015-08-05 whoop.tex
Done with Sec 3

@@
 %
-Currently, \whoop supports Linux kernel mutex and spinlock operations.  In addition, it is easy to enhance it with knowledge of other locking primitives.
+Currently, \whoop supports Linux kernel mutex and spinlock operations; it is easy to enhance it with knowledge of other locking primitives.
 %
-If \whoop cannot reliably infer a unique lock instance (e.g., because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning.
+If \whoop cannot reliably infer a unique lock instance (e.g.\ because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning.
 However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
@@
 
-To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.
-Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $\mathit{MR}$.
+To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from previous work~\cite{bardsley2014engineering}.
+Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region $\mathit{MR}$.
 When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_{\mathit{MR}}$, representing some unspecified index into $\mathit{MR}$; we call this the \emph{watched offset} for $\mathit{MR}$.
@@
 Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.
-The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.
-We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.
+This is done based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.
+We give two examples; for clarity we use notation from the simple concurrent programming model of \S\ref{sec:symbolicpairwise}.
 If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.
 These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.
-Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
+Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be incorrect, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.
 We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculate candidates that are later deemed to be incorrect, and thus discarded by Houdini.
@@
 %
-For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine.
+For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is given to the Boogie verification engine.
 For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.
-In particular, the verification for the root-level procedure, encoding the sequential program sketched in Fig.~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
+In particular, the verification for the root-level procedure, implementing the sequential program sketched in Fig.~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair.
 Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user.
@@
 %
-We have implemented various optimizations to increase the precision and performance of \whoop.  We comment on the two most effective optimizations.
+We have implemented various optimizations to increase the precision and performance of \whoop.  We comment on the two most effective ones.
 
-First, we enriched \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision.
+First, we enriched \whoop with information regarding \emph{kernel-imposed serialization} to increase precision.
 The Linux kernel can serialize calls to entry points, thus forcing them to run in sequence instead of an interleaved manner.
@@
 This is an ongoing manual effort: the more drivers we study, the more such properties we discover to make \whoop more precise.
-Currently, this information is hard-coded inside \whoop.
-In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.
+%Currently, this information is hard-coded inside \whoop. In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.
 
@@
 
-As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
+In addition to inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
 However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.
@@
 Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time.
-This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks.
+This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set, then the checking entry point executes, else it blocks.
 Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy.
-However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it ``inhibits user ingenuity'' (e.g.\ a user might expect that a driver can be accessed concurrently for performance).
+However, experts~\cite[p.\ 173]{corbet2005linux} advise against serializing drivers in this way, as it ``inhibits user ingenuity'' (e.g.\ a user might expect that a driver can be accessed concurrently for performance).
 
@@
 However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision.
-Making our model more precise is an ongoing manual effort, but requires Linux kernel expertise.
+Making our model more precise is an ongoing manual effort that requires Linux kernel expertise.
 We argue that further work on the model is orthogonal to the contributions of this paper.

### 2015-08-05 corral.tex
Removed vspace hacks for now.
Changed Figure into Fig.
Done with Sec 4

@@
 For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze Windows drivers~\cite{lal2014powering}.
-Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C).
+Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Fig.~\ref{fig:whoop}--C).
 Our technique, though, is general and capable in principle of accelerating any concurrency bug-finder that operates by interleaving threads at shared memory operations.
@@
 assert s == e; // check written value
-\end{boogie}\vspace{-2mm}
+\end{boogie}%\vspace{-2mm}
 %
@@
 assert x == s; // check read value
-\end{boogie}\vspace{-2mm}
+\end{boogie}%\vspace{-2mm}
 %
@@
 Our partial-order reduction is able in principle to accelerate \corral for \emph{arbitrary} bug-finding in concurrent programs.
-Although we did preliminary explorations in this direction, identifying useful safety properties to check was challenging since there are typically no assertions in drivers.
-Thus, in the remainder of this paper, we only regard data races as bugs and use \corral solely to look for data races.
+Although we did preliminary explorations in this direction, identifying useful safety properties to check was challenging since drivers typically contain no assertions.
+Thus, in this paper we use \corral solely to find data races.

### 2015-08-05 conclusion.tex
Done

@@
-In this paper we presented \whoop, a new fully automatic approach for detecting all possible data races in device drivers.
-Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which has the potential to scale well.
-Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrent bug-finder.
+In this paper we presented \whoop, a new automated approach for detecting all possible data races in device drivers.
+Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which scales well.
+Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrency bug-finder.
 Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.

### 2015-08-05 evaluation.tex
Done

@@
 \caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
-The symbols $+$, $\circ$ and $\times$, represent a context-switch bound of 2, 5 and 9, respectively.}
+The symbols $+$, $\circ$, and $\times$ represent a context-switch bound of 2, 5, and 9, respectively.}
 \label{fig:plot}
@@
 We evaluate our methodology against \sizeOfBenchmarks drivers from the Linux 4.0 kernel.
-We chose non-trivial drivers of several types: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog.
+We chose non-trivial drivers of several types: block, char, ethernet, near field communication (nfc), universal serial bus (usb), and watchdog.
 For all these drivers, we had to understand their environment and manually model it; this required approximately two months of work.
@@
 %
-Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral, because \whoop might find that a memory region is racy, but the same memory region might race in many points in the program.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of $\leq$ 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
+Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
 
@@
 
-As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
 This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
@@
 %
-Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
+Table~\ref{tab:results} presents runtime results from using \whoop, \corral, and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
 All reported times are in seconds and averaged over three runs.
 \corral was configured with a time budget of 10 hours (T.O.
-denotes timeout) and a csb of 2, 5 and 9.
+denotes timeout) and a csb of 2, 5, and 9.
 Standalone \corral uses \yieldall, which instruments context-switches (i.e.
@@
 For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
-We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might cause \corral to get stuck.
+We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which puts burden on \corral's stratified inlining.
 This is not an issue for \whoop, which uses procedure summarization.
@@
 Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr.
-Fig.~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9.
-We noticed that a higher csb typically results into greater speedups when exploiting \whoop.
-This is expected, as a higher csb translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
+Fig.~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; in ssu100 and pc8736x\_gpio with a csb of 9 we achieved a speedup of 12$\times$ and 20$\times$, respectively.
+We noticed that a higher csb typically results in greater speedups when exploiting \whoop.
+This is expected as complexity grows exponentially with csb, and hence \whoop's reduction helps more at a higher csb.
 %
-However, there are some cases where \whoop might slow down \corral.
+However, there are cases where \whoop might slow down \corral.
 We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9).
@@
 We tried to compare \whoop with other similar approaches (see \S\ref{related}).
-However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.
+However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007); we also could not find source code or binaries of other tools~\cite{kahlon2007fast, kahlon2009semantic, das2015section}.

### 2015-08-05 related.tex
Done

@@
 Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, hinting at its limited scalability.
-\whoop aims to achieve scalability \emph{and} precision: the first, via novel symbolic pairwise lockset analysis; and the second, by accelerating \corral, an industrial-strength precise bug-finder.
+\whoop aims to achieve scalability \emph{and} precision: the first via novel symbolic pairwise lockset analysis, and the second by accelerating \corral, an industrial-strength precise bug-finder.
 
-Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speed up dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
-Our tool, \whoop, uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then uses \corral to filter out false races.
+Prior works on static race checking also include the following.
+Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection.
+Kahlon et al.~\cite{kahlon2007fast} use a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; more recently, they used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings~\cite{kahlon2009semantic}.
+Das et al.~\cite{das2015section} employ inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection.
+\whoop uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then employs \corral to filter out false races.
 

### 2015-08-06 background.tex
Complete pass.

@@
 Fig.~\ref{fig:lock_example} shows how to use locking to eliminate the races in Fig.~\ref{fig:data_race_example}.
-Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
+\COULDCUT{Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.}
 %
@@
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program.
-Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start.
+Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks initially.
 In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock.
@@
 In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.
-The technique, though, suffers from imprecision (i.e.\ can report false bugs) since a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
+The technique, though, can report false alarms since a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}.
 Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler.
-Since the tool runs the program in a truly concurrent environment, it only gets to see what the thread did on the particular schedule that was followed.
 

### 2015-08-06 conclusion.tex
Complete pass.

@@
 Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrency bug-finder.
-Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.
+\COULDCUT{Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.}

### 2015-08-06 corral.tex
Complete pass.

@@
 \corral performs bounded exploration of a concurrent program in two steps.
-First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.
+First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program up to the context bound~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.
 Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program.
@@
 \whoop takes advantage of this precision to report only feasible data races.
-Also note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
+Also note that our instrumentation conveniently tolerates some benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
 
@@
 %
-The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair as race free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
+The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair (EPP) as race-free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
 %
-\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock.
+\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region (MR) that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock.
 In our experiments (see \S\ref{evaluation}), \yieldmr is significantly faster than \yieldcoarse.

### 2015-08-06 evaluation.tex
Complete pass.

@@
 Standalone \corral did not discover any data races that \whoop did not already report.
-This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.
+This is expected, as \whoop aims for soundness, and this increases our confidence in the implementation of \whoop.
 
@@
 This might be for performance, and indeed a lot of the races we found might be benign.
-Even benign races, though, lead to undefined behavior according to the C standard, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
+Even benign races, though, lead to undefined behavior according to the C standard\COULDCUT{, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations}.
 The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock.

### 2015-08-06 intro.tex
Complete pass.

@@
-Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer.
+Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software \COULDCUT{to provide an interface between the operating system and hardware devices that are attached to a computer}.
 Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}.

### 2015-08-06 related.tex
Complete pass.

@@
 
-Prior works on static race checking also include the following.
-Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection.
+\COULDCUT{Prior works on static race checking also include the following.} Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection.
 Kahlon et al.~\cite{kahlon2007fast} use a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; more recently, they used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings~\cite{kahlon2009semantic}.

### 2015-08-06 whoop.tex
Complete pass.

@@
 %
-Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.
+\COULDCUT{Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.}
 
@@
 
-Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.
-Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged to an automated theorem prover.
+Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence.
+A post-condition for the sequential program asserts, for every shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.
+Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions (VCs)~\cite{barnett2005weakest} are generated and discharged to an automated theorem prover.
 
@@
 At any given point of execution a certain number of threads are active, each thread executing a particular template.
-In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
+In the context of device drivers, a thread template corresponds to a driver entry point.
+Multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
 Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers, or dynamic memory allocation.
@@
 In particular, shared variables may not appear in arbitrary expressions.
-This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.
+This restriction simplifies our presentation of lockset instrumentation below and is trivial to enforce by pre-processing.
 We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.
@@
 Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.
-The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.
-Each thread consists of an id (drawn from $\ids$), a thread template $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.
+The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{thread state}.
+Each thread state has an associated thread template $T$, the type of the thread (and multiple threads may have the same associated template), and consistes of an id (drawn from $\ids$), an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.
 If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.
 
-Initially, the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
+Initially, the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of thread states is empty.
 %
-At any point of execution, a new thread may be added to the list of threads.
-This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
+At any point of execution, a new thread state may be added to the list of thread states.
+This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread state to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
 %
@@
 
-A thread terminates if it reaches the end of $\mainforthread{T}$, and is removed from the list of threads.
+A thread terminates if it reaches the end of $\mainforthread{T}$, in which case its state is removed from the list of thread states.
 Since our interest is in analysis of device drivers, which are reactive, we do not consider the notion of global program termination.
@@
 %
-Renaming is straightforward: all occurrences of each private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
+Renaming is straightforward: each occurrence of a private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.
 The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.
@@
 %
-We sketch an argument that if the program of Fig.~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
+We sketch an argument that if the program of Fig.~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds invariantly at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate.
 Let us assume that the program of Fig.~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.
@@
 The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
-Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).
+Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Boogie is a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ the Boogie verifier and \corral).
 Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine.
@@
 An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
-Thus, \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts.
+Thus, \whoop does not need to directly deal with pointers and alias analysis\COULDCUT{, a hard problem in its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts}.
 
@@
 When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant.
-Representing all locks statically, instead of their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
+Representing all locks statically, instead of using their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
 %
@@
 However, this requires the use of quantifiers to express properties related to set contents.
-For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.
+For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A\;.\; \neg X[a]$.
 It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.
@@
 Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.
-We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined.
+We found that this did not scale to large drivers, and some drivers do exhibit recursion, e.g.\ the r8169 ethernet driver (see \S\ref{evaluation}).
 
@@
 
-In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.
+\COULDCUT{
+In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.}
 
@@
 \whoop also uses over-approximation, which can be another source of imprecision.
-Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks.
+Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the latter could be addressed by providing a mechanism for users to declare custom locks.
 We also do not currently treat interrupt handlers in a special way; we just assume that they can execute concurrently with any entry point.

### 2015-08-06 evaluation.tex
Added a blurb how SMACK is now more sound

@@
 %
-We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5, Z3~4.3.2, Boogie rev.
+We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5.0, Z3~4.3.2, Boogie rev.
 4192 and \corral rev.

### 2015-08-06 whoop.tex
Added a blurb how SMACK is now more sound

@@
 
-In addition to inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
+In addition to inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) the 1.5.0 release of SMACK we used is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses;\footnote{Note that currently bit-vectors and unaligned byte-level memory accesses are soundly and precisely handled by SMACK (as of release 1.5.1).} and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.
 However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.

### 2015-08-07 background.tex
some fixes

@@
 \end{lstlisting}
-%\vspace{-2mm}
-\caption{Racy entry point in the generic\_nvram Linux 4.0 driver}
+\vspace{-2mm}
+\caption{Racy entry point in the generic\_nvram Linux 4.0 driver.}
 \label{fig:data_race_example}
@@
 \end{lstlisting}
-%\vspace{-2mm}
-\caption{Introducing a lock to eliminate the data races in the previous example}
+\vspace{-2mm}
+\caption{Introducing a lock to eliminate the races in the previous example.}
 \label{fig:lock_example}
@@
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program.
-Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks initially.
+Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start.
 In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock.
@@
 %
-When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
+When $T$ accesses $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
 %
@@
 %\vspace{-5mm}
-\caption{Applying lockset analysis on a concurrent program}
+\caption{Applying lockset analysis on a concurrent program.}
 \label{fig:locksets}

### 2015-08-07 whoop.tex
some fixes

@@
 In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.
-%
-\COULDCUT{Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.}
 
@@
 \end{tabular}
-\caption{The allowed statements in our simple programming model}
+\caption{The allowed statements in our simple programming model.}
 \label{fig:statements}
@@
 Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.
-The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{thread state}.
-Each thread state has an associated thread template $T$, the type of the thread (and multiple threads may have the same associated template), and consistes of an id (drawn from $\ids$), an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.
+The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{thread states}.
+Each thread state has an associated thread template $T$, the type of the thread (and multiple threads may have the same associated template), and consists of an id (drawn from $\ids$), an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.
 If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.
@@
 \end{tabular}
-\caption{Instrumenting statements for lockset analysis}
+\caption{Instrumenting statements for lockset analysis.}
 \label{fig:instrumentation}
@@
 \end{tabular}
-\caption{The sequential program to be analyzed to prove race-freedom for a pair of thread templates}
+\caption{The sequential program to be analyzed to prove race-freedom for a pair of thread templates.}
 \label{fig:sequentialization}
@@
 %\vspace{-2mm}
-\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
+\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools.}
 %\vspace{-3mm}
@@
 \end{lstlisting}
-%\vspace{-2mm}
-\caption{Entry point definitions in the generic\_nvram driver}
+\vspace{-2mm}
+\caption{Entry point definitions in the generic\_nvram driver.}
 \label{fig:entrypoints}
@@
 Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.
-We found that this did not scale to large drivers, and some drivers do exhibit recursion, e.g.\ the r8169 ethernet driver (see \S\ref{evaluation}).
+We found that this did not scale to large drivers, and that some drivers do exhibit recursion, e.g.\ the r8169 ethernet driver (see \S\ref{evaluation}).
 

### 2015-08-07 background.tex
fix

@@
 \label{definition:datarace}
-A \emph{data race} occurs when two distinct threads access a shared memory location, at least one of the accesses modifies the location, at least one of the accesses is non-atomic, and there is no mechanism in place to prevent these accesses from being simultaneous.
+A \emph{data race} occurs when two distinct threads access a shared memory location, at least one of the accesses is a write access, at least one of the accesses is non-atomic, and there is no mechanism in place to prevent these accesses from being simultaneous.
 \end{definition}

### 2015-08-10 corral.tex
fixes

@@
 \whoop takes advantage of this precision to report only feasible data races.
-Also note that our instrumentation conveniently tolerates some benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.
+%
+Note that our instrumentation conveniently tolerates some benign races: it does not report a read-write race if the write access updates the racy memory location with the same value that already existed; it also does not report a write-write race if the two write accesses update the racy memory location with the same value, which can be different from the pre-existing.
 
@@
 By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
-Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement.
+%
+Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement, and we also omit the \texttt{assert} that would check for a race.
+%
 This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.

### 2015-08-10 evaluation.tex
some fixes
fixes

@@
+We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}).
+We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral using different yield instrumentations and context-switch bounds.
+Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.
+
+%For reproducibility, we make all of our tools (including source) and benchmarks available online:
+
+%\begin{scriptsize}
+%\textbf{\texttt{\url{http://multicore.doc.ic.ac.uk/tools/Whoop/ASE2015}}}
+%\end{scriptsize}
+
+\noindent\textbf{Experimental Setup }
+%
+We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5.0, Z3~4.3.2, Boogie rev.
+4192 and \corral rev.
+534.
+We also used Mono~4.1.0 to run Boogie and \corral.
+
+\noindent\textbf{Benchmarks }
+%
+We evaluate our methodology against \sizeOfBenchmarks drivers from the Linux 4.0 kernel.
+We chose non-trivial drivers of several types: block, char, ethernet, near field communication (nfc), universal serial bus (usb), and watchdog.
+For all these drivers, we had to understand their environment and manually model it; this required approximately two months of work.
+
+\noindent\textbf{Race-Checking }
+%
+Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
+
+We can see that \whoop reports more races than \corral does.
+This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.
+On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.
+Another issue with \corral is loop coverage due to unsound loop unrolling.
+To tackle this, we enable the built-in loop over-approximation described in previous work~\cite{lal2014powering}.
+This can potentially lead \corral to report false bugs, but we have not seen this in practice.
+Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
+Standalone \corral did not discover any data races that \whoop did not already report.
+This is expected, as \whoop aims for soundness, and this increases our confidence in the implementation of \whoop.
+
+Most of the races that \whoop and \corral discovered fall into two categories.
+The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock.
+This might be for performance, and indeed a lot of the races we found might be benign.
+Even benign races, though, lead to undefined behavior according to the C standard.
+%
+%, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
+%
+The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock.
+This can lead to a race if another entry point simultaneously accesses the same field of the same struct.
+
+As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
+This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
+We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.
+
 \newcommand{\colspacing}{\hspace{1.8em}}
@@
 
-We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}).
-We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral using different yield instrumentations and context-switch bounds.
-Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.
-
-For reproducibility, we make all of our tools (including source) and benchmarks available online:
-
-\begin{scriptsize}
-\textbf{\texttt{\url{http://multicore.doc.ic.ac.uk/tools/Whoop/ASE2015}}}
-\end{scriptsize}
-
-\noindent\textbf{Experimental Setup }
-%
-We performed all experiments on a 3.40GHz Intel Core i7-2600 CPU with 16GB RAM running Ubuntu Linux 12.04.5 LTS, LLVM 3.5, SMACK 1.5.0, Z3~4.3.2, Boogie rev.
-4192 and \corral rev.
-534.
-We also used Mono~4.1.0 to run Boogie and \corral.
-
-\noindent\textbf{Benchmarks }
-%
-We evaluate our methodology against \sizeOfBenchmarks drivers from the Linux 4.0 kernel.
-We chose non-trivial drivers of several types: block, char, ethernet, near field communication (nfc), universal serial bus (usb), and watchdog.
-For all these drivers, we had to understand their environment and manually model it; this required approximately two months of work.
-
-\noindent\textbf{Race-Checking }
-%
-Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
-
-We can see that \whoop reports more races than \corral does.
-This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.
-On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.
-Another issue with \corral is loop coverage due to unsound loop unrolling.
-To tackle this, we enable the built-in loop over-approximation described in previous work~\cite{lal2014powering}.
-This can potentially lead \corral to report false bugs, but we have not seen this in practice.
-Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
-Standalone \corral did not discover any data races that \whoop did not already report.
-This is expected, as \whoop aims for soundness, and this increases our confidence in the implementation of \whoop.
-
-Most of the races that \whoop and \corral discovered fall into two categories.
-The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock.
-This might be for performance, and indeed a lot of the races we found might be benign.
-Even benign races, though, lead to undefined behavior according to the C standard\COULDCUT{, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations}.
-The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock.
-This can lead to a race if another entry point simultaneously accesses the same field of the same struct.
-
-As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
-This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
-We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.
-We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.
-
 \noindent\textbf{Accelerating \corral }

### 2015-08-10 intro.tex
fixes

--- 2015-08-05 intro.tex
@@
-Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer.
+Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and the devices that are attached to a computer.
 Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}.

### 2015-08-10 whoop.tex
fixes

@@
 A thread terminates if it reaches the end of $\mainforthread{T}$, in which case its state is removed from the list of thread states.
-Since our interest is in analysis of device drivers, which are reactive, we do not consider the notion of global program termination.
+Since our interest is in analysis of device drivers, which are reactive programs, we do not consider the notion of global program termination.
 
@@
 In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers.
-We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, automated tool for detecting data races in drivers.
+We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, automated tool for detecting data races in drivers.
 
@@
 The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
-Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Initially, \whoop uses three LLVM-based tools~\cite{lattner2004llvm}, Chauffeur, Clang and SMACK~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
 Boogie is a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ the Boogie verifier and \corral).
@@
 
-We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Fig.~\ref{fig:whoop}).
+We engineered the Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}} and \whoop\footnote{\url{https://github.com/pdeligia/whoop}} components of our toolchain (denoted with grey boxes in Fig.~\ref{fig:whoop}).
 For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.
@@
 %
-Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form.
+Next, the driver is compiled by Clang to LLVM-IR, a low-level assembly-like language in single static assignment (SSA) form.
 Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
@@
 An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL.
-Thus, \whoop does not need to directly deal with pointers and alias analysis\COULDCUT{, a hard problem in its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts}.
+Thus, \whoop does not need to directly deal with pointers and alias analysis.
+%, a hard problem in its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts.
 
@@
 
-\COULDCUT{
-In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.}
+%In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.
 
@@
 To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions.
-A limitation of our model is that it can, and will, ultimately result in false positives.
+%
+A limitation of our model is that it can be inaccurate, leading to semantic mismatches that can in turn lead to false positives and/or false negatives.
+%
 However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision.

### 2015-08-10 evaluation.tex
fixes

@@
 We also used Mono~4.1.0 to run Boogie and \corral.
+We configured \corral with a time budget of 10 hours (T.O.
+denotes timeout), a context-switch bound (csb) of 2, 5, and 9, and the default recursion depth bound of 1.
 
@@
 %
-Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
+Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a csb of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb than 2 did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
+We experimented with higher recursion depth bounds, but we were unable to discoverer any races that could not be found with the default one.
 
@@
 %
-Table~\ref{tab:results} presents runtime results from using \whoop, \corral, and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
-All reported times are in seconds and averaged over three runs.
-\corral was configured with a time budget of 10 hours (T.O.
-denotes timeout) and a csb of 2, 5, and 9.
+Table~\ref{tab:results} presents runtime results from using \whoop, standalone \corral, and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation.
 Standalone \corral uses \yieldall, which instruments context-switches (i.e.
@@
 The table also shows the number of context-switches per instrumentation (\#Y).
+All reported times are in seconds and averaged over three runs.
 

### 2015-08-11 background.tex
fixes

@@
 
-Fig.~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver.
-The driver can invoke the entry point concurrently from two threads, with the same \texttt{file} struct as a parameter.
+Fig.~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 device driver.
+This entry point can be invoked concurrently from two threads, with the same \texttt{file} struct as a parameter.
 This can lead to multiple possible data races because the threads can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order.
@@
 Fig.~\ref{fig:lock_example} shows how to use locking to eliminate the races in Fig.~\ref{fig:data_race_example}.
-\COULDCUT{Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.}
+%
+%Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
 %
@@
 Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program.
-Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start.
+%
+Every $\mathit{CLS}_T$ is initially empty because the threads do not hold any locks when they start executing.
 In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock.

### 2015-08-11 corral.tex
fixes

@@
 %
-To detect data races with \corral, \whoop outputs a Boogie program instrumented with a simple, but effective encoding of race checks.
+To detect data races with \corral, \whoop outputs a Boogie program instrumented with a simple, but effective encoding of race checks~\cite{erickson2010datacollider}.
 Whenever there is a write access to a shared variable $s$, \whoop instruments the program as follows:
@@
 However, \corral is precise and, assuming a precise environmental model, it will only report true races.
-\whoop takes advantage of this precision to report only feasible data races.
+\whoop takes advantage of this precision to report only feasible races.
 %
-Note that our instrumentation conveniently tolerates some benign races: it does not report a read-write race if the write access updates the racy memory location with the same value that already existed; it also does not report a write-write race if the two write accesses update the racy memory location with the same value, which can be different from the pre-existing.
+Note that our instrumentation conveniently tolerates some benign races: it does not report a read-write race if the write access updates the shared memory location with the same value that already existed; it also does not report a write-write race if the two write accesses update the shared memory location with the same value (which can be different from the pre-existing value).
 
@@
 %
-\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region (MR) that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock.
+\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region (MR) that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock statement.
 In our experiments (see \S\ref{evaluation}), \yieldmr is significantly faster than \yieldcoarse.

### 2015-08-11 evaluation.tex
fixes

@@
 %
-Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a csb of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb than 2 did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.
-We experimented with higher recursion depth bounds, but we were unable to discoverer any races that could not be found with the default one.
+Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a csb of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb than 2 did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its default recursion depth bound of 1 before discovering a deeper bug.
+Although we experimented with higher recursion depth bounds, we were unable to discover any races that could not be exposed with the default recursion depth bound.
 
-We can see that \whoop reports more races than \corral does.
+We can see in Table~\ref{tab:stats} that \whoop reports more races than \corral does.
 This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.
@@
 Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.
-Standalone \corral did not discover any data races that \whoop did not already report.
-This is expected, as \whoop aims for soundness, and this increases our confidence in the implementation of \whoop.
+Note that standalone \corral did not discover any races that \whoop did not already report.
+This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.
 
@@
 As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}).
-This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.
-We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.
+This causes a race if the kernel invokes \texttt{llseek} from another thread, while passing the same \texttt{file} object as a parameter.
+We observed that another char driver, using the same APIs, \emph{does} use a lock to protect the offset access in its respective \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.
 
@@
 \whoop uses over-approximation to scale and, as expected, executes significantly faster than \corral in all our benchmarks.
-For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds.
-We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which puts burden on \corral's stratified inlining.
+For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the same driver in 144.5 seconds.
+We believe that the reason behind this is that the r8169 driver has deeply-nested recursion in some of its entry points, which puts burden on \corral's stratified inlining.
 This is not an issue for \whoop, which uses procedure summarization.
@@
 We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9).
-We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.
+We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldcoarse or \yieldmr.
 

### 2015-08-11 related.tex
some fixes
fixes

--- 2015-08-05 related.tex
@@
 
-Prior works on static race checking also include the following.
+%Prior works on static race checking also include the following.
 Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection.
@@
 
-Our pairwise approach to analysing driver entry points, employing abstraction to model additional threads, was inspired by the \emph{two thread reduction} used by the GPUVerify tool in the analysis of data-parallel OpenCL and CUDA kernels~\cite{gpuverify,bardsley2014engineering}.
+Our pairwise approach to analyzing driver entry points, employing abstraction to model additional threads, was inspired by the \emph{two thread reduction} used by the GPUVerify tool in the analysis of data-parallel OpenCL and CUDA kernels~\cite{gpuverify,bardsley2014engineering}.
 The idea of pairwise analysis of components in a concurrent system has been broadly applied, notably in model checking of cache coherence protocols~\cite{mcmillan1999verification}.

### 2015-08-11 whoop.tex
fixes

@@
 Multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.
-Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers, or dynamic memory allocation.
+Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not typically occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers, or dynamic memory allocation.
 These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.
@@
 %
-For templates $T$ and $U$, which can be equal, we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrary many further concurrently executing threads.
+For templates $T$ and $U$, which can be equal, we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
 %
@@
 
-To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from previous work~\cite{bardsley2014engineering}.
-Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region $\mathit{MR}$.
+To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog} race-checking instrumentation, adapted from previous work~\cite{bardsley2014engineering}.
+Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie IVL these entry points share a common memory region $\mathit{MR}$.
 When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_{\mathit{MR}}$, representing some unspecified index into $\mathit{MR}$; we call this the \emph{watched offset} for $\mathit{MR}$.
 We then attempt to prove that it is impossible for $T$ and $U$ to race on $\mathit{MR}$ at index $\mathit{watched}_{\mathit{MR}}$.
-If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $\mathit{MR}$, since the watched offset was arbitrary.
+If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} $\mathit{MR}$, since the watched offset was arbitrary.
 This technique of choosing an arbitrary index to analyze for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $\mathit{MR}$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $\mathit{MR}$.

### 2015-08-11 whoop.tex
Added references to web pages

@@
 The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races.
-Initially, \whoop uses three LLVM-based tools~\cite{lattner2004llvm}, Chauffeur, Clang and SMACK~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
+Initially, \whoop uses three LLVM-based tools~\cite{lattner2004llvm,llvm}, Chauffeur, Clang~\cite{clang}, and SMACK~\cite{rakamaric2014smack,smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.
 Boogie is a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ the Boogie verifier and \corral).
